### 2001-03-02 technical/proposal/proposal.tex
Created the project proposal for the NLP toolkit. The following sections have been written:   - overview   - design criteria   - toolkit design and implementation
The following sections have not been written:
  - documentation
  - preliminary design and implementation


svn/trunk@6

%
% Natural Language Toolkit for Python
% Edward Loper
% 
% Created [03/02/01 12:02 AM]
% $Id$
%

\documentclass{article}
\usepackage{fullpage}

\begin{document}
\title{Proposal: A Natural Language Programming Toolkit for Python}
\author{Edward Loper, Steven Bird}
\maketitle

%################################################################
%#  OVERVIEW
%################################################################
\section{Overview}

CIS-530 is an introductory course in natural language processing
(NLP), which is intended to be accessable to both linguists and
computer scientists \cite{cis530}. In such a class, it is instructive
for students to build real NLP systems.  By building real systems,
students can gain an in-depth understanding of the issues involved in
particular aspects of NLP.

However, the students do not necessarily have any experience
programming or designing software systems.  Thus, when students
attempt to build NLP systems, they may spend a signifigant amount of
their time on programming tasks that are not directly related to the
course of the content.

In order to try to remove some of the time overhead associated with
these programming tasks, we propose to design and implement a toolkit
for building natural language systems using the Python
language \cite{python}. This toolkit will help students in a variety
of ways:

\begin{itemize}
  \item It will provide students with basic tools for manipulating
  data and performing tasks that are related to NLP.

  \item It will allow the students to experiment in building NLP
  systems at a variety of levels.  At one extreme, they can implement
  individual components for use with the toolkit.  At the other
  extreme, they can create large systems by combining toolkit
  components.

  \item It will provide the students with an infrastructure for
  building natural language systems, by providing consistant and
  well-defined interfaces for different components.  Thus, students
  will not have to spend their time thinking about system design
  issues.
\end{itemize}

%################################################################
%#  DESIGN GOALS/CRITERIA
%################################################################
\section{Design Criteria}

Several criteria must be considered in the design and implementation
of the toolkit.  The design criteria are divided into primary and
secondary criteria, and within those sections are listed in the order
of their importance.  A set of explicit non-requirements is also
given.  The toolkit is not expected to satisfy any of these
properties.

\subsection{Primary Design Criteria}

\begin{description}

\item[Ease of Use] The primary purpose of the toolkit is to allow
students to concentrate on building NLP systems.  The more time
students must spend learning to use the toolkit, the less useful it
is.

\item[Consistancy] The toolkit should use consistant data structures
and interfaces.

\item[Extensibility] The toolkit should easily accomodate new
components, whether those components replicate or extend the toolkit's
existing functionality.  Thus, the toolkit's design should be modular,
with simple and well-defined interfaces between modules.  The toolkit
should also be structured in such a way that it is obvious where new
extensions would fit into the toolkit's infrastructure.

\item[Documentation] The toolkit, its data structures, and its
implementation all need to be carefully and thouroughly documented.

\end{description}

\subsection{Secondary Design Criteria}

\begin{description}
\item[Simplicity] The toolkit should structure the complexities of
building NLP systems, not hide them.  Therefore, each class defined by 
the toolkit should be simple enough that a student could implement it
by the time they finish the course.

\item[Modularity] The interaction between different components of the
toolkit should be kept to a minimum.  In particular, it should be
possible to use complete individaul projects using small parts of the
toolkit, without worrying about how they interact with the rest of the
toolkit.  This will allow students to learn to use the toolkit
incrementally throughout the course.  Modularity also makes it easier
to change or extend the toolkit.
\end{description}

\subsection{Non-Requirements}

\begin{description}
\item[Comprehensiveness] The toolkit is not intended to provide a
comprehensive set of tools.  Indeed, there should be a wide variety of 
ways in which students can extend the toolkit.

\item[Efficiency] The toolkit doe not need to be highly optimized for
runtime performance.  However, it should be efficient enough that
students can use their NLP systems to perform real tasks.

\item[Cleverness] Clear designs and implementations are far preferable 
to ingenious yet indecipherable ones.
\end{description}

%################################################################
%#  TOOLKIT DESIGN AND IMPLEMENTATION
%################################################################
\section{Toolkit Design and Implementation}

\subsection{Infrastructure Design}

The natural langauge processing toolkit will be implemented by a
collection of independant modules\footnote{These modules do not
necessarily correspond to Python \texttt{module}s.}.  These modules are
organized according to two primary axes:

\begin{itemize}
  \item \textbf{Task:} Toolkit modules are generally associated with
  one or more NLP tasks, and several different modules may be
  associated with the same task.

  \item \textbf{Orientation:} Toolkit modules are either
  \emph{data-oriented} or \emph{task-oriented}.

  \begin{itemize}
    \item Data-oriented modules are used to store different types of
    information that are relevant to natural language processing.
    \item Task-oriented modules are used to perform a variety of tasks 
    that are relevant to natural language processing.
  \end{itemize}
\end{itemize}

\subsubsection{Dependancies}

In general, data-oriented modules should be self-contained, and should
exhibit very little inter-dependancy.  Data-oriented modules never
depend on task-oriented modules.

Task-oriented modules should also be relatively self-contained, with
very little inter-dependancy.  However, each task-oriented modules
will usually depend on a variety of data-oriented modules.  When
possible, the number of data-oriented modules that a task-oriented
module depends on should be kept low.  This allows students to use
task-oriented modules without needing to learn how to use a large
number of other modules.

\subsection{Module Design}

Each module define a small set of interfaces, which give
specifications for the types of classes that can be implemented by the
module.  Each module also contains classes that implement each
interface.  Often, a module will contain more than one implementation
for the same interface.

\subsubsection{Example Module: Tokens}

For the sake of concreteness, I will describe the design of a
data-driven module for handling tokens.

The token module defines two interfaces:

\vspace{1.2mm}\noindent
\begin{tabular}{||p{.21\textwidth}p{.7\textwidth}}
  \textbf{TokenType} & A unit of text, such as a word or a
  punctuation mark. \\

  \textbf{TextLocation} & The location of an entity within a
  text.\\
\end{tabular}
\vspace{1mm}

\noindent
Each of these interfaces specifies a set of methods that classes must
implement if they are to be considered \texttt{TokenType}s or
\texttt{TextLocation}s.  If a class properly implements every method specified
by an interface, then it is said to \emph{implement} that interface.

The token module also defines one top-level class:

\vspace{1.2mm}\noindent
\begin{tabular}{||p{.21\textwidth}p{.7\textwidth}}
  \textbf{Token} & An Occurance of a \texttt{TokenType}
  within a text.  A \texttt{token} consists of a \texttt{TokenType}
  and a \texttt{TextLocation}. \\
\end{tabular}
\vspace{1mm}

\noindent
This is somewhat unusual; most modules do not implement any top-level
classes.  However, when a data structure has only one reasonable
implementation, a top-level class may be used in place of an interface 
and interface specification.  This is the case for \texttt{Token}s,
which are basically just the result of combining a \texttt{TokenType} 
with a \texttt{TextLocation}.

The only method that \texttt{TokenType}s are required to implement is
the comparison operator.  This allows users of \texttt{TokenType}s to
decide whether two \texttt{TokenType}s represent the same unit of
text.  The token module currently defines two implementations for the
\texttt{TokenType} interface:

\vspace{1.2mm}\noindent
\begin{tabular}{||p{.21\textwidth}p{.7\textwidth}}
  \textbf{SimpleTokenType} & A \texttt{TokenType}s that is
  represented by a single Python \texttt{string}.  Examples of
  \texttt{SimpleTokenType} values are ``dog'' and ``walking.'' \\

  \textbf{TaggedTokenType} & A \texttt{TokenType} that is
  represented by a \texttt{string} and a part-of-speech tag.  Examples 
  of \texttt{TaggedTokenType} values are ``dog/NN,'' ``bank/NN,'' and
  ``bank/VB.'' \\
\end{tabular}
\vspace{1mm}

Like \texttt{TokenType}s, the only method that \texttt{TextLocation}s
are required to implement is the comparison operator.  This allows
users of \texttt{TextLocation}s to decide whether two
\texttt{TextLocation}s represent the same location in a text.  The
token module currently defines one implementation for the
\texttt{TextLocation} interface:

\vspace{1.2mm}\noindent
\begin{tabular}{||p{.21\textwidth}p{.7\textwidth}}
  \textbf{IndexTextLocation} & The location of an
  entity within a text, using a numerical index. \\
\end{tabular}
\vspace{1mm}

\subsection{Programming Concepts}

The structure of the Python language directly affects the design and
implementation of the toolkit.  This section discusses how the design
of the toolkit is affected by the Python language.

\subsubsection{Interfaces}

Python does not directly implement interfaces.  The toolkit will
therefore implement interfaces as simple classes, all of whose methods 
raise \texttt{AssertionError} exceptions.  A description of the
functionality provided by the interface will be contained in the
class's documentation string.  Each method's documentation string will 
contain a specification of the behavior for that method.

Occasionally, interfaces will define optional methods.  These are
methods that may be defined by classes that implement the interface,
but need not be.  Instead of raising \texttt{AssertionError}
exceptions, these methods will raise \texttt{NotImplementedError}
exceptions.

In order to implement an interface, a class should include that
interface as one of its bases, and should override every
(non-optional) method defined by the interface.

Since Python supports multiple inheritance, this implementation of
interfaces will allow a single class to implement multiple
interfaces.  In general, interfaces should be listed after true bases
in the base list, to ensure proper inheritance of methods.

In principle, an interface hierarchy could be built by defining
interfaces whose bases are interfaces.  However, I do not currently
plan to use this ability.

Classes that define interfaces are named with a trailing ``I,'' such
as \texttt{TokenizerI} or \texttt{EventI}.

\subsubsection{Typing}

Type checking is an imporant safety device, that allows programmers to
quickly detect the locations of errors in their code.  This is
especially important for novice programmers, who have less experience
with debugging.

Therefore, every function and method defined by the toolkit will
perform a complete type check on all of its arguments.  By default,
container types (lists, tuples, and dictionaries) will be checked in a 
``deep'' fashion: every element will be checked to ensure that it is
the correct type.  For example, if a procedure expects a list of
integers, it will check every element of the list to ensure that they
are integers before it begins executing.

However, such ``deep'' type checking can have a serious performance
cost: every time a procedure is called, the type of every element of a
potentially very large data structure must be checked.  Therefore, the 
toolkit will provide a function which can adjust the level of type
checking performed by all toolkit functions.  This function can be
used to temporarily lower the type checking level when processing
large amounts of data.  However, it is recommended that the highest
level of type checking be used when possible.

\subsubsection{Language Features}

The use of advanced language features and programming styles could
potentially make the toolkit much more difficult to use or to
understand.  The fewer language features that a student must learn in
order to use the toolkit, the faster they can start developing NLP
systems.  This section discusses the advantages and disadvantages of
using a variety of language features.  The language features used by
the toolkit are summarized by figure \ref{table:feature}.

\begin{figure}
\begin{centering}
\begin{tabular}{|l|l|}
\hline
\textbf{Language Feature} & \textbf{Use} \\
\hline
Classes & used \\
Exceptions & error conditions only \\
Operator Overloading & used \\
Python 2.0 Features & used \\
Function Arguments & minimal use \\
Lambda Functions & minimal use \\
Mapping and Filtering & not used \\
List Comprehensions & not used \\
\hline
\end{tabular}\\
\end{centering}
\caption{Summary of which advanced language features are used by the
toolkit.}
\label{table:feature}
\end{figure}

\begin{description}

  \item[Classes] Classes are an integral part of the design of the
  toolkit; they will be used.

  \item[Exceptions] Exceptions are used by the core Python language to
  report error conditions.  Exceptions can also be used to report
  exceptional conditions.  The toolkit will only use exceptions to
  report error conditions.  As a result, students who are using the
  toolkit need not learn anything about raising or catching
  exceptions.  Students who are extending the toolkit should learn how
  to raise exceptions, but need not learn anything about catching
  exceptions.

  \item[Operator Overloading] Python allows programmers to redefine
  the behavior of operators when used with class instances.  The use
  of operator overloading can make classes easier to use, since they
  can behave more like the built-in classes.  However, extensive use
  of operator overloading can make classes difficult to understand and
  difficult to program.  Classes and interfaces defined by the toolkit
  will use operator overloading.

  \item[Python 2.0 Features] Version 2.0 of Python defines a number of
  useful features, such as augmented assignment; string methods; and
  containership operator overloading.  In general, these features will
  make the toolkit easier to use and understand.  However, if the
  toolkit uses these features, then it cannot be used with earlier
  versions of Python.  The toolkit will use Python 2.0 features, and
  thus will require Python 2.0.

  \item[Function Arguments] Allowing functions to act as arguments can
  be a very powerful tool.  For example, by passing a function to the
  built-in function \texttt{sort}, a programmer can sort a list
  according to any arbitrary ordering.  However, the use of functions
  as arguments can be difficult to understand.  The toolkit will
  therefore make minimal use of functions as arguments.  It should
  never be necessary to use function arguments to accomplish a task.

  \item[Lambda Functions] Lambda functions allow simple functions to
  be defined where they are used.  When used in conjunction with
  functional arguments, lambda functions can be a powerful tool.
  However, their use is not always intuitive.  The toolkit will
  therefore make minimal use of lambda functions.  It will only use
  them in the context of simple and well-defined idioms, such as event
  creation: \texttt{FuncEvent(lambda~x:x>3)}.

  \item[Mapping and Filtering] Mapping and filtering provide a
  consise way of expressing transformations on lists.  However, they
  are fairly difficult to learn, and generally involve the use of
  \texttt{lambda} functions, so the toolkit will not use them.

  \item[List Comprehensions] List comprehensions provide an even more
  concise way of expressing transformations on lists.  However, in
  order to reduce the number of new language features that students
  must learn to understand the toolkit, they will not be used.

\end{description}

%################################################################
%#  DOCUMENTATION
%################################################################
\section{Documentation}

Two main types of documentation:
  * tutorial docs
  * reference docs

reference docs are generated from inline comments by epydoc.

%################################################################
%#  PRELININARY DESIGN AND IMPLEMENTATION
%################################################################
\section{Preliminary Design and Implementation}

As part of the preliminary design process, I desinged and implemented
six prototype modules.

%################################################################
%#  BiIBLIOGRAPHY
%################################################################
\begin{thebibliography}{1}

\bibitem{python}{The Python programming language. \\
    \texttt{http://www.python.org/}}

\bibitem{cis530}{CIS 530 class homepage.  
                 University of Pennsylvania. \\ 
    \texttt{http://www.cis.upenn.edu/~cis530/}}

\bibitem{epydoc}{Epydoc: Edloper's Python Documentation Tool.
    Contact \texttt{ed@loper.org} for more information.}

\end{thebibliography}

\end{document}

### 2001-03-02 technical/proposal/proposal.tex
Added content:   * Wrote documentation section   * Wrote about half of the preliminary design & impl section
svn/trunk@9

@@
 \usepackage{fullpage}
+\usepackage{boxedminipage}
 
@@
 using a variety of language features.  The language features used by
-the toolkit are summarized by figure \ref{table:feature}.
+the toolkit are summarized by figure \ref{fig:feature}.
 
@@
 toolkit.}
-\label{table:feature}
+\label{fig:feature}
 \end{figure}
@@
 
-Two main types of documentation:
-  * tutorial docs
-  * reference docs
+The toolkit will be accompanied by documentation that explains the
+toolkit, and describes how to use and extend it.  This documentation
+will be divided into three primary categories:
 
-reference docs are generated from inline comments by epydoc.
+\begin{itemize}
+  \item \textbf{Technical Reports}, which explain and justify the
+  toolkit's design and implementation.
+
+  \item \textbf{Tutorials}, which teach students how to use the
+  toolkit, in the context of performing specific tasks.
+
+  \item \textbf{Reference Documentation}, which describes every
+  module, interface, class, method, function, and variable in the
+  toolkit.
+\end{itemize}
+
+Technical reports will be available as latex files, dvi files, and
+postscript files.  Tutorials and reference documentation will be
+available as HTML files, accessble over the web.
+
+\subsection{Technical Reports}
+
+Throughout the course of developing the toolkit, I will write a number
+of technical reports, which are intended to explain and justify
+specific aspects of the toolkit's design or implementation.  This
+document is an example of a technical report.  Students should not
+need to read the technical reports to use the toolkit.  However,
+students can consult these reports if they would like further
+information about how the tookit is designed and why it is designed
+that way.
+
+\subsection{Tutorials}
+
+Tutorials are the primary type of documentation that students will use
+to learn how to use the toolkit.  Each tutorial will explain how to
+use some aspect of the toolkit, using specific examples.  In general,
+the tutorials will explain how the student can perform some class of
+tasks, such as using tokenizers and taggers to tag text with
+part-of-speech tags.  Some tutorials may be associated with specific
+toolkit modules.
+
+Different tutorials will assume different amounts of background
+knowledge about the toolkit.  This assumed background knowledge will
+be specified by listing ``prerequisite'' tutorials for each tutorial.
+At least one tutorial will assume no background knowledge about the
+toolkit.
+
+\subsection{Reference Documentation}
+
+Reference documentation gives definitions for each module, interface,
+class, method, function, and variable defined by the toolkit.  These
+definitions should be both precise and consise.  I currently plan to
+produce the reference documentation using Epydoc \cite{epydoc}.
+Epydoc is a tool for automatically converting python documentation
+strings to HTML pages.  Epydoc will generate the following HTML pages:
+
+\begin{itemize}
+  \item A page for each python \texttt{module}.  This page includes an
+  overview of the \texttt{module}; a list of the \texttt{classs}es
+  defined by the \texttt{module}; a list of the \texttt{function}s
+  defined by the \texttt{module}, and a detailed description of each
+  \texttt{function}; a list of the \texttt{variable}s defined by the
+  \texttt{module}, and a detailed description of each
+  \texttt{variable}.
+
+  \item A page for each python \texttt{class}.  This page includes an
+  overview of the \texttt{classs}; a list of the \texttt{method}s
+  defined by the \texttt{classs}, and a detailed description of each
+  \texttt{method}; a list of the instance \texttt{variable}s defined
+  by the \texttt{classs}, and a detailed description of each instance
+  \texttt{variable}; a list of the \texttt{classs} \texttt{variable}s
+  defined by the \texttt{classs}, and a detailed description of each
+  \texttt{classs} \texttt{variable}.
+
+  \item A page containing the toolit's complete \texttt{classs}
+  hierarchy.
+
+  \item A page containing an index of every \texttt{module},
+  \texttt{classs}, \texttt{method}, \texttt{function}, and
+  \texttt{variable} defined by the toolkit.
+
+\end{itemize}
+
+Epydoc is currently under development.  Contact Edward Loper for more
+information about Epydoc.  For examples of the HTML pages generated by 
+Epydoc, see the reference documentation for the toolkit's prototype
+modules:
+
+ \texttt{http://www.cis.upenn.edu/$\sim$edloper/??}
 
 %################################################################
-%#  PRELININARY DESIGN AND IMPLEMENTATION
+%#  PRELIMINARY DESIGN AND IMPLEMENTATION
 %################################################################
@@
 As part of the preliminary design process, I desinged and implemented
-six prototype modules.
+six prototype modules.  These modules are summarized in figures
+\ref{fig:mod-data} and \ref{fig:mod-task}, and described in the
+following sections.  Most of these summaries are copied directly from
+the modules' reference documentation.  For more information, see the
+reference documentation at:
+
+ \texttt{http://www.cis.upenn.edu/$\sim$edloper/??}
+
+\begin{figure}
+\noindent
+\begin{centering}
+\begin{boxedminipage}{.95\textwidth}
+\begin{itemize}
+  \item \textbf{Sets}: Encodes the mathmatical notion of a ``finite set.''
+  \begin{itemize}
+    \item \textit{Set}: A finite set.
+  \end{itemize}
+
+  \item \textbf{Tokens}: Encodes units of text such as words.
+  \begin{itemize}
+    \item \textit{TokenTypeI}:
+         A unit of text.
+    \item \textit{TextLocationI}:
+         A location within a text.
+    \item \textit{Token}:
+         An occurance of a unit of text.
+         Consists of a TokenType and a TextLocation.
+  \end{itemize}
+
+  \item \textbf{Syntax Trees}: Encodes syntax trees.  Not fully designed 
+        yet.
+
+  \item \textbf{Probability}: Encodes data structures associated with
+        the mathmatical notion of probability, such as events and
+        frequency distributions.
+  \begin{itemize}
+    \item Sample: Encodes the mathmatical notion of a
+          ``sample.''  This is actually not implemented as a class or
+          an interface -- (almost) anything can be a Sample.
+    \item \textit{EventI}:
+         A (possibly infinite) set of samples.
+    \item \textit{FreqDistI}:
+          The frequency distribution of a collection of samples.
+    \item \textit{ProbDistI}:
+          A probability distribution, typically derived from a
+          frequency distribution (e.g., using ELE).
+  \end{itemize}
+\end{itemize}
+\end{boxedminipage}\\
+\end{centering}
+ \caption[Data-oriented module prototypes]{Data-oriented module
+ prototypes.  Under each module, the interfaces and top-level classes
+ defined by that module are listed.}
+\label{fig:mod-data}
+\end{figure}
+
+\begin{figure}
+\noindent
+\begin{centering}
+\begin{boxedminipage}{.95\textwidth}
+\begin{itemize}
+  \item \textbf{Tokenizers}: Separate a string of text into a list of
+       Tokens. 
+  \begin{itemize}
+     \item \textit{TokenizerI}
+  \end{itemize}
+
+  \item \textbf{Taggers}: Assign tags to each Token in a list of Tokens.
+  \begin{itemize}
+     \item \textit{TaggerI}
+  \end{itemize}
+\end{itemize}
+\end{boxedminipage}\\
+\end{centering}
+ \caption[Task-oriented module prototypes]{Task-oriented module
+ prototypes.  Under each module, the interfaces and top-level classes
+ defined by that module are listed.}
+\label{fig:mod-task}
+\end{figure}
+
+\subsection{Sets}
+
+The set module contains a single class, \texttt{Set}.
+
+\subsubsection{\texttt{Set} Class}
+
+    An unordered container class that contains no duplicate elements.
+    In particular, a set contains no elements $e_1$ and $e_2$ such
+    that $e_1=e_2$.  Currently, the \texttt{Set} class is given a
+    fairly minimal implementation.  However, more members may be
+    defined in the future.
+
+    Although the \texttt{Set} class attempts to ensure that it
+    contains no duplicate elements, it can only do so under the
+    following circumstances:
+
+    \begin{itemize}
+      \item For all elements $e_i$, $e_j$ added to the \texttt{Set},
+           $e_i=e_j$ if and only if $e_j=e_i$.  This should always be the
+           case as long as the elements in the \texttt{Set} use
+           well-defined comparison functions.  An example where it
+           would not be the case would be if ei defined
+           \texttt{\_\_cmp\_\_()} to always return 0, and ej
+           defined \texttt{\_\_cmp\_\_()} to always return $-1$.
+
+      \item Mutable elements inserted in the \texttt{Set} are not
+           modified after they are inserted.
+    \end{itemize}
+
+    If these circumstances are not met, the \texttt{Set} will
+    continue to function, but it will no longer guarantee that it
+    contains no duplicate elements.
+
+\subsection{Tokens}
+
+A Token is an occurance of a single unit of text, such as a word or a
+punctuation mark.  The token module defines the \texttt{TokenTypeI}
+interface, the \texttt{TextLocationI} interface, and the \texttt{Token
+class}.
+
+\subsubsection{\texttt{TokenTypeI} Interface}
+
+    A \texttt{TokenTypeI} is single unit of text, such as a word or a
+    punctuation mark.  The precise defintion of what counts as a token
+    type will vary for different analyses.  For example, ``bank'' and
+    ``run'' might be tokens for one analysis, while ``bank/N'' and
+    ``bank/V'' might be tokens for another analysis.
+
+    Token types can be compared for equality, and should be equal if
+    and only if they refer to the same token type.
+
+\vspace{2mm}\noindent
+\textit{The \texttt{TokenTypeI} interface is currently implemented by
+the classes \texttt{SimpleTokenType} and \texttt{TaggedTokenType}.}
+
+\subsubsection{\texttt{TextLocationI} Interface}
+
+    A \texttt{TextLocationI} is the location of an entity within a
+    text.  Text locations can be compared for equality, and should be
+    equal if and only if they refer to the same location.
+
+\vspace{2mm}\noindent
+\textit{The \texttt{TextLocationI} interface is currently implemented
+by the \texttt{IndexTextLocation} class.}
+
+\subsubsection{\texttt{Token} Class}
+
+    A \texttt{Token} is an occurance of a single unit of text, such as
+    a word or a punctuation mark.  A Token consists of a token type
+    and a source.  The token type is the unit of text (e.g., a
+    specific word).  The source is the position at which this token
+    occured in the text.
+
+    The precise defintion of what counts as a token type, or unit of
+    text, will vary for different analyses.  For example, ``bank'' and 
+    ``run'' might be tokens for one analysis, while ``bank/N'' and
+    ``bank/V'' might be tokens for another analysis.
+
+\subsection{Syntax Trees}
+
+The syntax tree module defines data structures for storing syntax
+trees.  It is still under development.  Currently, it defines one
+interface: \texttt{SyntaxNodeI}
+
+\subsubsection{\texttt{SyntaxNodeI} Interface}
+
+\emph{This interface is still under development}
+
+A syntax node is a single internal node of a syntax tree.  Each syntax
+node is characterized by a tag and an ordered list of children.  The
+tag specifies the type of the syntax Node (e.g., ``NP'' or ``V'').
+Each child must either be a \texttt{Token}, or implement the
+\texttt{SyntaxNodeI} interface.  Access to the children is implemented
+by emulating a sequence type.  The \texttt{SyntaxNodeI} does
+\emph{not} perform any semantic or syntactic checks on its arguments.
+Figure \ref{fig:syntaxnode} shows how syntax nodes can be used:
+
+\begin{figure}
+\noindent
+\begin{centering}
+\begin{boxedminipage}{\textwidth}
+\begin{tabbing}
+\texttt{>>> cat = Token(\ldots)}\\
+\texttt{>>> dog = Token(\ldots)}\\
+\texttt{>>> the = Token(\ldots)}\\
+\texttt{>>> chases = Token(\ldots)}\\
+\\
+\texttt{>>> cat\_np = SyntaxNode('NP', [the, cat])}\\
+\texttt{>>> dog\_np = SyntaxNode('NP', [the, dog])}\\
+\texttt{>>> vp = SyntaxTree('VP', [chases, dog\_np])}\\
+\texttt{>>> s = SyntaxTree('S', [cat\_np, vp])}\\
+\\
+\texttt{>>> print s}\\
+\texttt{\textit{[S [NP the dog] [VP chases [NP the dog]]]}}\\
+\texttt{>>> print s[0]}\\
+\texttt{\textit{[NP the dog]}}\\
+\texttt{>>> print s[1].tag()}\\
+\texttt{\textit{VP}}\\
+\\
+\texttt{>>> s[0] = dog\_np}\\
+\texttt{>>> vp[:] = [cat, cat, cat]}\\
+\end{tabbing}
+\end{boxedminipage}\\
+\end{centering}
+ \caption[Example \texttt{SyntaxNode} use]
+ {A session showing how \texttt{SyntaxNode}s can be used.}
+\label{fig:syntaxnode}
+\end{figure}
+
+\subsection{Probability}
+
+The probability module provides data structures associated with the
+mathmatical notion of probability.  The following mathmatical terms
+are useful for thinking about probability:
+
+\begin{itemize}
+  \item An \emph{experiment} is an occurance we observe whose result 
+        is uncertain.
+  \item An \emph{outcome} is some aspect of the result of an
+        experiment which we wish to consider.
+  \item A \emph{sample} is a possible outcome of an experiment.
+  \item The \emph{sample space} of an experiment is the set of its
+        possible outcomes.
+  \item An \emph{event} is some subset of the sample space of an
+        experiment. 
+\end{itemize}
+
+The toolkit does not currently encode the notion of a sample with an
+interface; instead, it allows (almost) any object to function as a
+sample.  Thus, for example, the sample space corresponding to a roll
+of a die might be the set of integers from 1 to 6.  The only
+requirement on samples is that they properly define the comparison
+operator.
+
+The toolkit currently defines the \texttt{CFSample} class for the
+specific purpose of acting as a sample.  \texttt{CFSample}s consist of 
+a (\emph{context}, \emph{feature}) pair.
+
+The toolkit currently defines three interfaces for handling the
+mathmatical notion of probability: the \texttt{EventI} interface; the
+\texttt{FreqDistI} interface; and the \texttt{ProbDistI} interface.
+
+\subsubsection{\texttt{EventI} Interface}
+
+    An event is a subset of some sample space.  Note that this subset
+    need not be finite.  Events are typically written as the set of
+    samples they contain, or as a function in first order logic.
+    Examples are:
+
+\begin{verbatim}
+{1,2,3}
+{x:x>0}
+\end{verbatim}
+
+    The only method that events are required to implement is
+    \texttt{\_\_contains\_\_()}, which tests whether a sample is a
+    contained by the event.  However, when possible, events should
+    also define the following methods:
+    \begin{itemize}
+      \item \texttt{\_\_cmp\_\_()}, which tests whether this event is
+           equal to another event.
+      \item \texttt{subset()}, which tests whether this event is a
+           subset of another event.
+      \item \texttt{superset()}, which tests whether this event is
+           a superset of another event.
+      \item \texttt{union()}, which returns an event containing the
+           union of this event's samples and another event's samples.
+      \item \texttt{intersection()}, which returns an event
+           containing the intersection of this event's samples and
+           another event's samples.
+      \item \texttt{samples()}, which returns a \texttt{Set}
+           containing all of the samples that are contained by this
+           event. 
+      \item \texttt{\_\_len\_\_()}, which returns the number of samples 
+           contained by this event.
+    \end{itemize}
+
+    Classes implementing the \texttt{EventI} interface may choose
+    to only support certain classes of samples, or may choose to only
+    support certain types of events as arguments to the optional
+    methods (\texttt{\_\_cmp\_\_}, \texttt{subset}, etc.).  If a
+    method is unable to return a correct result because it is given an 
+    unsupported type of sample or event, it should raise a
+    NotImplementedError.
+
+\vspace{2mm}\noindent
+\textit{The \texttt{EventI} interface is currently implemented by the
+classes: \texttt{UniversalEvent}, \texttt{NullEvent},
+\texttt{SampleEvent}, \texttt{SetEvent}, \texttt{FuncEvent},
+\texttt{ContextEvent}.}
+
+\subsubsection{\texttt{FreqDistI} Interface}
+
+(still writing documentation)
+
+\subsubsection{\texttt{ProbDistI} Interface}
+
+(still implementing)
+
+\subsection{Tokenizers}
+
+The process of separating a string of text into Tokens is called
+\emph{tokenization}.  The tokenizer module defines the
+\texttt{TokenizerI} interface, which should be implemented by classes
+that tokenize text strings.
+
+\subsubsection{\texttt{TokenizerI} Interface}
+
+    A tokenizer is a processing class responsible for separating a
+    string of text into a list of \texttt{Token}s.  This process
+    is also known as \emph{tokenizing} the string of text.  Particular
+    \texttt{Tokenizer}s may split the text at different points, or
+    may produce Tokens with different types of \texttt{TokenType}.
+
+\vspace{2mm}\noindent
+\textit{The \texttt{TokenizerI} interface is currently implemented by
+the classes \texttt{SimpleTokenizer} and \texttt{TaggedTokenizer}.}
+
+\subsection{Taggers}
+
+The process of assigning a tag to each Token in a
+
+\subsubsection{\texttt{TaggerI} Interface}
 

### 2001-03-03 technical/proposal/proposal.tex
Moved the task-oriented module prototype table.  Minor wording change.
svn/trunk@11
Added "content" and "nomenclature" subsubsections to the module design subsection.  Gave URL for reference docs.  Minor doc changes.  Added docs for FreqDistI and ProbDistI.  Added docs for TaggerI.  Added bibliography items for some NLP books.
svn/trunk@13

@@
 
+\subsubsection{Content}
+
+The decision of what modules to include in the toolkit will be
+initially guided by the past content of CIS-530 \cite{cis530}.  Later,
+the content of several Natural Language Processing text books may be
+used to guide decisions about what modules to write \cite{allen1995},
+\cite{gazdar1989}, \cite{jurafsky2000}, \cite{manning2000}.
+
+\subsubsection{Nomenclature}
+
+When learning to use the toolkit, students intuitions about each
+entity in the toolkit will be guided by the entities names'.  It is
+therefore important to use consistant, correct, and intuitive
+nomenclature when naming the entities defined by the toolkit.
+The documentation should be careful to always use correct
+nomenclature; in particular, it should never conflate two concepts
+with the same name.
+
 \subsection{Programming Concepts}
@@
 Epydoc, see the reference documentation for the toolkit's prototype
-modules:
-
- \texttt{http://www.cis.upenn.edu/$\sim$edloper/??}
+modules 
+at ``\texttt{http://www.cis.upenn.edu/$\sim$edloper/nltk/nltk.html}''.
 
@@
 \ref{fig:mod-data} and \ref{fig:mod-task}, and described in the
-following sections.  Most of these summaries are copied directly from
+following sections.  Most of these descriptions are copied directly from
 the modules' reference documentation.  For more information, see the
-reference documentation at:
-
- \texttt{http://www.cis.upenn.edu/$\sim$edloper/??}
+reference documentation 
+at ``\texttt{http://www.cis.upenn.edu/$\sim$edloper/nltk/nltk.html}''.
 
@@
 
-\begin{figure}
-\noindent
-\begin{centering}
-\begin{boxedminipage}{.95\textwidth}
-\begin{itemize}
-  \item \textbf{Tokenizers}: Separate a string of text into a list of
-       Tokens. 
-  \begin{itemize}
-     \item \textit{TokenizerI}
-  \end{itemize}
-
-  \item \textbf{Taggers}: Assign tags to each Token in a list of Tokens.
-  \begin{itemize}
-     \item \textit{TaggerI}
-  \end{itemize}
-\end{itemize}
-\end{boxedminipage}\\
-\end{centering}
- \caption[Task-oriented module prototypes]{Task-oriented module
- prototypes.  Under each module, the interfaces and top-level classes
- defined by that module are listed.}
-\label{fig:mod-task}
-\end{figure}
-
 \subsection{Sets}
@@
 
-    An unordered container class that contains no duplicate elements.
-    In particular, a set contains no elements $e_1$ and $e_2$ such
-    that $e_1=e_2$.  Currently, the \texttt{Set} class is given a
-    fairly minimal implementation.  However, more members may be
-    defined in the future.
+    A \texttt{Set} is an unordered container class that contains no
+    duplicate elements.  In particular, a set contains no elements
+    $e_1$ and $e_2$ such that $e_1=e_2$.  Currently, the \texttt{Set}
+    class is given a fairly minimal implementation.  However, more
+    members may be defined in the future.
 
@@
 
-A Token is an occurance of a single unit of text, such as a word or a
+A \texttt{Token} is an occurance of a single unit of text, such as a word or a
 punctuation mark.  The token module defines the \texttt{TokenTypeI}
-interface, the \texttt{TextLocationI} interface, and the \texttt{Token
-class}.
+interface, the \texttt{TextLocationI} interface, and the \texttt{Token}
+class.
 
@@
 
-    Token types can be compared for equality, and should be equal if
-    and only if they refer to the same token type.
+    \texttt{TokenType}s can be compared for equality, and should be
+    equal if and only if they refer to the same token type.
 
@@
 
-\subsubsection{\texttt{Token} Class}
+\subsubsection{Token Class}
 
     A \texttt{Token} is an occurance of a single unit of text, such as
-    a word or a punctuation mark.  A Token consists of a token type
+    a word or a punctuation mark.  A \texttt{Token} consists of a token type
     and a source.  The token type is the unit of text (e.g., a
@@
 
+This module is still under development.  Several open questions
+remain, such as whether \texttt{SyntaxTree}s should correspond to
+individual occurances of syntax trees, or to ``types'' of syntax trees
+(c.f. the distinction between \texttt{Token} and TokenType).
+
 \subsubsection{\texttt{SyntaxNodeI} Interface}
@@
 \emph{not} perform any semantic or syntactic checks on its arguments.
-Figure \ref{fig:syntaxnode} shows how syntax nodes can be used:
+Figure \ref{fig:syntaxnode} shows how syntax nodes can be used.
 
@@
 
-    An event is a subset of some sample space.  Note that this subset
-    need not be finite.  Events are typically written as the set of
-    samples they contain, or as a function in first order logic.
-    Examples are:
+    An \texttt{Event} is a subset of some sample space.  Note that
+    this subset need not be finite.  Events are typically written as
+    the set of samples they contain, or using a function.  Examples
+    are:
 
@@
     unsupported type of sample or event, it should raise a
-    NotImplementedError.
+    \texttt{NotImplementedError}.
 
@@
 
-(still writing documentation)
+    A \texttt{FreqDist} represents the frequency distribution for the
+    outcomes of an experiment.  A frequency distribution records the
+    number of times each outcome of an experiment has occured.  For
+    example, a frequency distribution could be used to record the
+    frequency of each token in a document.  Formally, a frequency
+    distribution can be defined as a function mapping from samples to
+    the number of times that sample occured as an outcome.
+
+    Frequency distributions are generally constructed by running a
+    number of experiments, and incrementing the count for a sample
+    every time it is an outcome of an experiment.  For example, the
+    following code will produce a frequency distribution that encodes
+    how often each word occurs in a text:
+
+\begin{verbatim}
+freqDist = SimpleFreqDist()
+for word in document:
+    freqDist.inc(word)
+\end{verbatim}
+
+    Classes implementing the \texttt{FreqDistI} interface may
+    choose to only support certain classes of samples or events.  If a
+    method is unable to return a correct result because it is given an
+    unsupported type of sample or event, it should raise a
+    \texttt{NotImplementedError}.
+
+    Since several methods defined by \texttt{FreqDistI} can accept
+    either events or samples, classes that implement the EventI
+    interface should never be used as samples for a frequency
+    distribution.
+
+    Frequency distributions are required to implement the methods
+    \texttt{inc()}, \texttt{N()}, \texttt{count()},
+    \texttt{freq()}, \texttt{max()}, \texttt{cond\_freq()}, 
+    and \texttt{cond\_max()}.  In the future, this list may be
+    exapanded, and optional methods may be added.
+
+\vspace{2mm}\noindent
+\textit{The \texttt{FreqDistI} interface is currently implemented by the
+classes \texttt{SimpleFreqDistI} and \texttt{CFFreqDistI}.}
 
@@
 
-(still implementing)
+    A probability distribution for the outcomes of an experiment.  A
+    probability distribution specifies how likely it is that an
+    experiment will have any given outcome.  For example, a
+    probability distribution could be used to predict the probability
+    that a given word will appear in a given context.  Formally, a
+    probability distribution can be defined as a function mapping from
+    samples to nonnegative real numbers, such that the sum of every
+    number in the function's range is 1.0.  \texttt{ProbDist}s are
+    often used to model the probability distribution underlying a
+    frequency distribution.
+
+    Classes implementing the \texttt{ProbDistI} interface may
+    choose to only support certain classes of samples or events.  If a
+    method is unable to return a correct result because it is given an
+    unsupported type of sample or event, it should raise a
+    NotImplementedError.
+
+    Since several methods defined by \texttt{ProbDistI} can accept
+    either events or samples, classes that implement the EventI
+    interface should never be used as samples for a probability
+    distribution.
+
+    Probability distributions are required to implement the methods
+    \texttt{prob()} and \texttt{cond\_prob()},
+    \texttt{max()}, and \texttt{cond\_max()}.  In the future,
+    this list may be exapanded, and optional methods may be added.
+
+\vspace{2mm}\noindent
+\textit{The \texttt{ProbDistI} interface is not currently implemented
+any class.}
+
+\begin{figure}
+\noindent
+\begin{centering}
+\begin{boxedminipage}{.95\textwidth}
+\begin{itemize}
+  \item \textbf{Tokenizers}: Separate a string of text into a list of
+       \texttt{Token}s. 
+  \begin{itemize}
+     \item \textit{TokenizerI}
+  \end{itemize}
+
+  \item \textbf{Taggers}: Assign tags to each \texttt{Token} in a list of \texttt{Token}s.
+  \begin{itemize}
+     \item \textit{TaggerI}
+  \end{itemize}
+\end{itemize}
+\end{boxedminipage}\\
+\end{centering}
+ \caption[Task-oriented module prototypes]{Task-oriented module
+ prototypes.  Under each module, the interfaces and top-level classes
+ defined by that module are listed.}
+\label{fig:mod-task}
+\end{figure}
 
@@
 
-The process of separating a string of text into Tokens is called
+The process of separating a string of text into \texttt{Token}s is called
 \emph{tokenization}.  The tokenizer module defines the
@@
 
-    A tokenizer is a processing class responsible for separating a
-    string of text into a list of \texttt{Token}s.  This process
-    is also known as \emph{tokenizing} the string of text.  Particular
-    \texttt{Tokenizer}s may split the text at different points, or
-    may produce Tokens with different types of \texttt{TokenType}.
+    A \texttt{Tokenizer} is a processing class responsible for
+    separating a string of text into a list of \texttt{Token}s.  This
+    process is also known as \emph{tokenizing} the string of text.
+    Particular \texttt{Tokenizer}s may split the text at different
+    points, or may produce \texttt{Token}s with different types of
+    \texttt{TokenType}.
 
@@
 
-The process of assigning a tag to each Token in a
+The process of assigning a descriptive tag to each \texttt{Token} in a
+list of \texttt{Token}s is called \emph{tagging}.  Often, tagging is
+used to assign a part-of-speech tag to each \texttt{Token} in a list
+of \texttt{Token}s.  Particular \texttt{Taggers} may attatch different
+types of tags to the \texttt{Token}s in the list, or may use different
+algorithms for deciding which tag to attatch.
 
 \subsubsection{\texttt{TaggerI} Interface}
+
+    A \texttt{Tagger} is a processing class for assigning a tag to
+    each token in an ordered list of tokens.  Taggers are required to
+    define one function, \texttt{tag}, which takes a list of
+    \texttt{Token}s, and returns a list of \texttt{Token}s.
+    Typically, the input tokens will be \texttt{SimpleToken}s, and the
+    output tokens will be \texttt{TaggedToken}s.  However, taggers may
+    also be written to map between other types of tokens, as long as
+    they are still performing the same conceptual task.
+
+    Classes implementing the \texttt{TaggerI} interface may choose to
+    only support certain classes of tokens.  If a method is unable to
+    return a correct result because it is given an unsupported class
+    of token, then it should raise a \texttt{NotImplementedError}.
 
@@
 
-\bibitem{python}{The Python programming language. \\
-    \texttt{http://www.python.org/}}
+\bibitem{allen1995}
+{Allen, James.  1995.  
+ \emph{Natural Language Understanding}.
+ Redwood, CA: Benjamin Cummings.}
 
@@
 
+\bibitem{gazdar1989}
+{Gazdar, Gerald and Mellish, Chris.  1989.
+ \emph{Natural Language Processing in LISP : An Introduction to
+       Computational Linguistics}. 
+ Reading, MA: Addison-Wesley Pub. Co.}
+
+\bibitem{jurafsky2000}
+{Jurafsky, Daniel and Martin, James H.  2000.
+ \emph{Speech and Language Processing: An Introduction to Natural
+       Language Processing, Computational Linguistics, and Speech
+       Recognition}.
+ Upper Saddle River, NJ: Prentice Hall.}
+
+\bibitem{manning2000}
+{Manning, Christopher D. and Schutze, Hinrich.  2000.  
+ \emph{Foundations of Statistical Natural Language Processing}.
+ Cambridge, MA: MIT Press.}
+
+\bibitem{python}{The Python programming language. \\
+    \texttt{http://www.python.org/}}
+
 \end{thebibliography}

### 2001-03-03 technical/proposal/proposal.tex
Spell-checked.
svn/trunk@16

@@
 CIS-530 is an introductory course in natural language processing
-(NLP), which is intended to be accessable to both linguists and
+(NLP), which is intended to be accessible to both linguists and
 computer scientists \cite{cis530}. In such a class, it is instructive
@@
 programming or designing software systems.  Thus, when students
-attempt to build NLP systems, they may spend a signifigant amount of
+attempt to build NLP systems, they may spend a significant amount of
 their time on programming tasks that are not directly related to the
@@
   \item It will provide the students with an infrastructure for
-  building natural language systems, by providing consistant and
+  building natural language systems, by providing consistent and
   well-defined interfaces for different components.  Thus, students
@@
 
-\item[Consistancy] The toolkit should use consistant data structures
+\item[Consistency] The toolkit should use consistent data structures
 and interfaces.
 
-\item[Extensibility] The toolkit should easily accomodate new
+\item[Extensibility] The toolkit should easily accommodate new
 components, whether those components replicate or extend the toolkit's
@@
 \item[Documentation] The toolkit, its data structures, and its
-implementation all need to be carefully and thouroughly documented.
+implementation all need to be carefully and thoroughly documented.
 
@@
 toolkit should be kept to a minimum.  In particular, it should be
-possible to use complete individaul projects using small parts of the
+possible to use complete individual projects using small parts of the
 toolkit, without worrying about how they interact with the rest of the
@@
 
-The natural langauge processing toolkit will be implemented by a
-collection of independant modules\footnote{These modules do not
+The natural language processing toolkit will be implemented by a
+collection of independent modules\footnote{These modules do not
 necessarily correspond to Python \texttt{module}s.}.  These modules are
@@
 
-\subsubsection{Dependancies}
+\subsubsection{Dependencies}
 
 In general, data-oriented modules should be self-contained, and should
-exhibit very little inter-dependancy.  Data-oriented modules never
+exhibit very little inter-dependency.  Data-oriented modules never
 depend on task-oriented modules.
@@
 Task-oriented modules should also be relatively self-contained, with
-very little inter-dependancy.  However, each task-oriented modules
+very little inter-dependency.  However, each task-oriented modules
 will usually depend on a variety of data-oriented modules.  When
@@
 \begin{tabular}{||p{.21\textwidth}p{.7\textwidth}}
-  \textbf{Token} & An Occurance of a \texttt{TokenType}
+  \textbf{Token} & An occurance of a \texttt{TokenType}
   within a text.  A \texttt{token} consists of a \texttt{TokenType}
@@
 entity in the toolkit will be guided by the entities names'.  It is
-therefore important to use consistant, correct, and intuitive
+therefore important to use consistent, correct, and intuitive
 nomenclature when naming the entities defined by the toolkit.
@@
 
-Type checking is an imporant safety device, that allows programmers to
+Type checking is an important safety device, that allows programmers to
 quickly detect the locations of errors in their code.  This is
@@
   \item[Mapping and Filtering] Mapping and filtering provide a
-  consise way of expressing transformations on lists.  However, they
+  concise way of expressing transformations on lists.  However, they
   are fairly difficult to learn, and generally involve the use of
@@
 postscript files.  Tutorials and reference documentation will be
-available as HTML files, accessble over the web.
+available as HTML files, accessible over the web.
 
@@
 students can consult these reports if they would like further
-information about how the tookit is designed and why it is designed
+information about how the toolkit is designed and why it is designed
 that way.
@@
 class, method, function, and variable defined by the toolkit.  These
-definitions should be both precise and consise.  I currently plan to
+definitions should be both precise and concise.  I currently plan to
 produce the reference documentation using Epydoc \cite{epydoc}.
@@
   \item A page for each python \texttt{module}.  This page includes an
-  overview of the \texttt{module}; a list of the \texttt{classs}es
+  overview of the \texttt{module}; a list of the \texttt{class}es
   defined by the \texttt{module}; a list of the \texttt{function}s
@@
   \item A page for each python \texttt{class}.  This page includes an
-  overview of the \texttt{classs}; a list of the \texttt{method}s
-  defined by the \texttt{classs}, and a detailed description of each
+  overview of the \texttt{class}; a list of the \texttt{method}s
+  defined by the \texttt{class}, and a detailed description of each
   \texttt{method}; a list of the instance \texttt{variable}s defined
-  by the \texttt{classs}, and a detailed description of each instance
-  \texttt{variable}; a list of the \texttt{classs} \texttt{variable}s
-  defined by the \texttt{classs}, and a detailed description of each
-  \texttt{classs} \texttt{variable}.
+  by the \texttt{class}, and a detailed description of each instance
+  \texttt{variable}; a list of the \texttt{class} \texttt{variable}s
+  defined by the \texttt{class}, and a detailed description of each
+  \texttt{class} \texttt{variable}.
 
-  \item A page containing the toolit's complete \texttt{classs}
+  \item A page containing the toolkit's complete \texttt{class}
   hierarchy.
@@
   \item A page containing an index of every \texttt{module},
-  \texttt{classs}, \texttt{method}, \texttt{function}, and
+  \texttt{class}, \texttt{method}, \texttt{function}, and
   \texttt{variable} defined by the toolkit.
@@
 
-As part of the preliminary design process, I desinged and implemented
+As part of the preliminary design process, I designed and implemented
 six prototype modules.  These modules are summarized in figures
@@
 \begin{itemize}
-  \item \textbf{Sets}: Encodes the mathmatical notion of a ``finite set.''
+  \item \textbf{Sets}: Encodes the mathematical notion of a ``finite set.''
   \begin{itemize}
@@
   \item \textbf{Probability}: Encodes data structures associated with
-        the mathmatical notion of probability, such as events and
+        the mathematical notion of probability, such as events and
         frequency distributions.
   \begin{itemize}
-    \item Sample: Encodes the mathmatical notion of a
+    \item Sample: Encodes the mathematical notion of a
           ``sample.''  This is actually not implemented as a class or
@@
            well-defined comparison functions.  An example where it
-           would not be the case would be if ei defined
-           \texttt{\_\_cmp\_\_()} to always return 0, and ej
+           would not be the case would be if $e_i$ defined
+           \texttt{\_\_cmp\_\_()} to always return 0, and $e_j$
            defined \texttt{\_\_cmp\_\_()} to always return $-1$.
@@
     A \texttt{TokenTypeI} is single unit of text, such as a word or a
-    punctuation mark.  The precise defintion of what counts as a token
+    punctuation mark.  The precise definition of what counts as a token
     type will vary for different analyses.  For example, ``bank'' and
@@
     specific word).  The source is the position at which this token
-    occured in the text.
+    occurred in the text.
 
-    The precise defintion of what counts as a token type, or unit of
+    The precise definition of what counts as a token type, or unit of
     text, will vary for different analyses.  For example, ``bank'' and 
@@
 The probability module provides data structures associated with the
-mathmatical notion of probability.  The following mathmatical terms
+mathematical notion of probability.  The following mathematical terms
 are useful for thinking about probability:
@@
 The toolkit currently defines three interfaces for handling the
-mathmatical notion of probability: the \texttt{EventI} interface; the
+mathematical notion of probability: the \texttt{EventI} interface; the
 \texttt{FreqDistI} interface; and the \texttt{ProbDistI} interface.
@@
     outcomes of an experiment.  A frequency distribution records the
-    number of times each outcome of an experiment has occured.  For
+    number of times each outcome of an experiment has occurred.  For
     example, a frequency distribution could be used to record the
@@
     distribution can be defined as a function mapping from samples to
-    the number of times that sample occured as an outcome.
+    the number of times that sample occurred as an outcome.
 
@@
     and \texttt{cond\_max()}.  In the future, this list may be
-    exapanded, and optional methods may be added.
+    expanded, and optional methods may be added.
 
@@
     \texttt{max()}, and \texttt{cond\_max()}.  In the future,
-    this list may be exapanded, and optional methods may be added.
+    this list may be expanded, and optional methods may be added.
 
@@
 used to assign a part-of-speech tag to each \texttt{Token} in a list
-of \texttt{Token}s.  Particular \texttt{Taggers} may attatch different
+of \texttt{Token}s.  Particular \texttt{Taggers} may attach different
 types of tags to the \texttt{Token}s in the list, or may use different
-algorithms for deciding which tag to attatch.
+algorithms for deciding which tag to attach.
 
@@
 %################################################################
-%#  BiIBLIOGRAPHY
+%#  BIBLIOGRAPHY
 %################################################################

### 2001-05-11 misc/concepts.tex
A short description of some open issues on representation
svn/trunk@22

%
% Natural Language Toolkit for Python
% Catalog of Concepts/Datatypes
% Edward Loper
% 
% Created [03/02/01 12:02 AM]
% $Id$
%

\documentclass{article}
\usepackage{fullpage}

\begin{document}
\title{Notes on the NL Toolkit}
\author{Edward Loper}
\maketitle

This document catalogs some of the different conceptual entities which
the tooklit should deal with.  These conceptual entities sometimes
draw more distinctions than is convenient when actually working with
objects.

An open question is how strict the toolkit should be about enforcing
the distinctions between conceptually distinct types of entities.  On
one hand, strictly enforcing these distinctions is conceptually
``simpler,'' and may promote clearer thinking.  On the other hand, it
is often cumbersome to make these distinctions -- in many
circumstances, they don't matter, so they get dropped in practice.
For example, most people will not distinguish ``tokens'' from ``token
types'' unless they need to.  In a sense, this question boils down to
whether it will be easier to learn/use a set of rules/conventions that
are simpler, but somewhat less intuitive; or more complex, but more
intuitive.  I think that one danger is that the more complex rules
will only be more intuitive some of the time.  And it may be dangerous
to be inconsistantly intuitive..

Another open question is whether the current names should be kept.  In
particular, names of conceptual entities should be both \emph{correct}
and \emph{easy to use/remember}.

Also, it may be possible that some of these conceptual entities could
be merged or done away with altogther.  Are all of these entities
serving useful purposes?

\section{Token Stuff}

{\it (Currently, many of these definitions
are a bit hand-wavey.  That partially stems from the fact that they
are not really as well-defined as we would like (esp the more
primitive ones).  But I'll try to clean them up eventually.)}

\begin{itemize}

  \item \textbf{Lexeme:} An identifier for a word-like thing.
  Normally, this would be a string.  But if you wanted to process text
  more efficiently, it could be an integer.  It might be stemmed, not
  stemmed, case-normalized, not case-normalized, etc.

  \item \textbf{Token Type:} A word-like unit of text.  Includes a
  lexeme, and may also include other info, like part of speech tag,
  etc.

  \item \textbf{Token:} An occurance of a token type

  \item \textbf{Location:} The location of something in text.  There
  are actually a number of ways of representing locations: as indexes,
  as spans, etc.  Also, indexes can refer to word number, character
  number, etc.

\end{itemize}

\section{Syntax Tree Stuff}

\begin{itemize}

  \item \textbf{SyntaxNode:} The annotation that appears at a node
  in a syntax tree.  E.g., this might hold a phrase type (``NP'' or
  ``V'').

  \item \textbf{SyntaxTree:} An occurance of a tree or subtree.
  Consists of a SyntaxNode and an ordered list of zero or more
  children.  Children must be SyntaxTrees or Tokens.

  \item \textbf{SyntaxTreeType:} The ``type'' of a syntax tree (as
  opposed to an occurance of a syntax tree).  Consists of a SyntaxNode
  and an ordered list of zero or more children.  Children must be
  SyntaxTreeTypes or TokenTypes.

\end{itemize}

\section{How much to enforce?}

So basically, the question of deciding how strictly to enforce these
distinctions comes down to two questions:

\begin{enumerate}
  \item Should any entities/distinctions be eliminated?
  \item Should it be possible to use entities in environments that
        are not \emph{strictly speaking} correct?
\end{enumerate}

I'll leave the first question for later, and concentrate on the
second.  There are a number of ways that objects could try to be more
friendly when you use them in ways that that are not strictly speaking
correct:

\begin{itemize}
  \item Equality tests: What happens when you compare a string to a
  token?  A string to a TokenType?  A Token to a TokenType?  etc.?  A
  related question is how to handle comparisons within one class
  hierarchy.  E.g., equality tests between a tagged token and a simple
  token.  The strict view would say that any of these comparisons
  should raise an exception.  However, if we decide not to be strict,
  there are a number of places where these equality tests could be
  well-defined.  Examples that come to mind are:
  \begin{itemize}
    \item equality between a string and a SimpleTokenType
    \item equality between a string and a Token (ignore location)
    \item equality between an integer and a Location
    \item equality between a list and a SyntaxTree(Type)
  \end{itemize}

  \item Arguments to methods: Should methods attempt to handle
  arguments that are not quite the expected type?  E.g., when Steven
  passed strings to a FreqDist, when it was really (sort of) expecting
  TokenTypes.

  \item Arguments to constructors: Should constructors automatically
  ``wrap'' objects if they're not the right type?  E.g., should
  Token(``cat'') automatically create a SimpleTokenType containing
  ``cat''?
\end{itemize}

\section{Printing}

One thing that guides peoples intuitions about what an object is is
how it looks when it prints.  If two things look the same when you
print them, then intuitively they should be the same.  I think this
may have been a source of confusion in the first problem set, where
TokenTypes look just like strings when you print them, so users will
be tempted to do things like use ``=='' to test whether a TokenType
is equal to a string.

\newpage
\section{Implementation details}

\begin{itemize}
  \item \textbf{Lexeme}: normally an internal Python type, like string
  or int.

  \item \textbf{Token Type}: a class with TokenTypeI as a base.

  \item \textbf{Token}: the class Token

  \item \textbf{Location}: a class with LocationI as a base.

  \item \textbf{SyntaxNode}: a class with SyntaxNodeI as a base.

  \item \textbf{SyntaxTree}: the class SyntaxTree

  \item \textbf{SyntaxTreeType}: the class SyntaxTreeType

\end{itemize}

\noindent
Accessing the lexeme of a token:\\
\begin{tabular}{|l|ccccc|}
\hline
Code    &my\_token&. & type()&. & lexeme() \\
\hline
Entity  &Token && Token Type && Lexeme \\
Type    &\tt Token && \tt TokenTypeI && \tt string \\
\hline
\end{tabular}

\vspace{2mm}\noindent
Accessing the location of a token:\\
\begin{tabular}{|l|ccccc|}
\hline
Code    &my\_token&. & source()&. & index() \\
\hline
Entity  &Token && Location && Index \\
Type    &\tt Token && \tt LocationI && \tt int \\
\hline
\end{tabular}\\
(should that be .loc() instead of .source()?)

\vspace{2mm}\noindent
Accessing the tag of a syntax tree:\\
\begin{tabular}{|l|ccccc|}
\hline
Code    &my\_stree&. & node()&. & tag() \\
\hline
Entity  &SyntaxTree && SyntaxNode && Tag \\
Type    &\tt SyntaxTree && \tt SyntaxNodeI && \tt string \\
\hline
\end{tabular}

\vspace{2mm}\noindent
Accessing the children of a syntax tree:\\
\begin{tabular}{|l|cc|}
\hline
Code    &my\_stree& [2] \\
\hline
Entity  &SyntaxTree & SyntaxTree or Token \\
Type    &\tt SyntaxTree & {\tt SyntaxTree} or {\tt Token}\\
\hline
\end{tabular}

\end{document}

### 2001-05-11 technical/reptypes/reptypes.tex
A technical report describing the basic representation types used by the toolkit, and why they were chosen.
svn/trunk@26

%
% Natural Language Toolkit for Python
% Technical Report: Representation Types
% Edward Loper
% 
% Created [03/02/01 12:02 AM]
% $Id$
%

\newcommand{\concept}[1]{\textsf{#1}}
\newcommand{\code}[1]{\texttt{#1}}

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}

\begin{document}
\title{Basic Representation Types\\
\Large Natural Language Processing Toolkit: Technical Report}
\author{Edward Loper}
\maketitle

\section{Introduction}

  The natural language toolkit allows students to manipulate a number
  of conceptual objects, such as words, sentences, and syntax trees.
  Before we can build a consistant framework for manipulating such
  objects, we must decide how they should be represented.  This
  technical report explains basic types used by the toolkit to
  represent a number of conceptual objects.  Furthermore, it attempts
  to justify why this set of basic types was chosen.

  Section \ref{sec:concepts} describes the basic concepts which this
  technical report attempts to address.  Sections \ref{sec:locations}
  through \ref{sec:trees} describe the basic types used to encapsulate
  these concepts.  Section \ref{sec:summary} provides a summary of the
  basic types, and discusses how they interact.

\newpage
\section{Abstract Concepts}
\label{sec:concepts}

  The following list of abstract concepts is representative of the
  concepts which should be captured by the basic types defined in this
  techincal report.\footnote{The types used to represent many other
  abstract concepts (such as frequency distributions) are beyond the
  scope of this technical report.}  Where the meaning of the concepts
  are not obvious, I included explanations.

  \begin{itemize}

    \item \concept{Document}

    \item \concept{Sentence}

    \item \concept{Word Type}: A single word (as opposed to a
        \concept{Word Occurance}).

    \item \concept{Tagged Word Type}: A word that has been tagged with
        additional information, such as part-of-speech information.

    \item \concept{Stemmed Word Type}: A word that has been stripped of
        inflectional suffixes and prefixes.

    \item \concept{Case-Normalized Word Type}: A word whose case has been
        normalized (usually to either upper case or lower case).

    \item \concept{Word Occurance}: A single occurance of a word type.

    \item \concept{Tagged Word Occurance}
    \item \concept{Stemmed Word Occurance}
    \item \concept{Case-Normalized Word Occurance}

    \item \concept{Tokenized Document}: A document that has been divided
        into individual word occurances.

    \item \concept{Tokenized Sentence}: A sentence that has been divided
        into individual word occurances.

    \item \concept{Location}: The location of one unit of text within
        another unit of text.  A typical location might give the
        location of a word within a document or a sentence.

    \item \concept{Tagged Document} or \concept{Tagged Sentence}: A
        document or sentence that has been divided into individual word
        occurances, each of which has been tagged with additional
        information.

    \item \concept{Stemmed Document} or \concept{Stemmed Sentence}: A
        document or sentence that has been divided into individual word
        occurances, each of which has been stripped of inflectional
        affixes.

    \item \concept{Case-Normalized Document} or \concept{Case-Normalized
        Sentence}: A document or sentence that has been divided into
        individual word occurances, each of which has had its case
        normalized.

    \item \concept{Syntax Tree}: A hierarchical representation of the
        structure of a sentence.

    \item \concept{Syntax Node}: The annotation appearing at a node in a
        syntax tree.  Typically, this will be a tag, such as ``NP.''

    \item \concept{Syntax Tree Occurance}: A single occurance of a syntax
        tree.

  \end{itemize}

\section{Locations}
\label{sec:locations}
%  - location

  \subsection{Design}

  The location of an element of text is represented using a span-based
  data structure.  A \emph{span} consists of a \emph{start index} and
  an \emph{end index}, both of which are integers.  A span with start
  index $s$ and end index $e$ is written ``$[s,e)$,'' and specifies
  the location of the text beginning at $s$, and including everything
  up to (but not including) the text at $e$.  It is always true that
  $s<e$.\footnote{Should this be changed to $s\leq e$?}

  The start and end indices can be based on a variety of different
  \emph{units}, such as character number, word number, or sentence
  number.  However, for any two indeces $x$ and $y$, it should be true
  that $x<y$ if and only if $x$ indicates a location earlier in the
  text than $y$.  The location of an element of text may be explicitly
  tagged with information about what unit its indices use.  Unit
  labels take the form of case-insensitive \code{string}s.  Typical
  examples of unit labels are the strings \code{"character"} and
  \code{"word"}.

  The location of an element of text may also be tagged with its
  \emph{source}.  This is an object that gives an indication of where
  the text was derived from.  A typical example of a source would be a
  \code{string} containing the name of the file from which the element
  of text was read.  Sources may also be represented as locations; for
  example, the source for the location of a word might be the location
  of the sentence that contained it.  This allows locations to be
  specified in a hierarchical fashion.

  \subsection{Implementation}

  Locations are implemented using the \code{Location} class.
  \code{Location}s consist of:

  \begin{itemize}

    \item A \emph{start index} (accessed via the \code{start()} member
    function).  \\Type: \code{int}

    \item An \emph{end index} (accessed via the \code{end()} member
    function).  \\Type: \code{int}

    \item An optional \emph{unit} (accessed via the \code{unit()}
    member function).  \\Type: \code{string} \emph{(normalized to
    lower case)}

    \item An optional \emph{source} (accessed via the \code{source()}
    member function).  \\Type: \emph{(any)}

  \end{itemize}

  \noindent
  Locations are immutable objects.

  Locations can be compared for equality and ordering.  If two
  locations with different units or different sources are compared,
  then an exception will be raised.  When location comparisons do not
  raise exceptions, they are defined as follows:
  \begin{align*}
     [s_1,e_1) = [s_2,e_2) &\iff s_1=s_2 \quad \land \quad e_1=e_2 \\
     [s_1,e_1) < [s_2,e_2) &\iff e_1 < s_2 \\
     [s_1,e_1) > [s_2,e_2) &\iff s_1 > e_2 \\
     [s_1,e_1) \leq [s_2,e_2) &\iff [s_1,e_1) < [s_2,e_2) \quad \lor \quad
                             [s_1,e_1) = [s_2,e_2) \\
     [s_1,e_1) \geq [s_2,e_2) &\iff [s_1,e_1) > [s_2,e_2) \quad \lor \quad
                             [s_1,e_1) = [s_2,e_2) 
  \end{align*}

  \noindent
  Note that if two locations $l_1$ and $l_2$ overlap, then it will
  never be the case that $l_1<l_2$ or $l_1>l_2$.

  \subsection{Explanation and Comments}

  Originally, I proposed a general, user-extensible notion of
  location.  However, the span-based representation described here
  seems to be sufficiently powerful to represent almost any location.
  We decided that the additional power granted by making the notion of
  location user-extensible did not compensate for the added
  complexity.  Also, if the notion of location built-in, then other
  classes can rely on it for computation.  For example, syntax trees
  can compute their locations as a function of the locations of their
  leaves.  If locations were user-extensible objects, then there would
  be no general way to perform such computations.

  A simpler notion of location, based on a single index, was
  considered.  However, it was decided that the added power provided
  by representing locations as ranges was worthwhile, especially in
  the context of multi-word structures like syntax trees.

  The optional unit and source fields were added to help distinguish
  different sets of locations, and to help ensure that unrelated
  locations are not accidentally compared.  For example, it may be
  useful for a student to keep track of the location of the words in
  two different files; but comparing these locations would be
  meaningless.  We considered simply returning false when locations
  with non-matching units or sources are compared, but decided that
  raising exceptions would be more likely to help students find
  errors.  If students wish to compare two locations that may have
  different sources or units, then they must explicitly compare the
  location's sources or units, or catch and process the exceptions
  raised by the comparison methods.

  We decided to restrict location indices to integers because it
  simplifies the interface without signifigantly reducing its power.
  Other alternatives we considered were floating point numbers and
  arbitrary comperable objects.  Floating point numbers were rejected
  because rounding errors can invalidate tests for equality and
  ordering.  If a floating-point index seems appropriate, it is
  recommended that users either multiply the floating point numbers by
  a fixed constant and round them to integers, or create a
  bidirectional mapping between floating-point indices and integer
  indices.  The use of arbitrary comperable objects as indices was
  rejected because it signifigantly complicates the location
  interfaces, without adding signifigant power to the system.  Also,
  arbitrary comperable objects may incur some of the difficulties that
  would be encountered with floating point numbers.

\section{Words}
\label{sec:words}
%  - (type)
%  - token

\section{Collections of Words}
\label{sec:collections}
%  - string (sentence/document)
%  - list of token

\section{Syntax Trees}
\label{sec:trees}
%  - (node)
%  - tree (type of token)
%      - (node pytype)
%      - (leaf pytype)
%  - treetype

\section{Summary}
\label{sec:summary}
% Dependancies:
%     token \to location
%     tree \to token
%     tree \to location
%     token \to (type)
%     tree \to (type)
%     tree \to (node)

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The abstract concepts which

This document catalogs some of the different conceptual entities which
the tooklit should deal with.  These conceptual entities sometimes
draw more distinctions than is convenient when actually working with
objects.

An open question is how strict the toolkit should be about enforcing
the distinctions between conceptually distinct types of entities.  On
one hand, strictly enforcing these distinctions is conceptually
``simpler,'' and may promote clearer thinking.  On the other hand, it
is often cumbersome to make these distinctions -- in many
circumstances, they don't matter, so they get dropped in practice.
For example, most people will not distinguish ``tokens'' from ``token
types'' unless they need to.  In a sense, this question boils down to
whether it will be easier to learn/use a set of rules/conventions that
are simpler, but somewhat less intuitive; or more complex, but more
intuitive.  I think that one danger is that the more complex rules
will only be more intuitive some of the time.  And it may be dangerous
to be inconsistantly intuitive..

Another open question is whether the current names should be kept.  In
particular, names of conceptual entities should be both \emph{correct}
and \emph{easy to use/remember}.

Also, it may be possible that some of these conceptual entities could
be merged or done away with altogther.  Are all of these entities
serving useful purposes?

\section{Token Stuff}

{\it (Currently, many of these definitions
are a bit hand-wavey.  That partially stems from the fact that they
are not really as well-defined as we would like (esp the more
primitive ones).  But I'll try to clean them up eventually.)}

\begin{itemize}

  \item \textbf{Lexeme:} An identifier for a word-like thing.
  Normally, this would be a string.  But if you wanted to process text
  more efficiently, it could be an integer.  It might be stemmed, not
  stemmed, case-normalized, not case-normalized, etc.

  \item \textbf{Token Type:} A word-like unit of text.  Includes a
  lexeme, and may also include other info, like part of speech tag,
  etc.

  \item \textbf{Token:} An occurance of a token type

  \item \textbf{Location:} The location of something in text.  There
  are actually a number of ways of representing locations: as indexes,
  as spans, etc.  Also, indexes can refer to word number, character
  number, etc.

\end{itemize}

\section{Syntax Tree Stuff}

\begin{itemize}

  \item \textbf{SyntaxNode:} The annotation that appears at a node
  in a syntax tree.  E.g., this might hold a phrase type (``NP'' or
  ``V'').

  \item \textbf{SyntaxTree:} An occurance of a tree or subtree.
  Consists of a SyntaxNode and an ordered list of zero or more
  children.  Children must be SyntaxTrees or Tokens.

  \item \textbf{SyntaxTreeType:} The ``type'' of a syntax tree (as
  opposed to an occurance of a syntax tree).  Consists of a SyntaxNode
  and an ordered list of zero or more children.  Children must be
  SyntaxTreeTypes or TokenTypes.

\end{itemize}

\section{How much to enforce?}

So basically, the question of deciding how strictly to enforce these
distinctions comes down to two questions:

\begin{enumerate}
  \item Should any entities/distinctions be eliminated?
  \item Should it be possible to use entities in environments that
        are not \emph{strictly speaking} correct?
\end{enumerate}

I'll leave the first question for later, and concentrate on the
second.  There are a number of ways that objects could try to be more
friendly when you use them in ways that that are not strictly speaking
correct:

\begin{itemize}
  \item Equality tests: What happens when you compare a string to a
  token?  A string to a TokenType?  A Token to a TokenType?  etc.?  A
  related question is how to handle comparisons within one class
  hierarchy.  E.g., equality tests between a tagged token and a simple
  token.  The strict view would say that any of these comparisons
  should raise an exception.  However, if we decide not to be strict,
  there are a number of places where these equality tests could be
  well-defined.  Examples that come to mind are:
  \begin{itemize}
    \item equality between a string and a SimpleTokenType
    \item equality between a string and a Token (ignore location)
    \item equality between an integer and a Location
    \item equality between a list and a SyntaxTree(Type)
  \end{itemize}

  \item Arguments to methods: Should methods attempt to handle
  arguments that are not quite the expected type?  E.g., when Steven
  passed strings to a FreqDist, when it was really (sort of) expecting
  TokenTypes.

  \item Arguments to constructors: Should constructors automatically
  ``wrap'' objects if they're not the right type?  E.g., should
  Token(``cat'') automatically create a SimpleTokenType containing
  ``cat''?
\end{itemize}

### 2001-05-11 technical/proposal/proposal.tex
minor edits; added a section on "Statistical Processing and Visualization"
svn/trunk@28

@@
 their time on programming tasks that are not directly related to the
-course of the content.
+course content.
 
@@
 
-\item[Efficiency] The toolkit doe not need to be highly optimized for
+\item[Efficiency] The toolkit does not need to be highly optimized for
 runtime performance.  However, it should be efficient enough that
@@
 
-Each module define a small set of interfaces, which give
+Each module defines a small set of interfaces, which give
 specifications for the types of classes that can be implemented by the
@@
 \begin{tabular}{||p{.21\textwidth}p{.7\textwidth}}
-  \textbf{SimpleTokenType} & A \texttt{TokenType}s that is
+  \textbf{SimpleTokenType} & A \texttt{TokenType} that is
   represented by a single Python \texttt{string}.  Examples of
@@
 When learning to use the toolkit, students intuitions about each
-entity in the toolkit will be guided by the entities names'.  It is
+entity in the toolkit will be guided by the entities names.  It is
 therefore important to use consistent, correct, and intuitive
@@
 
+\subsection{Statistical Processing and Visualization}
+
+The Python standard library contains minimal support for statistical
+processing and visualization.  A number of third-party Python
+libraries provide basic statistical and visualization tools
+\cite{numpy, statspy, scipy, vtk}.  These tools could be used, either
+by the toolkit, or directly by students, to perform statistical
+processing and to visualize the results.
+
+However, most of these tools have limited capabilities, and may not be
+flexible enough for use with the toolkit.  Another option is to use a
+well-established system for statistical processing and visualization,
+that can be accessed indirectly via Python.  One such tool is the R
+language \cite{r}.  R is an open-source version of the S system for
+statistical processing and visualization \cite{s}.  This system
+provides the advanced statistical tools and well-designed
+visualization systems.  The R language can be accesed from Python via
+the R/S Python interface \cite{rspython}.
+
+The decision of what third-party modules and tools to be used by the
+toolkit will be made after a more thourough investigation of the
+advantages and distadvantages of respective tools.  Furthermore,
+third-party modules will only be incorperated into the toolkit once
+the toolkit develops enough to warrent their use.  However, tools such
+as R seem like very promising candidates for adding advanced features
+to the toolkit.
+
 %################################################################
@@
 Epydoc, see the reference documentation for the toolkit's prototype
-modules 
-at ``\texttt{http://www.cis.upenn.edu/$\sim$edloper/nltk/nltk.html}''.
+modules at 
+``\texttt{http://www.cis.upenn.edu/$\sim$edloper/nltk/prototype/nltk.html}''.
 
@@
 
+\bibitem{numpy}{Numerical Python \\
+    \texttt{http://numpy.sourceforge.net/}}
+
 \bibitem{python}{The Python programming language. \\
     \texttt{http://www.python.org/}}
+
+\bibitem{r}{The R Project for Statistical Computing \\
+    \texttt{http://www.r-project.org/}}
+
+\bibitem{rspython}{The R/S-Plus Python Interface \\
+    \texttt{http://www.omegahat.org/RSPython/}}
+
+\bibitem{s}{The S System \\
+    \texttt{http://cm.bell-labs.com/cm/ms/departments/sia/S/}}
+
+\bibitem{scipy}{Scientific Python \\
+    \texttt{http://starship.python.net/crew/hinsen/scientific.html}}
+
+\bibitem{statspy}{stats.py: A Python Statistical Module \\
+    \texttt{http://www.vex.net/parnassus/apyllo.py/684222876.912038798}}
+
+\bibitem{vtk}{VTK: The Visualization Toolkit \\
+    \texttt{http://www.kitware.com/vtk.html}}
 

### 2001-05-12 technical/reptypes/reptypes.tex
Added lots of content.. a paragraph about locations, wrote the section on words, most of the section on collectino of words..
svn/trunk@30

@@
 
+% Note: be careful to say ``x is a kind of y'' instead of ``x is a type 
+% of y,'' since we're giving ``type'' special meaning.
+
 \newcommand{\concept}[1]{\textsf{#1}}
@@
 
-\documentclass{article}
+\documentclass[11pt]{article}
 \usepackage{fullpage}
@@
   technical report attempts to address.  Sections \ref{sec:locations}
-  through \ref{sec:trees} describe the basic types used to encapsulate
-  these concepts.  Section \ref{sec:summary} provides a summary of the
-  basic types, and discusses how they interact.
+  through \ref{sec:trees} describe the basic types that were chosen to
+  encapsulate these concepts, and the rational behind choosing those
+  types.  Section \ref{sec:summary} provides a summary of the basic
+  types, and discusses how they interact.
 
@@
   data structure.  A \emph{span} consists of a \emph{start index} and
-  an \emph{end index}, both of which are integers.  A span with start
-  index $s$ and end index $e$ is written ``$[s,e)$,'' and specifies
-  the location of the text beginning at $s$, and including everything
-  up to (but not including) the text at $e$.  It is always true that
-  $s<e$.\footnote{Should this be changed to $s\leq e$?}
+  an \emph{end index}, both of which are non-negative integers.  A
+  span with start index $s$ and end index $e$ is written ``$[s,e)$,''
+  and specifies the location of the text beginning at $s$, and
+  including everything up to (but not including) the text at $e$.  It
+  is always true that $s<e$.\footnote{Should this be changed to $s\leq
+  e$?  If $s=e$, then the span would be empty -- does it make sense to
+  have empty locations?}
 
@@
     \item A \emph{start index} (accessed via the \code{start()} member
-    function).  \\Type: \code{int}
+    function).  \\
+    Type: \code{int}
 
     \item An \emph{end index} (accessed via the \code{end()} member
-    function).  \\Type: \code{int}
+    function).  \\
+    Type: \code{int}
 
     \item An optional \emph{unit} (accessed via the \code{unit()}
-    member function).  \\Type: \code{string} \emph{(normalized to
-    lower case)}
+    member function).  \\
+    Type: \code{string} \emph{(normalized to lower case)}
 
     \item An optional \emph{source} (accessed via the \code{source()}
-    member function).  \\Type: \emph{(any)}
+    member function).  \\
+    Type: \emph{(any)}
 
@@
   \noindent
-  Locations are immutable objects.
-
-  Locations can be compared for equality and ordering.  If two
-  locations with different units or different sources are compared,
-  then an exception will be raised.  When location comparisons do not
-  raise exceptions, they are defined as follows:
+  \code{Location}s are immutable objects.  \code{Location}s can be
+  compared for equality and ordering.  If two \code{Location}s with
+  different units or different sources are compared, then an exception
+  will be raised.  When \code{Location} comparisons do not raise
+  exceptions, they are defined as follows:
   \begin{align*}
@@
   \noindent
-  Note that if two locations $l_1$ and $l_2$ overlap, then it will
-  never be the case that $l_1<l_2$ or $l_1>l_2$.
+  Note that if two \code{Location}s $l_1$ and $l_2$ overlap, then it
+  will never be the case that $l_1<l_2$ or $l_1>l_2$.
 
-  \subsection{Explanation and Comments}
+  \subsection{Discussion}
 
@@
 
-  We decided to restrict location indices to integers because it
-  simplifies the interface without signifigantly reducing its power.
-  Other alternatives we considered were floating point numbers and
-  arbitrary comperable objects.  Floating point numbers were rejected
-  because rounding errors can invalidate tests for equality and
-  ordering.  If a floating-point index seems appropriate, it is
-  recommended that users either multiply the floating point numbers by
-  a fixed constant and round them to integers, or create a
+  We decided to restrict location indices to non-negative integers
+  because it simplifies the interface without signifigantly reducing
+  its power.  Other alternatives we considered were floating point
+  numbers and arbitrary comperable objects.  Floating point numbers
+  were rejected because rounding errors can invalidate tests for
+  equality and ordering.  If a floating-point index seems appropriate,
+  it is recommended that users either multiply the floating point
+  numbers by a fixed constant and round them to integers, or create a
   bidirectional mapping between floating-point indices and integer
@@
 
+  A span $[s,e)$ includes the text at $s$ but not the text at $e$ in
+  order to be consistant with Python slice notation.  Also, all
+  indices are required to be non-negative for consistancy with Python
+  slice notation, where negative indices have special meaning.
+
 \section{Words}
@@
 
+  \subsection{Design}
+
+  The term ``word'' is used to refer to a wide variety of different
+  abstract concepts, as can be seen by looking through the list of
+  concepts from Section \ref{sec:concepts}.  One important distinction
+  that is easy to overlook is the difference between a word and an
+  individual occurance of a word.  For example, the sentence ``my dog
+  likes his dog'' contains four words but five occurances of words
+  (since the word ``dog'' appears twice).  In order to make this
+  distinction clear, we will use the words \emph{type} and
+  \emph{token} to refer to words and to occurances of words,
+  respectively.  For example, in the sentence ``my dog likes his
+  dog,'' we can say that the second and fifth words are different
+  tokens, but that they have the same types.
+
+  Individual tokens extracted from the same text can always be
+  distinguished based upon their locations.  This fact provides us
+  with a convenient mechanism for encoding tokens: a token is simply a
+  pair consisting of a type and a location.  Two tokens are only equal
+  if both their location and thier type are equal.
+
+  The notions of token and type are actually more general than the
+  notions of word and word occurance.  In the context of the toolkit,
+  types can refer to any object derived from text, and tokens can
+  refer to any instances of objects derived from text.  For example,
+  syntax trees are actually special kinds of tokens (see Section
+  \ref{sec:trees}).
+
+  The toolkit does not define a specific encoding for individual kinds
+  of tokens or token types, like simple words, tagged words, or
+  stemmed words.  Instead, it defines the general encoding for tokens,
+  and allows students to use almost any object as a type.  For
+  example, \code{string}s could be used to encode the types of simple
+  words, stemmed words, and case-normalized words; and pairs of
+  strings could be used to encode tagged word types.
+
+  However, the toolkit does provide a number of simple classes that
+  are intended to be used as types.  Currently, the only class
+  provided by the toolkit is the \code{TaggedType} class, which can be
+  used to encode the type of a tagged token.  A \code{TaggedType}
+  simply consists of a base type and a tag, both of which are
+  \code{string}s.
+
+  \subsection{Implementation}
+
+  \subsubsection{Types}
+      A Type can be any immutable object that properly implements
+      equality comparison.
+
+  \subsubsection{Tokens}
+      Tokens are implemented using the \code{Token} class.
+      \code{Token}s consist of:
+
+      \begin{itemize}
+
+        \item A \emph{location} (accessed via the \code{loc()} member
+        function).  \\
+        Type: \code{Location}
+
+        \item A \emph{type} (accessed via the \code{type()} member
+        function).  \\
+        Type: \emph{(any)}
+
+      \end{itemize}
+
+      \noindent
+      \code{Token}s are immutable objects.
+
+  \subsubsection{TaggedTypes}
+      \vspace{3mm}\noindent
+      The types of tagged tokens can be encoded using the
+      \code{TaggedType} class.  A \code{TaggedType} consists of:
+
+      \begin{itemize}
+
+        \item A \emph{base type} (accessed via the \code{base()}
+        member function).  \\
+        Type: \code{string}
+
+        \item A \emph{tag} (accessed via the \code{tag()} member
+        function).  \\
+        Type: \code{string}
+
+      \end{itemize}
+
+      \noindent
+      \code{TaggedType}s are immutable objects.  Two tagged types are
+      equal if their bases are equal and their tags are equal.  Note
+      that equality of strings is defined to be case-sensitive.
+
+  \subsection{Discussion}
+  % - why not a class/interface for type?
+  % - why location+type?
+  % - why not classes to distinguish stemmed/case-normalized/etc?
+
+  We considered implementing an interface or class hierarchy for
+  types.  Using an interface, rather than allowing arbitrary Python
+  objects, has a number of benefits:
+
+  \begin{description}
+
+    \item[Exception Semantics] If an interface is used to define what
+    type objects must do, then types can exhibit more precise
+    exception semantics.  In particular, it would be possible to
+    ensure that comparing two type objects with different classes
+    would always raise an exception.  In constrast, it is impossible
+    to make such guarantees when comparing arbitrary Python objects.
+
+    \item[Type-kind Distinctions] If each kind of type (e.g., stemmed
+    types, case-normalized types, tagged types, etc.) is implemented
+    with a separate class, then it is easier to maintain distinctions
+    between different kinds of types.  But if built-in Python objects
+    such as strings and tuples are used to represent types, then the
+    distinctions between different kinds of type will become blurred.
+    For example, since strings are likely to be used for simple words,
+    stemmed words, and case-normalized words, it will be impossible to
+    determine what kind of type a string contains by inspecting the
+    string object.
+
+  \end{description}
+
+  However, requiring that all types implement an interface often
+  introduces an extra layer of indirection, and complicates the system
+  used to represent words.  Since words are fundamental to the
+  toolkit, we decided that they should be kept as simple as is
+  practical.  Some of the exception semantics can be regained at a
+  different level.  For example, even though testing two types for
+  equality may not raise an exception if they have different Python
+  types, it would be possible to implement the \code{type\_eq()}
+  method of tokens to perform this check.\footnote{The
+  \code{type\_eq()} method is also included because it can be more
+  efficient than extracting the tokens' types and comparing them,
+  especially for some special kinds of tokens like trees (see Section
+  \ref{sec:trees}).}  Also, although type-kind distinctions cannot be
+  derived directly from the type object, these distinctions are
+  usually implicit in the history of the object, and it should be rare
+  to have a type object without knowing what kind of type it contains.
+  Therefore, we decided that the benefits derived from maintaining a
+  class hierarchy for types did not justify the increase in complexity
+  for the toolkit's set of basic type.
+
 \section{Collections of Words}
@@
 %  - list of token
+
+  \subsection{Design}
+
+  The toolkit does not implement any special encodings for collections
+  of words.  Instead, it relies on Python's built-in types.  However,
+  there are a number of standard ways to use Python's built-in types
+  to encode collections of words:
+
+  \begin{description}
+
+    \item[Strings] Unprocessed text is usually represented as a single
+    string.  For example, when data is read from a file, it is
+    typically read into a string.  Tokenizers can be used to convert
+    strings into lists of tokens.
+
+    \item[List of Tokens] A piece of text can be represented as an
+    ordered list of the tokens that it contains.
+
+    \begin{itemize}
+
+        \item A list of tokens $l$ is said to be ``properly ordered''
+        if, for every $i<j$, $l[i].loc()<l[j].loc()$.
+
+        \item A list of tokens $l$ is said to be ``exhaustive'' if:
+
+        \begin{itemize}
+          \item for every $i$, $l[i].end() = l[i+1].start()$; 
+          \item $l[0].start()$ is the start of the text.
+          \item $l[-1].end()$ is the end of the text.
+        \end{itemize}
+
+    \end{itemize}
+
+    \item[List of Types] When the locations of individual tokens is
+    unimportant, a text can be represented as an ordered list of its
+    token's types.
+
+    \item[Set of Tokens] When the order of the tokens in a text is
+    unimportant, the text can be represented as a set of tokens.
+    \footnote{Python does not (yet) provide a built-in set type;
+    however, the toolkit defines a generic set class.}  Since each
+    token's location serves to distinguish it from all other tokens,
+    the set will not contain any duplicate elements.
+
+    \item[Set of Types] When the order of tokens in a text and the
+    presence of multiple tokens with the same type are unimportant,
+    the text can be represented as a set of types.
+
+    \item[Dictionary Mappings] Since both tokens and types are
+    immutable objects, they can be used as keys in dictionaries.  A
+    typical use of a dictionary would be to encode a multiset of types
+    as a dictionary mapping types to counts.
+
+  \end{description}
+
+%  \subsection{Implementation}
+%  There is no real implementation, is there??
+
+  \subsection{Discussion}
+
+  % Why not explicit encodings?  Advantage of explicit encodings:
+  % check that everything in a list/set/whatever has the same pytype.
 
@@
 %  - treetype
+
+  \subsection{Design}
+
+  \subsection{Implementation}
+
+  \subsection{Discussion}
 
@@
 
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The abstract concepts which
-
-This document catalogs some of the different conceptual entities which
-the tooklit should deal with.  These conceptual entities sometimes
-draw more distinctions than is convenient when actually working with
-objects.
-
-An open question is how strict the toolkit should be about enforcing
-the distinctions between conceptually distinct types of entities.  On
-one hand, strictly enforcing these distinctions is conceptually
-``simpler,'' and may promote clearer thinking.  On the other hand, it
-is often cumbersome to make these distinctions -- in many
-circumstances, they don't matter, so they get dropped in practice.
-For example, most people will not distinguish ``tokens'' from ``token
-types'' unless they need to.  In a sense, this question boils down to
-whether it will be easier to learn/use a set of rules/conventions that
-are simpler, but somewhat less intuitive; or more complex, but more
-intuitive.  I think that one danger is that the more complex rules
-will only be more intuitive some of the time.  And it may be dangerous
-to be inconsistantly intuitive..
-
-Another open question is whether the current names should be kept.  In
-particular, names of conceptual entities should be both \emph{correct}
-and \emph{easy to use/remember}.
-
-Also, it may be possible that some of these conceptual entities could
-be merged or done away with altogther.  Are all of these entities
-serving useful purposes?
-
-\section{Token Stuff}
-
-{\it (Currently, many of these definitions
-are a bit hand-wavey.  That partially stems from the fact that they
-are not really as well-defined as we would like (esp the more
-primitive ones).  But I'll try to clean them up eventually.)}
-
+%% Meta-section
+\newpage
+\section{Meta-Questions}
 \begin{itemize}
-
-  \item \textbf{Lexeme:} An identifier for a word-like thing.
-  Normally, this would be a string.  But if you wanted to process text
-  more efficiently, it could be an integer.  It might be stemmed, not
-  stemmed, case-normalized, not case-normalized, etc.
-
-  \item \textbf{Token Type:} A word-like unit of text.  Includes a
-  lexeme, and may also include other info, like part of speech tag,
-  etc.
-
-  \item \textbf{Token:} An occurance of a token type
-
-  \item \textbf{Location:} The location of something in text.  There
-  are actually a number of ways of representing locations: as indexes,
-  as spans, etc.  Also, indexes can refer to word number, character
-  number, etc.
-
+  \item Are the terms used in this document correct?  Could better
+  terms be chosen?  Especially: properly ordered, exhaustive,
+  type-kind.
+  \item Are the interfaces presented here reasonable?
+  \item Do any decisions need more justification?
 \end{itemize}
 
-\section{Syntax Tree Stuff}
-
-\begin{itemize}
-
-  \item \textbf{SyntaxNode:} The annotation that appears at a node
-  in a syntax tree.  E.g., this might hold a phrase type (``NP'' or
-  ``V'').
-
-  \item \textbf{SyntaxTree:} An occurance of a tree or subtree.
-  Consists of a SyntaxNode and an ordered list of zero or more
-  children.  Children must be SyntaxTrees or Tokens.
-
-  \item \textbf{SyntaxTreeType:} The ``type'' of a syntax tree (as
-  opposed to an occurance of a syntax tree).  Consists of a SyntaxNode
-  and an ordered list of zero or more children.  Children must be
-  SyntaxTreeTypes or TokenTypes.
-
-\end{itemize}
-
-\section{How much to enforce?}
-
-So basically, the question of deciding how strictly to enforce these
-distinctions comes down to two questions:
-
-\begin{enumerate}
-  \item Should any entities/distinctions be eliminated?
-  \item Should it be possible to use entities in environments that
-        are not \emph{strictly speaking} correct?
-\end{enumerate}
-
-I'll leave the first question for later, and concentrate on the
-second.  There are a number of ways that objects could try to be more
-friendly when you use them in ways that that are not strictly speaking
-correct:
-
-\begin{itemize}
-  \item Equality tests: What happens when you compare a string to a
-  token?  A string to a TokenType?  A Token to a TokenType?  etc.?  A
-  related question is how to handle comparisons within one class
-  hierarchy.  E.g., equality tests between a tagged token and a simple
-  token.  The strict view would say that any of these comparisons
-  should raise an exception.  However, if we decide not to be strict,
-  there are a number of places where these equality tests could be
-  well-defined.  Examples that come to mind are:
-  \begin{itemize}
-    \item equality between a string and a SimpleTokenType
-    \item equality between a string and a Token (ignore location)
-    \item equality between an integer and a Location
-    \item equality between a list and a SyntaxTree(Type)
-  \end{itemize}
-
-  \item Arguments to methods: Should methods attempt to handle
-  arguments that are not quite the expected type?  E.g., when Steven
-  passed strings to a FreqDist, when it was really (sort of) expecting
-  TokenTypes.
-
-  \item Arguments to constructors: Should constructors automatically
-  ``wrap'' objects if they're not the right type?  E.g., should
-  Token(``cat'') automatically create a SimpleTokenType containing
-  ``cat''?
-\end{itemize}
+\end{document}

### 2001-05-25 technical/reptypes/reptypes.tex
Added "None" location for Tokens (re)wrote the Tree section. Wrote the summary
svn/trunk@33

@@
 
+% To do:
+%   - Figures
+%   - SyntaxTree discussion
+%   - Summary
+
 \newcommand{\concept}[1]{\textsf{#1}}
@@
 \usepackage{amsmath}
+\usepackage{epsfig}
 
@@
 
+%============= INTRODUCTION =============%
 \section{Introduction}
 
+% We will explain the toolkit's basic types
   The natural language toolkit allows students to manipulate a number
@@
 
+% Description of sections
   Section \ref{sec:concepts} describes the basic concepts which this
@@
 \newpage
+%============= CONCEPTS =============%
 \section{Abstract Concepts}
@@
 
+% Abstract concepts we wish to address
   The following list of abstract concepts is representative of the
@@
 
+%============= LOCATIONS =============%
 \section{Locations}
 \label{sec:locations}
-%  - location
+% Should the standard location notation be changed to 
+% [s:e], like slices?
 
@@
 
+% Location = <start, end> (where start<end)
   The location of an element of text is represented using a span-based
@@
 
+% Units encode what unit the indices use
   The start and end indices can be based on a variety of different
@@
 
+% Source encodes where the text came from
   The location of an element of text may also be tagged with its
@@
   of the sentence that contained it.  This allows locations to be
-  specified in a hierarchical fashion.
+  specified in a hierarchical fashion. % as shown in figure foo?
 
@@
     function).  \\
-    Type: \code{int}
+    Python Type: \code{int}
 
@@
     member function).  \\
-    Type: \code{string} \emph{(normalized to lower case)}
+    Python Type: \code{string} \emph{(normalized to lower case)}
 
@@
     member function).  \\
-    Type: \emph{(any)}
+    Python Type: \emph{(any)}
 
@@
 
+% Ordering of Locations
   \noindent
@@
   exceptions, they are defined as follows:
+  %
   \begin{align*}
@@
 
+% Why not more complex/general?
   Originally, I proposed a general, user-extensible notion of
@@
 
+% Why not simpler?
   A simpler notion of location, based on a single index, was
@@
 
+% Why units and sources?
   The optional unit and source fields were added to help distinguish
@@
 
+% Why integer indices?
   We decided to restrict location indices to non-negative integers
@@
 
+% Why [s,e)?
   A span $[s,e)$ includes the text at $s$ but not the text at $e$ in
@@
 
+%============= WORDS =============%
 \section{Words}
@@
 
+% Tokens versus Types
   The term ``word'' is used to refer to a wide variety of different
@@
 
+% Distinguish Tokens via locations
   Individual tokens extracted from the same text can always be
@@
 
+% Special Location: None
+  Token locations are typically instances of the \code{Location}
+  class.  However, Tokens may also have the special location
+  \code{None}, which can be used when a token's location is unknown or
+  unimportant.  A Token with a location of \code{None} is not
+  considered to be equal to any other token.  In particular, even if
+  two tokens have the same type, and both have a location of
+  \code{None}, they are not considered equal.
+
+% Tokens are more general than words
   The notions of token and type are actually more general than the
@@
 
+% Types
   The toolkit does not define a specific encoding for individual kinds
@@
 
+% Special Types
   However, the toolkit does provide a number of simple classes that
@@
       A Type can be any immutable object that properly implements
-      equality comparison.
+      equality comparison and hashing.
 
@@
         function).  \\
-        Type: \code{Location}
+        Python Type: \code{Location}
 
@@
         function).  \\
-        Type: \emph{(any)}
+        Python Type: \emph{(any)}
 
@@
   \subsubsection{TaggedTypes}
-      \vspace{3mm}\noindent
       The types of tagged tokens can be encoded using the
@@
         member function).  \\
-        Type: \code{string}
+        Python Type: \code{string}
 
@@
         function).  \\
-        Type: \code{string}
+        Python Type: \code{string}
 
@@
 
+% Why we considered a class hierarchy
   We considered implementing an interface or class hierarchy for
@@
 
+% Why we rejected a class hierarchy
   However, requiring that all types implement an interface often
@@
 
+%============= COLLECTIONS =============%
 \section{Collections of Words}
@@
 
+% No special collection types
   The toolkit does not implement any special encodings for collections
@@
 
+% Standard ways to represent collections
   \begin{description}
@@
     \item[Strings] Unprocessed text is usually represented as a single
-    string.  For example, when data is read from a file, it is
+    string.  For example, when a document is read from a file, it is
     typically read into a string.  Tokenizers can be used to convert
@@
 
-        \item A list of tokens $l$ is said to be ``exhaustive'' if:
+        \item A list of tokens $l$ is said to be ``exhaustive''
+        if:\footnote{I might want to change this definition -- this
+        one implies properly ordered.  We might want exhaustive and
+        properly ordered to be mutually exclusive.}
 
@@
 
-    \item[Set of Tokens] When the order of the tokens in a text is
-    unimportant, the text can be represented as a set of tokens.
-    \footnote{Python does not (yet) provide a built-in set type;
-    however, the toolkit defines a generic set class.}  Since each
-    token's location serves to distinguish it from all other tokens,
-    the set will not contain any duplicate elements.
+    \item[Set of Tokens]\footnote{Python does not (yet) provide a
+    built-in set type; however, the toolkit defines a generic set
+    class.} When the order of the tokens in a text is unimportant, the
+    text can be represented as a set of tokens.  Since each token's
+    location serves to distinguish it from all other tokens, the set
+    will not contain any duplicate elements.
 
@@
 
-  % Why not explicit encodings?  Advantage of explicit encodings:
-  % check that everything in a list/set/whatever has the same pytype.
+% Why we considered special collection classes
+  The wide variety of ways in which students might want to arrange
+  collections of words would make it very difficult to design a
+  specialized ``word collection'' encoding mechanism.  However, it
+  would be possible to create several specific class-based encodings
+  for the more common kinds of word collection.  If collections of
+  words were encoded using a class, then it would be possible to check
+  or enforce a number of restrictions on the members of the
+  collections.  For example, for some types of collections, we might
+  want to check that:
 
+  \begin{itemize}
+
+    \item All tokens in the collection have the same kind of type.
+
+    \item The tokens in the collection are properly ordered.
+
+    \item The tokens in the collections are exhaustive.
+
+  \end{itemize}
+
+% Why we rejected special collection classes
+  However, adding new classes to encapsulate various types of
+  collection would complicate the toolkit, making it more difficult to
+  learn and to understand.  Furthermore, the fact that only a limited
+  set of arrangements would be available would create a division
+  between the ``supported'' arrangements and the ``unsupported ones.''
+  Given how easy Python's native collection types are to use, we
+  decided not to include any explicit ``word collection'' classes or
+  interfaces in the toolkit.
+
+%============= TREES =============%
 \section{Syntax Trees}
@@
 
+% Text-tree is like a rooted ordered tree
+  Syntax trees are represented using the \emph{text-tree} data
+  structure, a general structure for representing homogenous
+  hierarchical structures spanning text.  Figure~\ref{fig:tree} shows
+  the structure of an example text-tree.  Conceptually, text-trees are
+  very similar to the mathematical notion of a \emph{rooted ordered
+  tree}.  In particular, a text-tree contains of a set of
+  \emph{elements}, including a special \emph{root} element.  These
+  elements are connected by \emph{edges}, such that there is exactly
+  one path from the root element to each other element.  In addition,
+  children of each element are ordered.
+
+% Figure: example of Tree data structure
+  \begin{figure}
+  \begin{centering}
+    \epsfbox{tree.eps}\\
+  \end{centering}
+
+   \caption{Example of the \emph{text-tree} data structure used to
+   represent a syntax tree.  In this text-tree, the nodes are strings
+   containing phrase tags, such as \code{"S"} and \code{"VP"}; and the
+   leaves are strings containing word types, such as \code{"John"} and
+   \code{"Mary"}.}
+
+  \label{fig:tree}
+  \end{figure}
+
+\subsubsection{Leaves and Nodes}
+% Leaves and nodes are mutually exclusive sets
+  However, there are a few signifigant differences between text-trees
+  and mathematical trees.  The most important difference is that the
+  elements of a text-tree are explicitly divided into two separate
+  classes, which we will call \emph{leaves} and \emph{nodes}.  The
+  \emph{leaves} represent the individual elements of text spanned by
+  the text-tree.  Leaves may not have children.  The \emph{nodes}
+  associate information with the text-tree's hierarchical
+  constituants.  For example, nodes could be used to label syntactic
+  constituants with phrase tags, such as ``NP'' and ``VP.''
+
+% ``Leaf'' and ``node'' don't have their standard definitions
+  It is important to realize that these are \emph{not} the standard
+  definitions for the terms \emph{leaf} and \emph{node}: in the
+  context of a rooted ordered tree, node is synonymous with element,
+  and a leaf is any element with no children.  But in the context of
+  text-trees, leaf elements and node elements are mutually exclusive,
+  and it is possible for node elements to have no children.
+
+% Use standard terminology 
+  However, much of the standard terminology for rooted ordered trees
+  applies to text-trees.  In particular, the terms \emph{child},
+  \emph{descendant}, \emph{parent}, \emph{ancestor}, \emph{sister},
+  \emph{root}, \emph{subtree}, \emph{height}, and \emph{depth} have
+  their typical meanings.
+
+%\subsubsection{Homogeneity}
+%% Text-trees are homogenous
+%  Another difference between text-trees and general tree structures is
+%  that text-trees must be homogenous.  Before we can formalize this
+%  property, we must introduce the term ``pytype'': a Python object's
+%  \emph{pytype} is the object's Python type if the object has a
+%  built-in type; and is its class if the object is an instance.  The
+%  homogeneity requiremenet for text-trees states that:
+%
+%  \begin{itemize}
+%
+%    \item Every node in a tree must have the same pytype.
+%
+%    \item Each leaf in a tree must have the same pytype.
+%
+%  \end{itemize}
+
+\subsubsection{Proper Use}
+% Trees are more general than syntax trees
+  Although text-trees were designed to handle syntax trees, they are
+  capable of encoding a more general class of objects.  For example,
+  text-trees could be used to encode parser rules.  In fact, it would
+  be possible to use the text-tree class to represent almost any
+  homogenous tree-based data structure.  However, the text-tree class
+  is primarily intended to encode structures that exist over spans of
+  text, and it should not be used in unrelated contexts.
+
+\subsubsection{Text-Tree Tokens}
+% Tokens vs. Types
+  As was the case with words, there is an important distinction to be
+  made between \emph{occurances} of syntax trees and syntax tree
+  \emph{types}.  Occurances of text-trees are represented with the
+  \emph{text-tree token} data structure.  This data structure is
+  parallel to the text-tree data structure; but instead of containing
+  types, the leaves contain tokens.  Figure \ref{fig:treetoken} shows
+  the structure of an example text-tree token.
+
+% Figure: example of Tree data structure
+  \begin{figure}
+  \begin{centering}
+    \epsfbox{treetoken.eps}\\
+  \end{centering}
+
+   \caption{Example of the \emph{text-tree token} data structure used
+   to represent an occurance of a syntax tree.  This text-tree token's
+   type is the text-tree shown in figure \ref{fig:tree}; and its
+   location is [8, 11).}
+
+  \label{fig:treetoken}
+  \end{figure}
+
+% Text-tree tokens are tokens.
+  Text-tree tokens are special kinds of tokens; in particular, every
+  text-tree token has a type and a location.  A text-tree token's type
+  is the text-tree that would be obtained by replacing all the leaf
+  tokens with their types.  A text-tree's location spans from the
+  beginning of its first leaf token to the end of its last leaf token,
+  and has the same unit and source as its tokens.  If a text-tree
+  token does not have any leaves, then its location is the special
+  value \code{None}.  As with other tokens, a text-tree tokens whose
+  locations is \code{None} is never equal to any other text-tree
+  token.
+
+% Tree != Type+Location
+  Note that the location of a text-tree token contains only the overall
+  location of the tree, and not the locations of its individual
+  leaves.  As a result, a text-tree token can not be treated as consisting
+  of just its location and its type.  In particular, it is possible to
+  construct two text-tree tokens that are not equal, even though their
+  types and locations are equal (See Figure~\ref{fig:badtree}).
+  However, it is highly unlikely that such pairs of text-tree tokens will
+  occur in practice.
+
+% Figure: Tree != Type+Location
+  % (the [b] puts it at the bottom of the page -- makes it look better)
+  \begin{figure}[b]
+  \begin{centering}
+    \epsfbox{badtree.eps}\\
+  \end{centering}
+
+   \caption{(a) and (b) show two text-tree tokens that are not equal,
+   even though their types and locations (shown in (c)) are equal.
+   However, this should not occur in practice, since it can only occur
+   if two tokens derived from the same source have overlapping
+   locations.}
+  \label{fig:badtree}
+  \end{figure}
+
+  \subsubsection{Consistancy Conditions}
+  \label{sec:consistancy}
+
+% Consistancy Conditions
+  Text-trees and text-tree tokens are constrained by a number of
+  consistancy conditions.  These conditions are designed to simplify
+  procedures written by students to process text-trees and text-tree
+  tokens.  They are also designed to help prevent some common bugs.
+  Several of the consistancy conditions require that different
+  elements have the ``same kind of value.''  We can formalize this
+  notion of ``kind of value'' with the term ``pytype'': a Python
+  object's \emph{pytype} is the object's Python type if the object has
+  a built-in type; and is its class if the object is an instance.  The
+  consistancy conditions are:
+
+  \begin{enumerate}
+
+%Trees must be proper (this is implied by def. of text-tree & token)
+%    \item Every text-tree and text-tree token must be a proper tree.
+%    In particular, their structure must be acyclic, and may contain no
+%    merge points.
+
+% Node values must be consistant
+    \item Every node in a text-tree or text-tree token must have the
+    same kind of value.  In particular, for any text-tree or text-tree
+    token $T$ with nodes $n_1, n_2, \ldots, n_k$, it must be the case that
+    $pytype(n_i) = pytype(n_j)$ for all $i$ and $j$, $1\leq i\leq j\leq n$.
+
+% Leaf tokens must be consistant (text-tree)
+    \item Every leaf in a text-tree must have the same pytype.  In
+    particular, for any text-tree with leaves $l_1, l_2, \ldots, l_n$, it
+    must be the case that $pytype(l_i) = pytype(type(l_j))$ for all
+    $i$ and $j$, $1\leq i\leq j\leq n$.
+
+% Leaf tokens types must be consistant (text-tree token)
+    \item The type field of every leaf token in a text-tree token must
+    have the same pytype.  In particular, for any text-tree token with
+    leaves $l_1, l_2, \ldots, l_n$, it must be the case that
+    $pytype(type(l_i)) = pytype(type(l_j))$ for all $i$ and $j$, $1\leq
+    i\leq j\leq n$.
+
+% Leaves must be properly ordered
+    \item Every text-tree token's leaves must be properly ordered.  In
+    particular, for every text-tree token with children $c_1, c_2, \ldots,
+    c_n$, it must be the case that $location(c_i)<location(c_j)$
+    whenever $i<j$, $location(c_i)\neq\code{none}$, and
+    $location(c_j)\neq\code{none}$.
+
+% Leaf locations must be consistant
+    \item The location of every leaf in a text-tree token must have the
+    same source and the same unit.  (This condition is actually implied
+    by the preceeding condition, since ordering comparison of two
+    locations is only defined when two locations have the same source
+    and unit.)
+
+  \end{enumerate}
+
   \subsection{Implementation}
+
+    \subsubsection{Text-Tree}
+
+      Text-trees are represented with recursive data structures based
+      on the \code{Tree} class.  The \code{Tree} instances used to
+      represent an example text-tree are shown in
+      Figure~\ref{fig:tree-impl}.  Each \code{Tree} instance
+      encodes the local relationships between a node and its children.
+      A \code{Tree} consists of:
+
+% Figure: Implementation of a Tree
+      \begin{figure}
+      \begin{centering}
+        \epsfbox{tree-impl.eps}\\
+      \end{centering}
+
+       \caption{The \code{Tree} instances used to encode an example
+       text-tree.  The example text-tree is shown in (a), and the
+       \code{Tree}-based data structure is shown in (b).}
+      \label{fig:tree-impl}
+      \end{figure}
+
+      \begin{itemize}
+
+        \item A \emph{node} (accessed via the \code{node()}
+        member function).  \\
+        Python Type: \emph{(nodetype)}, where \emph{(nodetype)} can be
+        any (single) Python type.
+
+        \item A list of \emph{children}.  (accessed via standard
+        sequence operations, such as indexing and slicing) \\
+        Python Type: \code{list} of \code{Tree} and
+        \emph{(leaftype)}, where \emph{(leaftype)} can be any (single)
+        Python type.
+
+      \end{itemize}
+
+      \noindent Several auxilliary accessors can be used to return
+      various information about a \code{Tree}:
+
+      \begin{itemize}
+
+        \item \code{leaves()} will return an ordered list of the
+        \code{Tree}'s leaves. \\
+        Python Type: list of \emph{(leaftype)}, where
+        \emph{(leaftype)} can be any (single) Python type.
+
+        \item \code{nodes()} will a new \code{Tree} that contains all
+        of the \code{Tree}'s nodes, but no leaves. \\
+        Python Type: \code{Tree}
+
+      \end{itemize}
+
+      \noindent \code{Tree}s are immutable objects.
+
+    \subsubsection{Text-Tree Tokens}
+
+      Text-trees tokens are represented with recursive data structures
+      based on the \code{TreeToken} class.  The \code{TreeToken}
+      instances used to represent an example text-tree occurance are
+      shown in Figure~\ref{fig:treetoken-impl}.  Each \code{TreeToken}
+      instance encodes the local relationships between a node and its
+      children.  A \code{TreeToken} consists of:
+
+% Figure: Implementation of a TreeToken
+      \begin{figure}
+      \begin{centering}
+        \epsfbox{treetoken-impl.eps}\\
+      \end{centering}
+
+       \caption{The \code{TreeToken} instances used to encode an
+       example text-tree token.  This text-tree token's type is the
+       text-tree shown in Figure \ref{fig:tree-impl}.  Its location is
+       [10, 14).}
+      \label{fig:treetoken-impl}
+      \end{figure}
+
+      \begin{itemize}
+
+        \item A \emph{node} (accessed via the \code{node()}
+        member function).  \\
+        Python Type: \emph{(nodetype)}, where \emph{(nodetype)} can be
+        any (single) Python type.
+
+        \item A list of \emph{children}.  (accessed via standard
+        sequence operations, such as indexing and slicing) \\
+        Python Type: \code{list} of \code{TreeToken} and \code{Token}.
+
+      \end{itemize}
+
+      \noindent Several auxilliary accessors can be used to return
+      various information about a \code{TreeToken}:
+
+      \begin{itemize}
+
+        \item \code{type()} will return a \code{Tree} encoding the
+        \code{TreeToken}'s type. \\
+%        In particular, it will
+%        return \textit{type}(self), where \textit{type}() is defined as:
+%        \begin{align*}
+%         \textit{type}\big(\code{TreeToken}(n, c_1, c_2, \ldots, c_n)\big) &= 
+%              \code{Tree}\big(n, \textit{type}(c_1), 
+%              \textit{type}(c_2), \ldots, \textit{type}(c_n)\big)\\
+%         \textit{type}\big(\code{Token}(t, l)\big) &= t
+%        \end{align*}
+%        \noindent ($n$ is a \code{TreeToken}'s node value, $c_i$
+%        are a \code{TreeToken}'s children, $t$ is a \code{Token}'s
+%        type, and $l$ is a \code{Token}'s location.)\\
+        Python Type: \code{Tree}
+
+        \item \code{loc()} will return the \code{TreeToken}'s
+        location. \\
+        Python Type: \code{Location}
+
+        \item \code{leaves()} will return an ordered list of the
+        \code{TreeToken}'s leaves. \\
+        Python Type: list of \emph{(leaftype)}, where
+        \emph{(leaftype)} can be any (single) Python type.
+
+        \item \code{nodes()} will a new \code{TreeToken} that contains
+        all of the \code{TreeToken}'s nodes, but no leaves. \\
+        Python Type: \code{TreeToken}
+
+      \end{itemize}
+
+      \noindent \code{TreeToken}s are immutable objects.  The
+      \code{TreeToken} class is derived from the \code{Token} class.
+
+    \subsubsection{Consistancy Conditions}
+
+      Since many of the consistancy conditions are difficult to
+      verify, it is the user's responsibility not to violate any of
+      the conditions; however, most of the consistancy conditions will
+      be checked at runtime, to help prevent bugs.
 
@@
 
+% Why encode Trees explicitly?
+  We decided to encode syntax trees explicitly, rather than relying on
+  the built-in Python types, because the built-in types are not well
+  suited for handling homogenous trees.  In particular, there are a
+  number of operations over homogenous trees, such as finding the set
+  of leaves, which are unnecessarily difficult to perform with data
+  structures built from the built-in Python types.  Furthermore,
+  creating an explicit representation allows us to enforce consistancy
+  conditions.  As discussed in Section \ref{sec:consistancy}, these
+  conditions can simplify procedures written by students, and can help
+  prevent common bugs.
+
+% What other reps did we consider?
+  \vspace{4mm}\noindent We considered two alternative proposals for
+  representing text-tree tokens:
+
+  \begin{itemize}
+
+% Why not simply use Tokens, TreeTypes, and more simple Locations?
+    \item Occurances of text-trees could be represented using simple
+    tokens and locations, as described in Sections \ref{sec:locations}
+    and \ref{sec:words}.  This approach has the advantage of
+    simplicity.  However, it does provide any mechanism with which
+    text-trees can store the locations of their constituant elements.
+    As a result, it is not possible to create tokens representing the
+    constituants.
+
+% Why not simply use Tokens, TreeTypes, and more complex Locations?
+    % This really needs more explanation!
+    \item Occurances of text-trees could be represented using simple
+    tokens, but complex location objects could be used to store the
+    position of each of the text-tree's elements.  This approach
+    solves some of the problems with the first proposal.  But if
+    occurances of text-trees were represented using the standard
+    \code{Token} class, then it would be unnecessarily difficult to
+    perform a number of tree-specific operations, such as constructing
+    a list of the tokens representing a text-tree token's leaves.
+
+  \end{itemize}
+
+% Why make TreeTokens a subclass of Token?
+  \noindent Note that even though occurances of text-trees are not
+  directly encoded with the \code{Token} class, the fact that they are
+  special types of tokens is represented by the fact that the
+  \code{TreeToken} class is a subclass of the \code{Token} class.
+
+% Why immutable?
+  \vspace{1mm}\noindent We decided to make text-trees and text-tree
+  tokens immutable for two reasons:
+
+  \begin{itemize}
+
+    \item Immutability allows text-trees and text-tree tokens to be
+    used as keys in dictionaries and elements of dictionary-based
+    sets.
+
+    \item Immutability allows sub-trees to be safely shared.  This
+    eliminates an important class of bugs having to do with modifying
+    a tree without realizing that it has accidentally been shared.
+
+  \end{itemize}
+
+% Why the constraints?
+  We decided to enforce the constraints described in
+  \ref{sec:consistancy} because we believe that they will help prevent
+  errors in student code.  We may decide to strengthen or weaken the
+  consistancy conditions in the future, once we have gained more
+  experience with using the system.
+
+%============= SUMMARY =============%
+\newpage
 \section{Summary}
 \label{sec:summary}
-% Dependancies:
-%     token \to location
-%     tree \to token
-%     tree \to location
-%     token \to (type)
-%     tree \to (type)
-%     tree \to (node)
 
-%% Meta-section
+  This document defined four classes, which can be used together with
+  Python's built-in types to represent a number of conceptual objects
+  related to natural language processing.  Figure
+  \ref{fig:concept-reprs} shows which classes and types are typically
+  used to represent different conceptual objects.
+
+  Figure \ref{fig:containership} shows the internal structure of each
+  of the classes defined by this document.  Figure
+  \ref{fig:dependencies} shows the dependencies between these classes.
+
+% Overall-picture
+% refer to the figures, and explain them.
+
+% Typical representations for concepts from sec:concepts.
+  \begin{figure}
+  \begin{centering}
+  \begin{tabular}{ll}
+    Concept & Typical Representation \\
+    \hline
+
+    \concept{Document} & \code{String} \\
+
+    \concept{Sentence} & \code{String} \\
+
+    \concept{Word Type} & \code{String} \\
+
+    \concept{Tagged Word Type} & \code{TaggedType} \\
+
+    \concept{Stemmed Word Type} & \code{String} \\
+
+    \concept{Case-Normalized Word Type} & \code{String} \\
+
+    \concept{Word Occurance} & \code{Token} (type=\code{String})\\
+
+    \concept{Tagged Word Occurance} & \code{Token}
+    (type=\code{TaggedType})\\
+
+    \concept{Stemmed Word Occurance} & \code{Token}
+    (type=\code{String})\\
+
+    \concept{Case-Normalized Word Occurance} & \code{Token}
+    (type=\code{String})\\
+
+    \concept{Tokenized Document} & \code{list} of \code{Token}\\
+
+    \concept{Tokenized Sentence} & \code{list} of \code{Token}\\
+
+    \concept{Location} & \code{Location} \\
+
+    \concept{Tagged Document/Sentence} &
+    \code{list} of \code{Token} (type=\code{String})\\
+
+    \concept{Stemmed Document/Sentence} &
+    \code{list} of \code{Token} (type=\code{String})\\
+
+    \concept{Case-Normalized Document/Sentence} &
+    \code{list} of \code{Token} (type=\code{String})\\
+
+    \concept{Syntax Tree} & \code{Tree} \\
+
+    \concept{Syntax Node} & \code{String} \\
+
+    \concept{Syntax Tree Occurance} & \code{TreeToken} \\
+
+  \end{tabular}\\
+  \end{centering}
+
+  \caption{The mechanisms typically used to encode each of the
+  abstract concepts that was listed in section \ref{sec:concepts}.
+  (Note, though, that it may sometimes be adventageous to use other
+  representations.)}
+  \label{fig:concept-reprs}
+  \end{figure}
+
+% Containership relations
+  \begin{figure}
+  \begin{centering}
+    \epsfbox{containership.eps}\\
+  \end{centering}
+  \caption{The constituants that compose each representation class
+  described in this document.}
+  \label{fig:containership}
+  \end{figure}
+
+% Dependancies
+  \begin{figure}
+  \begin{centering}
+    \epsfbox{dependencies.eps}\\
+  \end{centering}
+  \caption{The dependancies between the basic representation classes
+  described in this document.}
+  \label{fig:dependencies}
+  \end{figure}
+
 \newpage
+%============= META-QUESTIONS =============%
 \section{Meta-Questions}
@@
   \item Do any decisions need more justification?
+  \item Other comments are always appreciated.
 \end{itemize}

### 2001-05-28 technical/printing/printing.tex
Initial import: Representations for printing
svn/trunk@47

--- 2001-05-11 technical/reptypes/reptypes.tex
@@
 % Natural Language Toolkit for Python
-% Technical Report: Representation Types
+% Representations for Printing
 % Edward Loper
 % 
-% Created [03/02/01 12:02 AM]
+% Created [05/27/01 11:44 PM]
 % $Id$
@@
 
-\newcommand{\concept}[1]{\textsf{#1}}
-\newcommand{\code}[1]{\texttt{#1}}
-
-\documentclass{article}
+\documentclass[11pt]{article}
 \usepackage{fullpage}
-\usepackage{amsmath}
 
 \begin{document}
-\title{Basic Representation Types\\
+\title{Representations for Printing\\
 \Large Natural Language Processing Toolkit: Technical Report}
@@
 
+%============= INTRODUCTION =============%
 \section{Introduction}
 
-  The natural language toolkit allows students to manipulate a number
-  of conceptual objects, such as words, sentences, and syntax trees.
-  Before we can build a consistant framework for manipulating such
-  objects, we must decide how they should be represented.  This
-  technical report explains basic types used by the toolkit to
-  represent a number of conceptual objects.  Furthermore, it attempts
-  to justify why this set of basic types was chosen.
-
-  Section \ref{sec:concepts} describes the basic concepts which this
-  technical report attempts to address.  Sections \ref{sec:locations}
-  through \ref{sec:trees} describe the basic types used to encapsulate
-  these concepts.  Section \ref{sec:summary} provides a summary of the
-  basic types, and discusses how they interact.
-
-\newpage
-\section{Abstract Concepts}
-\label{sec:concepts}
-
-  The following list of abstract concepts is representative of the
-  concepts which should be captured by the basic types defined in this
-  techincal report.\footnote{The types used to represent many other
-  abstract concepts (such as frequency distributions) are beyond the
-  scope of this technical report.}  Where the meaning of the concepts
-  are not obvious, I included explanations.
-
-  \begin{itemize}
-
-    \item \concept{Document}
-
-    \item \concept{Sentence}
-
-    \item \concept{Word Type}: A single word (as opposed to a
-        \concept{Word Occurance}).
-
-    \item \concept{Tagged Word Type}: A word that has been tagged with
-        additional information, such as part-of-speech information.
-
-    \item \concept{Stemmed Word Type}: A word that has been stripped of
-        inflectional suffixes and prefixes.
-
-    \item \concept{Case-Normalized Word Type}: A word whose case has been
-        normalized (usually to either upper case or lower case).
-
-    \item \concept{Word Occurance}: A single occurance of a word type.
-
-    \item \concept{Tagged Word Occurance}
-    \item \concept{Stemmed Word Occurance}
-    \item \concept{Case-Normalized Word Occurance}
-
-    \item \concept{Tokenized Document}: A document that has been divided
-        into individual word occurances.
-
-    \item \concept{Tokenized Sentence}: A sentence that has been divided
-        into individual word occurances.
-
-    \item \concept{Location}: The location of one unit of text within
-        another unit of text.  A typical location might give the
-        location of a word within a document or a sentence.
-
-    \item \concept{Tagged Document} or \concept{Tagged Sentence}: A
-        document or sentence that has been divided into individual word
-        occurances, each of which has been tagged with additional
-        information.
-
-    \item \concept{Stemmed Document} or \concept{Stemmed Sentence}: A
-        document or sentence that has been divided into individual word
-        occurances, each of which has been stripped of inflectional
-        affixes.
-
-    \item \concept{Case-Normalized Document} or \concept{Case-Normalized
-        Sentence}: A document or sentence that has been divided into
-        individual word occurances, each of which has had its case
-        normalized.
-
-    \item \concept{Syntax Tree}: A hierarchical representation of the
-        structure of a sentence.
-
-    \item \concept{Syntax Node}: The annotation appearing at a node in a
-        syntax tree.  Typically, this will be a tag, such as ``NP.''
-
-    \item \concept{Syntax Tree Occurance}: A single occurance of a syntax
-        tree.
-
-  \end{itemize}
-
-\section{Locations}
-\label{sec:locations}
-%  - location
-
-  \subsection{Design}
-
-  The location of an element of text is represented using a span-based
-  data structure.  A \emph{span} consists of a \emph{start index} and
-  an \emph{end index}, both of which are integers.  A span with start
-  index $s$ and end index $e$ is written ``$[s,e)$,'' and specifies
-  the location of the text beginning at $s$, and including everything
-  up to (but not including) the text at $e$.  It is always true that
-  $s<e$.\footnote{Should this be changed to $s\leq e$?}
-
-  The start and end indices can be based on a variety of different
-  \emph{units}, such as character number, word number, or sentence
-  number.  However, for any two indeces $x$ and $y$, it should be true
-  that $x<y$ if and only if $x$ indicates a location earlier in the
-  text than $y$.  The location of an element of text may be explicitly
-  tagged with information about what unit its indices use.  Unit
-  labels take the form of case-insensitive \code{string}s.  Typical
-  examples of unit labels are the strings \code{"character"} and
-  \code{"word"}.
-
-  The location of an element of text may also be tagged with its
-  \emph{source}.  This is an object that gives an indication of where
-  the text was derived from.  A typical example of a source would be a
-  \code{string} containing the name of the file from which the element
-  of text was read.  Sources may also be represented as locations; for
-  example, the source for the location of a word might be the location
-  of the sentence that contained it.  This allows locations to be
-  specified in a hierarchical fashion.
-
-  \subsection{Implementation}
-
-  Locations are implemented using the \code{Location} class.
-  \code{Location}s consist of:
-
-  \begin{itemize}
-
-    \item A \emph{start index} (accessed via the \code{start()} member
-    function).  \\Type: \code{int}
-
-    \item An \emph{end index} (accessed via the \code{end()} member
-    function).  \\Type: \code{int}
-
-    \item An optional \emph{unit} (accessed via the \code{unit()}
-    member function).  \\Type: \code{string} \emph{(normalized to
-    lower case)}
-
-    \item An optional \emph{source} (accessed via the \code{source()}
-    member function).  \\Type: \emph{(any)}
-
-  \end{itemize}
-
-  \noindent
-  Locations are immutable objects.
-
-  Locations can be compared for equality and ordering.  If two
-  locations with different units or different sources are compared,
-  then an exception will be raised.  When location comparisons do not
-  raise exceptions, they are defined as follows:
-  \begin{align*}
-     [s_1,e_1) = [s_2,e_2) &\iff s_1=s_2 \quad \land \quad e_1=e_2 \\
-     [s_1,e_1) < [s_2,e_2) &\iff e_1 < s_2 \\
-     [s_1,e_1) > [s_2,e_2) &\iff s_1 > e_2 \\
-     [s_1,e_1) \leq [s_2,e_2) &\iff [s_1,e_1) < [s_2,e_2) \quad \lor \quad
-                             [s_1,e_1) = [s_2,e_2) \\
-     [s_1,e_1) \geq [s_2,e_2) &\iff [s_1,e_1) > [s_2,e_2) \quad \lor \quad
-                             [s_1,e_1) = [s_2,e_2) 
-  \end{align*}
-
-  \noindent
-  Note that if two locations $l_1$ and $l_2$ overlap, then it will
-  never be the case that $l_1<l_2$ or $l_1>l_2$.
-
-  \subsection{Explanation and Comments}
-
-  Originally, I proposed a general, user-extensible notion of
-  location.  However, the span-based representation described here
-  seems to be sufficiently powerful to represent almost any location.
-  We decided that the additional power granted by making the notion of
-  location user-extensible did not compensate for the added
-  complexity.  Also, if the notion of location built-in, then other
-  classes can rely on it for computation.  For example, syntax trees
-  can compute their locations as a function of the locations of their
-  leaves.  If locations were user-extensible objects, then there would
-  be no general way to perform such computations.
-
-  A simpler notion of location, based on a single index, was
-  considered.  However, it was decided that the added power provided
-  by representing locations as ranges was worthwhile, especially in
-  the context of multi-word structures like syntax trees.
-
-  The optional unit and source fields were added to help distinguish
-  different sets of locations, and to help ensure that unrelated
-  locations are not accidentally compared.  For example, it may be
-  useful for a student to keep track of the location of the words in
-  two different files; but comparing these locations would be
-  meaningless.  We considered simply returning false when locations
-  with non-matching units or sources are compared, but decided that
-  raising exceptions would be more likely to help students find
-  errors.  If students wish to compare two locations that may have
-  different sources or units, then they must explicitly compare the
-  location's sources or units, or catch and process the exceptions
-  raised by the comparison methods.
-
-  We decided to restrict location indices to integers because it
-  simplifies the interface without signifigantly reducing its power.
-  Other alternatives we considered were floating point numbers and
-  arbitrary comperable objects.  Floating point numbers were rejected
-  because rounding errors can invalidate tests for equality and
-  ordering.  If a floating-point index seems appropriate, it is
-  recommended that users either multiply the floating point numbers by
-  a fixed constant and round them to integers, or create a
-  bidirectional mapping between floating-point indices and integer
-  indices.  The use of arbitrary comperable objects as indices was
-  rejected because it signifigantly complicates the location
-  interfaces, without adding signifigant power to the system.  Also,
-  arbitrary comperable objects may incur some of the difficulties that
-  would be encountered with floating point numbers.
-
-\section{Words}
-\label{sec:words}
-%  - (type)
-%  - token
-
-\section{Collections of Words}
-\label{sec:collections}
-%  - string (sentence/document)
-%  - list of token
-
-\section{Syntax Trees}
-\label{sec:trees}
-%  - (node)
-%  - tree (type of token)
-%      - (node pytype)
-%      - (leaf pytype)
-%  - treetype
-
-\section{Summary}
-\label{sec:summary}
-% Dependancies:
-%     token \to location
-%     tree \to token
-%     tree \to location
-%     token \to (type)
-%     tree \to (type)
-%     tree \to (node)
-
 \end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The abstract concepts which
-
-This document catalogs some of the different conceptual entities which
-the tooklit should deal with.  These conceptual entities sometimes
-draw more distinctions than is convenient when actually working with
-objects.
-
-An open question is how strict the toolkit should be about enforcing
-the distinctions between conceptually distinct types of entities.  On
-one hand, strictly enforcing these distinctions is conceptually
-``simpler,'' and may promote clearer thinking.  On the other hand, it
-is often cumbersome to make these distinctions -- in many
-circumstances, they don't matter, so they get dropped in practice.
-For example, most people will not distinguish ``tokens'' from ``token
-types'' unless they need to.  In a sense, this question boils down to
-whether it will be easier to learn/use a set of rules/conventions that
-are simpler, but somewhat less intuitive; or more complex, but more
-intuitive.  I think that one danger is that the more complex rules
-will only be more intuitive some of the time.  And it may be dangerous
-to be inconsistantly intuitive..
-
-Another open question is whether the current names should be kept.  In
-particular, names of conceptual entities should be both \emph{correct}
-and \emph{easy to use/remember}.
-
-Also, it may be possible that some of these conceptual entities could
-be merged or done away with altogther.  Are all of these entities
-serving useful purposes?
-
-\section{Token Stuff}
-
-{\it (Currently, many of these definitions
-are a bit hand-wavey.  That partially stems from the fact that they
-are not really as well-defined as we would like (esp the more
-primitive ones).  But I'll try to clean them up eventually.)}
-
-\begin{itemize}
-
-  \item \textbf{Lexeme:} An identifier for a word-like thing.
-  Normally, this would be a string.  But if you wanted to process text
-  more efficiently, it could be an integer.  It might be stemmed, not
-  stemmed, case-normalized, not case-normalized, etc.
-
-  \item \textbf{Token Type:} A word-like unit of text.  Includes a
-  lexeme, and may also include other info, like part of speech tag,
-  etc.
-
-  \item \textbf{Token:} An occurance of a token type
-
-  \item \textbf{Location:} The location of something in text.  There
-  are actually a number of ways of representing locations: as indexes,
-  as spans, etc.  Also, indexes can refer to word number, character
-  number, etc.
-
-\end{itemize}
-
-\section{Syntax Tree Stuff}
-
-\begin{itemize}
-
-  \item \textbf{SyntaxNode:} The annotation that appears at a node
-  in a syntax tree.  E.g., this might hold a phrase type (``NP'' or
-  ``V'').
-
-  \item \textbf{SyntaxTree:} An occurance of a tree or subtree.
-  Consists of a SyntaxNode and an ordered list of zero or more
-  children.  Children must be SyntaxTrees or Tokens.
-
-  \item \textbf{SyntaxTreeType:} The ``type'' of a syntax tree (as
-  opposed to an occurance of a syntax tree).  Consists of a SyntaxNode
-  and an ordered list of zero or more children.  Children must be
-  SyntaxTreeTypes or TokenTypes.
-
-\end{itemize}
-
-\section{How much to enforce?}
-
-So basically, the question of deciding how strictly to enforce these
-distinctions comes down to two questions:
-
-\begin{enumerate}
-  \item Should any entities/distinctions be eliminated?
-  \item Should it be possible to use entities in environments that
-        are not \emph{strictly speaking} correct?
-\end{enumerate}
-
-I'll leave the first question for later, and concentrate on the
-second.  There are a number of ways that objects could try to be more
-friendly when you use them in ways that that are not strictly speaking
-correct:
-
-\begin{itemize}
-  \item Equality tests: What happens when you compare a string to a
-  token?  A string to a TokenType?  A Token to a TokenType?  etc.?  A
-  related question is how to handle comparisons within one class
-  hierarchy.  E.g., equality tests between a tagged token and a simple
-  token.  The strict view would say that any of these comparisons
-  should raise an exception.  However, if we decide not to be strict,
-  there are a number of places where these equality tests could be
-  well-defined.  Examples that come to mind are:
-  \begin{itemize}
-    \item equality between a string and a SimpleTokenType
-    \item equality between a string and a Token (ignore location)
-    \item equality between an integer and a Location
-    \item equality between a list and a SyntaxTree(Type)
-  \end{itemize}
-
-  \item Arguments to methods: Should methods attempt to handle
-  arguments that are not quite the expected type?  E.g., when Steven
-  passed strings to a FreqDist, when it was really (sort of) expecting
-  TokenTypes.
-
-  \item Arguments to constructors: Should constructors automatically
-  ``wrap'' objects if they're not the right type?  E.g., should
-  Token(``cat'') automatically create a SimpleTokenType containing
-  ``cat''?
-\end{itemize}

### 2001-06-22 technical/printing/printing.tex
Wrote a first draft of the printing report
svn/trunk@53

@@
 
+% One thing that guides peoples intuitions about what an object is is
+% how it looks when it prints. If two things look the same when you
+% print them, then intuitively they should be the same. I think this may
+% have been a source of confusion in the first problem set, where Tokens
+% look just like strings when you print them, so users will be tempted
+% to do things like use == to test whether a TokenType is equal to a
+% string.
+
 %============= INTRODUCTION =============%
@@
 
+  The printed representations of an object guides users' intuitions
+  about the content, function, and purpose of the object.  Often, a
+  user's first encounter with an object is with its printed
+  representation.  Users rely on this representation to let them know
+  what type of object it is, and what its important features are.  The
+  printed representations of each type of object defined by the NLP
+  toolkit must be carefully considered.
+
+%============= GOALS =============%
+\section{Goals}
+
+  The following goals guide the design of printed representations.
+
+  \begin{description}
+
+    \item[Compactness] Printed representations should be simple and
+    compact.  Simpler representations are easier to read and easier to
+    understand, especially when the object is included in the
+    representation of a larger data structure, such as a list of
+    objects.
+
+    \item[Consistancy] The printed representation of an object should
+    be consistant between different environments.  Inconsistant
+    representations can produce confusion about object values and
+    object types.
+
+    \item[Composition] The representation of an object should indicate
+    what data fields it abstractly contains.  In particular, for each
+    data field abstractly contained by an object, its representation
+    should include some indication that the field is contained within
+    it.
+
+    \item[Type Indication] The representation should give some direct
+    or indirect indication of its type.  In particular, it should be
+    easy to tell the type of an object from its printed
+    representation.
+
+    \item[Value Indication] If two objects have the same value, they
+    should have the same printed representations; if two objects have
+    different values, then they should have different printed
+    representations.
+
+    \item[Intuitiveness] Representations should rely on the
+    conventional meanings of formats and symbols (e.g., the use of
+    ``@'' to denote ``at'').  This will make it easier for users to
+    understand what the representations stand for.
+
+    \item[Standard Representations] When possible, established
+    standard representations (such as ``dog/NP'' for tagging) should
+    be used.
+
+  \end{description}
+
+%============= ISSUES =============%
+\section{Issues}
+
+The representations are still under development.  This section
+describes some of the issues involved in designing representations.  I
+may not use the solutions discussed here.
+
+\subsection{Context-Dependant Representations}
+
+  Representations that show the complete state of an object are useful
+  for certain tasks, but they can be overly verbose for complex
+  objects, or for lists of objects, making it hard to understand the
+  structures shown by the printed representations.  In particular,
+  complete representations of complex objects are not \emph{compact}.
+
+  The principle of \emph{composition} requires that the representation
+  display each field of an object; but it may be useful to display a
+  ``summary'' of some of the fields, instead of a complete
+  representation.  For example, a \texttt{Location} whose unit is
+  \texttt{"token"} and whose source is \texttt{"foo.txt"} might
+  normally be represented as:
+
+  \begin{quote}
+    \texttt{@[token 1:token 3]@foo.txt}
+  \end{quote}
+
+  But if a large list of tokens is displayed, each with a location,
+  then this representation may be overly verbose.  A ``summary'' of
+  this location might be:
+
+  \begin{quote}
+    \texttt{@[1:3]}
+  \end{quote}
+
+  Note that this solution is somewhat contrary to \emph{consistancy}:
+  we are using different representations for the same object in
+  different environments.  However, if summary representations are
+  chosen carefully enough, I don't believe that this will be a
+  problem.
+
+\subsubsection{\texttt{str} and \texttt{repr}}
+
+  The \texttt{str} and \texttt{repr} functions could be used to
+  implement the difference between full representations and summary
+  representations.  Officially, \texttt{str} gives the ``informal
+  representation'' of an object and \texttt{repr} gives its ``formal
+  representation.''  However, one of the main differences between them
+  in practice is that \texttt{str} is used when a single object is
+  printed, but \texttt{repr} is used when an object is printed as part
+  of a list, tuple, or dictionary.
+
+  We could extend this distinction to the toolkit's objects, using
+  \texttt{str} to encode an object's top-level representation, and
+  \texttt{repr} to encode its summary.  The advantage of this approach
+  is that it guarantees that each object has exactly one summary
+  representation and one full representation.  For some objects, these
+  may be the same representation.
+
+\subsection{Refactoring Information}
+
+  Unfortunately, summary representations don't always seem like the
+  right tool to make representations more compact.  In a number of
+  cases, the representation of a complex object may contain a large
+  amount of redundant information.  It may make sense to display this
+  information exactly once.  e.g., syntax trees: how to attatch
+  locations?
+
+%============= PRINTING REPS =============%
+\section{Printing Representations}
+
+Here's a list of how I'm currently representing things:
+
+\newpage
+\begin{tabbing}
+\tt
+\hspace{3in}\=\\
+\textbf{$>>>$ print Location(1, 3)}\\
+@[1:3]\\
+\textbf{$>>>$ print Location(1, 2)}\\
+@[1]\\
+\textbf{$>>>$ print Location(1, 3, source='foo.txt')}\\
+@[1:3]@'foo.txt'\\
+\textbf{$>>>$ print Location(1, 3, unit='word')}\\
+@[word 1:word 3]\\
+\textbf{$>>>$ print Location(1, 2, unit='word', source='foo.txt')}\\
+@[word 1:word 2]\\
+\\
+\textbf{$>>>$ print Token('hi', Location(1, 2))}\\
+'hi'@[1]\\\
+\textbf{$>>>$ print Token('hi', Location(1, 3))}\\
+'hi'@[1:3]\\\
+\textbf{$>>>$ print Token('hi', Location(1, 3, unit='word', source='foo.txt'))}\\
+'hi'@[1:3] \>\textrm{Loc's unit and source are not shown}\\
+\textbf{$>>>$ }\\
+\\
+\textbf{$>>>$ print Tree('np', 'the', 'boy')}\\
+('np': 'the' 'boy') \>\textrm{No commas and presence of ':' distinguises from tuple}\\
+\textbf{$>>>$ print Tree('vp', 'likes', Tree('np', 'the', 'boy'))}\\
+('vp': 'likes' ('np': 'the' 'boy'))\\
+\textbf{$>>>$ print TreeToken('vp', Token('likes',1), Tree('np', Token('John',2)))}\\
+('vp': 'likes' ('np': 'John))@[1:3]\\
+
+\end{tabbing}
+
+%============= RANDOM NOTES =============%
+\section{Random Notes}
+
+\subsection{Issues}
+
+There are a number of considerations when deciding how to represent
+objects:
+
+\begin{itemize}
+
+  \item Users should be able to tell what type of object something is
+    by printing it.  In particular:
+    \begin{itemize}
+      \item It should be possible to distinguish an object from any
+        built-in type (string, list, etc)
+      \item It should be possible to distinguish each of the different
+        classes defined by the toolkit from each other.
+    \end{itemize}
+
+    Of course, this is within reason -- a string can look like any
+    object, for example.
+
+  \item How much to display?  Representations that show the complete
+  state of an object can be useful, but they can be overly verbose for
+  large objects, or for lists of objects, making it hard to understand
+  the structures shown by the printed representations.
+
+  \item One rep per object, or not?  In particular, should there be a
+  different embedded representation?  Difference between str and repr
+  as embedding.
+
+  \item Representations should be intuitive
+
+  \item Representations should be simple
+
+  \item Should representations be readable?
+
+  \item Should standard reps be used (eg, treebank)
+
+  \item Should some classes be treated specially (eg, quote strings)
+
+\end{itemize}
+
 \end{document}

### 2001-12-07 technical/printing/printing.tex
Added copyright/license notices to all files
svn/trunk@82
- Revised design goals - Added section discussing str vs repr, etc. - Added a list of guidelines (incomplete)
svn/trunk@403

@@
+% Natural Language Toolkit Technical Report:
+% Representations for Printing
 %
-% Natural Language Toolkit for Python
-% Representations for Printing
-% Edward Loper
+% Copyright (C) 2001 University of Pennsylvania
+% Author: Edward Loper <edloper@gradient.cis.upenn.edu>
+% URL: <http://nltk.sf.net>
+% For license information, see LICENSE.TXT
 %
-% Created [05/27/01 11:44 PM]
 % $Id$
-%
 
@@
 
+  \subsection{Primary Goals}
+
   \begin{description}
 
-    \item[Compactness] Printed representations should be simple and
-    compact.  Simpler representations are easier to read and easier to
-    understand, especially when the object is included in the
-    representation of a larger data structure, such as a list of
-    objects.
+    \item[Type Indication] The representation should give some direct
+    or indirect indication of its type.  In particular, it should be
+    easy to tell the type of an object from its printed
+    representation.
+
+    \item[Intuitiveness] Representations should rely on the
+    conventional meanings of formats and symbols (e.g., the use of
+    ``@'' to denote ``at'').  This will make it easier for users to
+    understand what the representations stand for.
+
+    \item[Standard Representations] When possible, established
+    standard representations (such as ``dog/NP'' for tagging) should
+    be used.
 
@@
 
-    \item[Type Indication] The representation should give some direct
-    or indirect indication of its type.  In particular, it should be
-    easy to tell the type of an object from its printed
-    representation.
+  \end{description}
+
+  \subsection{Secondary Goals}
+
+  \begin{description}
 
@@
 
-    \item[Intuitiveness] Representations should rely on the
-    conventional meanings of formats and symbols (e.g., the use of
-    ``@'' to denote ``at'').  This will make it easier for users to
-    understand what the representations stand for.
-
-    \item[Standard Representations] When possible, established
-    standard representations (such as ``dog/NP'' for tagging) should
-    be used.
+    \item[Compactness] Printed representations should be simple and
+    compact.  Simpler representations are easier to read and easier to
+    understand, especially when the object is included in the
+    representation of a larger data structure, such as a list of
+    objects.
 
   \end{description}
+
+%============= PYTHON =============%
+\section{Python Constructs}
+
+  Python classes may define two different string representations: one
+  is returned by \texttt{str}, and the other by \texttt{repr}.  The
+  differences between these two represenetations are:
+
+  \begin{itemize}
+
+    \item The ``Official'' Definitions
+    (\texttt{http://www.python.org/doc/current/ref/customization.html})
+    \begin{itemize}
+      \item 
+        \texttt{repr}: If at all possible, this should look like a
+        valid Python expression that could be used to recreate an
+        object with the same value (given an appropriate
+        environment). If this is not possible, a string of the form
+        "$<$...some useful description...$>$" should be returned.
+        This is typically used for debugging, so it is important that
+        the representation is information-rich and unambiguous.
+
+      \item
+        \texttt{str}: This differs from \_\_repr\_\_() in that it does not
+        have to be a valid Python expression: a more convenient or
+        concise representation may be used instead. 
+    \end{itemize}
+
+    \item Usage Contexts
+    \begin{itemize}
+
+      \item \texttt{repr}
+      \begin{itemize}
+        \item The result of \texttt{repr()}
+        \item The result of backquotes (e.g., \texttt{`x`})
+        \item The result of the string formatting operator
+            ``\texttt{\%r}''
+        \item Used when an object representation is ``contained'' in
+            another object representation (e.g., to display an element
+            of a list or a key of a dictionary)
+      \end{itemize}
+
+      \item \texttt{str}
+      \begin{itemize}
+        \item The result of \texttt{str()}
+        \item The result of the string formatting operator 
+            ``\texttt{\%s}''
+        \item The output of print
+      \end{itemize}
+
+    \end{itemize}
+
+    \item Builtin Type Usage Conventions
+    \begin{itemize}
+      \item \texttt{repr(x)} contains no newlines.
+      \item \texttt{repr(x)} contains only simple ASCII characters.
+    \end{itemize}
+
+    \item Default Values
+    \begin{itemize}
+      \item The default value for \texttt{repr} is a string of the form:\\
+        \texttt{$<$\_\_\textit{module}\_\_.\textit{class} instance at \textit{address}$>$}
+      \item The default value for \texttt{str} is the value returned
+        by \texttt{repr}
+    \end{itemize}
+
+  \end{itemize}
+
+  \subsection{The Official Definition}
+
+    After consideration of the contexts in which \texttt{repr} and
+    \texttt{str} are used, we have decided \emph{not} to make use of
+    the ``official'' definitions.  This decision is primarily
+    motivated by the fact that \texttt{repr} is always used when an
+    object representation is ``contained'' in another object
+    representation.  We believe that this would lead to overly verbose
+    and difficult to read representations of complex objects.  For
+    example, if we adopted the ``official'' definition of
+    \texttt{repr}, then the string representation (using \texttt{repr}
+    \emph{or} \texttt{str}) of a list of tokens would look something
+    like:
+
+\begin{verbatim}
+"[Token('The', Location(0, 1, unit='w', source='foo.txt')), Token('dog', Lo
+cation(1, 2, unit='w', source='foo.txt')), Token('saw', Location(2, 3, unit
+='w', source='foo.txt')), Token('the', Location(3, 4, unit='w', source='foo
+.txt')), Token('man', Location(4, 5, unit='w', source='foo.txt'))]"
+\end{verbatim}
+
+    We believe that this is signifigantly less readable than the
+    result of using a \texttt{repr} which does not evaluate to return
+    its original value:
+
+\begin{verbatim}
+"['The'@[0w]@'foo.txt', 'dog'@[1w]@'foo.txt', 'saw'@[2w]@'foo.txt', 'the'@[
+3w]@'foo.txt', 'man'@[4w]@'foo.txt']"
+\end{verbatim}
+
+  \subsection{str vs repr}
+
+    Instead, we will make the following distinction between the
+    \texttt{str} and \texttt{repr} string representations:
+
+    \begin{itemize}
+
+      \item \texttt{repr} gives a single-line representation, which
+          would be suitable for inclusion in the representation of
+          another object (i.e., it is delimited by quotes, brackets,
+          braces, etc., when appropriate).  This representation should
+          contain only standard ASCII characters.  \texttt{repr}
+          should be defined for all classes.
+
+      \item \texttt{str} can optionally give a more human-readable
+          representation of an object.  This representation may span
+          multiple lines.
+    \end{itemize}
+
+\section{Guidelines}
+
+  The following guidelines are intended to help design string
+  representations for new nltk objects.  In these guidelines, the
+  string representation of an object is refered to as \texttt{s}.
+
+  \subsection{General Guidelines}
+    \begin{itemize}
+      \item All classes should implement \texttt{repr}.  When the
+        output of \texttt{repr} may be difficult to read, \texttt{str}
+        should be implemented.
+      \item If a field has no value (or the default value), it may be
+        ommited from the printed representation.  (E.g.,
+        \texttt{Location}s with no source don't display the source.
+      \item When \texttt{s} needs to contain the representation of an
+        embedded object, and the object's type is variable,
+        \texttt{repr} should always be used (not \texttt{str}).  No
+        special type-checking should be applied (e.g., to treat
+        strings specially).
+      \item It should always be possible to tell the type of an object
+          from its printed representation.
+    \end{itemize}
+
+  \subsection{Repr}
+    \begin{itemize}
+      \item \texttt{s} may not span multiple lines (i.e., it may not
+        contain newlines)
+    \end{itemize}
+
+  \subsection{Str}
+    \begin{itemize}
+      \item \texttt{s} may span multiple lines.
+      \item Multiline \texttt{str} representations shoud not contain any
+        trailing newlines.
+      \item \texttt{s} may use indentation to show internal structure.
+      \item \texttt{s} may be a simple table.
+      \item If \texttt{s} is a multiline representation, it should try 
+        very hard not to be more than 75 characters wide.
+      \item It should always be possible to distinguish the
+          \texttt{s} from the \texttt{str} for any built-in object
+          (except strings, which can have arbitrary representations);
+          or for any other class.
+      \item For multiline \texttt{str} representations, \texttt{s}
+        should begin with an unindented line, whose first word is the
+        name of the class.  All subsequent lines should be indented.
+    \end{itemize}
+
+  \subsection{Data-Oriented Classes}
+    \begin{itemize}
+      \item When practical, \texttt{s} should specify the complete
+          state of the object.  In particular, if it is possible to
+          give the complete state of a typical instance in less than
+          75 characters, then \texttt{s} should give the complete
+          state.  
+      \item When it is not practical to show the complete state
+          of a data-oriented instance, this should be indicated with
+          elipsis dots (``...'').
+    \end{itemize}
+
+  \subsection{Task-Oriented Classes}
+    \begin{itemize}
+      \item For repr, \texttt{s} should have the form:\\
+        \texttt{$<$\textit{class} \textit{info}$>$}\\
+        where \texttt{\textit{info}} is a description of the
+        parameters and the state of the class.
+    \end{itemize}
+
+%============= PP =============%
+\section{Pretty-Printing Methods}
+
+  Classes may provide pretty-printing methods.  These methods should
+  have the following signature:
+
+  \texttt{pp(self, left, right)}
+
+  Where \texttt{left} and \texttt{right} are left and right margins.
+  \texttt{pp} should assume that it is already at position
+  \texttt{left}.  E.g., a pp function to draw a vertical line, when
+  given \texttt{left}=5, would return:
+
+  \texttt{"|" + ("~~~~~|"*lineheight)}
+
+%============= EXAMPLES =============%
+\section{Examples}
+
+\subsection{Locations}
+
+  \texttt{str} and \texttt{repr} representations are identical.  Unit
+  \& Loc are not displayed if they are not specified.  End index is not
+  displayed if it has the default value (=start+1).
+
+{\tt
+\begin{tabbing}
+\emph{$>>>$ Location(1, 3)}\\
+@[1:3]\\
+\emph{$>>>$ Location(1, 2)}\\
+@[1]\\
+\emph{$>>>$ Location(1, 3, source='foo.txt')}\\
+@[1:3]@'foo.txt'\\
+\emph{$>>>$ Location(1, 3, unit='w')}\\
+@[1w:3w]\\
+\emph{$>>>$ Location(1, 2, unit='w', source='foo.txt')}\\
+@[1w]@'foo.txt'\\
+\emph{$>>>$ Location(1, 3, unit='w', source='foo.txt')}\\
+@[1w:3w]@'foo.txt'
+\end{tabbing}
+}
+
+\subsection{Tokens}
+
+  \texttt{str} and \texttt{repr} representations are identical.
+  Location is included using \texttt{repr}.  If location is
+  unspecified, it is displayed as \texttt{@[?]}.  (Note that this
+  would never be a valid location by itself.)
+
+{\tt
+\begin{tabbing}
+\emph{$>>>$ Token('x', Location(1, 3, unit='w', source='foo.txt')}\\
+'x'@[1w:3w]@'foo.txt'\\
+\emph{$>>>$ Token('x')}\\
+'x'@[?]
+\end{tabbing}
+}
+
+\subsection{Trees}
+
+  Lack of commas, and presence of ':', distinguises trees from tuples.
+
+{\tt
+\begin{tabbing}
+\emph{$>>>$ Tree('np', 'the', 'boy')}\\
+('np': 'the' 'boy') \\
+\emph{$>>>$ Tree('vp', 'likes', Tree('np', 'the', 'boy'))}\\
+('vp': 'likes' ('np': 'the' 'boy'))\\
+\emph{$>>>$ TreeToken('vp', Token('likes',1), TreeToken('np', Token('John',2)))}\\
+('vp': 'likes' ('np': 'John))@[1:3]\\
+\\
+\textrm{Trees have one-line repr representations, \& multiline str representations:}\\
+\emph{$>>>$ Tree('x', 'a'*40, Tree('b', 'c'*40, Tree('d', 'e'*40)))}\\
+('x': 'aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa' ('b': 'cccccccccccccccccccccccc\\
+cccccccccccccccc' ('d': 'eeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeee')))\\
+\emph{$>>>$ print Tree('x', 'a'*40, Tree('b', 'c'*40, Tree('d', 'e'*40)))}\\
+('x':\\
+\qquad  'aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa'\\
+\qquad  ('b':\\
+\qquad\qquad    'cccccccccccccccccccccccccccccccccccccccc'\\
+\qquad\qquad    ('d': 'eeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeeee')))
+\end{tabbing}
+}
+
+\subsection{SimpleFreqDists}
+
+{\tt
+\begin{tabbing}
+\emph{$>>>$ s = SimpleFreqDist(\ldots)}\\
+\{SimpleFreqDist $<$CFSample 1, '7'$>$:1, $<$CFSample 5, '2'$>$:1, $<$CFSample 3, '2'$>$\\
+:1, $<$CFSample 2, '7'$>$:1, $<$CFSample 7, '2'$>$:2, $<$CFSample 9, '7'$>$:2, $<$CFSampl\\
+e 6, '7'$>$:3, $<$CFSample 5, '7'$>$:1, $<$CFSample 3, '7'$>$:1, $<$CFSample 9, '9'$>$:2,\\
+ $<$CFSample 7, '9'$>$:3, $<$CFSample 8, '5'$>$:5\}\\
+\emph{$>>>$ print s}\\
+\{SimpleFreqDist\\
+\qquad $<$CFSample 1, '7'$>$: 1, $<$CFSample 5, '2'$>$: 1, $<$CFSample 3, '2'$>$: 1, \\
+\qquad $<$CFSample 2, '7'$>$: 1, $<$CFSample 7, '2'$>$: 2, $<$CFSample 9, '7'$>$: 2, \\
+\qquad $<$CFSample 6, '7'$>$: 3, $<$CFSample 5, '7'$>$: 1, $<$CFSample 3, '7'$>$: 1, \\
+\qquad $<$CFSample 9, '9'$>$: 2, $<$CFSample 7, '9'$>$: 3, $<$CFSample 8, '5'$>$: 5\}
+\end{tabbing}
+}
 
@@
 
-%============= PRINTING REPS =============%
-\section{Printing Representations}
-
-Here's a list of how I'm currently representing things:
-
-\newpage
-\begin{tabbing}
-\tt
-\hspace{3in}\=\\
-\textbf{$>>>$ print Location(1, 3)}\\
-@[1:3]\\
-\textbf{$>>>$ print Location(1, 2)}\\
-@[1]\\
-\textbf{$>>>$ print Location(1, 3, source='foo.txt')}\\
-@[1:3]@'foo.txt'\\
-\textbf{$>>>$ print Location(1, 3, unit='word')}\\
-@[word 1:word 3]\\
-\textbf{$>>>$ print Location(1, 2, unit='word', source='foo.txt')}\\
-@[word 1:word 2]\\
-\\
-\textbf{$>>>$ print Token('hi', Location(1, 2))}\\
-'hi'@[1]\\\
-\textbf{$>>>$ print Token('hi', Location(1, 3))}\\
-'hi'@[1:3]\\\
-\textbf{$>>>$ print Token('hi', Location(1, 3, unit='word', source='foo.txt'))}\\
-'hi'@[1:3] \>\textrm{Loc's unit and source are not shown}\\
-\textbf{$>>>$ }\\
-\\
-\textbf{$>>>$ print Tree('np', 'the', 'boy')}\\
-('np': 'the' 'boy') \>\textrm{No commas and presence of ':' distinguises from tuple}\\
-\textbf{$>>>$ print Tree('vp', 'likes', Tree('np', 'the', 'boy'))}\\
-('vp': 'likes' ('np': 'the' 'boy'))\\
-\textbf{$>>>$ print TreeToken('vp', Token('likes',1), Tree('np', Token('John',2)))}\\
-('vp': 'likes' ('np': 'John))@[1:3]\\
-
-\end{tabbing}
-
 %============= RANDOM NOTES =============%

### 2001-12-13 technical/classification/classification.tex
- Initial import: text classification technical report
svn/trunk@442

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{amsmath}

\title{NLTK: Text Classification}
\author{Edward Loper}

\begin{document}
\maketitle

\section{Introduction}

  The Natural Language Toolkit is a pedagogical package designed to
  help students get hands-on experience with using, designing, and
  implementing natural language processing systems.  It is organized
  as a suite of modules, each of which provides basic data types,
  tools, and interfaces for a particular natural language processing
  task.

  This report gives an overview of the design and implementation of
  the text classification module.  This module is used to implement
  systems that classify texts into categories.  It focuses on the
  problem of \emph{single-category text classification}, in which:

  \begin{itemize}
    \item The set of categories is known.
    \item The number of categories is finite.
    \item Each text belongs to exactly one category.
  \end{itemize}

  This report assumes that the reader is familiar with text
  classification, feature-based classification, and feature selection.

\section{Documentation}

  The text classification module is accompanied by extensive
  documentation.

  \begin{itemize}

    \item The \emph{Text Classification Tutorial} provides a gentle
    introduction to the text classification module.  It uses concrete
    examples to explain and motivate the data types, tools, and
    interfaces that are used for text classification.

    \texttt{http://nltk.sourceforge.net/tutorial/classifying/t1.html}

    \item The \emph{Text Classification Reference Documentation} gives
    precise and detailed descriptions of every sub-module, interface,
    class, method, and function defined by the text classification
    module.

    \texttt{http://nltk.sourceforge.net/ref/nltk.classifier.html}

  \end{itemize}

\section{Architecture}

  The text classification module currently consists of five
  sub-modules:

  \begin{itemize}

    \item \texttt{nltk.classifier} defines the basic data types and
    interfaces that are used to define the text classification task
    and the classifier training task.

    \item \texttt{nltk.classifier.feature} defines data types,
    classes, and interfaces used for feature-based classification.
    Features provide a standard way of encoding the information used
    to make classification decisions.  This standard encoding allows
    the same classifier to be used to solve many different problems.

    \item \texttt{nltk.classifier.featureselection} defines classes
    and interfaces used to select which features are relevant for a
    given classification task.

    \item \texttt{nltk.classifier.naivebayes} defines a text
    classifier model based on the Naive Bayes assumption; and a
    classifier trainer to build Naive Bayes classifiers from training
    data.

    \item \texttt{nltk.classifier.maxent} defines a text classifier
    model based on the maximum entropy modeling framework; and a
    classifier trainer to build maximum entropy classifiers from
    training data.

  \end{itemize}

  This report focuses on the text classification module's \emph{core
  architecture}, consisting of the first three modules.  For more
  information on \texttt{nltk.classifier.naivebayes} or on
  \texttt{nltk.classifier.maxent}, see the tutorial documentation and
  the reference documentation.

  \subsection{Basic Data Types}

    The classification module defines two basic data types:

    \begin{itemize}

      \item A \textit{Label} is a unique identifier for a category.

      \item A \textit{LabeledText} pairs a \textit{Label} with a
      \textit{TextType}.

    \end{itemize}

    \textit{Label} is an abstract data type: it is not directly
    encoded in Python.  The only restrictions on \textit{Labels} are
    that they must be immutable; and they must be comparable for
    equality and inequality.  For example, numbers, strings, tuples,
    and instances can all be used as labels.  Defining \textit{Labels}
    as abstract types provides the user with maximal flexibility in
    their use of the classification module.  The restriction that
    \textit{Labels} be immutable allows them to be used as dictionary
    keys; and ensures that \textit{LabeledTexts} can be used as
    \texttt{Token} types\footnote{\texttt{Token} types are required to
    be immutable.}.  The restriction that \textit{Labels} be
    comparable allows us to differentiate labels.

    The \textit{LabeledText} data type is implemented with the
    \texttt{LabeledText} class, which pairs a \textit{TextType} with
    a \textit{Label}.  The following list summarizes the methods
    provided by the \texttt{LabeledText} class:

    \begin{itemize}

      \item The \textit{LabeledText} constructor builds a
      \texttt{LabeledText} from a given \textit{TextType} and
      \textit{Label}.

      \item The \texttt{text} method returns the
      \texttt{LabeledText}'s \textit{TextType}.

      \item The \texttt{label} method returns the
      \texttt{LabeledText}'s \textit{Label}.

      \item Comparison operators test \texttt{LabeledText}s for
      equality and inequality.

      \item The hash operator allows \texttt{LabeledText}s to be used
      as dictionary keys.

      \item The string representation operator is used to print
      \texttt{LabeledText}s.

    \end{itemize}

    \noindent
    \textit{LabeledTexts} are a considered a sub-type of the abstract
    data type \textit{TextType}.

    We chose not to define a data type to represent categories
    themselves; instead, categories are identified via
    \textit{Labels}.  This decision was based on the fact that we
    could think of no common interface for categories that would
    provide more functionality than labels.  However, it remains
    possible to add a \textit{Category} data type, if it becomes
    apparent that such a data type would be useful.  Since the only
    restriction on \textit{Labels} is that they be immutable, it would
    be possible to use \textit{Labels} containing pointers to their
    categories; these could be used to navigate from a
    \textit{LabeledText} to its \textit{Category}.

  \subsection{Text Classification}

    The text classification module defines the \textit{Classifier}
    interface, which requires that text classifiers implement two
    methods:

    \begin{itemize}

      \item \texttt{labels} returns the list of category labels that
      can be assigned by the \textit{Classifier}.

      \item \texttt{classify} determines which label is most
      appropriate for a given text token, and return a
      \texttt{LabeledText} token constructed from the given text token
      and the chosen label.

    \end{itemize}

    \noindent
    \textit{Classifiers} are also encouraged to implement the
    following methods:

    \begin{itemize}

      \item \texttt{prob} takes a \texttt{LabeledText} token, and
      returns the probability that the text is assigned to the correct
      category.

      \item \texttt{distribution} takes an unlabeled token, and
      returns a probability distribution over \texttt{LabeledText}
      tokens, indicating the likelihood of each potential
      classification decision.

      \item \texttt{distribution\_dictionary} takes an unlabeled token,
      and returns a dictionary mapping from each category's label to
      the probability that the token is a member of that category.

      \item \texttt{distribution\_list} takes an unlabeled token, and
      returns a list specifying the likelihood that it is a member of
      each category.  The $i^{th}$ element of this list corresponds to
      the $i^{th}$ label returned by \texttt{labels}.
    \end{itemize}

    The two required methods were chosen because they represent the
    functionality which we believe that any \textit{Classifier} should be
    capable of providing.  The remaining methods are optional, since
    some classification algorithms cannot generate probability
    estimates.

    The methods \texttt{prob}, \texttt{distribution}, and
    \texttt{distribution\_dictionary} are provided ensure that users
    have convenient access to probability estimates.
    \texttt{distribution\_list} is included to allow efficient access
    to these estimates (e.g., for classifier training algorithms).

    Although the \textit{Classifer} interface defines a large number
    of methods, each \textit{Classifier} does not need to implement
    each method.  Instead, an abstract base class can be used to
    provide default definitions for most methods.  In particular, the
    methods \texttt{labels} and \texttt{distribution\_list} can be used
    to define default definitions for every other method.  These
    methods only need to be implemented directly when desired (e.g, to
    improve efficiency).

    We are currently considering adding another optional method, which
    would take an unlabeled token, and return a corresponding list of
    \texttt{LabeledText} tokens, sorted in descending order of
    likelihood.  This method would be useful for classifiers that can
    decide which categories are more likely than which other
    categories; but which cannot generate probability estimates.  An
    abstract base class could provide a default definition for this
    method, based on \texttt{distribution\_list}.

  \subsection{Classifier Training}

    Typically, \textit{Classifiers} encode specific classifier models;
    but do not include the algorithms for training the classifiers.
    Instead, \textit{ClassifierTrainers} are used to generate
    classifiers from training data.  The \textit{ClassifierTrainer}
    interface requires that classifier trainers implement a single
    method: \texttt{train}.  This method takes a list of
    \texttt{LabeledText} tokens, and returns a new
    \textit{Classifier}.

    The \texttt{train} method should also accept an optional keyword
    argument ``\texttt{labels},'' which specifies the set of category
    labels that the classifier should use.  When this keyword argument
    is not used, the \textit{ClassifierTrainer} should use the set of
    labels that are attested in the given list of \texttt{LabeledText}
    tokens.

    \textit{ClassifierTrainers} can use arguments to their
    constructors to provide additional information for the training
    algorithm.  For example, the constructor for the naive Bayes
    classifier trainer expects a feature detector list.
    \textit{ClassifierTrainers} can also define optional keyword
    arguments to the \texttt{train} method to control parameters of
    the training algorithm.  For example, the maximum entropy
    classifier trainers accept keyword arguments that determine how
    long the algorithm should iterate.

    We deliberately kept the \textit{ClassifierTrainer} interface
    simple, to ensure that it can be used to define a wide variety of
    classifier training algorithms.

  \subsection{Evaluation Metrics}

    The text classification module defines two functions for
    evaluating the performance of a \textit{Classifier}:

    \begin{itemize}

      \item \texttt{accuracy} takes a \textit{Classifier} and a list
      of \texttt{LabeledText} tokens, and returns the percentage of
      the texts which the \textit{Classifier} correctly categorizes.

      \item \texttt{log\_likelihood} takes a \textit{Classifier} and a
      list of \texttt{LabeledText} tokens, and evaluates the per-token
      log likelihood of the list of tokens for the \textit{Classifer}.
      This value $LL(classifier, toks)$ is defined as:

      $$LL(classifier, toks) = \frac{1}{|toks|}\sum_{tok \in toks} 
        \log P_{classifier}(label_{tok}|type_{tok})$$
    \end{itemize}

    We may add more evaluation metrics in the future.

  \subsection{Feature-Based Classifiers}

    Most text classification algorithms do not depend on the specific
    details of the task being performed. By defining a standard
    encoding for the information used to make classification
    decisions, we can allow \textit{Classifiers} to abstract away from
    the details of particular tasks.  This allows us to use the same
    \textit{Classifier} to perform many different classification
    tasks; and to solve the same classification task with many
    different \textit{Classifiers}.

    \subsubsection{Features}

      There are many possible ways of encoding the information that is
      relevant to classification.  We decided to use ``features'' to
      encode this information, since they are simple, flexible, and
      fairly standard.  A \emph{feature} specifies some aspect of a
      \texttt{LabeledText} that is relevant to deciding how likely
      that \texttt{LabeledText} is to occur.  These features can be
      used by classification algorithms to examine how likely
      different labels are for a given text.  Features can be
      expressed as functions mapping from \texttt{LabeledTexts} to
      values.

    \subsubsection{Contexts}

      We considered several alternative encodings for the information
      that is relevant to classification.  One promising possibility
      is sometimes called ``contexts'' or ``aspects.''\footnote{To my
      knowledge, this encoding does not have a standard name.}  A
      \emph{context} specifies some aspect of a \textit{TextType} that
      is relevant to classification.  Contexts can be expressed as
      functions mapping from \textit{TextTypes} to values.

      For many problems, contexts can be expressed in terms of
      features.  In particular, if $c$ is a context function, then for
      some problems we can think of $c$ as shorthand for a set of 
      features $f_1, f_2, \ldots, f_L$ (where $L$ is the set of labels):

      \begin{equation}
        f_i(lt) = \bigg\{ \begin{array}{cl} 
                 c(lt_{text}) & \qquad \textrm{if } lt_{label}=L_i \\
                 0 & \qquad \textrm{otherwise} \end{array}
      \end{equation}

      We decided not to use contexts for our standard information
      encoding, since features are more powerful.  However, we leave
      open the possibility of introducing alternative encodings (such
      as contexts) to the classification module in the future.

    \subsubsection{Feature Data Types}

      The feature sub-module defines three data types:

      \begin{itemize}

        \item A \textit{FeatureID} is a unique identifier for a feature.

        \item A \textit{FeatureValue} is a value that can be produced
        by a feature.

        \item An \textit{Assignment} pairs a \textit{FeatureID} with a
        \textit{FeatureValue}.
      \end{itemize}

      \textit{FeatureIDs} are bounded non-negative integer
      identifiers.  These identifiers are used to indicate which
      feature a \textit{FeatureValue} or \textit{FeatureDetector}
      corresponds to.  The mapping from \textit{FeatureIDs} to
      \textit{FeatureValues} and \textit{FeatureDetectors} is
      specified using \textit{FeatureValueLists} and
      \textit{FeatureDetectorLists}, which are discussed below.

      \textit{FeatureValue} is an abstract data type.  The only
      restrictions that the text classification module places on
      \textit{FeatureValues} are that they must be immutable, and they
      must be comparable for equality and inequality.  However,
      feature-based \textit{Classifiers} usually put additional
      restrictions on the kinds of \textit{FeatueValues} that they
      support.  The most commonly supported kinds of
      \textit{FeatureValue} are booleans and integers.

      \textit{Assignments} pair \textit{FeatureIDs} with
      \textit{FeatureValues}.  They are used to efficiently process
      sparse \textit{FeatureValueLists}.  \textit{Assignments} are
      currently represented as two-element \texttt{tuples}, whose
      first element is a \textit{FeatureID}, and whose second element
      is a \textit{FeatureValue}.  However, we are considering
      implementing them as a class.

    \subsubsection{Feature Interfaces}

      The feature sub-module defines three interfaces:

      \begin{itemize}
        \item A \textit{FeatureDetector} encapsulates a feature.

        \item A \textit{FeatureDetectorList} provides an efficient way
        of grouping feature detectors together, and associating each
        detector with a \textit{FeatureID}.

        \item A \textit{FeatureValueList} provides an efficient way of
        grouping feature values together, and associating each value
        with a \textit{FeatureID}.
      \end{itemize}

      \paragraph{FeatureDetectors}

      \textit{FeatureDetectors} can be used to find the value of a
      feature for a given \texttt{LabeledText}.
      \textit{FeatureDetectors} are required to implement a single
      method: \texttt{detect}.  This method returns the
      \textit{FeatureValue} for a given \texttt{LabeledText}.

      Originally, \textit{FeatureDetectors} were simply called
      \textit{Features}.  However, the bare term ``feature'' can be
      confusing, since it is used in the literature to refer to both
      \textit{FeatureDetectors} and \textit{FeatureValues}.  We
      therefore decided to use the more explicit name
      \textit{FeatureDetector} instead.

      \paragraph{FeatureDetectorLists}

      \textit{FeatureDetectorLists} are data structures that represent
      a set of features.  Abstractly, a \textit{FeatureDetectorList}
      can be thought of as a \texttt{list} of
      \textit{FeatureDetectors}.  The $i^{th}$ element of this list is
      the \textit{FeatureDetector} corresponding to the
      \textit{FeatureID} $i$.  \textit{FeatureDetectorLists} are
      required to implement four methods:

      \begin{itemize}

        \item \texttt{detect} takes a \texttt{LabeledText},
        and returns a \textit{FeatureValueList} specifying the
        \textit{FeatureValue} for each feature.

        \item The length operator returns the number of features
        represented by the \textit{FeatureDetectorList}:

        \item The indexing operator allows
        \textit{FeatureDetectorLists} to be treated as \texttt{lists}
        of \textit{FeatureDetectors}.

        \item The addition operator merges the features defined by two
        \textit{FeatureDetectorLists}.

      \end{itemize}

      \noindent \textit{FeatureDetectorLists} serve three important
      functions:

      \begin{itemize}
        \item They provide a mechanism for grouping
        \textit{FeatureDetectors} together.

        \item They provide a means of associating \textit{FeatureIDs}
        with each \textit{FeatureDetector}.

        \item They allow for efficient implementations for sets of
        related \textit{FeatureDetectors}.
      \end{itemize}

      Although \textit{FeatureDetectorLists} can be abstractly thought
      of as lists of independent \textit{FeatureDetectors}, they are
      not usually implemented that way, for efficiency reasons.  Using
      a complex \textit{FeatureDetectorList} to find the
      \textit{FeatureValues} for a \texttt{LabeledText} allows us to
      examine all features in parallel.  For example, consider a
      bag-of-words feature set for a document.  Implementing this
      feature set with a list of independent \textit{FeatureDetectors}
      would require making one pass through the document for each
      feature.  But if we build a single complex
      \textit{FeatureDetectorList}, we can make one pass through the
      document, checking for all relevant words at the same time.

      In fact, the \textit{FeatureDetector} interface will almost
      never be used in practice; instead, we will always use
      \textit{FeatureDetectorLists}.  The \textit{FeatureDetector}
      interface is kept mainly for pedagogical purposes.

      \paragraph{FeatureValueLists}

      \textit{FeatureValueLists} are data structures that represent a
      set of features.  Abstractly, a \textit{FeatureValueList} can be
      thought of as a \texttt{list} of \textit{FeatureValues}.  The
      $i^{th}$ element of this list is the \textit{FeatureValue}
      corresponding to the \textit{FeatureID} $i$.

      For many classification tasks, \textit{FeatureValueLists} are
      very sparse; in other words, most of the \textit{FeatureValues}
      have some \emph{default value} (usually, zero).

      \textit{FeatureValueLists} are required to implement four
      methods:

      \begin{itemize}

        \item \texttt{assignments} returns a list of the
        \textit{Assignments} for each feature whose
        \textit{FeatureValue} is not the default value.

        \item \texttt{default} returns the \textit{FeatureValueList}'s
        default value.

        \item The length operator returns the number of features
        represented by the \textit{FeatureValueList}:

        \item The indexing operator allows
        \textit{FeatureValueLists} to be treated as \texttt{lists}
        of \textit{FeatureValues}.

      \end{itemize}

      \noindent
      \textit{FeatureValueLists} serve three important functions:

      \begin{itemize}

        \item They provide a mechanism for grouping
        \textit{FeatureValues} together.

        \item They provide a means of associating \textit{FeatureIDs}
        with each \textit{FeatureValue}.

        \item They allow for efficient encoding of sparse sets of
        \textit{FeatureValues}.

      \end{itemize}

      It is possible to process \textit{FeatureValueLists} by
      accessing each feature individually.  However, for sparse
      \textit{FeatureValueLists}, it can be significantly more
      efficient to process the list of \textit{Assignments} returned
      by the \texttt{assignments} method.  This approach is used by
      both the Naive Bayes classifier and the maximum entropy
      classifier.

      Currently, the default value is an attribute of the
      \textit{FeatureValueList}, and not of the
      \textit{FeatureDetectorList}.  However, we are considering
      adding a default value attribute to
      \textit{FeatureDetectorLists}.  This would prevent a single
      \textit{FeatureDetectorList} from producing different
      \textit{FeatureValueLists} with different default values.  It
      would also allow \textit{Classifiers} and
      \textit{ClassifierTrainers} to perform compatibility checks on
      any \textit{FeatureDetectorLists} that they are given.

    \subsubsection{AbstractFeatureClassifier}

      The feature sub-module defines
      \texttt{AbstractFeatureClassifier}, an abstract base class for
      building feature-driven classifiers.
      \texttt{AbstractFeatureClassifier} provides default definitions
      for all of the \texttt{Classifier} methods.  The only method
      that subclasses need to implement is \texttt{fvlist\_likelihood},
      which returns a float indicating the likelihood of a given
      \textit{FeatureValueList}.  This base significantly simplifies
      the design of feature-based \textit{Classifiers}.  See the
      reference documentation or the tutorial documentation for more
      details.

  \subsection{Feature Selection}

    The feature selection sub-module defines the
    \textit{FeatureSelector} interface, which requires that feature
    selectors define one method: \texttt{select}.  This method takes a
    \textit{FeatureDetectorList}, and decides which features are
    relevant for classification. It returns a new
    \textit{FeatureDetectorList} that includes the
    \textit{FeatureDetectors} for the relevant features.

    \textit{FeatureSelectors} can use arguments to their constructors
    to provide additional information for feature selection, such as a
    set of training data.  \textit{FeatureSelectors} can also define
    optional keyword arguments to the \texttt{select} method, to
    control parameters of feature selection.

    The \texttt{SelectedFDList} class provides
    \textit{FeatureSelectors} with a convenient way of building
    \textit{FeatureDetectorLists} containing the
    \textit{FeatureDetectors} for the relevant features.  A
    \texttt{SelectedFDList} consists of a subset of the feature
    detectors defined by a base \textit{FeatureDetectorList}.  See the
    reference documentation or the tutorial documentation for more
    details.

\section{Classifier Issues}

  This section describes some open issues concerning the text
  classifiers currently implemented by the text classification module.

  \subsection{Smoothing}

    The Naive Bayes classifier (\texttt{NBClassifier}) is
    paramaterized by a single probability distribution, whose samples
    are \textit{FeatureValueLists}.  It uses this probability
    distribution to find the probabilities of individual
    \textit{Assignments}.

    This probability distribution is usually constructed by first
    creating a frequency distribution over \textit{FeatureValueLists};
    and then using that frequency distribution to create a probability
    distribution class, such as an \texttt{MLEProbDist} or a
    \texttt{LaplaceProbDist}.

    However, this does not produce the desired effect.  Most of the
    probability distribution classes are designed to perform smoothing
    over \emph{samples}.  But we really want to be using smoothing
    over \emph{events}.  For example, consider a
    \texttt{LaplaceProbDist}.  This probability distribution uses the
    following formula to smooth frequency estimates:

      $P(s) = \frac{count(s) + 1}{N+B}$

    \noindent Where $s$ is a sample, $N$ is the total number of
    samples, and $B$ is the number of possible outcomes.  Using this
    estimate to find the probability of an event gives:

      $P(e) = \frac{count(e) + |e|}{N+B}$

    \noindent Now, consider the case of a frequency distribution over
    \textit{FeatureValueLists}, with 1,000 binary features.  The
    number of possible outcomes is $2^{1000}$.  As a result, the
    estimated for each assignment will be almost identical.

    NLTK's probability framework needs to be redesigned to allow
    probability distributions to perform smoothing over events, and
    not just over samples.

  \subsection{Zero Probabilities}

    Some classification algorithms begin by calculating un-normalized
    estimates of $P(l|t)$ for each label; and then normalize them by
    dividing by $\sum_lP(l|t)$.  However, for some model settings and
    some inputs, the un-normalized estimates of $P(l|t)$ may be zero
    for each label.  In this case, it is not always clear what the
    \textit{Classifier} should do.  Some alternatives are:

    \begin{itemize}
      \item Return a uniform distribution.

      \item Return a distribution that assigns a probability of zero
      to every label.

      \item ``Cancel'' zeros until at least one un-normalized estimate
      is nonzero.  For example, if the un-normalized estimates are
      $0\times2\times3$, $0\times1.8\times4$, and $0\times0\times3$, then replace these estimates with
      $2\times3$, $1.8\times4$, and $0\times3$.
    \end{itemize}

    For some classifiers, it is apparent which choice is correct.  For
    example, the maximum entropy classifier should return a uniform
    distribution, since this maximizes entropy.  However, for other
    classifiers, the choice is not so clear.

\end{document}

### 2002-03-11 technical/acl2002/acl02sub.tex
Files for ACL-02 workshop submission
svn/trunk@487

%
% File acl02sub.tex
%
% Contact: acl02-submission@cs.ualberta.ca

\documentclass[11pt]{article}
\usepackage{acl02sub}
\title{Instructions for ACL-02 Submission}
\author{
\begin{tabular}[t]{c}
Eugene Charniak\\
Computer Science Department\\
Brown University\\
Providence, RI, USA, 02912 
\end{tabular}
\hspace{6pt}
\begin{tabular}[t]{c}
Dekang Lin\\
Department of Computing Science\\
University of Alberta\\
Edmonton, Alberta, Canada, T6G 2E8
\end{tabular}
}
\summary{%
  This document is an example of what your submission manuscript to
  ACL-02 should look like.  Authors are asked to conform to the
  directions reported in this document.  See the ACL-02 call for
  papers for further instructions about the submission.}
\paperid{Pxxxx}
\keywords{keyword1, keyword2, keyword3, keyword4, keyword5}
\contact{author of record (for correspondence)}
\conference{submission to other conference(s)}
\date{}

\begin{document}
\makeidpage
\maketitle

\section{Introduction}

This document is a slightly revised version of the instructions
for preparing copies for the final ACL-02 proceedings. The
format here described allows for a graceful transition to the
style required for that publication.

This document has been elaborated from similar documents used for
previous editions of the ACL and EACL annual meetings. Those
versions were written by several people, including John Chen,
Henry S. Thompson and Donald Walker. This document is an example
of what your submission manuscript to ACL-02 should look like.
Authors are asked to conform to the directions reported in this
document.

\section{General Instructions}

As reviewing will be blind, a separate identification page will be
required reporting authors' names, affiliations, and email addresses
(see below for additional information on the identification page).
The paper itself should not include the authors' names and
affiliations. Furthermore, self-references that reveal the authors'
identities, e.g., ``We previously showed \cite{dawidskene79} ...'',
should be avoided. Instead, use citations such as ``Dawid and Skene
\shortcite{dawidskene79} previously showed ...''  In general, authors
should avoid including any information in the body of the paper, in
the acknowledgments and in the reference list, that would identify the
authors themselves or their institutions. Such information can be
added to the final camera-ready version for publication.

The maximum length of a submission manuscript is $8$ pages,
printed single-sided.  Print all text, including section titles
and figures, in two-column format where each column is 7.6cm by
23.5cm (3.0in by 9.25in) and there is a 0.6cm (0.25in) space
between the two columns. Exceptions to the two-column format
include the title at the top of the first page and any full-width
figures or tables. (See the guidelines later regarding formatting
the first page of the submission manuscript.)  Start all pages
directly under the top margin. Text should be centered on each
page. On A4 paper, this roughly means leaving 2.5cm (1in) margins
on left and right sides of each page as well as a 3.0cm (1.2in)
margin on the top of each page. Type single spaced. Indent when
starting a new paragraph. Use standard fonts such as Times Roman
or Computer Modern Roman, with 11 points for text and 14 to 16
points for headings and title.

\subsection{Electronically-available resources}

This file ({\tt acl02sub.tex}) along with the LaTeX style file ({\tt
  acl02sub.sty}) and ACL bibliography style ({\tt acl.bst}) are
available from\\
http://www.cs.ualberta.ca/$\sim$lindek/acl02/style/\\
A
Microsoft Word Style file ({\tt acl02sub.doc}) is also present at the
same URL.  We strongly recommend the use of these style files, that
have been appropriately tailored for the ACL-02 submission.

\subsection{The identification page}

As reviewing will be blind, a separate identification page is required.
The identification page should include the paper title, authors' names,
affiliations, and email addresses, the paper ID code generated upon
paper registration (see the ACL-02 call for papers for instructions
about the electronic registration procedure for submitted papers), up to
5 keywords specifying the subject area, name and address of the contact
author, and a short summary (up to 5 lines).  The identification page
should also specify whether the paper is under consideration for other
conferences.

\subsection{The First Page}

Center the title across both columns.  As already mentioned above,
do not include authors' names and affiliations.  Use the
two-column format only when you begin the abstract.

{\bf Title}: Place the title at the top of the first page, followed by
the paper ID code as generated upon paper registration.  Long title
should be typed on two lines without a blank line
intervening. Approximately, put the title at 2.7cm (1.1in) from the top
of the page, and leave 2.5cm (1in) between the title and the body of the
first page.

{\bf Abstract}: Type the abstract at the beginning of the first column.
The width of the abstract text should be smaller than the width of the
columns for the text in the body of the paper by about 0.7cm (0.3in) on
each side.  Center the word {\bf Abstract} in bold form above the body
of the abstract. The abstract should be no longer than 200 words.

{\bf Text}: Begin typing the main body of the text immediately
after the abstract, observing the two-column format as shown in
this example.

\subsection{Sections}

{\bf Headings}: Type and label section and subsection headings in
the style shown on these pages.  Use numbered sections, in order
to facilitate cross references.

{\bf References}: Follow the ``Guidelines for Formatting
Submissions'' to {\em Computational Linguistics} that appears in
the first issue of each volume, if possible.  That is, citations
within the text appear in parentheses as (Author, year) or, if the
author's name appears in the text itself, as Author (year). Gather
the full set of references together under the heading {\bf
References}; place the section before any {\bf Appendices}, unless
they contain references. Arrange the references alphabetically by
first author, rather than by order of occurrence in the text.
Provide as complete a citation as possible, using a consistent
format, such as the one for CL or the one in the {\em
Publication Manual of the American Psychological Association}
(American Psychological Association, 1983).  Use of full names for
authors rather than initials is preferred.  A list of
abbreviations for common computer science journals can be found in
the ACM {\em Computing Reviews} (Association for Computing
Machinery, 1983).

The LaTeX and BibTeX style files provided roughly fit the APA format,
allowing regular citations (Author, year), short citations for text
that mentions the Author (year), and multiple citations (Author, year;
Author, year; Author, year).

{\bf Appendixes}: Appendixes, if any, directly follow the text and the
references (but see above).  Letter them in sequence and provide an
informative title: {\bf Appendix A. Title of Appendix.}

\subsection{Footnotes}

{\bf Footnotes}: Put footnotes at the bottom of the page. They may
be numbered or referred to by asterisks or other
symbols.\footnote{This is how a footnote should appear.} Footnotes
should be separated from the text by a line.\footnote{Note the
line separating the footnotes from the text.}

\subsection{Graphics}

{\bf Illustrations}: Place figures, tables, and photographs in the paper
near where they are first discussed, rather than at the end, if
possible.  Wide illustrations may run across both columns.

{\bf Captions}: Provide a caption for every illustration; number each one
sequentially in the form:  ``Figure 1. Caption of the Figure.'' ``Table 1.
Caption of the Table.''  Type the captions for figures below the
figures.  Type the captions for tables above the tables.
%check that the .sty does so

\section{Length of Submission}

Eight pages ($8$) is the length limit for ACL-02 submissions. All
illustrations, references, and appendices must be accommodated within
these page limits, also observing the formatting instructions given in
the present document. This will also be the number of pages allocated
for each paper in the ACL-02 proceedings.

Papers that do not conform to the specified length
and formatting requirements are subject to be rejected without
review.

\section{Submission Procedure}

\paragraph{Paper registration:} You must submit a notification of
submission by filling out the form at\\
\centerline{http://www.cs.ualberta.ca/$\sim$lindek/acl02}\\
The authors should fill in the title of the paper, the authors' names,
affiliations, and email addresses, one or two general topic areas, up
to 5 keywords specifying the subject area, and a short summary (up to
200 words).  The authors should also specify whether the paper is
under consideration for other conferences or workshops, and if so,
which ones.

Each submission will be assigned an identification number. Please use
it on all correspondence with the program committee.

\paragraph{Paper submission:} All papers must be submitted electronically
at the same web address. The first page of your paper must include the
identification number obtained from paper registration. The paper must
be submitted no later than 12 noon Mountain Time (7PM GMT) on Feb. 1
2002. Papers submitted after that time will not be reviewed. Papers
must be in PDF format. The submission web page includes information
about converting different types of documents to PDF. The program
committee will make every attempt to print out your paper
successfully, but cannot take responsibility if they do not. Authors
are strongly encouraged to submit papers 48 hours before the
submission deadline so that unprinting formats can be detected and
corrected by the submission deadline. If for some reason an author is
not able to submit electronically, or if we have discovered in advance
that there is a problem with the PDF file, authors should contact
Dekang Lin (lindek@cs.ualberta.ca) concerning hard-copy submission.

\bibliographystyle{acl}

\bibliography{sample}

%\section*{References}
%
%American Psychological Association.  1983.  {\em Publications Manual}.
%Washington, D.C.: American Psychological Association.
%
%Association for Computing Machinery.  1983.  {\em Computing Reviews}, 24(11):
%503-512.

\end{document}

### 2002-03-11 technical/acl2002/submission.tex
first commit of ACL submission draft - with abstract, introduction, and overall structure.
svn/trunk@489

\documentclass[11pt]{article}
\usepackage{acl02sub,url}
\title{NLTK: The Natural Language Tool Kit}
\author{
\begin{tabular}[t]{c}
Edward Loper\\
Department of Computer and Information Science\\
University of Pennsylvania\\
Philadelphia, PA 19104-6389, USA
\end{tabular}
\hspace{6pt}
\begin{tabular}[t]{c}
Steven Bird\\
Linguistic Data Consortium\\
University of Pennsylvania\\
Philadelphia, PA 19104-2608, USA
\end{tabular}
}
\summary{%
NLTK, the Natural Language Tool Kit, is a suite of program modules,
tutorials and problem sets.  NLTK covers symbolic and statistical
NLP, and is interfaced to annotated corpora.
Students augment and replace existing components, learning
structured programming by example, and manipulating sophisticated
models from the outset.  Along with extensive documentation and
problem sets, NLTK provides self-contained, ready-to-use CL
courseware.}
\paperid{Pxxxx}
\keywords{courseware, Python, corpora, keyword4?, keyword5?}
\contact{Edward Loper}
\conference{this paper has not been submitted to any other conferences}
\date{\today}

\begin{document}
\makeidpage
\maketitle

\begin{abstract}
Students in computational linguistics courses must often learn
a new programming language.  Where courses are offered in linguistics
departments, the students may be learning to program for the first time.
Many low-level ``housekeeping'' tasks must
be accomplished in order to do interesting projects.  At the same
time, teachers of computational linguistics courses sometimes feel
that they spend too much time teaching students to program, and not
enough time teaching the subject itself.  They may even avoid
programming assignments altogether.  However, we believe that it is
crucial for a first computational linguistics course to include a
strong practical component, in which students develop real programs to
solve real problems with real data.

Python is a new object-oriented scripting language which runs on all
platforms.  Python has been praised as "executable pseudocode", since
programs are so easy to write.  Recently, we have been developing
NLTK, an open-source Natural Language Toolkit written in Python.
In this presentation, we will motivate, describe and demonstrate NLTK.

NLTK, the Natural Language Tool Kit, is a suite of program modules,
tutorials and problem sets.  NLTK covers symbolic and statistical
natural language processing, and is interfaced to annotated corpora.
Students augment and replace existing NLTK components, learning
structured programming by example, and manipulating sophisticated
models from the outset.  Along with extensive documentation and
problem sets, NLTK provides self-contained, ready-to-use CL
courseware.
\end{abstract}

\section{Introduction}

Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is difficult not
just because of the variety of data structures, but also because of
the diverse range of topics which may need to be included in the
syllabus.  A widespread practice is to teach multiple programming
languages, where each provides data structures and native functions
that are a good fit for the task at hand (e.g. Prolog for parsing,
Perl for corpus processing, a finite-state toolkit for morphological
analysis).  By relying on the built-in features of various languages,
the teacher avoids having to develop a lot of software infrastructure.
An unfortunate consequence is that a significant part of the course
must be devoted to teaching these languages.  Further, many
interesting projects require the languages to be bridged, e.g. a
project that involved syntactic parsing of corpus data from a
morphologically rich language would involve all three languages
mentioned above (Perl for file I/O, format conversions and overall
program control, with calls out to a finite state toolkit for
morphological analysis and to a Prolog engine for parsing).  It is
clear that these considerable overheads and shortcomings warrant a
fresh approach.

Apart from the practical component, computational linguistics courses
may depend on appropriate software in another way, namely for in-class
demonstrations.  This context calls for highly interactive graphical
user interfaces making it possible to view program state (e.g. the
chart of a chart parser), observe program execution step-by-step
(e.g. execution of a finite-state machine), and even make minor
modifications to programs in response to ``what if'' questions from
the class.  Because of these difficulties it is common to avoid live
demonstrations, and keep classes for theoretical presentations only.
Apart from being dull, this approach leaves students to solve
important practical problems on their own, or to deal with them less
efficiently in office hours.

In this paper we describe a new approach to the above challenges,
a streamlined and flexible way of organizing the practical component
of an introductory computational linguistics course.  We describe
NLTK, the Natural Language Tool Kit, which we have developed in
conjunction with a course we have taught at the University of Pennsylvania.

OVERVIEW OF THE PAPER

All materials discussed here are available under an open
source license from \url{nltk.sf.net}.

\section{Choice of programming language}

Requirements on a Programming Language

shallow learning curve
new programmers must get immediate rewards
support for rapid prototyping
we want to avoid the compilation step
self-documenting code
programs must be immediately comprehensible
support for good programming style
it must be easy to write well-structured programs
graphical user interface
the language must have a good, easy-to-use GUI

Python: object-oriented scripting

shallow learning curve
Python was designed to be easily learnt by children
support for rapid prototyping
Python is interpreted, with no compilation step
self-documenting code
Python has been called "executable pseudocode"
support for good programming style
Python is object-oriented (but not punitively so)
graphical user interface
Python has an interface to the tk GUI toolkit

\section{NLTK design principles}

\section{Structure of the toolkit}

Code, documentation, ...

\section{Basic modules}

\subsection{Token}

\subsection{Tree}

\subsection{Tagger}

\subsection{Probability}

\section{Advanced modules}

\subsection{ChunkParser}

\subsection{ChartParser}

\subsection{FSA}

\subsection{Classifier}

\section{Evaluation}

How we used NLTK at Penn (including the competition).

Strengths and weaknesses

\section{Other approaches}

Java: \cite{Hammond02}
-- no obligation to cite this as it is unpublished
(and I haven't seen a copy).

\section{Conclusion}

Overview of the paper.

Upbeat conclusion.

Invitation to participate.

\bibliographystyle{acl}
\bibliography{submission}

\end{document}

### 2002-03-13 technical/acl2002/submission.tex
drafted section on choice of programming language
svn/trunk@490

@@
 \author{
-\begin{tabular}[t]{c}
-Edward Loper\\
-Department of Computer and Information Science\\
-University of Pennsylvania\\
-Philadelphia, PA 19104-6389, USA
-\end{tabular}
-\hspace{6pt}
-\begin{tabular}[t]{c}
-Steven Bird\\
-Linguistic Data Consortium\\
-University of Pennsylvania\\
-Philadelphia, PA 19104-2608, USA
-\end{tabular}
+Edward Loper and Steven Bird\\
+Department of Computer and Information Science and Linguistic Data Consortium\\
+University of Pennsylvania, Philadelphia, PA 19104-6389, USA
 }
@@
 NLTK, the Natural Language Tool Kit, is a suite of program modules,
-tutorials and problem sets.  NLTK covers symbolic and statistical
+tutorials and problem sets, providing ready-to-use CL courseware.
+NLTK covers symbolic and statistical
 NLP, and is interfaced to annotated corpora.
@@
 structured programming by example, and manipulating sophisticated
-models from the outset.  Along with extensive documentation and
-problem sets, NLTK provides self-contained, ready-to-use CL
-courseware.}
+models from the outset.}
 \paperid{Pxxxx}
@@
 Students in computational linguistics courses must often learn
-a new programming language.  Where courses are offered in linguistics
-departments, the students may be learning to program for the first time.
-Many low-level ``housekeeping'' tasks must
-be accomplished in order to do interesting projects.  At the same
+a new programming language.  In some cases, such as when courses
+are offered in linguistics departments,
+the students may be learning to program for the very time.
+In order to do interesting projects, it is usually necessary for
+students to do many low-level ``housekeeping'' tasks.  At the same
 time, teachers of computational linguistics courses sometimes feel
@@
 Python is a new object-oriented scripting language which runs on all
-platforms.  Python has been praised as "executable pseudocode", since
+platforms.  Python has been praised as ``executable pseudocode,'' since
 programs are so easy to write.  Recently, we have been developing
@@
 syllabus.  A widespread practice is to teach multiple programming
-languages, where each provides data structures and native functions
+languages, where each language provides native data structures and functions
 that are a good fit for the task at hand (e.g. Prolog for parsing,
@@
 Apart from the practical component, computational linguistics courses
-may depend on appropriate software in another way, namely for in-class
+may also depend on software for in-class
 demonstrations.  This context calls for highly interactive graphical
@@
 
-Requirements on a Programming Language
+The most basic step in setting up a practical component is choosing a
+suitable programming language.  Here we list the desiderata that influence
+our choice.
 
-shallow learning curve
-new programmers must get immediate rewards
-support for rapid prototyping
-we want to avoid the compilation step
-self-documenting code
-programs must be immediately comprehensible
-support for good programming style
-it must be easy to write well-structured programs
-graphical user interface
-the language must have a good, easy-to-use GUI
+First, the language must have a shallow learning curve, so that novice
+programmers get immediate rewards for their efforts.
+Second, the language must support rapid prototyping and a short develop/test cycle;
+an obligatory compilation step is a serious detraction.
+Third, the code should be self-documenting,
+with a transparent syntax and semantics.
+Fourth, it should be easy to write structured programs, ideally object-oriented
+but without the burden associated with languages like C++.
+Finally, the language must have an easy-to-use graphics library to support
+the development of simple graphical user interfaces.
 
-Python: object-oriented scripting
-
-shallow learning curve
-Python was designed to be easily learnt by children
-support for rapid prototyping
-Python is interpreted, with no compilation step
-self-documenting code
-Python has been called "executable pseudocode"
-support for good programming style
-Python is object-oriented (but not punitively so)
-graphical user interface
-Python has an interface to the tk GUI toolkit
+In surveying the available languages, we believe that Python offers an especially good
+fit to the above requirements.  Python is an object-oriented scripting language
+developed by Guido van Rossum [REF] and available on all platforms
+\url{www.python.org}.
+Python offers a shallow learning curve by virtue of the
+fact that it was designed to be easily learnt by children [REF].
+As an interpreted language, Python is suitable for rapid prototyping.
+Python code is exceptionally readable, and it has been praised as
+``executable pseudocode.''
+Python is an object-oriented language, but not punitively so, and it
+is easy to encapsulate data and methods inside Python classes.
+Finally, Python has an interface to the ``tk'' graphics toolkit, and
+writing graphical interfaces is straightforward.
 

### 2002-03-16 technical/acl2002/submission.tex
Added abridged discussion of design criteria from NLTK proposal
svn/trunk@492

@@
 \usepackage{acl02sub,url}
-\title{NLTK: The Natural Language Tool Kit}
+\title{NLTK: The Natural Language Toolkit}
 \author{
@@
 \summary{%
-NLTK, the Natural Language Tool Kit, is a suite of program modules,
+NLTK, the Natural Language Toolkit, is a suite of program modules,
 tutorials and problem sets, providing ready-to-use CL courseware.
@@
 
-NLTK, the Natural Language Tool Kit, is a suite of program modules,
+NLTK, the Natural Language Toolkit, is a suite of program modules,
 tutorials and problem sets.  NLTK covers symbolic and statistical
@@
 of an introductory computational linguistics course.  We describe
-NLTK, the Natural Language Tool Kit, which we have developed in
+NLTK, the Natural Language Toolkit, which we have developed in
 conjunction with a course we have taught at the University of Pennsylvania.
@@
 
-\section{NLTK design principles}
+\section{Design criteria}
 
-\section{Structure of the toolkit}
+Several criteria must be considered in the design and implementation
+of the toolkit.  The design criteria are divided into primary and
+secondary criteria, and within those sections are listed in the order
+of their importance.  A set of explicit non-requirements is also
+given.  The toolkit is not expected to satisfy any of these
+properties.
 
-Code, documentation, ...
+\subsection{Primary Design Criteria}
+
+\textbf{Ease of Use.} The primary purpose of the toolkit is to allow
+students to concentrate on building NLP systems.  The more time
+students must spend learning to use the toolkit, the less useful it
+is.
+
+\textbf{Consistency.} The toolkit should use consistent data structures
+and interfaces.
+
+\textbf{Extensibility.} The toolkit should easily accommodate new
+components, whether those components replicate or extend the toolkit's
+existing functionality.  Thus, the toolkit's design should be modular,
+with simple and well-defined interfaces between modules.  The toolkit
+should also be structured in such a way that it is obvious where new
+extensions would fit into the toolkit's infrastructure.
+
+\textbf{Documentation.} The toolkit, its data structures, and its
+implementation all need to be documented.  Three documentation types
+are necessary: technical reports, to explain the toolkit's
+design and implementation; tutorials, to teach students how to use the
+toolkit to perform specific tasks; and reference documentation, to describe
+every module, interface, class and method.  All nomenclature must be
+carefully chosen and consistently used.
+
+\subsection{Secondary Design Criteria}
+
+\textbf{Simplicity.} The toolkit should structure the complexities of
+building NLP systems, not hide them.  Therefore, each class defined by 
+the toolkit should be simple enough that a student could implement it
+by the time they finish the course.
+
+\textbf{Modularity.} The interaction between different components of the
+toolkit should be kept to a minimum.  In particular, it should be
+possible to use complete individual projects using small parts of the
+toolkit, without worrying about how they interact with the rest of the
+toolkit.  This will allow students to learn to use the toolkit
+incrementally throughout the course.  Modularity also makes it easier
+to change or extend the toolkit.
+
+\subsection{Non-Requirements}
+
+\textbf{Comprehensiveness.} The toolkit is not intended to provide a
+comprehensive set of tools.  Indeed, there should be a wide variety of 
+ways in which students can extend the toolkit.
+
+\textbf{Efficiency.} The toolkit does not need to be highly optimized for
+runtime performance.  However, it should be efficient enough that
+students can use their NLP systems to perform real tasks.
+
+\textbf{Cleverness.} Clear designs and implementations are far preferable 
+to ingenious yet indecipherable ones.
+
+\section{Python Implementation}
+
+The structure of the Python language directly affects the design and
+implementation of the toolkit.  This section discusses how the design
+of the toolkit is affected by the Python language.
+
+\textbf{Interfaces.}
+Python does not directly implement interfaces.  The toolkit
+therefore implements interfaces as simple classes, all of whose methods 
+raise \texttt{AssertionError} exceptions.  A description of the
+functionality provided by the interface is contained in the
+class's documentation string.  Each method's documentation string
+specifies the behavior of that method.
+In order to implement an interface, a class should include that
+interface as one of its bases, and should override every
+(non-optional) method defined by the interface.
+Classes that define interfaces are named with a trailing ``I,'' such
+as \texttt{TokenizerI} or \texttt{EventI}.
+
+\textbf{Typing.}
+Type checking is an important safety device, allowing programmers to
+quickly locate errors in their code.  This is especially important for
+novice programmers, who have less experience with debugging.  Every
+function and method defined by the toolkit will check the types of its
+its arguments.  Given the impact on performance, the level of type checking
+will be adjustable.
+
+\textbf{Advanced Language Features.}
+The use of advanced language features and programming styles could
+potentially make the toolkit much more difficult to use or to
+understand.  The fewer language features that a student must learn in
+order to use the toolkit, the faster they can start developing NLP
+systems.  The language features used by
+the toolkit are summarized by figure \ref{fig:feature}.
+
+\begin{figure}
+\begin{centering}
+\begin{tabular}{|l|l|}
+\hline
+\textbf{Language Feature} & \textbf{Use} \\
+\hline
+Classes & used \\
+Exceptions & error conditions only \\
+Operator Overloading & used \\
+Python 2.0 Features & used \\
+Python 2.1 Features & not used \\
+Function Arguments & minimal use \\
+Lambda Functions & minimal use \\
+Mapping and Filtering & not used \\
+List Comprehensions & used \\
+Tkinter & used \\
+\hline
+\end{tabular}\\
+\end{centering}
+\caption{Advanced language features used by NLTK.}
+\label{fig:feature}
+\end{figure}
 
 \section{Basic modules}
-
-\subsection{Token}
 

### 2002-03-18 docbook/sgml/docbook/dsssl-stylesheets-1.59/test/cases/makemlang.pl
Added files used by (open)Jade
svn/trunk@494

@@ -1,0 +1,67 @@
+#!/usr/bin/perl -- # -*- Perl -*-
+
+use strict;
+my %languages = ('ca' => 'Catalan',
+		 'cs' => 'Czech',
+		 'da' => 'Danish',
+		 'de' => 'German',
+		 'el' => 'Greek',
+		 'en' => 'English',
+		 'es' => 'Spanish',
+		 'et' => 'Estonian',
+		 'fi' => 'Finnish',
+		 'fr' => 'French',
+		 'hu' => 'Hungarian',
+		 'id' => 'Indonesian',
+		 'it' => 'Italian',
+		 'ja' => 'Japanese',
+		 'ko' => 'Korean',
+		 'nl' => 'Dutch',
+		 'no' => 'Norwegian',
+		 'pl' => 'Polish',
+		 'pt' => 'Portuguese',
+		 'pt_br' => 'Portuguese (Brazil)',
+		 'ro' => 'Romanian',
+		 'ru' => 'Russian',
+		 'sk' => 'Slovak',
+		 'sl' => 'Slovenian',
+		 'sv' => 'Swedish',
+		 'zh_cn' => 'Chinese (Continental)');
+
+foreach my $lang (sort keys %languages) {
+    my $langid = $lang;
+    $langid =~ s/_//g;
+
+    print <<EOF;
+<chapter lang='$lang' id='${langid}chap'>
+<chapterinfo>
+<author id='${langid}author'>
+  <surname>Surname</surname>
+  <firstname>Firstname</firstname>
+</author>
+</chapterinfo>
+<title>$languages{$lang}</title>
+<note>
+<para>
+This paragraph would be in $languages{$lang}, if I knew any.
+<quote>Quote test.</quote>
+</para>
+</note>
+<para>Author formatting test: <xref linkend='${langid}author'>.</para>
+<para>Cross reference tests:
+EOF
+
+    foreach my $otherlang (sort keys %languages) {
+	my $otherlangid = $otherlang;
+	$otherlangid =~ s/_//g;
+	if ($otherlang ne $lang) {
+	    print "See <xref linkend='${otherlangid}chap'>.\n";
+	}
+    }
+
+    print <<EOF2;
+</para>
+</chapter>
+
+EOF2
+}

### 2002-03-18 technical/acl2002/submission.tex
Added summary of NLTK modules from LSA handout
svn/trunk@496

@@
 source license from \url{nltk.sf.net}.
+NLTK runs on most platforms, including Windows, OS X, Linux, and
+UNIX.
 
@@
 
+\section{Modules}
+
+NLTK currently includes the following modules.
+
+\subsection*{Basics}
+
+\begin{description}
+\item[token:] Basic classes for processing individual
+      elements of text, such as words or sentences. 
+\item[tree:] Classes for representing tree structures
+      over text (such as syntax trees and morphological trees).
+\item[probability:] Classes that encode frequency
+      distributions and probability distributions.
+\end{description}
+
+\subsection*{Tagging}
+
+\begin{description}
+\item[tagger:] A standard interface to tag each token of
+      a text with supplementary information, such as its part of
+      speech; and several implementations of that interface.
+\end{description}
+
+\subsection*{Parsing}
+
+\begin{description}
+\item[parser:] A standard interface to produce trees
+      representing the structure of texts.
+\item[chartparser:] A flexible parser implementation
+      that uses a \emph{chart} to record hypotheses about
+      syntactic constituents.
+\item[srparser\_template:] A partial implementation of a
+      shift-reduce parser; completing the implementation was a
+      student exercise.
+\item[chunkparser:] A standard interface for robust
+      parsers used to identify non-overlapping linguistic groups
+      (such as noun phrases) in unrestricted text.
+\item[rechunkparser:] A regular-expression based
+      implementation of the chunk parser interface.
+\end{description}
+
+\subsection*{Text Classification}
+
+\begin{description}
+\item[classifier:] A standard interface for classifying
+      texts into categories.
+\item[classifier.feature:] A standard way of encoding
+      the information used to make classification decisions.
+\item[classifier.naivebayes:] A text classifier
+      implementation based on the Naive Bayes assumption.
+\item[classifier.maxent:] An implementation of the
+      maximum entropy model for text classification; and
+      implementations of the GIS and IIS algorithms for training
+      the classifier.
+\item[classifier.featureselection:] A standard interface
+      for choosing which features are relevant for making
+      classification decisions.
+\end{description}
+
+\subsection*{Visualization}
+
+\begin{description}
+\item[draw.tree:] A graphical representation for tree
+      structures, such as syntax trees and morphological trees. 
+\item[draw.tree\_edit:] A graphical interface used to
+      build and modify tree structures.
+\item[draw.plot\_graph:] A graphical tool to graph
+      arbitrary functions.
+\item[draw.chart:] An interactive graphical tool used to
+      experiment with chart parsers.
+\end{description}
+
 \section{Basic modules}
@@
 
-Upbeat conclusion.
-
-Invitation to participate.
+NLTK is an open source project, and we welcome any contributions.  We
+deliberately structured NLTK to facilitate parallel development.
+Readers who are interested in contributing to NLTK, or who have any
+suggestions for improvements, are encouraged to contact the authors.
 

### 2002-03-18 technical/acl2002/submission.tex
Added section on the chunking project, based on LSA report
svn/trunk@497

@@
 \documentclass[11pt]{article}
-\usepackage{acl02sub,url}
+\usepackage{acl02sub,url,alltt,epsfig}
 \title{NLTK: The Natural Language Toolkit}
@@
 \paperid{Pxxxx}
-\keywords{courseware, Python, corpora, keyword4?, keyword5?}
+\keywords{teaching, courseware, Python, corpora}
 \contact{Edward Loper}
@@
 \date{\today}
+
+\newenvironment{sv}{\scriptsize\begin{alltt}}{\end{alltt}\normalsize}
 
@@
 
-\section{Basic modules}
+\section{Uses of NLTK}
 
-\subsection{Tree}
+\subsection{Assigments and projects}
 
-\subsection{Tagger}
+NLTK can be used for simple and advanced projects, as well as
+everything in between.  In the simplest assignments, students
+tweak an existing module to alter its behavior.  The variety of
+existing modules makes it easy to create simple assignments.
+A more challenging task is to develop a new module.  Here, NLTK provides
+some useful starting points: predefined interfaces and data structures, and
+existing modules that implement the same interface.  Advanced projects
+typically involve the development of entirely new functionality for a
+previously unsupported NLP task, or the development of a complete system
+out of existing and new modules.
 
-\subsection{Probability}
+Here we illustrate the use of NLTK in an assignment of moderate
+difficulty, for the case of chunk parsing.  This is a kind of parsing
+where we only identify the main constituents of a phrase, a technique
+with diverse applications.  The NLTK \texttt{chunk} module offers
+a variety of rule formats.  For example,
+\texttt{ChunkRule('<NN.*>')} builds chunks over consecutive noun tags;
+\texttt{ChinkRule('<VB.>')} excises verbs from existing chunks;
+\texttt{SplitRule('<NN>', '<DT>')} splits into two pieces any existing
+chunk that contains a noun followed by a determiner; and
+\texttt{MergeRule('<JJ>', '<JJ>')} combines two adjacent chunks
+where the first chunk ends and the second chunk starts with \texttt{JJ}.
+The goal of the assignment is to come up with a cascade of these rules,
+and to create a set of chunks that are as similar as possible to the test set.
 
-\section{Advanced modules}
+The chunking tutorial provides all the necessary code except for the
+chunk rules.  The provided code is responsible for loading the
+chunked, tagged text using an existing tokenizer, creating an
+unchunked version of the text, applying the chunk rules to the
+unchunked text, and scoring the result.  Students focus on the
+NLP task only -- providing a ruleset with the best coverage.
 
-\subsection{ChunkParser}
+In the remainder of this section we reproduce some of the cascades
+created by the students.  The first example illustrates a combination
+of several rule types:
 
-\subsection{ChartParser}
+\begin{sv}
+cascade = [
+  ChunkRule('<DT><NN.*><VB.><NN.*>'),
+  ChunkRule('<DT><VB.><NN.*>'),
+  ChunkRule('<.*>'),
+  UnChunkRule('<IN|VB.*|CC|MD|RB.*>'),
+  UnChunkRule("<,|\.|``|''>"),
+  MergeRule('<NN.*|DT|JJ.*|CD>', '<NN.*|DT|JJ.*|CD>'),
+  SplitRule('<NN.*>', '<DT|JJ>')
+]
+\end{sv}
 
-\subsection{FSA}
+This next example illustrates a brute-force approach.  All tags most
+commonly found inside of chunks are chunked, while other tags are excluded.
 
-\subsection{Classifier}
+\begin{sv}
+cascade = [
+  ChunkRule('<\$|CD|DT|EX|PDT
+             |PRP.*|WP.*|\#|FW
+             |JJ.*|NN.*|POS|RBS|WDT>*')
+]
+\end{sv}
+
+The third example involves putting everything in a single chunk,
+then excising elements that do not belong.
+
+\begin{sv}
+cascade = [
+  ChunkRule('<.*>+')
+  ChinkRule('<VB.*|IN|CC|R.*|MD|WRB|TO|.|,>+')
+]
+\end{sv}
+
+The main program which surrounds the student's work is moderately
+simple as well.  The line replaced by students is underlined.
+
+\begin{sv}
+from nltk.token import *
+from nltk.chunkparser import *
+from nltk.rechunkparser import *
+import os
+
+path = "/cis530/data/wsj/tagged/"
+
+sent_tok = RETokenizer('(=======+)?\\n\\n+', 0)
+sentences = []
+files = os.listdir(path)
+for file in files:
+    print 'Loading %s...' % file
+    text = open(path+file).read()
+    sentences += sent_tok.tokenize(text, source=file, unit='s')
+
+chunkscore = ChunkScore()
+
+\underline{cascade = [ ??? ]}
+chunkparser = REChunkParser(cascade)
+
+ctt = ChunkedTaggedTokenizer()
+for s in sentences:
+    correct = ctt.tokenize(s.type(), source=s.loc())
+    unchunked = unchunk(correct)
+    guess = chunkparser.parse(unchunked)
+    chunkscore.score(correct, guess)
+
+print chunkscore
+\end{sv}
+
+The score for each project is the precision, recall and F-measure.
+Figure~\ref{contest} shows the results; students with the best two scores
+were presented with prizes.
+
+\begin{figure}
+\centerline{\epsfig{figure=contest.ps,width=\linewidth}}
+\caption{Precision/Recall Graph for Chunking Competition}\label{contest}
+\vspace*{2ex}\hrule
+\end{figure}
+
+\subsection{Class demonstrations}
+
+[chart parsing example, including screenshot]
 
@@
 
-How we used NLTK at Penn (including the competition).
+About CIS-530: mixture of CS and linguistics students.
+Approx 40 hours of classes.  All class materials (slides,
+assignments) are posted on the course website
+\url{http://www.cis.upenn.edu/~cis530/}.
 
-Strengths and weaknesses
+The Language: Python was new to 95\% of students.  Some were initially
+bothered about its block structuring, which is sensitive to white
+space.  All quickly learnt the language, and reported that any initial
+reservations were unfounded.  Many students said that they enjoyed
+writing Python code.
+
+CD-ROM: put NLTK and data on CD-ROM and gave it out at the start of class.
+
+The toolkit: students liked being able to do interesting projects from
+the outset, liked all the documentation, liked being able to run
+everything on their machine at home, ...
+
+Problems: available corpus data, ...
+
+Other strengths / weaknesses
 
@@
 
+NOTES: Many CL textbooks provide good exercises at the end of each
+chapter.  This isn't enough for many situations.  E.g. students spend
+lots of time writing interface code (e.g. to process a corpus).
+E.g. teachers spend lots of time creating do-able assignments...
+
 NLTK is an open source project, and we welcome any contributions.  We

### 2002-03-20 technical/handout/handout.tex
Short overview of NLTK
svn/trunk@505

\documentclass[12pt]{article}
\usepackage{fullpage}

\def\nogap{\setlength\itemsep{.0in}\setlength{\parskip}{0in}}

\pagestyle{empty}
\begin{document}
\title{NLTK: A Natural Language Toolkit}
\author{Steven Bird \& Edward Loper}

% clunch:
%\date{Thursday December 12, 2001}

% lsa:
\date{Sunday January 6, 2002}

\maketitle

\section*{Overview}
\thispagestyle{empty}

The Natural Language Toolkit is a Python package that simplifies the
construction of programs that process natural language.  In
particular:

% Note: can I say NLP, or do I have to say natural language
% processing? 
\begin{itemize}\nogap

  \item It provides basic tools for manipulating data and performing
  tasks related to NLP.

  \item It defines standard interfaces between the different components
  of an NLP system.

  \item It provides an infrastructure for building new NLP systems.

\end{itemize}

 \noindent The toolkit's primary aim is to serve as a pedagogical
 tool; but it is also useful as a framework for implementing
 research-related programs.

% Say something about what platforms are supported..
\vspace{3mm}\noindent NLTK runs on most platforms, including Windows, OS X, Linux, and
UNIX.

\section*{Toolkit Contents}
\thispagestyle{empty}

\begin{itemize}\nogap
    \item \textbf{Python Modules} implement the basic data types,
      tools, and interfaces that make up the toolkit.

    \item \textbf{Tutorials} teach students how to use the toolkit, in the
      context of performing specific tasks.

    \item \textbf{Exercises and Problem Sets} help students learn more about
      various aspects of natural language processing.

    \item \textbf{Reference Documentation} provides precise definitions of the
      behavior of each module, interface, class, method, function, and
      variable defined by the toolkit.

    \item \textbf{Technical Documentation} explains and justifies the
      toolkit's design and implementation. 
\end{itemize}

\section*{Contributing}
\thispagestyle{empty}

NLTK is an open source project, and we welcome any contributions.  We
deliberately structured NLTK to facilitate parallel development.  If
you are interested in contributing to NLTK, or have any ideas for
improvements, please talk to us, or send us email at
\texttt{edloper@gradient.cis.upenn.edu} and
\texttt{sb@unagi.cis.upenn.edu}.

\section*{Modules}
\thispagestyle{empty}

NLTK currently includes the following modules.

{\small
\begin{itemize}\nogap

  \item Basics
    \begin{itemize}\nogap
      \item \texttt{token}: Basic classes for processing individual
            elements of text, such as words or sentences. 
      \item \texttt{tree}: Classes for representing tree structures
            over text (such as syntax trees and morphological trees).
      \item \texttt{probability}: Classes that encode frequency
            distributions and probability distributions.
    \end{itemize}

  \item Tagging
    \begin{itemize}\nogap
      \item \texttt{tagger}: A standard interface to tag each token of
            a text with supplementary information, such as its part of
            speech; and several implementations of that interface.
    \end{itemize}

  \item Parsing
    \begin{itemize}\nogap
      \item \texttt{parser}: A standard interface to produce trees
            representing the structure of texts.
      \item \texttt{chartparser}: A flexible parser implementation
            that uses a \emph{chart} to record hypotheses about
            syntactic constituents.
      \item \texttt{srparser\_template}: A partial implementation of a
            shift-reduce parser; completing the implementation was a
            student exercise.
      \item \texttt{chunkparser}: A standard interface for robust
            parsers used to identify non-overlapping linguistic groups
            (such as noun phrases) in unrestricted text.
      \item \texttt{rechunkparser}: A regular-expression based
            implementation of the chunk parser interface.
    \end{itemize}

  \item Text Classification
    \begin{itemize}\nogap
      \item \texttt{classifier}: A standard interface for classifying
          texts into categories.
      \item \texttt{classifier.feature}: A standard way of encoding
          the information used to make classification decisions.
      \item \texttt{classifier.naivebayes}: A text classifier
          implementation based on the Naive Bayes assumption.
      \item \texttt{classifier.maxent}: An implementation of the
          maximum entropy model for text classification; and
          implementations of the GIS and IIS algorithms for training
          the classifier.
      \item \texttt{classifier.featureselection}: A standard interface
          for choosing which features are relevant for making
          classification decisions.
    \end{itemize}

  \item Visualization
    \begin{itemize}\nogap 
      \item \texttt{draw.tree}: A graphical representation for tree
            structures, such as syntax trees and morphological trees. 
      \item \texttt{draw.tree\_edit}: A graphical interface used to
            build and modify tree structures.
      \item \texttt{draw.plot\_graph}: A graphical tool to graph
            arbitrary functions.
      \item \texttt{draw.chart}: An interactive graphical tool used to
            experiment with chart parsers.
    \end{itemize}

\end{itemize}
}
\section*{URL}

For more information about NLTK, or to download a copy, please
visit our web page:

\texttt{http://nltk.sourceforge.net}

\end{document}

### 2002-03-20 technical/acl2002/submission.tex
- Rewrote modules section as plaintext - Minor changes - Fixed bugs in preformatted code (needed backslashes, etc)
svn/trunk@513

@@
 
+% Note: the maximum paper length is 6 pages.
+
+% Outline:
+%   - abstract
+%   - introduction
+%   - choice of programming language
+%   - design criteria
+%   - python implementation (??)
+%   - modules
+%   - uses
+%     - assignments/projects
+%     - class demonstrations
+%   - evaluation
+%   - other approaches
+%   - conclusion
+
+% To do:
+%   - FSA module description
+%   - visualization modules description
+%   - in-class demonstrations section
+%   - Get rid of Python implementation section, or at least shorten it
+%   - General clean-up
+
 \newenvironment{sv}{\scriptsize\begin{alltt}}{\end{alltt}\normalsize}
@@
 \begin{abstract}
-Students in computational linguistics courses must often learn
-a new programming language.  In some cases, such as when courses
-are offered in linguistics departments,
-the students may be learning to program for the very time.
-In order to do interesting projects, it is usually necessary for
-students to do many low-level ``housekeeping'' tasks.  At the same
-time, teachers of computational linguistics courses sometimes feel
-that they spend too much time teaching students to program, and not
-enough time teaching the subject itself.  They may even avoid
-programming assignments altogether.  However, we believe that it is
-crucial for a first computational linguistics course to include a
-strong practical component, in which students develop real programs to
-solve real problems with real data.
+Students in computational linguistics courses must often learn a new
+programming language.  In some cases, such as when courses are offered
+in linguistics departments, the students may be learning to program
+for the very time.  In order to do interesting projects, it is usually
+necessary for students to do many low-level ``housekeeping'' tasks.
+At the same time, teachers of computational linguistics courses
+sometimes feel that they spend too much time teaching students to
+program, and not enough time teaching the subject itself.  They may
+even avoid programming assignments altogether.  However, we believe
+that it is crucial for a first computational linguistics course to
+include a strong practical component, in which students develop real
+programs to solve real problems with real data.
 
@@
 
-\textbf{Ease of Use.} The primary purpose of the toolkit is to allow
-students to concentrate on building NLP systems.  The more time
-students must spend learning to use the toolkit, the less useful it
-is.
+\noindent\textbf{Ease of Use.} The primary purpose of the toolkit is
+to allow students to concentrate on building NLP systems.  The more
+time students must spend learning to use the toolkit, the less useful
+it is.
 
-\textbf{Consistency.} The toolkit should use consistent data structures
-and interfaces.
+\noindent\textbf{Consistency.} The toolkit should use consistent data
+structures and interfaces.
 
-\textbf{Extensibility.} The toolkit should easily accommodate new
-components, whether those components replicate or extend the toolkit's
-existing functionality.  Thus, the toolkit's design should be modular,
-with simple and well-defined interfaces between modules.  The toolkit
-should also be structured in such a way that it is obvious where new
-extensions would fit into the toolkit's infrastructure.
+\noindent\textbf{Extensibility.} The toolkit should easily
+accommodate new components, whether those components replicate or
+extend the toolkit's existing functionality.  Thus, the toolkit's
+design should be modular, with simple and well-defined interfaces
+between modules.  The toolkit should also be structured in such a way
+that it is obvious where new extensions would fit into the toolkit's
+infrastructure.
 
-\textbf{Documentation.} The toolkit, its data structures, and its
-implementation all need to be documented.  Three documentation types
-are necessary: technical reports, to explain the toolkit's
+\noindent\textbf{Documentation.} The toolkit, its data structures,
+and its implementation all need to be documented.  Three documentation
+types are necessary: technical reports, to explain the toolkit's
 design and implementation; tutorials, to teach students how to use the
-toolkit to perform specific tasks; and reference documentation, to describe
-every module, interface, class and method.  All nomenclature must be
-carefully chosen and consistently used.
+toolkit to perform specific tasks; and reference documentation, to
+describe every module, interface, class and method.  All nomenclature
+must be carefully chosen and consistently used.
 
@@
 
-\textbf{Simplicity.} The toolkit should structure the complexities of
-building NLP systems, not hide them.  Therefore, each class defined by 
-the toolkit should be simple enough that a student could implement it
-by the time they finish the course.
+\noindent\textbf{Simplicity.} The toolkit should structure the
+complexities of building NLP systems, not hide them.  Therefore, each
+class defined by the toolkit should be simple enough that a student
+could implement it by the time they finish the course.
 
-\textbf{Modularity.} The interaction between different components of the
-toolkit should be kept to a minimum.  In particular, it should be
-possible to use complete individual projects using small parts of the
-toolkit, without worrying about how they interact with the rest of the
-toolkit.  This will allow students to learn to use the toolkit
-incrementally throughout the course.  Modularity also makes it easier
-to change or extend the toolkit.
+\noindent\textbf{Modularity.} The interaction between different
+components of the toolkit should be kept to a minimum.  In particular,
+it should be possible to complete individual projects using small
+parts of the toolkit, without worrying about how they interact with
+the rest of the toolkit.  This will allow students to learn to use the
+toolkit incrementally throughout the course.  Modularity also makes it
+easier to change or extend the toolkit.
 
@@
 
-\textbf{Comprehensiveness.} The toolkit is not intended to provide a
-comprehensive set of tools.  Indeed, there should be a wide variety of 
-ways in which students can extend the toolkit.
+\noindent\textbf{Comprehensiveness.} The toolkit is not intended to
+provide a comprehensive set of tools.  Indeed, there should be a wide
+variety of ways in which students can extend the toolkit.
 
-\textbf{Efficiency.} The toolkit does not need to be highly optimized for
-runtime performance.  However, it should be efficient enough that
-students can use their NLP systems to perform real tasks.
+\noindent\textbf{Efficiency.} The toolkit does not need to be highly
+optimized for runtime performance.  However, it should be efficient
+enough that students can use their NLP systems to perform real tasks.
 
-\textbf{Cleverness.} Clear designs and implementations are far preferable 
-to ingenious yet indecipherable ones.
+\noindent\textbf{Cleverness.} Clear designs and implementations are
+far preferable to ingenious yet indecipherable ones.
 
@@
 
-\textbf{Interfaces.}
+% [EL] This is the only section that I think we might want to keep
+% from python impl.  But I'm not sure we even need to keep this one.
+\noindent\textbf{Interfaces.}
 Python does not directly implement interfaces.  The toolkit
@@
 
-\textbf{Typing.}
+% [EL] Mention that Python doesn't provide native type checking?  Or
+% ditch this section? :)
+\noindent\textbf{Typing.}
 Type checking is an important safety device, allowing programmers to
@@
 
-\textbf{Advanced Language Features.}
-The use of advanced language features and programming styles could
-potentially make the toolkit much more difficult to use or to
-understand.  The fewer language features that a student must learn in
-order to use the toolkit, the faster they can start developing NLP
-systems.  The language features used by
-the toolkit are summarized by figure \ref{fig:feature}.
+%\noindent\textbf{Advanced Language Features.}
+%The use of advanced language features and programming styles could
+%potentially make the toolkit much more difficult to use or to
+%understand.  The fewer language features that a student must learn in
+%order to use the toolkit, the faster they can start developing NLP
+%systems.  The language features used by
+%the toolkit are summarized by figure \ref{fig:feature}.
 
-\begin{figure}
-\begin{centering}
-\begin{tabular}{|l|l|}
-\hline
-\textbf{Language Feature} & \textbf{Use} \\
-\hline
-Classes & used \\
-Exceptions & error conditions only \\
-Operator Overloading & used \\
-Python 2.0 Features & used \\
-Python 2.1 Features & not used \\
-Function Arguments & minimal use \\
-Lambda Functions & minimal use \\
-Mapping and Filtering & not used \\
-List Comprehensions & used \\
-Tkinter & used \\
-\hline
-\end{tabular}\\
-\end{centering}
-\caption{Advanced language features used by NLTK.}
-\label{fig:feature}
-\end{figure}
+%\begin{figure}
+%\begin{centering}
+%\begin{tabular}{|l|l|}
+%\hline
+%\noindent\textbf{Language Feature} & \textbf{Use} \\
+%\hline
+%Classes & used \\
+%Exceptions & error conditions only \\
+%Operator Overloading & used \\
+%Python 2.0 Features & used \\
+%Python 2.1 Features & not used \\
+%Function Arguments & minimal use \\
+%Lambda Functions & minimal use \\
+%Mapping and Filtering & not used \\
+%List Comprehensions & used \\
+%Tkinter & used \\
+%\hline
+%\end{tabular}\\
+%\end{centering}
+%\caption{Advanced language features used by NLTK.}
+%\label{fig:feature}
+%\end{figure}
 
 \section{Modules}
+% [EL]: I think that this section works better as a textual
+% description, rather than a long list.  Also, we're going to be short
+% on space.. :)
+% What order should these be in?  Put more impressive stuff at the
+% beginning and end?  Maybe
+% parsing;tagging;visualization;classification?  or swap viz and
+% classification?
 
-NLTK currently includes the following modules.
+The toolkit is implemented as a collection of independant
+\emph{modules}, each of which defines a specific data structure or
+task.
 
-\subsection*{Basics}
+A set of core modules defines basic data types and processing tasks
+that are used througout the toolkit.  The \texttt{token} module
+defines basic classes for processing individual elements of text, such
+as words or sentences.  The \texttt{tree} module provides data
+structures for representing tree strucutres over text, such as syntax
+trees and morphological trees.  The \texttt{probability} module
+defines classes that encode frequency distributions and probability
+distributions, including a variety of smoothing techniques.
 
-\begin{description}
-\item[token:] Basic classes for processing individual
-      elements of text, such as words or sentences. 
-\item[tree:] Classes for representing tree structures
-      over text (such as syntax trees and morphological trees).
-\item[probability:] Classes that encode frequency
-      distributions and probability distributions.
-\end{description}
+The remaining modules define data structures and interfaces for
+performing specific NLP tasks.
 
-\subsection*{Tagging}
+\subsection*{Parsing Modules}
 
-\begin{description}
-\item[tagger:] A standard interface to tag each token of
-      a text with supplementary information, such as its part of
-      speech; and several implementations of that interface.
-\end{description}
+The \texttt{parser} module defines a high-level interface for
+producing trees that represent the structure of texts.  The
+\texttt{chunkparser} module defines a subinterface for parsers that
+identify non-overlapping linguistic groups (such as noun phrases) in
+unrestricted text.
 
-\subsection*{Parsing}
+Three modules provide implementations for these abstract interfaces.
+The \texttt{srparser} module implements a simple shift-reduce parser.
+The \texttt{chartparser} module provides a flexible parser that uses a
+\emph{chart} to record hypotheses about syntactic constituents.  And
+the \texttt{rechunkparser} module defines a transformational
+regular-expression based implementation of the chunk parser interface.
 
-\begin{description}
-\item[parser:] A standard interface to produce trees
-      representing the structure of texts.
-\item[chartparser:] A flexible parser implementation
-      that uses a \emph{chart} to record hypotheses about
-      syntactic constituents.
-\item[srparser\_template:] A partial implementation of a
-      shift-reduce parser; completing the implementation was a
-      student exercise.
-\item[chunkparser:] A standard interface for robust
-      parsers used to identify non-overlapping linguistic groups
-      (such as noun phrases) in unrestricted text.
-\item[rechunkparser:] A regular-expression based
-      implementation of the chunk parser interface.
-\end{description}
+\subsection*{Tagging Modules}
+
+The \texttt{tagger} module defines a standard interface for augmenting
+each token of a text with supplementary information, such as its part
+of speech or its WordNet synset tag; and provides several
+different implementations for that interface.
+
+\subsection*{Finite State Automata}
+
+\texttt{fsa}.  Planning to add HMMs??
+
+\subsection*{Visualization}
+
+\texttt{draw.tree}
+\texttt{draw.tree\_edit}
+\texttt{draw.plot\_graph}
+\texttt{draw.chart}
+\texttt{draw.fsa}
 
@@
 
-\begin{description}
-\item[classifier:] A standard interface for classifying
-      texts into categories.
-\item[classifier.feature:] A standard way of encoding
-      the information used to make classification decisions.
-\item[classifier.naivebayes:] A text classifier
-      implementation based on the Naive Bayes assumption.
-\item[classifier.maxent:] An implementation of the
-      maximum entropy model for text classification; and
-      implementations of the GIS and IIS algorithms for training
-      the classifier.
-\item[classifier.featureselection:] A standard interface
-      for choosing which features are relevant for making
-      classification decisions.
-\end{description}
+The \texttt{classifier} module defines a standard interface for
+classifying texts into categories.  This interface is currently
+implemented by two modules: the \texttt{classifier.naivebayes} module
+defines a text classifier based on the Naive Bayes assumption; and the
+\texttt{classifier.maxent} module defines the maximum entropy model
+for text classification, and includes two algorithms for training the
+model: Generalized Iterative Scaling and Improved Iterative Scaling.
 
-\subsection*{Visualization}
+The \texttt{classifier.feature} module provides a standard encoding
+for the information that is used to make decisions for a particular
+classification task.  This standard encoding allows students to
+experiment with the differences between different text classification
+algorithms, using identical feature sets.
 
-\begin{description}
-\item[draw.tree:] A graphical representation for tree
-      structures, such as syntax trees and morphological trees. 
-\item[draw.tree\_edit:] A graphical interface used to
-      build and modify tree structures.
-\item[draw.plot\_graph:] A graphical tool to graph
-      arbitrary functions.
-\item[draw.chart:] An interactive graphical tool used to
-      experiment with chart parsers.
-\end{description}
+The \texttt{classifier.featureselection} module defines a standard
+interface for choosing which features are relevant for a particular
+classification task.  Good feature selection can signifigantly improve
+classification performance.
+
+%NLTK currently includes the following modules.
+
+%\subsection*{Basics}
+
+%\begin{description}
+%\item[token:] Basic classes for processing individual
+%      elements of text, such as words or sentences. 
+%\item[tree:] Classes for representing tree structures
+%      over text (such as syntax trees and morphological trees).
+%\item[probability:] Classes that encode frequency
+%      distributions and probability distributions.
+%\end{description}
+
+%\subsection*{Tagging}
+
+%\begin{description}
+%\item[tagger:] A standard interface to tag each token of
+%      a text with supplementary information, such as its part of
+%      speech; and several implementations of that interface.
+%\end{description}
+
+%\subsection*{Parsing}
+
+%\begin{description}
+%\item[parser:] A standard interface to produce trees
+%      representing the structure of texts.
+%\item[chartparser:] A flexible parser implementation
+%      that uses a \emph{chart} to record hypotheses about
+%      syntactic constituents.
+%\item[srparser\_template:] A partial implementation of a
+%      shift-reduce parser; completing the implementation was a
+%      student exercise.
+%\item[chunkparser:] A standard interface for robust
+%      parsers used to identify non-overlapping linguistic groups
+%      (such as noun phrases) in unrestricted text.
+%\item[rechunkparser:] A regular-expression based
+%      implementation of the chunk parser interface.
+%\end{description}
+
+%\subsection*{Text Classification}
+
+%\begin{description}
+%\item[classifier:] A standard interface for classifying
+%      texts into categories.
+%\item[classifier.feature:] A standard way of encoding
+%      the information used to make classification decisions.
+%\item[classifier.naivebayes:] A text classifier
+%      implementation based on the Naive Bayes assumption.
+%\item[classifier.maxent:] An implementation of the
+%      maximum entropy model for text classification; and
+%      implementations of the GIS and IIS algorithms for training
+%      the classifier.
+%\item[classifier.featureselection:] A standard interface
+%      for choosing which features are relevant for making
+%      classification decisions.
+%\end{description}
+
+%\subsection*{Visualization}
+
+%\begin{description}
+%\item[draw.tree:] A graphical representation for tree
+%      structures, such as syntax trees and morphological trees. 
+%\item[draw.tree\_edit:] A graphical interface used to
+%      build and modify tree structures.
+%\item[draw.plot\_graph:] A graphical tool to graph
+%      arbitrary functions.
+%\item[draw.chart:] An interactive graphical tool used to
+%      experiment with chart parsers.
+%\end{description}
 
@@
 NLTK can be used for simple and advanced projects, as well as
-everything in between.  In the simplest assignments, students
-tweak an existing module to alter its behavior.  The variety of
-existing modules makes it easy to create simple assignments.
-A more challenging task is to develop a new module.  Here, NLTK provides
-some useful starting points: predefined interfaces and data structures, and
+everything in between.  In the simplest assignments, students tweak an
+existing module to alter its behavior.  The variety of existing
+modules makes it easy to create simple assignments.  A more
+challenging task is to develop a new module.  Here, NLTK provides some
+useful starting points: predefined interfaces and data structures, and
 existing modules that implement the same interface.  Advanced projects
 typically involve the development of entirely new functionality for a
-previously unsupported NLP task, or the development of a complete system
-out of existing and new modules.
+previously unsupported NLP task, or the development of a complete
+system out of existing and new modules.
 
@@
   UnChunkRule('<IN|VB.*|CC|MD|RB.*>'),
-  UnChunkRule("<,|\.|``|''>"),
+  UnChunkRule("<,|{\textbackslash}.|``|''>"),
   MergeRule('<NN.*|DT|JJ.*|CD>', '<NN.*|DT|JJ.*|CD>'),
@@
 cascade = [
-  ChunkRule('<\$|CD|DT|EX|PDT
-             |PRP.*|WP.*|\#|FW
+  ChunkRule('<{\textbackslash}\$|CD|DT|EX|PDT
+             |PRP.*|WP.*|{\textbackslash}\#|FW
              |JJ.*|NN.*|POS|RBS|WDT>*')
@@
 
+% [EL] I'm not sure this is necessary:
 The main program which surrounds the student's work is moderately
 simple as well.  The line replaced by students is underlined.
-
+% [EL] I find the undeline hard to read; change to italics or bold or
+% vbar in the margin?  Or maybe a \Rightarrow of some sort in the margin?
 \begin{sv}
@@
 
-sent_tok = RETokenizer('(=======+)?\\n\\n+', 0)
+sent_tok = RETokenizer('(=======+)?{\textbackslash}{\textbackslash}n{\textbackslash}{\textbackslash}n+', 0)
 sentences = []
@@
 for file in files:
-    print 'Loading %s...' % file
+    print 'Loading \%s...' \% file
     text = open(path+file).read()

### 2002-03-20 technical/acl2002/submission.tex
- Minor revisions for "design criteria" intro - commented out "python implementation" section - Added 1st draft text for modules:fsa - Added 1st draft text for modules:visualization - Revised intro to "assignments and projects" - Put chunk parsing in a subsubsection* - Commented out explanation of what code they added for chunkparser - Added 1st draft text for class demonstrations
svn/trunk@514

@@
 %   - General clean-up
+%   - ispell
 
@@
 Apart from the practical component, computational linguistics courses
-may also depend on software for in-class
-demonstrations.  This context calls for highly interactive graphical
-user interfaces making it possible to view program state (e.g. the
-chart of a chart parser), observe program execution step-by-step
-(e.g. execution of a finite-state machine), and even make minor
-modifications to programs in response to ``what if'' questions from
-the class.  Because of these difficulties it is common to avoid live
-demonstrations, and keep classes for theoretical presentations only.
-Apart from being dull, this approach leaves students to solve
-important practical problems on their own, or to deal with them less
-efficiently in office hours.
+may also depend on software for in-class demonstrations.  This context
+calls for highly interactive graphical user interfaces making it
+possible to view program state (e.g. the chart of a chart parser),
+observe program execution step-by-step (e.g. execution of a
+finite-state machine), and even make minor modifications to programs
+in response to ``what if'' questions from the class.  Because of these
+difficulties it is common to avoid live demonstrations, and keep
+classes for theoretical presentations only.  Apart from being dull,
+this approach leaves students to solve important practical problems on
+their own, or to deal with them less efficiently in office hours.
 
@@
 
-Several criteria must be considered in the design and implementation
-of the toolkit.  The design criteria are divided into primary and
+Several criteria were considered in the design and implementation of
+the toolkit.  The design criteria are divided into primary and
 secondary criteria, and within those sections are listed in the order
-of their importance.  A set of explicit non-requirements is also
-given.  The toolkit is not expected to satisfy any of these
-properties.
+of their importance.  It was also important to decide what goals the
+toolkit would \emph{not} attempt to accomplish; we therefore include
+an explicit set of non-requirements, which the toolkit is not expected
+to satisfy.
 
@@
 
-\section{Python Implementation}
-
-The structure of the Python language directly affects the design and
-implementation of the toolkit.  This section discusses how the design
-of the toolkit is affected by the Python language.
-
-% [EL] This is the only section that I think we might want to keep
-% from python impl.  But I'm not sure we even need to keep this one.
-\noindent\textbf{Interfaces.}
-Python does not directly implement interfaces.  The toolkit
-therefore implements interfaces as simple classes, all of whose methods 
-raise \texttt{AssertionError} exceptions.  A description of the
-functionality provided by the interface is contained in the
-class's documentation string.  Each method's documentation string
-specifies the behavior of that method.
-In order to implement an interface, a class should include that
-interface as one of its bases, and should override every
-(non-optional) method defined by the interface.
-Classes that define interfaces are named with a trailing ``I,'' such
-as \texttt{TokenizerI} or \texttt{EventI}.
-
-% [EL] Mention that Python doesn't provide native type checking?  Or
-% ditch this section? :)
-\noindent\textbf{Typing.}
-Type checking is an important safety device, allowing programmers to
-quickly locate errors in their code.  This is especially important for
-novice programmers, who have less experience with debugging.  Every
-function and method defined by the toolkit will check the types of its
-its arguments.  Given the impact on performance, the level of type checking
-will be adjustable.
-
+%\section{Python Implementation}
+%
+%The structure of the Python language directly affects the design and
+%implementation of the toolkit.  This section discusses how the design
+%of the toolkit is affected by the Python language.
+%
+%% [EL] This is the only section that I think we might want to keep
+%% from python impl.  But I'm not sure we even need to keep this one.
+%\noindent\textbf{Interfaces.}
+%Python does not directly implement interfaces.  The toolkit
+%therefore implements interfaces as simple classes, all of whose methods 
+%raise \texttt{AssertionError} exceptions.  A description of the
+%functionality provided by the interface is contained in the
+%class's documentation string.  Each method's documentation string
+%specifies the behavior of that method.
+%In order to implement an interface, a class should include that
+%interface as one of its bases, and should override every
+%(non-optional) method defined by the interface.
+%Classes that define interfaces are named with a trailing ``I,'' such
+%as \texttt{TokenizerI} or \texttt{EventI}.
+%
+%% [EL] Mention that Python doesn't provide native type checking?  Or
+%% ditch this section? :)
+%\noindent\textbf{Typing.}
+%Type checking is an important safety device, allowing programmers to
+%quickly locate errors in their code.  This is especially important for
+%novice programmers, who have less experience with debugging.  Every
+%function and method defined by the toolkit will check the types of its
+%its arguments.  Given the impact on performance, the level of type checking
+%will be adjustable.
+%
 %\noindent\textbf{Advanced Language Features.}
@@
 %the toolkit are summarized by figure \ref{fig:feature}.
-
+%
 %\begin{figure}
@@
 % What order should these be in?  Put more impressive stuff at the
-% beginning and end?  Maybe
-% parsing;tagging;visualization;classification?  or swap viz and
-% classification?
+% beginning and end?
 
@@
 The remaining modules define data structures and interfaces for
-performing specific NLP tasks.
+performing specific NLP tasks.  This list of modules will grow over
+time, as we add new tasks and algorithms to the toolkit.
 
@@
 
-\texttt{fsa}.  Planning to add HMMs??
+% Steven: do you want to add anything here?  Say anything about HMMS? 
+% Will you have implemented more interfaces by the time of ACL?
+The \texttt{fsa} module defines a data type for encoding finite state
+automata; and an interface for creating automata from
+regular expressions.
 
@@
 
-\texttt{draw.tree}
-\texttt{draw.tree\_edit}
-\texttt{draw.plot\_graph}
-\texttt{draw.chart}
-\texttt{draw.fsa}
+A set of visualization modules define graphical interfaces for viewing
+and editing data structures; and graphical tools for experimenting
+with tasks.  The \texttt{draw.tree} module provides a
+simple graphical interface for displaying tree structures; and the
+\texttt{draw.tree\_edit} module provides an interface for building and
+modifying tree structures.  The \texttt{draw.plot\_graph} module can be
+used to graph mathematical functions.  The \texttt{draw.fsa} module
+provides a graphical tool for displaying and simulating finite state
+automata.  The \texttt{draw.chart} module provides an interactive
+graphical tool for experimenting with chart parsers.
 
@@
 
+%% Old content for this section:
 %NLTK currently includes the following modules.
-
+%
 %\subsection*{Basics}
-
+%
 %\begin{description}
@@
 %\end{description}
-
+%
 %\subsection*{Tagging}
-
+%
 %\begin{description}
@@
 %\end{description}
-
+%
 %\subsection*{Parsing}
-
+%
 %\begin{description}
@@
 %\end{description}
-
+%
 %\subsection*{Text Classification}
-
+%
 %\begin{description}
@@
 %\end{description}
-
+%
 %\subsection*{Visualization}
-
+%
 %\begin{description}
@@
 
-NLTK can be used for simple and advanced projects, as well as
-everything in between.  In the simplest assignments, students tweak an
-existing module to alter its behavior.  The variety of existing
-modules makes it easy to create simple assignments.  A more
-challenging task is to develop a new module.  Here, NLTK provides some
-useful starting points: predefined interfaces and data structures, and
-existing modules that implement the same interface.  Advanced projects
-typically involve the development of entirely new functionality for a
-previously unsupported NLP task, or the development of a complete
-system out of existing and new modules.
+NLTK can be used to create student assignments of varying difficulty
+and scope.  
+% Use a module
+In the simplest assignments, students experiment with an existing
+module.  The wide variety of existing modules provide many opportunies
+for creating these simple assignments.
+% Edit/extend a module.
+Once students become more familiar with the toolkit, they can be asked
+to make minor changes or extensions to an existing module.
+% Develop a new module.
+A more challenging task is to develop a new module.  Here, NLTK
+provides some useful starting points: predefined interfaces and data
+structures, and existing modules that implement the same interface.
+% Advanced projects..
+Advanced projects typically involve the development of entirely new
+functionality for a previously unsupported NLP task, or the
+development of a complete system out of existing and new modules.
 
+\subsubsection*{Example: Chunk Parsing}
+
+% Akward
 Here we illustrate the use of NLTK in an assignment of moderate
@@
 
-% [EL] I'm not sure this is necessary:
-The main program which surrounds the student's work is moderately
-simple as well.  The line replaced by students is underlined.
-% [EL] I find the undeline hard to read; change to italics or bold or
-% vbar in the margin?  Or maybe a \Rightarrow of some sort in the margin?
-\begin{sv}
-from nltk.token import *
-from nltk.chunkparser import *
-from nltk.rechunkparser import *
-import os
-
-path = "/cis530/data/wsj/tagged/"
-
-sent_tok = RETokenizer('(=======+)?{\textbackslash}{\textbackslash}n{\textbackslash}{\textbackslash}n+', 0)
-sentences = []
-files = os.listdir(path)
-for file in files:
-    print 'Loading \%s...' \% file
-    text = open(path+file).read()
-    sentences += sent_tok.tokenize(text, source=file, unit='s')
-
-chunkscore = ChunkScore()
-
-\underline{cascade = [ ??? ]}
-chunkparser = REChunkParser(cascade)
-
-ctt = ChunkedTaggedTokenizer()
-for s in sentences:
-    correct = ctt.tokenize(s.type(), source=s.loc())
-    unchunked = unchunk(correct)
-    guess = chunkparser.parse(unchunked)
-    chunkscore.score(correct, guess)
-
-print chunkscore
-\end{sv}
+%% [EL] I'm not sure this is necessary:
+%The main program which surrounds the student's work is moderately
+%simple as well.  The line replaced by students is underlined.
+%% [EL] I find the undeline hard to read; change to italics or bold or
+%% vbar in the margin?  Or maybe a \Rightarrow of some sort in the margin?
+%\begin{sv}
+%from nltk.token import *
+%from nltk.chunkparser import *
+%from nltk.rechunkparser import *
+%import os
+%
+%path = "/cis530/data/wsj/tagged/"
+%
+%sent_tok = RETokenizer('(=======+)?{\textbackslash}{\textbackslash}n{\textbackslash}{\textbackslash}n+', 0)
+%sentences = []
+%files = os.listdir(path)
+%for file in files:
+%    print 'Loading \%s...' \% file
+%    text = open(path+file).read()
+%    sentences += sent_tok.tokenize(text, source=file, unit='s')
+%
+%chunkscore = ChunkScore()
+%
+%\underline{cascade = [ ??? ]}
+%chunkparser = REChunkParser(cascade)
+%
+%ctt = ChunkedTaggedTokenizer()
+%for s in sentences:
+%    correct = ctt.tokenize(s.type(), source=s.loc())
+%    unchunked = unchunk(correct)
+%    guess = chunkparser.parse(unchunked)
+%    chunkscore.score(correct, guess)
+% 
+%print chunkscore
+%\end{sv}
 
 The score for each project is the precision, recall and F-measure.
-Figure~\ref{contest} shows the results; students with the best two scores
-were presented with prizes.
+Figure~\ref{fig:contest} shows the results; students with the best two
+scores were presented with prizes.
 
@@
 \centerline{\epsfig{figure=contest.ps,width=\linewidth}}
-\caption{Precision/Recall Graph for Chunking Competition}\label{contest}
+\caption{Precision/Recall Graph for Chunking Competition}\label{fig:contest}
 \vspace*{2ex}\hrule
@@
 
-[chart parsing example, including screenshot]
+NLTK provides graphical tools that can be used to give in-class
+demonstrations, which help to explain basic NLP concepts and
+algorithms.  These interactive tools can be used to display relevant
+data structures, and to show step-by-step executions of algorithms.
+Both data structures and control flow can be easily modified during
+the demonstration, in response questions from the class.
+
+Since these graphical tools are included with the toolkit, they can
+also be used by students.  This allows students to experiment with the
+algorithms that they have seen presented in class.
+
+\subsubsection*{Example: The Chart Parsing Tool}
+
+The chart parsing tool can be used to explain the basic concepts
+behind chart parsing, and to show how the algorithm works.  Chart
+parsing is a flexible parsing algorithm that uses a data structure
+called a \emph{chart} to record hypotheses about syntactic
+constituents.  Each hypothesis is represented by a single \emph{edge}
+on the chart.  A set of \emph{rules} determine when new edges can be
+added to the chart.  This set of rules controls the overall behavior
+of the parser (e.g., whether it parses top-down or bottom-up).
+
+The chart parsing tool demonstrates the process of parsing a single
+sentence, with a given grammar and lexicon.  Its display is divided
+into three sections: the bottom section displays the chart; the middle
+section displays the sentence; and the top section displays the
+partial syntax tree corresponding to the selected edge.  Buttons along
+the bottom of the window are used to control the execution of the
+algorithm.  The main display window for the chart parsing tool is
+shown in Figure~\ref{fig:chartparse}.
+
+This tool can be used to explain several different aspects of chart
+parsing.  First, it can be used to explain the basic chart data
+structure, and to show how edges can represent hypotheses about
+syntactic constituents.  It can then be used to demonstrate and
+explain the individual rules that the chart parsers uses to create new
+edges.  Finally, it can be used to show how these individual rules
+combine to find a complete parse for the sentence.
+
+% Is ``user'' a good word here?  ``lecturer''?
+To reduce the overhead of setting up demonstrations during lecture,
+the user can define a list of preset charts.  The tool can then be
+reset to any one of these charts at any time.
+
+The chart parsing tool allows for flexible control of the parsing
+algorithm.  At each step of the algorithm, the user can select which
+rule or strategy they wish to apply.  This allows the user to
+experiment with mixing different strategies (e.g., top-down and
+bottom-up).  The user can exercise fine-grained control over the
+algorithm by selecting which edge they wish to apply a rule to.  This
+flexibility allows lecturers to use the tool to respond to a wide
+variety of questions; and allows students to experiment with different
+variations on the chart parsing algorithm.
+
+\begin{figure}
+%\centerline{\epsfig{figure=chart.ps,width=\linewidth}}
+\caption{Chart Parsing Tool}\label{fig:chartparse}
+\vspace*{2ex}\hrule
+\end{figure}
 

### 2002-03-20 technical/acl2002/submission.tex
- Added misc comments - Italicized design criteria (so they don't look so much like   other headings).
svn/trunk@515

@@
 % To do:
-%   - FSA module description
-%   - visualization modules description
-%   - in-class demonstrations section
-%   - Get rid of Python implementation section, or at least shorten it
+%   - Fill in OVERVIEW in the intro
+%   - Write up evaluation
+%   - Write up other approaches
+%   - Write up conclusion
+%   - Get rid of Python implementation section, or at least shorten it?
 %   - General clean-up
 %   - ispell
+
+% We need more content on:
+%   - documentation
+%   - corpora?
 
@@
 
-\noindent\textbf{Ease of Use.} The primary purpose of the toolkit is
+\noindent\textit{\textbf{Ease of Use.}} The primary purpose of the toolkit is
 to allow students to concentrate on building NLP systems.  The more
@@
 
-\noindent\textbf{Consistency.} The toolkit should use consistent data
+\noindent\textit{\textbf{Consistency.}} The toolkit should use consistent data
 structures and interfaces.
 
-\noindent\textbf{Extensibility.} The toolkit should easily
+\noindent\textit{\textbf{Extensibility.}} The toolkit should easily
 accommodate new components, whether those components replicate or
@@
 
-\noindent\textbf{Documentation.} The toolkit, its data structures,
+\noindent\textit{\textbf{Documentation.}} The toolkit, its data structures,
 and its implementation all need to be documented.  Three documentation
@@
 
-\noindent\textbf{Simplicity.} The toolkit should structure the
+\noindent\textit{\textbf{Simplicity.}} The toolkit should structure the
 complexities of building NLP systems, not hide them.  Therefore, each
@@
 
-\noindent\textbf{Modularity.} The interaction between different
+\noindent\textit{\textbf{Modularity.}} The interaction between different
 components of the toolkit should be kept to a minimum.  In particular,
@@
 
-\noindent\textbf{Comprehensiveness.} The toolkit is not intended to
+\noindent\textit{\textbf{Comprehensiveness.}} The toolkit is not intended to
 provide a comprehensive set of tools.  Indeed, there should be a wide
@@
 
-\noindent\textbf{Efficiency.} The toolkit does not need to be highly
+\noindent\textit{\textbf{Efficiency.}} The toolkit does not need to be highly
 optimized for runtime performance.  However, it should be efficient
@@
 
-\noindent\textbf{Cleverness.} Clear designs and implementations are
+\noindent\textit{\textbf{Cleverness.}} Clear designs and implementations are
 far preferable to ingenious yet indecipherable ones.

### 2002-03-21 technical/acl2002/submission.tex
- Added section on documentation - General clean-up sweep - Commented out abstract
svn/trunk@519

@@
 NLTK, the Natural Language Toolkit, is a suite of program modules,
-tutorials and problem sets, providing ready-to-use CL courseware.
-NLTK covers symbolic and statistical
-NLP, and is interfaced to annotated corpora.
-Students augment and replace existing components, learning
-structured programming by example, and manipulating sophisticated
-models from the outset.}
+tutorials and problem sets, providing ready-to-use computational
+linguistics courseware.  NLTK covers symbolic and statistical natural
+language processing, and is interfaced to annotated corpora.  Students
+augment and replace existing components, learning structured
+programming by example, and manipulating sophisticated models from the
+outset.}
+
 \paperid{Pxxxx}
@@
 
+% To do:
+%   - Write up evaluation
+%   - Write up other approaches
+%   - Write up conclusion
+%   - General clean-up
+%   - ispell
+%   - Should URLs include http:// or not (we should be consistant)?
+
 % Outline:
@@
 %   - design criteria
-%   - python implementation (??)
 %   - modules
@@
 
-% To do:
-%   - Fill in OVERVIEW in the intro
-%   - Write up evaluation
-%   - Write up other approaches
-%   - Write up conclusion
-%   - Get rid of Python implementation section, or at least shorten it?
-%   - General clean-up
-%   - ispell
-
-% We need more content on:
-%   - documentation
-%   - corpora?
-
 \newenvironment{sv}{\scriptsize\begin{alltt}}{\end{alltt}\normalsize}
@@
 
-\begin{abstract}
-Students in computational linguistics courses must often learn a new
-programming language.  In some cases, such as when courses are offered
-in linguistics departments, the students may be learning to program
-for the very time.  In order to do interesting projects, it is usually
-necessary for students to do many low-level ``housekeeping'' tasks.
-At the same time, teachers of computational linguistics courses
-sometimes feel that they spend too much time teaching students to
-program, and not enough time teaching the subject itself.  They may
-even avoid programming assignments altogether.  However, we believe
-that it is crucial for a first computational linguistics course to
-include a strong practical component, in which students develop real
-programs to solve real problems with real data.
+% ===================== Abstract =====================
+% [EL] I think that this abstract should be left out, or shortened and
+% merged with the \summary.  The text in the 1st paragraph of the intro
+% seems much more direct and relevant to our project than the 1st
+% paragraph of this abstract (in particular, the 1st sentence seems like
+% a strange way to start). :) I don't think that the second paragraph
+% needs to be included in the abstract -- it's not a central point of
+% our paper; and the third paragraph is almost identical to the \summary
+% (but not quite).  For now, I'm just commenting out the abstract.
 
-Python is a new object-oriented scripting language which runs on all
-platforms.  Python has been praised as ``executable pseudocode,'' since
-programs are so easy to write.  Recently, we have been developing
-NLTK, an open-source Natural Language Toolkit written in Python.
-In this presentation, we will motivate, describe and demonstrate NLTK.
+%\begin{abstract}
+%Students in computational linguistics courses must often learn a new
+%programming language.  In some cases, such as when courses are offered
+%in linguistics departments, the students may be learning to program
+%for the very time.  In order to do interesting projects, it is usually
+%necessary for students to do many low-level ``housekeeping'' tasks.
+%At the same time, teachers of computational linguistics courses
+%sometimes feel that they spend too much time teaching students to
+%program, and not enough time teaching the subject itself.  They may
+%even avoid programming assignments altogether.  However, we believe
+%that it is crucial for a first computational linguistics course to
+%include a strong practical component, in which students develop real
+%programs to solve real problems with real data.
+%
+%Python is a new object-oriented scripting language which runs on all
+%platforms.  Python has been praised as ``executable pseudocode,'' since
+%programs are so easy to write.  Recently, we have been developing
+%NLTK, an open-source Natural Language Toolkit written in Python.
+%In this presentation, we will motivate, describe and demonstrate NLTK.
+%
+%NLTK, the Natural Language Toolkit, is a suite of program modules,
+%tutorials and problem sets.  NLTK covers symbolic and statistical
+%natural language processing, and is interfaced to annotated corpora.
+%Students augment and replace existing NLTK components, learning
+%structured programming by example, and manipulating sophisticated
+%models from the outset.  Along with extensive documentation and
+%problem sets, NLTK provides self-contained, ready-to-use CL
+%courseware.
+%\end{abstract}
 
-NLTK, the Natural Language Toolkit, is a suite of program modules,
-tutorials and problem sets.  NLTK covers symbolic and statistical
-natural language processing, and is interfaced to annotated corpora.
-Students augment and replace existing NLTK components, learning
-structured programming by example, and manipulating sophisticated
-models from the outset.  Along with extensive documentation and
-problem sets, NLTK provides self-contained, ready-to-use CL
-courseware.
-\end{abstract}
-
+% ===================== Introduction =====================
 \section{Introduction}
@@
 often faced with the challenge of setting up a practical programming
-component for student assignments and projects.  This is difficult not
-just because of the variety of data structures, but also because of
-the diverse range of topics which may need to be included in the
-syllabus.  A widespread practice is to teach multiple programming
-languages, where each language provides native data structures and functions
-that are a good fit for the task at hand (e.g. Prolog for parsing,
-Perl for corpus processing, a finite-state toolkit for morphological
-analysis).  By relying on the built-in features of various languages,
-the teacher avoids having to develop a lot of software infrastructure.
+component for student assignments and projects.  This is a difficult
+task because different comptational linguistics domains require a
+variety of different data structures and functionality; and because a
+diverse range of topics may need to be included in the syllabus.
+
+A widespread practice is to employ multiple programming languages,
+where each language provides native data structures and functions that
+are a good fit for the task at hand.  For example, a course might use
+Prolog for parsing, Perl for corpus processing, and a finite-state
+toolkit such for morphological analysis.  By relying on the built-in
+features of various languages, the teacher avoids having to develop a
+lot of software infrastructure.
+
 An unfortunate consequence is that a significant part of the course
 must be devoted to teaching these languages.  Further, many
-interesting projects require the languages to be bridged, e.g. a
-project that involved syntactic parsing of corpus data from a
-morphologically rich language would involve all three languages
-mentioned above (Perl for file I/O, format conversions and overall
-program control, with calls out to a finite state toolkit for
-morphological analysis and to a Prolog engine for parsing).  It is
-clear that these considerable overheads and shortcomings warrant a
-fresh approach.
+interesting projects span a variety of domains, and would require that
+multiple languages be bridged.  For example, a student project that
+involved syntactic parsing of corpus data from a morphologically rich
+language might involve all three of the languages mentioned above:
+Perl for string processing; a finite state toolkit for morphological
+analysis; and Prolog for parsing.
+% [EL] This was more precise, but harder to read/parse:
+%    (Perl for file I/O, format conversions and
+%    overall program control, with calls out to a finite state toolkit for
+%    morphological analysis and to a Prolog engine for parsing).  
+It is clear that these considerable overheads and shortcomings warrant
+a fresh approach.
 
+% [EL] ``difficulties'' doesn't seem like the right word here?
 Apart from the practical component, computational linguistics courses
 may also depend on software for in-class demonstrations.  This context
-calls for highly interactive graphical user interfaces making it
+calls for highly interactive graphical user interfaces, making it
 possible to view program state (e.g. the chart of a chart parser),
@@
 
-In this paper we describe a new approach to the above challenges,
-a streamlined and flexible way of organizing the practical component
-of an introductory computational linguistics course.  We describe
-NLTK, the Natural Language Toolkit, which we have developed in
-conjunction with a course we have taught at the University of Pennsylvania.
+% [EL] akward
+In this paper we describe a new approach to the above challenges, a
+streamlined and flexible way of organizing the practical component of
+an introductory computational linguistics course.  We describe NLTK,
+the Natural Language Toolkit, which we have developed in conjunction
+with a course we have taught at the University of Pennsylvania.
 
-OVERVIEW OF THE PAPER
+% Overview
+% [EL] Given that we're short on space, I decided that this wasn't
+% really necessary for a 6-page paper.  If we do include it, it needs
+% to be completed (I only went up through sec:uses).
+%%Section \ref{sec:python} discusses why we decided to implement the
+%%toolkit using Python.  Section \ref{sec:criteria} describes several
+%%important considerations and desiderata that guided the design and
+%%implementation of the toolkit.  Sections \ref{sec:modules} and
+%%\ref{sec:documentation} describe the current contents of the toolkit.
+%%Section \ref{sec:uses} explains how the toolkit can be used for both
+%%assignments and class demonstrations.
 
-All materials discussed here are available under an open
-source license from \url{nltk.sf.net}.
-NLTK runs on most platforms, including Windows, OS X, Linux, and
-UNIX.
+All materials discussed here are available under an open source
+license from \url{nltk.sf.net}.  NLTK runs on most platforms,
+including Windows, OS X, Linux, and UNIX.
 
-\section{Choice of programming language}
+% ===================== Python =====================
+\section{Choice of Programming Language}
+\label{sec:python}
 
 The most basic step in setting up a practical component is choosing a
-suitable programming language.  Here we list the desiderata that influence
-our choice.
+suitable programming language.  A number of desiderata influenced our
+choice.  First, the language must have a shallow learning curve, so
+that novice programmers get immediate rewards for their efforts.
+Second, the language must support rapid prototyping and a short
+develop/test cycle; an obligatory compilation step is a serious
+detraction.  Third, the code should be self-documenting, with a
+transparent syntax and semantics.  Fourth, it should be easy to write
+structured programs, ideally object-oriented but without the burden
+associated with languages like C++.  Finally, the language must have
+an easy-to-use graphics library to support the development of simple
+graphical user interfaces.
 
-First, the language must have a shallow learning curve, so that novice
-programmers get immediate rewards for their efforts.
-Second, the language must support rapid prototyping and a short develop/test cycle;
-an obligatory compilation step is a serious detraction.
-Third, the code should be self-documenting,
-with a transparent syntax and semantics.
-Fourth, it should be easy to write structured programs, ideally object-oriented
-but without the burden associated with languages like C++.
-Finally, the language must have an easy-to-use graphics library to support
-the development of simple graphical user interfaces.
-
-In surveying the available languages, we believe that Python offers an especially good
-fit to the above requirements.  Python is an object-oriented scripting language
-developed by Guido van Rossum [REF] and available on all platforms
-\url{www.python.org}.
-Python offers a shallow learning curve by virtue of the
-fact that it was designed to be easily learnt by children [REF].
-As an interpreted language, Python is suitable for rapid prototyping.
-Python code is exceptionally readable, and it has been praised as
-``executable pseudocode.''
+% I moved www.python.org to the \cite.
+In surveying the available languages, we believe that Python offers an
+especially good fit to the above requirements.  Python is an
+object-oriented scripting language developed by Guido van Rossum
+and available on all platforms \cite{python}.  Python offers a
+shallow learning curve; it was designed to be easily learnt by
+children \cite{rossum99}.  As an interpreted language, Python is
+suitable for rapid prototyping.  Python code is exceptionally
+readable, and it has been praised as ``executable pseudocode.''
+% Couldn't find a REF for ``executable pseudocode'' -- closest I got
+% was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
 Python is an object-oriented language, but not punitively so, and it
 is easy to encapsulate data and methods inside Python classes.
-Finally, Python has an interface to the ``tk'' graphics toolkit, and
-writing graphical interfaces is straightforward.
+Finally, Python has an interface to the ``Tk'' graphics toolkit
+\cite{tkinter}, and writing graphical interfaces is straightforward.
 
-\section{Design criteria}
+% ===================== Design Criteria =====================
+\section{Design Criteria}
+\label{sec:criteria}
 
@@
 the toolkit.  The design criteria are divided into primary and
-secondary criteria, and within those sections are listed in the order
-of their importance.  It was also important to decide what goals the
-toolkit would \emph{not} attempt to accomplish; we therefore include
-an explicit set of non-requirements, which the toolkit is not expected
-to satisfy.
+secondary criteria, and listed in the order of their importance.  It
+was also important to decide what goals the toolkit would \emph{not}
+attempt to accomplish; we therefore include an explicit set of
+non-requirements, which the toolkit is not expected to satisfy.
 
@@
 
-\noindent\textit{\textbf{Ease of Use.}} The primary purpose of the toolkit is
+\paragraph{\textit{Ease of Use.}} The primary purpose of the toolkit is
 to allow students to concentrate on building NLP systems.  The more
@@
 
-\noindent\textit{\textbf{Consistency.}} The toolkit should use consistent data
+\paragraph{\textit{Consistency.}} The toolkit should use consistent data
 structures and interfaces.
 
-\noindent\textit{\textbf{Extensibility.}} The toolkit should easily
+\paragraph{\textit{Extensibility.}} The toolkit should easily
 accommodate new components, whether those components replicate or
@@
 
-\noindent\textit{\textbf{Documentation.}} The toolkit, its data structures,
-and its implementation all need to be documented.  Three documentation
-types are necessary: technical reports, to explain the toolkit's
-design and implementation; tutorials, to teach students how to use the
-toolkit to perform specific tasks; and reference documentation, to
-describe every module, interface, class and method.  All nomenclature
-must be carefully chosen and consistently used.
+\paragraph{\textit{Documentation.}} The toolkit, its data structures,
+and its implementation all need to be carefully and thoroughly
+documented.  Three documentation types are necessary: tutorials, to
+teach students how to use the toolkit to perform specific tasks;
+technical reports, to explain the toolkit's design and implementation;
+and reference documentation, to describe every module, interface,
+class and method.  All nomenclature must be carefully chosen and
+consistently used.
 
@@
 
-\noindent\textit{\textbf{Simplicity.}} The toolkit should structure the
+\paragraph{\textit{Simplicity.}} The toolkit should structure the
 complexities of building NLP systems, not hide them.  Therefore, each
 class defined by the toolkit should be simple enough that a student
-could implement it by the time they finish the course.
+could implement it by the time they finish an introductory course in
+computational linguistics.
 
-\noindent\textit{\textbf{Modularity.}} The interaction between different
+\paragraph{\textit{Modularity.}} The interaction between different
 components of the toolkit should be kept to a minimum.  In particular,
@@
 parts of the toolkit, without worrying about how they interact with
-the rest of the toolkit.  This will allow students to learn to use the
-toolkit incrementally throughout the course.  Modularity also makes it
-easier to change or extend the toolkit.
+the rest of the toolkit.  This allows students to learn how to use the
+toolkit incrementally throughout a course.  Modularity also makes it
+easier to change and extend the toolkit.
 
@@
 
-\noindent\textit{\textbf{Comprehensiveness.}} The toolkit is not intended to
+\paragraph{\textit{Comprehensiveness.}} The toolkit is not intended to
 provide a comprehensive set of tools.  Indeed, there should be a wide
@@
 
-\noindent\textit{\textbf{Efficiency.}} The toolkit does not need to be highly
+\paragraph{\textit{Efficiency.}} The toolkit does not need to be highly
 optimized for runtime performance.  However, it should be efficient
@@
 
-\noindent\textit{\textbf{Cleverness.}} Clear designs and implementations are
+\paragraph{\textit{Cleverness.}} Clear designs and implementations are
 far preferable to ingenious yet indecipherable ones.
 
-%\section{Python Implementation}
-%
-%The structure of the Python language directly affects the design and
-%implementation of the toolkit.  This section discusses how the design
-%of the toolkit is affected by the Python language.
-%
-%% [EL] This is the only section that I think we might want to keep
-%% from python impl.  But I'm not sure we even need to keep this one.
-%\noindent\textbf{Interfaces.}
-%Python does not directly implement interfaces.  The toolkit
-%therefore implements interfaces as simple classes, all of whose methods 
-%raise \texttt{AssertionError} exceptions.  A description of the
-%functionality provided by the interface is contained in the
-%class's documentation string.  Each method's documentation string
-%specifies the behavior of that method.
-%In order to implement an interface, a class should include that
-%interface as one of its bases, and should override every
-%(non-optional) method defined by the interface.
-%Classes that define interfaces are named with a trailing ``I,'' such
-%as \texttt{TokenizerI} or \texttt{EventI}.
-%
-%% [EL] Mention that Python doesn't provide native type checking?  Or
-%% ditch this section? :)
-%\noindent\textbf{Typing.}
-%Type checking is an important safety device, allowing programmers to
-%quickly locate errors in their code.  This is especially important for
-%novice programmers, who have less experience with debugging.  Every
-%function and method defined by the toolkit will check the types of its
-%its arguments.  Given the impact on performance, the level of type checking
-%will be adjustable.
-%
-%\noindent\textbf{Advanced Language Features.}
-%The use of advanced language features and programming styles could
-%potentially make the toolkit much more difficult to use or to
-%understand.  The fewer language features that a student must learn in
-%order to use the toolkit, the faster they can start developing NLP
-%systems.  The language features used by
-%the toolkit are summarized by figure \ref{fig:feature}.
-%
-%\begin{figure}
-%\begin{centering}
-%\begin{tabular}{|l|l|}
-%\hline
-%\noindent\textbf{Language Feature} & \textbf{Use} \\
-%\hline
-%Classes & used \\
-%Exceptions & error conditions only \\
-%Operator Overloading & used \\
-%Python 2.0 Features & used \\
-%Python 2.1 Features & not used \\
-%Function Arguments & minimal use \\
-%Lambda Functions & minimal use \\
-%Mapping and Filtering & not used \\
-%List Comprehensions & used \\
-%Tkinter & used \\
-%\hline
-%\end{tabular}\\
-%\end{centering}
-%\caption{Advanced language features used by NLTK.}
-%\label{fig:feature}
-%\end{figure}
-
+% ===================== Modules =====================
 \section{Modules}
-% [EL]: I think that this section works better as a textual
-% description, rather than a long list.  Also, we're going to be short
-% on space.. :)
+\label{sec:modules}
 % What order should these be in?  Put more impressive stuff at the
@@
 
-A set of core modules defines basic data types and processing tasks
+A set of core modules defines basic data types and processing systems
 that are used througout the toolkit.  The \texttt{token} module
@@
 as words or sentences.  The \texttt{tree} module provides data
-structures for representing tree strucutres over text, such as syntax
+structures for representing tree structures over text, such as syntax
 trees and morphological trees.  The \texttt{probability} module
 defines classes that encode frequency distributions and probability
-distributions, including a variety of smoothing techniques.
+distributions, including a variety of statistical smoothing
+techniques.
 
@@
 The \texttt{parser} module defines a high-level interface for
-producing trees that represent the structure of texts.  The
+producing trees that represent the structures of texts.  The
 \texttt{chunkparser} module defines a subinterface for parsers that
@@
 each token of a text with supplementary information, such as its part
-of speech or its WordNet synset tag; and provides several
-different implementations for that interface.
+of speech or its WordNet synset tag; and provides several different
+implementations for that interface.
 
@@
 and editing data structures; and graphical tools for experimenting
-with tasks.  The \texttt{draw.tree} module provides a
-simple graphical interface for displaying tree structures; and the
+with tasks.  The \texttt{draw.tree} module provides a simple graphical
+interface for displaying tree structures; and the
 \texttt{draw.tree\_edit} module provides an interface for building and
@@
 \texttt{classifier.maxent} module defines the maximum entropy model
-for text classification, and includes two algorithms for training the
-model: Generalized Iterative Scaling and Improved Iterative Scaling.
+for text classification, and implements two algorithms for training
+the model: Generalized Iterative Scaling and Improved Iterative
+Scaling.
 
@@
 
-%% Old content for this section:
-%NLTK currently includes the following modules.
-%
-%\subsection*{Basics}
-%
-%\begin{description}
-%\item[token:] Basic classes for processing individual
-%      elements of text, such as words or sentences. 
-%\item[tree:] Classes for representing tree structures
-%      over text (such as syntax trees and morphological trees).
-%\item[probability:] Classes that encode frequency
-%      distributions and probability distributions.
-%\end{description}
-%
-%\subsection*{Tagging}
-%
-%\begin{description}
-%\item[tagger:] A standard interface to tag each token of
-%      a text with supplementary information, such as its part of
-%      speech; and several implementations of that interface.
-%\end{description}
-%
-%\subsection*{Parsing}
-%
-%\begin{description}
-%\item[parser:] A standard interface to produce trees
-%      representing the structure of texts.
-%\item[chartparser:] A flexible parser implementation
-%      that uses a \emph{chart} to record hypotheses about
-%      syntactic constituents.
-%\item[srparser\_template:] A partial implementation of a
-%      shift-reduce parser; completing the implementation was a
-%      student exercise.
-%\item[chunkparser:] A standard interface for robust
-%      parsers used to identify non-overlapping linguistic groups
-%      (such as noun phrases) in unrestricted text.
-%\item[rechunkparser:] A regular-expression based
-%      implementation of the chunk parser interface.
-%\end{description}
-%
-%\subsection*{Text Classification}
-%
-%\begin{description}
-%\item[classifier:] A standard interface for classifying
-%      texts into categories.
-%\item[classifier.feature:] A standard way of encoding
-%      the information used to make classification decisions.
-%\item[classifier.naivebayes:] A text classifier
-%      implementation based on the Naive Bayes assumption.
-%\item[classifier.maxent:] An implementation of the
-%      maximum entropy model for text classification; and
-%      implementations of the GIS and IIS algorithms for training
-%      the classifier.
-%\item[classifier.featureselection:] A standard interface
-%      for choosing which features are relevant for making
-%      classification decisions.
-%\end{description}
-%
-%\subsection*{Visualization}
-%
-%\begin{description}
-%\item[draw.tree:] A graphical representation for tree
-%      structures, such as syntax trees and morphological trees. 
-%\item[draw.tree\_edit:] A graphical interface used to
-%      build and modify tree structures.
-%\item[draw.plot\_graph:] A graphical tool to graph
-%      arbitrary functions.
-%\item[draw.chart:] An interactive graphical tool used to
-%      experiment with chart parsers.
-%\end{description}
+% ===================== Documentation =====================
+\section{Documentation}
+\label{sec:documentation}
 
+The toolkit is accompanied by extensive documentation that explains
+the toolkit, and describes how to use and extend it.  This
+documentation is divided into three primary categories:
+
+\paragraph{\textit{Tutorials}} teach students how to use the toolkit,
+in the context of performing specific tasks.  Each tutorial focuses on
+a single domain, such as tagging, probablistic systems, or text
+classification.  The tutorials include a high-level discussion that
+explains and motivates the domain, followed by a detailed
+walk-through that uses examples to show how NLTK can be used to
+perform specific tasks.
+
+\paragraph{\textit{Reference Documentation}} provides precise
+definitions for every module, interface, class, method, function, and
+variable in the toolkit.  It is automatically extracted from docstring
+comments in the Python source code, using Epydoc \cite{epydoc}.
+
+\paragraph{\textit{Technical Reports}} explain and justify the
+toolkit's design and implementation.  They are used by the developers
+of the toolkit to guide and document the toolkit's construction.
+Students can also consult these reports if they would like further
+information about how the toolkit is designed and why it is designed
+that way.
+
+% ===================== Uses =====================
 \section{Uses of NLTK}
+\label{sec:uses}
 
-\subsection{Assigments and projects}
+\subsection{Assigments and Projects}
 
@@
 
-% Akward
-Here we illustrate the use of NLTK in an assignment of moderate
-difficulty, for the case of chunk parsing.  This is a kind of parsing
-where we only identify the main constituents of a phrase, a technique
-with diverse applications.  The NLTK \texttt{chunk} module offers
-a variety of rule formats.  For example,
-\texttt{ChunkRule('<NN.*>')} builds chunks over consecutive noun tags;
-\texttt{ChinkRule('<VB.>')} excises verbs from existing chunks;
-\texttt{SplitRule('<NN>', '<DT>')} splits into two pieces any existing
-chunk that contains a noun followed by a determiner; and
-\texttt{MergeRule('<JJ>', '<JJ>')} combines two adjacent chunks
-where the first chunk ends and the second chunk starts with \texttt{JJ}.
-The goal of the assignment is to come up with a cascade of these rules,
-and to create a set of chunks that are as similar as possible to the test set.
+As an example of a moderately difficult assignment, we asked students
+to construct a chunk parser that correctly identifies base noun phrase
+chunks in a given text, by defining a cascade of transformational
+chunking rules.  The NLTK \texttt{rechunkparser} module provides a
+variety of rule types, which the students can instantiate to construct
+complete rules.  For example, \texttt{ChunkRule('<NN.*>')} builds
+chunks from sequences of consecutive nouns; \texttt{ChinkRule('<VB.>')}
+excises verbs from existing chunks; \texttt{SplitRule('<NN>', '<DT>')}
+splits any existing chunk that contains a singular noun followed by
+determiner into two pieces; and \texttt{MergeRule('<JJ>', '<JJ>')}
+combines two adjacent chunks where the first chunk ends and the second
+chunk starts with adjectives.
 
-The chunking tutorial provides all the necessary code except for the
-chunk rules.  The provided code is responsible for loading the
-chunked, tagged text using an existing tokenizer, creating an
-unchunked version of the text, applying the chunk rules to the
-unchunked text, and scoring the result.  Students focus on the
-NLP task only -- providing a ruleset with the best coverage.
+The chunking tutorial motivates chunk parsing, describes each rule
+type, and provides all the necessary code for the assignment.  The
+provided code is responsible for loading the chunked, part-of-speech
+tagged text using an existing tokenizer, creating an unchunked version
+of the text, applying the chunk rules to the unchunked text, and
+scoring the result.  Students focus on the NLP task only -- providing
+a ruleset with the best coverage.
 
@@
   UnChunkRule('<IN|VB.*|CC|MD|RB.*>'),
-  UnChunkRule("<,|{\textbackslash}.|``|''>"),
+  UnChunkRule("<,|{\textbackslash}{\textbackslash}.|``|''>"),
   MergeRule('<NN.*|DT|JJ.*|CD>', '<NN.*|DT|JJ.*|CD>'),
@@
 
-This next example illustrates a brute-force approach.  All tags most
-commonly found inside of chunks are chunked, while other tags are excluded.
+The next example illustrates a brute-force statistical approach.  The
+student calculated how often each part-of-speech tag was included in a
+noun phrase.  They then constructed chunks from any sequence of tags
+that occured in a noun phrase more than 50\% of the time.
 
@@
 cascade = [
-  ChunkRule('<{\textbackslash}\$|CD|DT|EX|PDT
-             |PRP.*|WP.*|{\textbackslash}\#|FW
+  ChunkRule('<{\textbackslash}{\textbackslash}\$|CD|DT|EX|PDT
+             |PRP.*|WP.*|{\textbackslash}{\textbackslash}\#|FW
              |JJ.*|NN.*|POS|RBS|WDT>*')
@@
 
-The third example involves putting everything in a single chunk,
-then excising elements that do not belong.
+In the third example, the student constructed a single chunk
+containing the entire text, and then excised all elements that did not
+belong.
 
@@
 
-%% [EL] I'm not sure this is necessary:
-%The main program which surrounds the student's work is moderately
-%simple as well.  The line replaced by students is underlined.
-%% [EL] I find the undeline hard to read; change to italics or bold or
-%% vbar in the margin?  Or maybe a \Rightarrow of some sort in the margin?
-%\begin{sv}
-%from nltk.token import *
-%from nltk.chunkparser import *
-%from nltk.rechunkparser import *
-%import os
+% [EL] This makes it sound like we GRADED them using their precision,
+% recall, and f-measure.  Either say something about the contest, or
+% leave this out:
+Each student's project was scored using its precision, recall, and
+F-measure.
+
+% [EL] Leave out the picture.
+%Figure~\ref{fig:contest} shows the results; students with the best two
+%scores were presented with prizes.
 %
-%path = "/cis530/data/wsj/tagged/"
-%
-%sent_tok = RETokenizer('(=======+)?{\textbackslash}{\textbackslash}n{\textbackslash}{\textbackslash}n+', 0)
-%sentences = []
-%files = os.listdir(path)
-%for file in files:
-%    print 'Loading \%s...' \% file
-%    text = open(path+file).read()
-%    sentences += sent_tok.tokenize(text, source=file, unit='s')
-%
-%chunkscore = ChunkScore()
-%
-%\underline{cascade = [ ??? ]}
-%chunkparser = REChunkParser(cascade)
-%
-%ctt = ChunkedTaggedTokenizer()
-%for s in sentences:
-%    correct = ctt.tokenize(s.type(), source=s.loc())
-%    unchunked = unchunk(correct)
-%    guess = chunkparser.parse(unchunked)
-%    chunkscore.score(correct, guess)
-% 
-%print chunkscore
-%\end{sv}
-
-The score for each project is the precision, recall and F-measure.
-Figure~\ref{fig:contest} shows the results; students with the best two
-scores were presented with prizes.
-
-\begin{figure}
-\centerline{\epsfig{figure=contest.ps,width=\linewidth}}
-\caption{Precision/Recall Graph for Chunking Competition}\label{fig:contest}
-\vspace*{2ex}\hrule
-\end{figure}
+%\begin{figure}
+%\centerline{\epsfig{figure=contest.ps,width=\linewidth}}
+%\caption{Precision/Recall Graph for Chunking Competition}\label{fig:contest}
+%\vspace*{2ex}\hrule
+%\end{figure}
 
@@
 NLTK provides graphical tools that can be used to give in-class
-demonstrations, which help to explain basic NLP concepts and
-algorithms.  These interactive tools can be used to display relevant
-data structures, and to show step-by-step executions of algorithms.
+demonstrations that help to explain basic NLP concepts and algorithms.
+These interactive tools can be used to display relevant data
+structures, and to show the step-by-step execution of algorithms.
 Both data structures and control flow can be easily modified during
@@
 Since these graphical tools are included with the toolkit, they can
-also be used by students.  This allows students to experiment with the
-algorithms that they have seen presented in class.
+also be used by students.  This allows students to experiment at home
+with the algorithms that they have seen presented in class.
 
@@
 
-The chart parsing tool can be used to explain the basic concepts
-behind chart parsing, and to show how the algorithm works.  Chart
-parsing is a flexible parsing algorithm that uses a data structure
-called a \emph{chart} to record hypotheses about syntactic
-constituents.  Each hypothesis is represented by a single \emph{edge}
-on the chart.  A set of \emph{rules} determine when new edges can be
-added to the chart.  This set of rules controls the overall behavior
-of the parser (e.g., whether it parses top-down or bottom-up).
+The chart parsing tool is an example of a graphical tool provided by
+NLTK.  This tool can be used to explain the basic concepts behind
+chart parsing, and to show how the algorithm works.  Chart parsing is
+a flexible parsing algorithm that uses a data structure called a
+\emph{chart} to record hypotheses about syntactic constituents.  Each
+hypothesis is represented by a single \emph{edge} on the chart.  A set
+of \emph{rules} determine when new edges can be added to the chart.
+This set of rules controls the overall behavior of the parser (e.g.,
+whether it parses top-down or bottom-up).
 
@@
 edges.  Finally, it can be used to show how these individual rules
-combine to find a complete parse for the sentence.
+combine to find a complete parse for a given sentence.
 
@@
 \begin{figure}
-%\centerline{\epsfig{figure=chart.ps,width=\linewidth}}
+\centerline{\epsfig{figure=chartparse.eps,width=\linewidth}}
 \caption{Chart Parsing Tool}\label{fig:chartparse}
@@
 
+% ===================== Evaluation =====================
 \section{Evaluation}
+\label{sec:evaluation}
 
@@
 
-\section{Other approaches}
+% ===================== Other Approaches =====================
+\section{Other Approaches}
+\label{sec:approaches}
 
-Java: \cite{Hammond02}
--- no obligation to cite this as it is unpublished
-(and I haven't seen a copy).
+\textit{This section will include a discussion of other approaches to
+implementing computational linguistics courseware, including:
+Programming for Linguistics: Java Technology for Language
+Researchers\cite{Hammond02}; The OpenNLP project\cite{opennlp};
+others?}
 
+% ===================== Conclusion =====================
 \section{Conclusion}
+\label{sec:conclusion}
 
@@
 
+% include URL (a second time; 1st was in the intro) in conclusion
+
 \bibliographystyle{acl}

### 2002-03-28 technical/acl2002/submission.tex
- Converted "Evaluation" section to text - Converted "Conclusion" section to prose - Minor changes
svn/trunk@523

@@
 % Outline:
-%   - abstract
+%   - abstract (summary, above.)
 %   - introduction
@@
 
-% Steven: do you want to add anything here?  Say anything about HMMS? 
+% [EL] Steven, do you want to add anything here?  Say anything about HMMS? 
 % Will you have implemented more interfaces by the time of ACL?
 The \texttt{fsa} module defines a data type for encoding finite state
-automata; and an interface for creating automata from
-regular expressions.
+automata; and an interface for creating automata from regular
+expressions.
 
@@
 
-A set of visualization modules define graphical interfaces for viewing
-and editing data structures; and graphical tools for experimenting
-with tasks.  The \texttt{draw.tree} module provides a simple graphical
-interface for displaying tree structures; and the
+% jb: this is broken up oddly.
+A set of visualization modules define graphical tools for viewing and
+editing data structures, and for experimenting with NLP tasks.  The
+\texttt{draw.tree} module provides a simple graphical interface for
+displaying tree structures; and the
 \texttt{draw.tree\_edit} module provides an interface for building and
@@
 
-% [EL] This makes it sound like we GRADED them using their precision,
-% recall, and f-measure.  Either say something about the contest, or
-% leave this out:
-Each student's project was scored using its precision, recall, and
-F-measure.
+%% [EL] This makes it sound like we GRADED them using their precision,
+%% recall, and f-measure.  Either say something about the contest, or
+%% leave this out:
+%Each student's project was scored using its precision, recall, and
+%F-measure.
 
@@
 syntactic constituents.  It can then be used to demonstrate and
-explain the individual rules that the chart parsers uses to create new
+explain the individual rules that the chart parser uses to create new
 edges.  Finally, it can be used to show how these individual rules
@@
 
-About CIS-530: mixture of CS and linguistics students.
-Approx 40 hours of classes.  All class materials (slides,
-assignments) are posted on the course website
-\url{http://www.cis.upenn.edu/~cis530/}.
+% Do we want to mention that there were 40h of lecture time?
+% Should we be more explicit about who took the class?
+We used NLTK as a basis for the assignments and student projects in
+CIS-530, an introductory computational linguistics class taught at the
+University of Pennsylvania.  CIS-530 is a graduate level class,
+although some advanced undergraduates were also enrolled.  Most
+students had either a computer science background or a linguistics
+backgroud.  Students were required to complete five assignments, two
+exams, and a final project.  All class materials are available from
+the corse website \url{http://www.cis.upenn.edu/~cis530/}.
 
-The Language: Python was new to 95\% of students.  Some were initially
-bothered about its block structuring, which is sensitive to white
-space.  All quickly learnt the language, and reported that any initial
-reservations were unfounded.  Many students said that they enjoyed
-writing Python code.
+% (Leaving this out for now\ldots)
+%The Language: Python was new to 95\% of students.  Some were initially
+%bothered about its block structuring, which is sensitive to white
+%space.  All quickly learnt the language, and reported that any initial
+%reservations were unfounded.  Many students said that they enjoyed
+%writing Python code.
 
-CD-ROM: put NLTK and data on CD-ROM and gave it out at the start of class.
+% Where does this go??
+%At the beginning of the semester, we distributed 2 CD-ROMs to the
+%students, containing NLTK and all of the corpera that we would use for
+%the class.
 
-The toolkit: students liked being able to do interesting projects from
-the outset, liked all the documentation, liked being able to run
-everything on their machine at home, ...
+The experience of using NLTK was very positive, both for us and for
+the students.  The students liked the fact that they could do
+interesting projects from the outset.  They also liked being able to
+run everything on their computer at home.  The students found the
+extensive documentation very helpful for learning to use the toolkit.
+They found the interfaces defined by NLTK intuitive, and appreciated
+the ease with which they could combine different components to create
+complete NLP systems.
 
-Problems: available corpus data, ...
+We did encounter a few difficulties during the semester.  One problem
+is finding large, clean, open-source corpera that the students can use
+for their assignments.  This became especially apparent when students
+started planning their final projects, since many projects required
+different types of corpera.  Another issue was the fact that we were
+actively developing NLTK during the semester; some modules were only
+completed one or two weeks before we used them.  As a result, students
+who worked at home needed to download new versions of the toolkit
+several times throughout the semester.  Luckily, Python has extensive
+support for installation scripts, which made these upgrades much
+simpler.  The students encountered a couple of bugs in the toolkit,
+but none were serious, and all were quickly corrected.
 
-Other strengths / weaknesses
+%Problems: available corpus data, ...
+%
+%Other strengths / weaknesses
 
@@
 \textit{This section will include a discussion of other approaches to
-implementing computational linguistics courseware, including:
-Programming for Linguistics: Java Technology for Language
-Researchers\cite{Hammond02}; The OpenNLP project\cite{opennlp};
-others?}
+implementing computational linguistics courseware, including
+\emph{Programming for Linguistics: Java Technology for Language
+Researchers}\cite{Hammond02} and \emph{the OpenNLP project}\cite{opennlp}.}
 
 % ===================== Conclusion =====================
-\section{Conclusion}
+\section{Conclusions and Future Work}
 \label{sec:conclusion}
 
-Overview of the paper.
+%Overview of the paper:
+% [EL] Pick better 3 adjectives :)
+NLTK provides a simple, extensible, uniform framework for student
+assignments and class demonstrations.  It is well documented, easy to
+learn, and simple to use.  We hope that NLTK will allow computational
+linguistics classes to include more hands-on experience building NLP
+components and systems.
 
-NOTES: Many CL textbooks provide good exercises at the end of each
-chapter.  This isn't enough for many situations.  E.g. students spend
-lots of time writing interface code (e.g. to process a corpus).
-E.g. teachers spend lots of time creating do-able assignments...
+% [EL] No room for this for now\ldots
+%NOTES: Many CL textbooks provide good exercises at the end of each
+%chapter.  This isn't enough for many situations.  E.g. students spend
+%lots of time writing interface code (e.g. to process a corpus).
+%E.g. teachers spend lots of time creating do-able assignments...
 
-NLTK is an open source project, and we welcome any contributions.  We
-deliberately structured NLTK to facilitate parallel development.
-Readers who are interested in contributing to NLTK, or who have any
+%Future work: 
+We plan to continue extending the breadth of materials covered by the
+toolkit.  We are currently working on NLTK modules for Hidden Markov
+Models, probablistic parsing, and tree adjoining grammars.  We also
+plan to increase the number of algorithms implemented by some existing
+modules, such as the text classification module.  Finally, we are
+putting together a standard collection of large, clean, open-source
+corpera, including corpera appropriate for every module defined by the
+toolkit.
+
+NLTK is an open source project, and we welcome any contributions.
+Readers who are interested in contributing to NLTK, or who have
 suggestions for improvements, are encouraged to contact the authors.
 
-% include URL (a second time; 1st was in the intro) in conclusion
+% include URL (a second time; 1st was in the intro) in conclusion??
 

### 2002-03-28 technical/acl2002/submission.tex
- Minor revisions - Merged "primary design criteria" and "secondary design criteria"   into a single "requirements" subsection
svn/trunk@530

@@
 %   - Should URLs include http:// or not (we should be consistant)?
+% We assume that the user is familiar with the acronym ``NLP;'' is
+% that ok?
 
@@
 
+% [EL] ``functionality'' isn't a great word, here.
 Teachers of introductory courses on computational linguistics are
@@
 % [EL] ``difficulties'' doesn't seem like the right word here?
+% Maybe ``challenging requirements''?
 Apart from the practical component, computational linguistics courses
@@
 
-% [EL] akward
-In this paper we describe a new approach to the above challenges, a
+In this paper we introduce a new approach to the above challenges, a
 streamlined and flexible way of organizing the practical component of
@@
 
-All materials discussed here are available under an open source
-license from \url{nltk.sf.net}.  NLTK runs on most platforms,
-including Windows, OS X, Linux, and UNIX.
+% [EL] This was too ambiguous for me (e.g., we discuss Python)
+%All materials discussed here are available under an open source
+%license from \url{nltk.sf.net}.  NLTK runs on most platforms,
+%including Windows, OS X, Linux, and UNIX.
+The Natural Language Toolkit is available under an open source license
+from \url{http://nltk.sf.net}.  NLTK runs on most platforms, including
+Windows, OS X, Linux, and UNIX.
 
@@
 
+% [EL] is ``desiderata'' an acceptable word here?
 The most basic step in setting up a practical component is choosing a
@@
 associated with languages like C++.  Finally, the language must have
-an easy-to-use graphics library to support the development of simple
+an easy-to-use graphics library to support the development of
 graphical user interfaces.
 
-% I moved www.python.org to the \cite.
+% [EL] ``punitive'' like a strange word.
+% [EL] I couldn't find a REF for ``executable pseudocode'' -- closest
+% I got was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
 In surveying the available languages, we believe that Python offers an
@@
 readable, and it has been praised as ``executable pseudocode.''
-% Couldn't find a REF for ``executable pseudocode'' -- closest I got
-% was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
 Python is an object-oriented language, but not punitively so, and it
@@
 
+% [EL] Is this ambiguous between descending/ascending?
 Several criteria were considered in the design and implementation of
-the toolkit.  The design criteria are divided into primary and
-secondary criteria, and listed in the order of their importance.  It
-was also important to decide what goals the toolkit would \emph{not}
-attempt to accomplish; we therefore include an explicit set of
-non-requirements, which the toolkit is not expected to satisfy.
+the toolkit.  These design criteria are listed in the order of their
+importance.  It was also important to decide what goals the toolkit
+would \emph{not} attempt to accomplish; we therefore include an
+explicit set of non-requirements, which the toolkit is not expected to
+satisfy.
 
-\subsection{Primary Design Criteria}
+\subsection{Requirements}
 
@@
 accommodate new components, whether those components replicate or
-extend the toolkit's existing functionality.  Thus, the toolkit's
-design should be modular, with simple and well-defined interfaces
-between modules.  The toolkit should also be structured in such a way
-that it is obvious where new extensions would fit into the toolkit's
-infrastructure.
+extend the toolkit's existing functionality.  The toolkit should be
+structured in such a way that it is obvious where new extensions would
+fit into the toolkit's infrastructure.
 
+% [EL] I commented out the list of documentation types, since we have
+% a section that talks about the documentation.
 \paragraph{\textit{Documentation.}} The toolkit, its data structures,
 and its implementation all need to be carefully and thoroughly
-documented.  Three documentation types are necessary: tutorials, to
-teach students how to use the toolkit to perform specific tasks;
-technical reports, to explain the toolkit's design and implementation;
-and reference documentation, to describe every module, interface,
-class and method.  All nomenclature must be carefully chosen and
-consistently used.
-
-\subsection{Secondary Design Criteria}
+documented.  
+%Three documentation types are necessary: tutorials, to
+%teach students how to use the toolkit to perform specific tasks;
+%technical reports, to explain the toolkit's design and implementation;
+%and reference documentation, to describe every module, interface,
+%class and method.  
+All nomenclature must be carefully chosen and consistently used.
 
@@
 \paragraph{\textit{Modularity.}} The interaction between different
-components of the toolkit should be kept to a minimum.  In particular,
-it should be possible to complete individual projects using small
-parts of the toolkit, without worrying about how they interact with
-the rest of the toolkit.  This allows students to learn how to use the
-toolkit incrementally throughout a course.  Modularity also makes it
-easier to change and extend the toolkit.
+components of the toolkit should be kept to a minimum, using simple,
+well-defined interfaces.  In particular, it should be possible to
+complete individual projects using small parts of the toolkit, without
+worrying about how they interact with the rest of the toolkit.  This
+allows students to learn how to use the toolkit incrementally
+throughout a course.  Modularity also makes it easier to change and
+extend the toolkit.
 
@@
 \label{sec:modules}
-% What order should these be in?  Put more impressive stuff at the
-% beginning and end?
+% What order should the subsections be in?  Put more impressive stuff
+% at the beginning and end?
 
+% [EL] is ``defines'' the right word?
 The toolkit is implemented as a collection of independant
@@
 that are used througout the toolkit.  The \texttt{token} module
-defines basic classes for processing individual elements of text, such
-as words or sentences.  The \texttt{tree} module provides data
+provides basic classes for processing individual elements of text,
+such as words or sentences.  The \texttt{tree} module defines data
 structures for representing tree structures over text, such as syntax
 trees and morphological trees.  The \texttt{probability} module
-defines classes that encode frequency distributions and probability
+implements classes that encode frequency distributions and probability
 distributions, including a variety of statistical smoothing
@@
 \texttt{chunkparser} module defines a subinterface for parsers that
-identify non-overlapping linguistic groups (such as noun phrases) in
-unrestricted text.
+identify non-overlapping linguistic groups (such as base noun phrases)
+in unrestricted text.
 
@@
 of speech or its WordNet synset tag; and provides several different
-implementations for that interface.
+implementations for this interface.
 
@@
 % jb: this is broken up oddly.
-A set of visualization modules define graphical tools for viewing and
-editing data structures, and for experimenting with NLP tasks.  The
-\texttt{draw.tree} module provides a simple graphical interface for
-displaying tree structures; and the
+
+Visualization modules define graphical interfaces for viewing and
+manipulating data structures, and graphical tools for experimenting
+with NLP tasks.  The \texttt{draw.tree} module provides a simple
+graphical interface for displaying tree structures; and the
 \texttt{draw.tree\_edit} module provides an interface for building and
@@
 classifying texts into categories.  This interface is currently
-implemented by two modules: the \texttt{classifier.naivebayes} module
-defines a text classifier based on the Naive Bayes assumption; and the
+implemented by two modules.  The \texttt{classifier.naivebayes} module
+defines a text classifier based on the Naive Bayes assumption. The
 \texttt{classifier.maxent} module defines the maximum entropy model
@@
 
+% [EL] Provide some indication of how much tutorial material is
+% available?  (Currently, 8 tutorials, mean=16pp, range=6pp-52pp.)
 \paragraph{\textit{Tutorials}} teach students how to use the toolkit,
@@
 Students can also consult these reports if they would like further
-information about how the toolkit is designed and why it is designed
+information about how the toolkit is designed, and why it is designed
 that way.
@@
 chunking rules.  The NLTK \texttt{rechunkparser} module provides a
-variety of rule types, which the students can instantiate to construct
-complete rules.  For example, \texttt{ChunkRule('<NN.*>')} builds
-chunks from sequences of consecutive nouns; \texttt{ChinkRule('<VB.>')}
-excises verbs from existing chunks; \texttt{SplitRule('<NN>', '<DT>')}
-splits any existing chunk that contains a singular noun followed by
-determiner into two pieces; and \texttt{MergeRule('<JJ>', '<JJ>')}
-combines two adjacent chunks where the first chunk ends and the second
-chunk starts with adjectives.
+variety of regular-expression based rule types, which the students can
+instantiate to construct complete rules.  For example,
+\texttt{ChunkRule('<NN.*>')} builds chunks from sequences of
+consecutive nouns; \texttt{ChinkRule('<VB.>')} excises verbs from
+existing chunks; \texttt{SplitRule('<NN>', '<DT>')} splits any
+existing chunk that contains a singular noun followed by determiner
+into two pieces; and \texttt{MergeRule('<JJ>', '<JJ>')} combines two
+adjacent chunks where the first chunk ends and the second chunk starts
+with adjectives.
 
@@
 
-NLTK provides graphical tools that can be used to give in-class
-demonstrations that help to explain basic NLP concepts and algorithms.
-These interactive tools can be used to display relevant data
-structures, and to show the step-by-step execution of algorithms.
-Both data structures and control flow can be easily modified during
-the demonstration, in response questions from the class.
+NLTK provides graphical tools that can be used in class demonstrations
+to help explain basic NLP concepts and algorithms.  These interactive
+tools can be used to display relevant data structures and to show the
+step-by-step execution of algorithms.  Both data structures and
+control flow can be easily modified during the demonstration, in
+response questions from the class.
 
@@
 although some advanced undergraduates were also enrolled.  Most
-students had either a computer science background or a linguistics
-backgroud.  Students were required to complete five assignments, two
-exams, and a final project.  All class materials are available from
-the corse website \url{http://www.cis.upenn.edu/~cis530/}.
+students had a background in either computer science or linguistics
+(but not both).  Students were required to complete five assignments,
+two exams, and a final project.  All class materials are available
+from the corse website \url{http://www.cis.upenn.edu/~cis530/}.
 
-% (Leaving this out for now\ldots)
+% (Leaving this out for now..)
 %The Language: Python was new to 95\% of students.  Some were initially
@@
 
-% Where does this go??
+% Where does this go??  Leaving it out for now..
 %At the beginning of the semester, we distributed 2 CD-ROMs to the
@@
 We did encounter a few difficulties during the semester.  One problem
-is finding large, clean, open-source corpera that the students can use
-for their assignments.  This became especially apparent when students
-started planning their final projects, since many projects required
-different types of corpera.  Another issue was the fact that we were
-actively developing NLTK during the semester; some modules were only
-completed one or two weeks before we used them.  As a result, students
-who worked at home needed to download new versions of the toolkit
-several times throughout the semester.  Luckily, Python has extensive
-support for installation scripts, which made these upgrades much
-simpler.  The students encountered a couple of bugs in the toolkit,
-but none were serious, and all were quickly corrected.
+was finding large clean corpera that the students could use for their
+assignments.  Several of the students needed assistance finding
+suitable corpera for their final projects.  Another issue was the fact
+that we were actively developing NLTK during the semester; some
+modules were only completed one or two weeks before the students used
+them.  As a result, students who worked at home needed to download new
+versions of the toolkit several times throughout the semester.
+Luckily, Python has extensive support for installation scripts, which
+made these upgrades much simpler.  The students encountered a couple
+of bugs in the toolkit, but none were serious, and all were quickly
+corrected.
 
@@
 learn, and simple to use.  We hope that NLTK will allow computational
-linguistics classes to include more hands-on experience building NLP
-components and systems.
+linguistics classes to include more hands-on experience with using and
+building NLP components and systems.
 
-% [EL] No room for this for now\ldots
+% [EL] No room for this for now...
 %NOTES: Many CL textbooks provide good exercises at the end of each
@@
 plan to increase the number of algorithms implemented by some existing
-modules, such as the text classification module.  Finally, we are
-putting together a standard collection of large, clean, open-source
-corpera, including corpera appropriate for every module defined by the
-toolkit.
+modules, such as the text classification module.
+
+Finding suitable corpera is a prerequisite for many studnet
+assignments and projects.  We are therefore putting together a
+standard collection of large, clean, open-source corpera, containing
+corpera appropriate for every module defined by the toolkit.
 

### 2002-03-28 technical/acl2002/submission.tex
- Spellcheck - Put mbox's around urls to prevent linewrap
svn/trunk@531

@@
 % To do:
-%   - Write up evaluation
 %   - Write up other approaches
-%   - Write up conclusion
 %   - General clean-up
-%   - ispell
-%   - Should URLs include http:// or not (we should be consistant)?
 % We assume that the user is familiar with the acronym ``NLP;'' is
@@
 component for student assignments and projects.  This is a difficult
-task because different comptational linguistics domains require a
+task because different computational linguistics domains require a
 variety of different data structures and functionality; and because a
@@
 The Natural Language Toolkit is available under an open source license
-from \url{http://nltk.sf.net}.  NLTK runs on most platforms, including
+from \mbox{\url{http://nltk.sf.net/}}.  NLTK runs on most platforms, including
 Windows, OS X, Linux, and UNIX.
@@
 % [EL] is ``defines'' the right word?
-The toolkit is implemented as a collection of independant
+The toolkit is implemented as a collection of independent
 \emph{modules}, each of which defines a specific data structure or
@@
 A set of core modules defines basic data types and processing systems
-that are used througout the toolkit.  The \texttt{token} module
+that are used throughout the toolkit.  The \texttt{token} module
 provides basic classes for processing individual elements of text,
@@
 producing trees that represent the structures of texts.  The
-\texttt{chunkparser} module defines a subinterface for parsers that
+\texttt{chunkparser} module defines a sub-interface for parsers that
 identify non-overlapping linguistic groups (such as base noun phrases)
@@
 interface for choosing which features are relevant for a particular
-classification task.  Good feature selection can signifigantly improve
+classification task.  Good feature selection can significantly improve
 classification performance.
@@
 in the context of performing specific tasks.  Each tutorial focuses on
-a single domain, such as tagging, probablistic systems, or text
+a single domain, such as tagging, probabilistic systems, or text
 classification.  The tutorials include a high-level discussion that
@@
 
-\subsection{Assigments and Projects}
+\subsection{Assignments and Projects}
 
@@
 In the simplest assignments, students experiment with an existing
-module.  The wide variety of existing modules provide many opportunies
+module.  The wide variety of existing modules provide many opportunities
 for creating these simple assignments.
@@
 scoring the result.  Students focus on the NLP task only -- providing
-a ruleset with the best coverage.
+a rule set with the best coverage.
 
@@
 noun phrase.  They then constructed chunks from any sequence of tags
-that occured in a noun phrase more than 50\% of the time.
+that occurred in a noun phrase more than 50\% of the time.
 
@@
 two exams, and a final project.  All class materials are available
-from the corse website \url{http://www.cis.upenn.edu/~cis530/}.
+from the course website \mbox{\url{http://www.cis.upenn.edu/~cis530/}}.
 
@@
 We did encounter a few difficulties during the semester.  One problem
-was finding large clean corpera that the students could use for their
+was finding large clean corpora that the students could use for their
 assignments.  Several of the students needed assistance finding
-suitable corpera for their final projects.  Another issue was the fact
+suitable corpora for their final projects.  Another issue was the fact
 that we were actively developing NLTK during the semester; some
@@
 toolkit.  We are currently working on NLTK modules for Hidden Markov
-Models, probablistic parsing, and tree adjoining grammars.  We also
+Models, probabilistic parsing, and tree adjoining grammars.  We also
 plan to increase the number of algorithms implemented by some existing
@@
 
-Finding suitable corpera is a prerequisite for many studnet
+Finding suitable corpora is a prerequisite for many student
 assignments and projects.  We are therefore putting together a
-standard collection of large, clean, open-source corpera, containing
-corpera appropriate for every module defined by the toolkit.
+standard collection of large, clean, open-source corpora, containing
+corpora appropriate for every module defined by the toolkit.
 

### 2002-03-29 technical/acl2002/submission.tex
- Very minor tweaks
svn/trunk@532

@@
+%
+% Submission for ``Effective Tools And Methodologies For Teaching
+% Natural Language Processing And Computational Linguistics'' workshop
+% at ACL 2002
+%
+% Authors: Edward Loper and Steven Bird
+% 
+% $Id$
+%
 \documentclass[11pt]{article}
@@
 \contact{Edward Loper}
-\conference{this paper has not been submitted to any other conferences}
-\date{\today}
+\conference{this paper has not been submitted to any other
+conferences}
+% In the submission example, they had \date{}; but \date{\today} is
+% helpful when printing draft copies, to keep track of versions.
+%\date{\today}
+\date{}
 
@@
 %   - Write up other approaches
-%   - General clean-up
 % We assume that the user is familiar with the acronym ``NLP;'' is
@@
 \paragraph{\textit{Ease of Use.}} The primary purpose of the toolkit is
-to allow students to concentrate on building NLP systems.  The more
-time students must spend learning to use the toolkit, the less useful
-it is.
+to allow students to concentrate on building natural language
+processing (NLP) systems.  The more time students must spend learning
+to use the toolkit, the less useful it is.
 
@@
 
-\textit{This section will include a discussion of other approaches to
+This section will include a discussion of other approaches to
 implementing computational linguistics courseware, including
 \emph{Programming for Linguistics: Java Technology for Language
-Researchers}\cite{Hammond02} and \emph{the OpenNLP project}\cite{opennlp}.}
+Researchers}\cite{Hammond02} and \emph{the OpenNLP
+project}\cite{opennlp}.
 

### 2002-05-06 technical/acl2002/acl02.tex
first commit of acl02.tex
svn/trunk@547
removed submission-specific macros
svn/trunk@549

--- 2002-03-29 technical/acl2002/submission.tex
@@
 \documentclass[11pt]{article}
-\usepackage{acl02sub,url,alltt,epsfig}
+\usepackage{acl2002,url,alltt,epsfig}
 \title{NLTK: The Natural Language Toolkit}
@@
 }
-\summary{%
-NLTK, the Natural Language Toolkit, is a suite of program modules,
-tutorials and problem sets, providing ready-to-use computational
-linguistics courseware.  NLTK covers symbolic and statistical natural
-language processing, and is interfaced to annotated corpora.  Students
-augment and replace existing components, learning structured
-programming by example, and manipulating sophisticated models from the
-outset.}
-
-\paperid{Pxxxx}
-\keywords{teaching, courseware, Python, corpora}
-\contact{Edward Loper}
-\conference{this paper has not been submitted to any other
-conferences}
-% In the submission example, they had \date{}; but \date{\today} is
-% helpful when printing draft copies, to keep track of versions.
-%\date{\today}
 \date{}
-
-% Note: the maximum paper length is 6 pages.
 
@@
 \begin{document}
-\makeidpage
 \maketitle
 
-% ===================== Abstract =====================
-% [EL] I think that this abstract should be left out, or shortened and
-% merged with the \summary.  The text in the 1st paragraph of the intro
-% seems much more direct and relevant to our project than the 1st
-% paragraph of this abstract (in particular, the 1st sentence seems like
-% a strange way to start). :) I don't think that the second paragraph
-% needs to be included in the abstract -- it's not a central point of
-% our paper; and the third paragraph is almost identical to the \summary
-% (but not quite).  For now, I'm just commenting out the abstract.
-
-%\begin{abstract}
-%Students in computational linguistics courses must often learn a new
-%programming language.  In some cases, such as when courses are offered
-%in linguistics departments, the students may be learning to program
-%for the very time.  In order to do interesting projects, it is usually
-%necessary for students to do many low-level ``housekeeping'' tasks.
-%At the same time, teachers of computational linguistics courses
-%sometimes feel that they spend too much time teaching students to
-%program, and not enough time teaching the subject itself.  They may
-%even avoid programming assignments altogether.  However, we believe
-%that it is crucial for a first computational linguistics course to
-%include a strong practical component, in which students develop real
-%programs to solve real problems with real data.
-%
-%Python is a new object-oriented scripting language which runs on all
-%platforms.  Python has been praised as ``executable pseudocode,'' since
-%programs are so easy to write.  Recently, we have been developing
-%NLTK, an open-source Natural Language Toolkit written in Python.
-%In this presentation, we will motivate, describe and demonstrate NLTK.
-%
-%NLTK, the Natural Language Toolkit, is a suite of program modules,
-%tutorials and problem sets.  NLTK covers symbolic and statistical
-%natural language processing, and is interfaced to annotated corpora.
-%Students augment and replace existing NLTK components, learning
-%structured programming by example, and manipulating sophisticated
-%models from the outset.  Along with extensive documentation and
-%problem sets, NLTK provides self-contained, ready-to-use CL
-%courseware.
-%\end{abstract}
+\begin{abstract}
+NLTK, the Natural Language Toolkit, is a suite of program modules,
+tutorials and problem sets, providing ready-to-use computational
+linguistics courseware.  NLTK covers symbolic and statistical natural
+language processing, and is interfaced to annotated corpora.  Students
+augment and replace existing components, learning structured
+programming by example, and manipulating sophisticated models from the
+outset.
+\end{abstract}
 
@@
 \bibliographystyle{acl}
-\bibliography{submission}
+\bibliography{nltk}
 

### 2002-05-06 technical/acl2002/acl02.tex
Several minor editorial corrections
svn/trunk@551

@@
 Edward Loper and Steven Bird\\
-Department of Computer and Information Science and Linguistic Data Consortium\\
+Department of Computer and Information Science \\
 University of Pennsylvania, Philadelphia, PA 19104-6389, USA
@@
 \begin{abstract}
-NLTK, the Natural Language Toolkit, is a suite of program modules,
+NLTK, the Natural Language Toolkit, is a suite of open source program modules,
 tutorials and problem sets, providing ready-to-use computational
@@
 
-% [EL] ``functionality'' isn't a great word, here.
 Teachers of introductory courses on computational linguistics are
@@
 task because different computational linguistics domains require a
-variety of different data structures and functionality; and because a
+variety of different data structures and functions; and because a
 diverse range of topics may need to be included in the syllabus.
@@
 Prolog for parsing, Perl for corpus processing, and a finite-state
-toolkit such for morphological analysis.  By relying on the built-in
+toolkit for morphological analysis.  By relying on the built-in
 features of various languages, the teacher avoids having to develop a
@@
 
-An unfortunate consequence is that a significant part of the course
-must be devoted to teaching these languages.  Further, many
+An unfortunate consequence is that a significant part of such courses
+must be devoted to teaching programming languages.  Further, many
 interesting projects span a variety of domains, and would require that
@@
 analysis; and Prolog for parsing.
-% [EL] This was more precise, but harder to read/parse:
-%    (Perl for file I/O, format conversions and
-%    overall program control, with calls out to a finite state toolkit for
-%    morphological analysis and to a Prolog engine for parsing).  
 It is clear that these considerable overheads and shortcomings warrant
@@
 
-% [EL] ``difficulties'' doesn't seem like the right word here?
-% Maybe ``challenging requirements''?
 Apart from the practical component, computational linguistics courses
@@
 
-% Overview
-% [EL] Given that we're short on space, I decided that this wasn't
-% really necessary for a 6-page paper.  If we do include it, it needs
-% to be completed (I only went up through sec:uses).
-%%Section \ref{sec:python} discusses why we decided to implement the
-%%toolkit using Python.  Section \ref{sec:criteria} describes several
-%%important considerations and desiderata that guided the design and
-%%implementation of the toolkit.  Sections \ref{sec:modules} and
-%%\ref{sec:documentation} describe the current contents of the toolkit.
-%%Section \ref{sec:uses} explains how the toolkit can be used for both
-%%assignments and class demonstrations.
-
-% [EL] This was too ambiguous for me (e.g., we discuss Python)
-%All materials discussed here are available under an open source
-%license from \url{nltk.sf.net}.  NLTK runs on most platforms,
-%including Windows, OS X, Linux, and UNIX.
 The Natural Language Toolkit is available under an open source license
 from \mbox{\url{http://nltk.sf.net/}}.  NLTK runs on most platforms, including
-Windows, OS X, Linux, and UNIX.
+Windows, OS X, Linux, and Unix.
 
@@
 is easy to encapsulate data and methods inside Python classes.
-Finally, Python has an interface to the ``Tk'' graphics toolkit
+Finally, Python has an interface to the Tk graphics toolkit
 \cite{tkinter}, and writing graphical interfaces is straightforward.
@@
 structures, and existing modules that implement the same interface.
-% Advanced projects..
-Advanced projects typically involve the development of entirely new
-functionality for a previously unsupported NLP task, or the
-development of a complete system out of existing and new modules.
 
@@
 control flow can be easily modified during the demonstration, in
-response questions from the class.
+response to questions from the class.
 
@@
 
+\subsection{Advanced projects}
+
+Advanced projects typically involve the development of entirely new
+functionality for a previously unsupported NLP task, or the
+development of a complete system out of existing and new modules.
+
 % ===================== Evaluation =====================
@@
 students had a background in either computer science or linguistics
-(but not both).  Students were required to complete five assignments,
+(and occasionally both).  Students were required to complete five assignments,
 two exams, and a final project.  All class materials are available
@@
 Luckily, Python has extensive support for installation scripts, which
-made these upgrades much simpler.  The students encountered a couple
+made these upgrades simple.  The students encountered a couple
 of bugs in the toolkit, but none were serious, and all were quickly
@@
 % [EL] Pick better 3 adjectives :)
-NLTK provides a simple, extensible, uniform framework for student
-assignments and class demonstrations.  It is well documented, easy to
+NLTK provides a simple, extensible, uniform framework for
+assignments, projects, and class demonstrations.  It is well documented, easy to
 learn, and simple to use.  We hope that NLTK will allow computational

### 2002-05-06 technical/acl2002/acl02.tex
- Added section on type checking - Added comments on visualization not needing to be simple - Moved chart parsing figure to put it on the same page as the   discussion of it - Minor changes
svn/trunk@552

@@
 %   - Write up other approaches
-% We assume that the user is familiar with the acronym ``NLP;'' is
-% that ok?
+%   - Write up advanced projects section
 
@@
 %   - uses
-%     - assignments/projects
+%     - assignments
 %     - class demonstrations
+%     - projects
 %   - evaluation
@@
 regular-expression based implementation of the chunk parser interface.
+% [EL] Add the PCFG parser(s) to this once I'm done with my 630 project
 
@@
 
+\subsection*{Type Checking}
+
+% [EL] This could use another pass or two. :)
+Debugging time is an important factor in the toolkit's ease of use.
+To reduce the amount of time students must spend debugging their code,
+we provide a type checking module, which can be used to ensure that
+functions are given valid arguments.  The type checking module is
+used by all of the basic data types and processing classes.
+
+Since type checking is done explicitly, it can slow the toolkit down.
+However, when efficiency is an issue, type checking can be easily
+turned off; and with type checking is disabled, there is no
+performance penalty.
+
 \subsection*{Visualization}
-
-% jb: this is broken up oddly.
 
@@
 with NLP tasks.  The \texttt{draw.tree} module provides a simple
-graphical interface for displaying tree structures; and the
+graphical interface for displaying tree structures.  The
 \texttt{draw.tree\_edit} module provides an interface for building and
@@
 graphical tool for experimenting with chart parsers.
+
+% [EL] This still needs work...
+The visualization modules provide interfaces for interaction and
+experimentation; they do not directly implement any NLP data
+structures or tasks.  Simplicity of implementation is therefore less
+of an issue for the visualization modules than it is for the rest of
+the toolkit.
 
@@
 
-\subsection{Assignments and Projects}
+\subsection{Assignments}
 
@@
 
+\begin{figure}
+\centerline{\epsfig{figure=chartparse.eps,width=\linewidth}}
+\caption{Chart Parsing Tool}\label{fig:chartparse}
+\vspace*{2ex}\hrule
+\end{figure}
+
 This tool can be used to explain several different aspects of chart
@@
 
-\begin{figure}
-\centerline{\epsfig{figure=chartparse.eps,width=\linewidth}}
-\caption{Chart Parsing Tool}\label{fig:chartparse}
-\vspace*{2ex}\hrule
-\end{figure}
-
-\subsection{Advanced projects}
+\subsection{Advanced Projects}
 
@@
 development of a complete system out of existing and new modules.
+
+% [EL] FILL THIS IN!!!!
 
@@
 
-%Problems: available corpus data, ...
-%
-%Other strengths / weaknesses
-
 % ===================== Other Approaches =====================
@@
 % [EL] Pick better 3 adjectives :)
-NLTK provides a simple, extensible, uniform framework for
-assignments, projects, and class demonstrations.  It is well documented, easy to
+NLTK provides a simple, extensible, uniform framework for assignments,
+projects, and class demonstrations.  It is well documented, easy to
 learn, and simple to use.  We hope that NLTK will allow computational
@@
 
+% [EL] Add something saying explicitly that there's still a lot to
+% do?  (In response to reviewer comment)
 %Future work: 
@@
 standard collection of large, clean, open-source corpora, containing
-corpora appropriate for every module defined by the toolkit.
+data appropriate for every module defined by the toolkit.
 

### 2002-05-07 technical/multivalue/multivalue.tex
- Initial import: technical report discussing what standards to use for   processing classes that return multiple alternative values.
svn/trunk@559

% Natural Language Toolkit Technical Report:
% Multiple Return Values
%
% Copyright (C) 2001 University of Pennsylvania
% Author: Edward Loper <edloper@gradient.cis.upenn.edu>
% URL: <http://nltk.sf.net>
% For license information, see LICENSE.TXT
%
% $Id$

\documentclass[11pt]{article}
\usepackage{fullpage}

\begin{document}
\title{Alternative Outputs\\
\Large Natural Language Processing Toolkit: Technical Report}
\author{Edward Loper}
\maketitle

% From the info file:
% Many processing modules can find more than one "possible" or
% "reasonable" result.  This report disucsses a set of conventions for
% these situations.

% Things to talk about:
%   - what are AO's used for?
%      - 1-input/1-output, multi-in/multi-out, etc.
%      - threading
%        - with probabilities (independence?)
%        - with 
%   - ProbablisticMixIn
%   - interfaces: parse, parse_n, parse_dist
%   - procedure names

%============= INTRODUCTION =============%
\section{Introduction}

  Many NLP algorithms can generate multiple ``possible'' outputs for a
  given input.  For example, a parser might produce several different
  syntax trees for a single sentence.  Since this pattern is so
  common, we have decided to adopt standards interfaces for encoding
  it in NLTK processing classes.  This techincal report explains and
  justifies these interfaces.

  Note that this document is concerned with algorithms that generate a
  variable number of \emph{mutually exclusive} possibilities for a
  single output value.  In particular, it is not meant to address
  tasks that generate multiple output values.  For example, this
  document does \emph{not} address what methods should be used to
  return the list of (overlapping) classes generated by a
  multiple-category text classifier.

  We will use the term \emph{alternative outputs} to refer to any set
  of mutually exclusive possibilities for a single output value.  A
  processing class that generates alternative outputs is an
  \emph{alternative output class}, or \emph{AO class} for short.

%============= ALTERNATIVE OUTPUTS =============%
\section{Alternative Output Types}

  Many AO classes can estimate information about the likelihood of
  their output alternatives.  We define three types of AO class, based
  on what kind of likelihood information they can estimate about each
  alternative:

  \begin{description}
    \item An \emph{AO-unordered class} has no information about the
    likelihood of each alternative.

    \item An \emph{AO-ordered class} can estimate the relative
    ordering of the alternative's likelihoods; however, it does not
    have direct information about the likelihood of any given
    alternative.

    \item An \emph{AO-distribution class} can estimate the probability
    of each alternative.
  \end{description}

  Other types of AO class might occasionally prove useful.  For
  example, we could define a type of AO class that generates a partial
  ordering of alternatives; or a type of AO class that generates
  likelihood estimates that are not probabilities.  However, in the
  interest of keeping our interfaces simple, we choose to restrict
  ourselves to these three types.  We believe that these types will be
  sufficient for NLTK.

%============= USE CASES =============%
\section{Use Cases}

  Before deciding what interfaces AO classes should implement, it is
  important to consider how they will be used.  This section describes
  and categorizes the different ways that we expect AO classes to be
  used.  First, we will examine the different ways that a single AO
  class might be used; then, we will examine typical ways of combining
  multiple AO classes.

  \subsection{AO Class Use Cases}

In section (?), we defined three different types of AO class

    An AO class can map a single input to a set of alternative
    outputs.  Depending on how we intend to use the task, we might
    want different kinds of information about this set.

    The kinds of information the task can recieve depend on what
    information we have about the outputs' relative likelihood.  The
    following sections list the types of information we might want,
    given different amounts of information about output likelihoods.

    \subsubsection{Alternative Sets}

      If the

      \begin{description}
        \item[any] We wish to have a single alternative.
        \item[all] We want all alternatives
        \item[any $n$] We wish to have at most $n$ alternatives.
      \end{description}

    \subsubsection{Alternative Lists}

      \begin{description}
        \item[best]
        \item[sorted]
        \item[best $n$]
      \end{description}

      (worst, worst $n$)

    \subsubsection{Alternative Distributions}
% 1. we want the best answer
% 2. we want the best n answers
% 3. we want the probability of a given answer
% 4.

      There are four foo:

      \begin{description}
        \item[best prob]
        \item[sorted+prob]
        \item[best $n$+prob]
      \end{description}

    \subsubsection{Complex Outputs}

      We might want P's on each element of a list of outputs..  But
      too bad! :)

    what info we want: best, best n, best n w/ probs, pdist

  \subsection{Multiple-Task}

%============= PROBABILITY =============%
\section{Modelling Probability}
  \section{Probability Distributions}

    one way to represent probs is pdists.

  \section{ProbablisticMixIn}

    another way is probablistic mixins.

%============= METHODS =============%
\section{Alternative Output Methods}
% - parse_n
% - parse_dist

  \subsection{Method Names}

  We considered a number of different names\ldots

\end{document}

### 2002-05-13 technical/acl2002/acl02.tex
Wrote survey section, new paragraph in conclusion
svn/trunk@612

@@
 % I got was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
+% [SB] changed Python citation to URL
 In surveying the available languages, we believe that Python offers an
@@
 object-oriented scripting language developed by Guido van Rossum
-and available on all platforms \cite{python}.  Python offers a
+and available on all platforms (\url{www.python.org}).  Python offers a
 shallow learning curve; it was designed to be easily learnt by
@@
 
-% (Leaving this out for now..)
-%The Language: Python was new to 95\% of students.  Some were initially
-%bothered about its block structuring, which is sensitive to white
-%space.  All quickly learnt the language, and reported that any initial
-%reservations were unfounded.  Many students said that they enjoyed
-%writing Python code.
-
-% Where does this go??  Leaving it out for now..
-%At the beginning of the semester, we distributed 2 CD-ROMs to the
-%students, containing NLTK and all of the corpera that we would use for
-%the class.
-
 The experience of using NLTK was very positive, both for us and for
@@
 
-This section will include a discussion of other approaches to
-implementing computational linguistics courseware, including
+The computational component of computational linguistics courses takes
+many forms.  In this section we briefly review a selection of approaches,
+classified according to the (original) target audience.
+
+{\bf Linguistics Students.}
+Various books introduce programming or computing to linguists.
+These are elementary on the computational side, providing a
+gentle introduction to students having no prior experience
+in computer science.  Examples of such books are:
+\emph{Using Computers in Linguistics}
+\cite{Lawler98}, and
 \emph{Programming for Linguistics: Java Technology for Language
-Researchers}\cite{Hammond02} and \emph{the OpenNLP
-project}\cite{opennlp}.
+Researchers} \cite{Hammond02}.
+
+{\bf Grammar Developers.}
+Infrastructure for grammar development has a long history in
+unification-based (or constraint-based) grammar frameworks, from DCG
+\cite{PereiraWarren80} to HPSG \cite{PollardSag94}.  Recent work includes
+\cite{Copestake00,Baldridge02}.  A concurrent development has been the
+finite state toolkits, such as the Xerox toolkit \cite{Beesley02}.  This
+work has found widespread pedagogical application.
+
+{\bf Other Researchers and Developers.}
+A variety of toolkits have been created for research or R\&D
+purposes.  Examples include
+the \emph{CMU-Cambridge Statistical Language Modeling Toolkit}
+\cite{Clarkson97},
+the \emph{EMU Speech Database System} \cite{Harrington99},
+the \emph{General Architecture for Text Engineering} \cite{Bontcheva02},
+the \emph{Maxent Package for Maximum Entropy Models} \cite{maxent},
+and the \emph{Annotation Graph Toolkit} \cite{MaedaBird02}.
+Although not originally motivated by pedagogical needs, all of these
+toolkits have pedagogical applications and many have already been
+used in teaching.
 
@@
 building NLP components and systems.
+
+NLTK is unique in its combination of three factors.
+First, it was deliberately designed as courseware and gives pedagogical
+goals primary status.  Second, its target audience consists of both
+linguists and computer scientists, and it is accessible and challenging
+at many levels of prior computational skill.  Finally, it is based on
+an object-oriented scripting language supporting rapid prototyping and
+literate programming.
 

### 2002-05-14 technical/acl2002/acl02.tex
- Added pcfgparser to parser modules - Wrote advanced projects section - Added acknowledgements section (still needs to be filled in!)
svn/trunk@616

@@
 % To do:
-%   - Write up other approaches
-%   - Write up advanced projects section
+%   - Add acknowledgements
 
@@
 
-% [EL] ``punitive'' like a strange word.
+% [EL] ``punitive'' seems like a strange word.
 % [EL] I couldn't find a REF for ``executable pseudocode'' -- closest
@@
 and its implementation all need to be carefully and thoroughly
-documented.  
-%Three documentation types are necessary: tutorials, to
-%teach students how to use the toolkit to perform specific tasks;
-%technical reports, to explain the toolkit's design and implementation;
-%and reference documentation, to describe every module, interface,
-%class and method.  
-All nomenclature must be carefully chosen and consistently used.
+documented.  All nomenclature must be carefully chosen and
+consistently used.
 
@@
 
-Three modules provide implementations for these abstract interfaces.
+Four modules provide implementations for these abstract interfaces.
 The \texttt{srparser} module implements a simple shift-reduce parser.
-The \texttt{chartparser} module provides a flexible parser that uses a
-\emph{chart} to record hypotheses about syntactic constituents.  And
-the \texttt{rechunkparser} module defines a transformational
-regular-expression based implementation of the chunk parser interface.
-% [EL] Add the PCFG parser(s) to this once I'm done with my 630 project
+The \texttt{chartparser} module defines a flexible parser that uses a
+\emph{chart} to record hypotheses about syntactic constituents.  The
+\texttt{pcfgparser} module provides a variety of different parsers for
+probabilistic grammars.  And the \texttt{rechunkparser} module defines
+a transformational regular-expression based implementation of the
+chunk parser interface.
 
@@
 The visualization modules provide interfaces for interaction and
-experimentation; they do not directly implement any NLP data
-structures or tasks.  Simplicity of implementation is therefore less
-of an issue for the visualization modules than it is for the rest of
-the toolkit.
+experimentation; they do not directly implement NLP data structures or
+tasks.  Simplicity of implementation is therefore less of an issue for
+the visualization modules than it is for the rest of the toolkit.
 
@@
 % [EL] Provide some indication of how much tutorial material is
-% available?  (Currently, 8 tutorials, mean=16pp, range=6pp-52pp.)
+% available?  (Currently, 9 tutorials, mean=16pp, range=6pp-52pp.)
 \paragraph{\textit{Tutorials}} teach students how to use the toolkit,
@@
 
-Advanced projects typically involve the development of entirely new
+NLTK provides students with a flexible framework for advanced
+projects.  Typical projects involve the development of entirely new
 functionality for a previously unsupported NLP task, or the
@@
 
-% [EL] FILL THIS IN!!!!
+% Broad coverage 
+The toolkit's broad coverage allows students to explore a wide variety
+of topics.  In our introductory computational linguistics course,
+topics for student projects included text generation, word sense
+disambiguation, collocation analysis, and morphological analysis.
+
+% Less grunt-work
+NLTK eliminates the tedious infrastructure-building that is typically
+associated with advanced student projects, by providing students with
+the basic data structures, tools, and interfaces that they need.  This
+allows the students to concentrate on the problems that interest them.
+
+% Meaningful contribution
+The collaborative, open-source nature of the toolkit can provide
+students with a sense that their projects are meaningful
+contributions, and not just exercises.  Several of the students in our
+course have expressed interest in incorporating their projects into
+the toolkit.
+
+% Good examples
+Finally, many of the modules included in the toolkit provide students
+with good examples of what projects should look like, with well
+thought out interfaces, clean code structure, and thorough
+documentation.
+
+\subsubsection*{Example: Probabilistic Parsing}
+% [EL] I'm a bit uncomfortable including this description without
+% mentioning that I'm the author -- seems a little misleading.  But at
+% the same time, it seems like it's much less interesting if I say
+% that I wrote it, given that I wrote the majority of the toolkit
+% anyway.  :-/  Perhaps we should just leave out this section.  Any
+% ideas?
+
+The probabilistic parsing module was created as a class project for a
+statistical NLP course.  The toolkit provided the basic data types and
+interfaces for parsing.  The project extended these, adding a new
+probabilistic parsing interface, and using subclasses to create a
+probabilistic version of the CFG data structure.  These new components
+were used in conjunction with several existing components, such as the
+chart data structure, to define two implementations of the
+probabilistic parsing interface.  Finally, a tutorial was written that
+explained the basic motivations and concepts behind probabilistic
+parsing, and described the new interfaces, data structures, and
+parsers.
 
@@
 %Future work: 
-We plan to continue extending the breadth of materials covered by the
-toolkit.  We are currently working on NLTK modules for Hidden Markov
-Models, probabilistic parsing, and tree adjoining grammars.  We also
-plan to increase the number of algorithms implemented by some existing
-modules, such as the text classification module.
+We plan to continue extending the breadth of materials covered by
+the toolkit.  We are currently working on NLTK modules for Hidden
+Markov Models, language modeling, and tree adjoining grammars.  We
+also plan to increase the number of algorithms implemented by some
+existing modules, such as the text classification module.
 
@@
 
+% ===================== Acknowledgements =====================
+\section{Acknowledgments}
+\label{sec:acknowledgments}
+
+\emph{Add acknowledgements: reviewers; funders; students of 530;
+others?}
+
 \bibliographystyle{acl}
@@
 \end{document}
+% LocalWords:  stevenbird Exp pcfgparser subclasses CFG DCG HPSG toolkits CMU

### 2002-05-14 technical/acl2002/acl02.tex
- Minor wording changes
svn/trunk@618

@@
 language processing, and is interfaced to annotated corpora.  Students
-augment and replace existing components, learning structured
-programming by example, and manipulating sophisticated models from the
+augment and replace existing components, learn structured
+programming by example, and manipulate sophisticated models from the
 outset.
@@
 task because different computational linguistics domains require a
-variety of different data structures and functions; and because a
+variety of different data structures and functions, and because a
 diverse range of topics may need to be included in the syllabus.
@@
 
-% [EL] is ``desiderata'' an acceptable word here?
 The most basic step in setting up a practical component is choosing a
-suitable programming language.  A number of desiderata influenced our
-choice.  First, the language must have a shallow learning curve, so
-that novice programmers get immediate rewards for their efforts.
+suitable programming language.  A number of considerations influenced
+our choice.  First, the language must have a shallow learning curve,
+so that novice programmers get immediate rewards for their efforts.
 Second, the language must support rapid prototyping and a short

### 2002-05-15 technical/acl2002/acl02.tex
ran ispell, added my hyphenation rules, enlarged cascade displays, other minor editorial corrections, acknowledgements
svn/trunk@620

@@
 \usepackage{acl2002,url,alltt,epsfig}
+
+% hyphenation control
+\pretolerance 250
+\tolerance 500
+\hyphenpenalty 200
+\exhyphenpenalty 100
+\doublehyphendemerits 7500
+\finalhyphendemerits 7500
+\brokenpenalty 10000
+\lefthyphenmin 3
+\righthyphenmin 3
+\widowpenalty 10000
+\clubpenalty 10000
+\displaywidowpenalty 10000
+\looseness 1
+
 \title{NLTK: The Natural Language Toolkit}
@@
 \date{}
-
-% To do:
-%   - Add acknowledgements
 
@@
 
-\newenvironment{sv}{\scriptsize\begin{alltt}}{\end{alltt}\normalsize}
+\newenvironment{sv}{\small\begin{alltt}}{\end{alltt}\normalsize}
 
@@
 The Natural Language Toolkit is available under an open source license
-from \mbox{\url{http://nltk.sf.net/}}.  NLTK runs on most platforms, including
-Windows, OS X, Linux, and Unix.
+from \mbox{\url{http://nltk.sf.net/}}.  NLTK runs on all platforms
+supported by Python, including Windows, OS X, Linux, and Unix.
 
@@
   UnChunkRule("<,|{\textbackslash}{\textbackslash}.|``|''>"),
-  MergeRule('<NN.*|DT|JJ.*|CD>', '<NN.*|DT|JJ.*|CD>'),
+  MergeRule('<NN.*|DT|JJ.*|CD>',
+            '<NN.*|DT|JJ.*|CD>'),
   SplitRule('<NN.*>', '<DT|JJ>')
@@
 NLTK eliminates the tedious infrastructure-building that is typically
-associated with advanced student projects, by providing students with
+associated with advanced student projects by providing students with
 the basic data structures, tools, and interfaces that they need.  This
@@
 with good examples of what projects should look like, with well
-thought out interfaces, clean code structure, and thorough
+thought-out interfaces, clean code structure, and thorough
 documentation.
@@
 
-Finding suitable corpora is a prerequisite for many student
-assignments and projects.  We are therefore putting together a
-standard collection of large, clean, open-source corpora, containing
-data appropriate for every module defined by the toolkit.
+Finding suitable corpora is a prerequisite for many student assignments and
+projects.  We are therefore putting together a collection of corpora
+containing data appropriate for every module defined by the toolkit.
 
@@
 
-\emph{Add acknowledgements: reviewers; funders; students of 530;
-others?}
+We are indebted to our students for feedback on the toolkit,
+and to anonymous reviewers and the workshop organizers for comments
+on an earlier version of this paper.  We are grateful to Mitch
+Marcus and the Department of Computer and Information Science
+at the University of Pennsylvania for sponsoring the work
+reported here.
 

### 2002-05-15 technical/acl2002/acl02.tex
-Spelled out CFG - Added Jee Bang to acknowledgements
svn/trunk@621

@@
 probabilistic parsing interface, and using subclasses to create a
-probabilistic version of the CFG data structure.  These new components
-were used in conjunction with several existing components, such as the
-chart data structure, to define two implementations of the
-probabilistic parsing interface.  Finally, a tutorial was written that
-explained the basic motivations and concepts behind probabilistic
-parsing, and described the new interfaces, data structures, and
-parsers.
+probabilistic version of the context free grammar data structure.
+These new components were used in conjunction with several existing
+components, such as the chart data structure, to define two
+implementations of the probabilistic parsing interface.  Finally, a
+tutorial was written that explained the basic motivations and concepts
+behind probabilistic parsing, and described the new interfaces, data
+structures, and parsers.
 
@@
 
-We are indebted to our students for feedback on the toolkit,
-and to anonymous reviewers and the workshop organizers for comments
-on an earlier version of this paper.  We are grateful to Mitch
-Marcus and the Department of Computer and Information Science
-at the University of Pennsylvania for sponsoring the work
-reported here.
+We are indebted to our students for feedback on the toolkit, and to
+anonymous reviewers, Jee Bang, and the workshop organizers for
+comments on an earlier version of this paper.  We are grateful to
+Mitch Marcus and the Department of Computer and Information Science at
+the University of Pennsylvania for sponsoring the work reported here.
 

### 2002-06-07 technical/handout/handout.tex
- Updated to include new modules; and revised descriptions of some   modules. - Moved URL section to page 1, to make room for more modules.
svn/trunk@713

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage{fullpage}
@@
 \author{Steven Bird \& Edward Loper}
+\date{}
 
-% clunch:
-%\date{Thursday December 12, 2001}
-
-% lsa:
-\date{Sunday January 6, 2002}
-
+%////////////////////////////////////////
+% Page 1: Overview
+%////////////////////////////////////////
 \maketitle
@@
 
+\section*{URL}
+
+For more information about NLTK, or to download a copy, please
+visit our web page:
+
+\texttt{http://nltk.sourceforge.net}
+
+%////////////////////////////////////////
+% Page 2: Modules
+%////////////////////////////////////////
 \section*{Modules}
@@
 
-{\small
+{%\small
 \begin{itemize}\nogap
@@
     \begin{itemize}\nogap
-      \item \texttt{token}: Basic classes for processing individual
-            elements of text, such as words or sentences. 
-      \item \texttt{tree}: Classes for representing tree structures
-            over text (such as syntax trees and morphological trees).
+      \item \texttt{token}: Basic classes for encoding and processing
+            individual elements of text, such as words or sentences. 
+      \item \texttt{tree}: Classes for representing hierarchical structures
+            over text, such as syntax trees.
       \item \texttt{probability}: Classes that encode frequency
@@
     \begin{itemize}\nogap
+      \item \texttt{cfg}: Basic data types for encoding context free
+            grammars. 
       \item \texttt{parser}: A standard interface to produce trees
-            representing the structure of texts.
-      \item \texttt{chartparser}: A flexible parser implementation
+            representing the structure of texts; and two
+            simple implementations of that interface.
+      \item \texttt{parser.chart}: A flexible parser implementation
             that uses a \emph{chart} to record hypotheses about
             syntactic constituents. 
-      \item \texttt{srparser\_template}: A partial implementation of a
-            shift-reduce parser; completing the implementation was a
-            student exercise.
-      \item \texttt{chunkparser}: A standard interface for robust
+      \item \texttt{parser.chunk}: A standard interface for robust
             parsers used to identify non-overlapping linguistic groups
-            (such as noun phrases) in unrestricted text.
-      \item \texttt{rechunkparser}: A regular-expression based
-            implementation of the chunk parser interface.
+            (such as base noun phrases) in unrestricted text; and a
+            regular-expression based implementation of that interface.
+      \item \texttt{parser.probabilistic}: A standard interface for
+            probabilistic parsers; and two implementations of that
+            interface. 
     \end{itemize}
@@
       \item \texttt{classifier.featureselection}: A standard interface
-          for choosing which features are relevant for making
-          classification decisions.
+          for choosing which features are relevant for a given
+          classification decision.
     \end{itemize}
 
-  \item Visualization
+  \item Finite State Automata
     \begin{itemize}\nogap 
-      \item \texttt{draw.tree}: A graphical representation for tree
-            structures, such as syntax trees and morphological trees. 
+      \item \texttt{fsa}: Classes for representing finite state
+          automata and regular expressions.
+    \end{itemize}
+
+  \item Visualization and Interactive Tools
+    \begin{itemize}\nogap 
+      \item \texttt{draw}: A Tk-based framework for building
+            graphical tools.
+      \item \texttt{draw.tree}: A graphical representation for
+            hierarchical structures.
       \item \texttt{draw.tree\_edit}: A graphical interface used to
             build and modify tree structures.
-      \item \texttt{draw.plot\_graph}: A graphical tool to graph
+      \item \texttt{draw.fsa}: A graphical representation for finite
+            state automata.
+      \item \texttt{draw.plot}: A tool for graphing
             arbitrary functions.
-      \item \texttt{draw.chart}: An interactive graphical tool used to
-            experiment with chart parsers.
+      \item \texttt{draw.chart}: An interactive graphical tool for
+            experimenting with the chart parser.
+      \item \texttt{draw.rdparser}: An interactive graphical tool for
+            exploring the recursive descent parser.
+      \item \texttt{draw.srparser}: An interactive graphical tool for
+            learning about the shift/reduce parser.
     \end{itemize}
@@
 }
-\section*{URL}
-
-For more information about NLTK, or to download a copy, please
-visit our web page:
-
-\texttt{http://nltk.sourceforge.net}
 

### 2002-07-06 technical/conventions/conventions.tex
- Added conventions technical report: describes design, coding,   documenting, and testing conventions for nltk.
svn/trunk@800

% Natural Language Toolkit Technical Report:
% Conventions
%
% Copyright (C) 2001 University of Pennsylvania
% Author: Edward Loper <edloper@gradient.cis.upenn.edu>
% URL: <http://nltk.sf.net>
% For license information, see LICENSE.TXT
%
% $Id$

\documentclass{article}
\usepackage{fullpage}
\usepackage{boxedminipage}

\begin{document}
\title{Conventions}
\author{Edward Loper}
\maketitle

%################################################################
%#  OVERVIEW
%################################################################
\section{Overview}

\section{Immutable vs Mutable Variables}

  Basic data types should generally be immutable.

\section{Nomenclature}

  \subsection{Variable Names}

    all lower case, with underscores

  \subsection{Class Names}

    WordInitialCaps

  \subsection{Module Names}

    short, lower case names.

\section{Packages and Modules}

\section{Data Classes}

\section{Interfaces}

  use AssertionError for methods that must be implemented.

  use NotImplementedError for optional methods.

\section{String Representations}

  str is for multiline; repr is single-line, embeddable.  e.g., str
  for production is something like ``A -> B C'' but for embeddable
  (repr), it's more like ``[production: A -> B C]''..  etc.

\section{Task Interfaces & Classes}

  \subsection{Multiple Return Values}

  \subsection{Trace Output}

    trace output is controlled by a single ``trace level'' instance
    variable.

    trace values: 0: no output; 1: less than a page for a typical
    run; 2 or higher: more verbose.

    setting trace output:
      - by default, trace=0
      - constructor takes a trace keyword argument
      - trace(n) method sets trace level; use default value for n.

\end{document}

### 2003-08-02 technical/conventions/conventions.tex
- Expanded on various sections.
svn/trunk@1032

@@
 \usepackage{fullpage}
-\usepackage{boxedminipage}
+\usepackage{parskip}
+\setlength{\parindent}{0ex}
+\setlength{\parskip}{1ex}
 
@@
 %################################################################
-%#  OVERVIEW
+%#  Overview
 %################################################################
@@
 
+This document lists design and coding conventions that are used
+throughout NLTK.  Whenever possible, these conventions should be
+followed.
+
+You should also look at the source code of existing NLTK modules, to
+get a feel for conventions that aren't explicitly documented here.
+
+%################################################################
+%#  Module Requirements
+%################################################################
+\section{Module Requirements}
+
+The following checklist lists requirements for a module to be added to
+NLTK.  (Some of these requirements may be fulfilled after the module
+is added to NLTK, but they should be fulfilled eventually.)
+
+\begin{itemize}
+\item Reference documentation
+  \begin{itemize}
+  \item The module is documented by a docstring, describing the
+    purpose of the module, and describing how the objects defined by
+    the class relate to that purpose.
+  \item Each interface is documented by a docstring, describing the
+    requirements imposed by the interface.
+  \item Each class is documented by a docstring, describing the class.
+  \item Each function or method is documented by a docstring,
+    including documentation for parameters, parameter types, return
+    value, and return value type.
+  \end{itemize}
+\item Tutorial documentation
+  \begin{itemize}
+  \item Provide motivation for the module's task.
+  \item Explain how to use the module to perform that task.
+  \item Explain how to create and manipulate any data structures
+    defined by the module.
+  \item Explain any algorithms used.
+  \end{itemize}
+
+\item Unit testing code
+  \begin{itemize}
+  \item Each function or method is tested
+  \end{itemize}
+\end{itemize}
+
+%################################################################
+%#  Mutability
+%################################################################
 \section{Immutable vs Mutable Variables}
 
-  Basic data types should generally be immutable.
+When possible, basic data types (such as \texttt{Token},
+\texttt{Tree}, and \texttt{CFGProduction}) should be immutable.  An
+\textit{immutable} object is one whose value cannot be changed, once
+the object is created.  In Python, integers and strings are immutable;
+but lists and dictionaries are mutable.
 
-\section{Nomenclature}
+Immutable objects have several useful properties, but for our
+purposes, the most important are:
+
+\begin{itemize}
+\item They can be used as types for tokens.
+\item They can be used as keys in dictionaries.
+\item They can be stored in Sets (\texttt{nltk.set.Set}).
+\end{itemize}
+
+The following guidelines should help you decide whether an object
+should be mutable or immutable:
+
+\begin{itemize}
+\item Any class that represents a text type \textit{must} be
+  immutable.
+\item Classes that represent fixed values, such as CFG productions or
+  rules for chunking, should be immutable.
+\item Container classes (such as parsing charts) are usually mutable.
+\item If there's not a good reason to make a class mutable, then keep
+  it immutable.
+\end{itemize}
+
+If you're unsure about whether a class should be mutable or immutable,
+contact \texttt{edloper@gradient.cis.upenn.edu}.
+
+\subsection{Making a Class Immutable}
+
+To make a class immutable:
+\begin{itemize}
+\item Do not define any methods that modify its contents.
+\item Do not define any methods that return a reference to a mutable
+  instance variable.  (E.g., don't return a list that's used
+  internally; otherwise, the caller might modify it).  If you need to
+  return a mutable value, then copy it.  But you should also consider
+  using immutable structures (e.g. tuples) for your instance
+  variables.
+\item Define the special method \texttt{\_\_cmp\_\_}, and optionally
+  define the two special methods \texttt{\_\_eq\_\_} and \texttt{\_\_ne\_\_}.
+\item Define the special method \texttt{\_\_hash\_\_}.
+\end{itemize}
+
+%################################################################
+%#  Naming Objects
+%################################################################
+\section{Naming Objects}
+
+Object names should be descriptive, but not overly long.
 
@@
 
-    all lower case, with underscores
+Name variables with lower case names.  Use underscores to separate
+words in the variable name.  Some conventional names for common types
+are:
+
+\begin{tabular}{l|l}
+Type & Conventional Names \\
+\hline
+\texttt{Token} & \texttt{token}, \texttt{tok}\\
+\texttt{Type} & \texttt{type}, \texttt{typ}, \texttt{word}\\
+\texttt{FreqDist} & \texttt{freqdist}, \texttt{fdist}\\
+\texttt{ProbDistI} & \texttt{probdist}, \texttt{pdist}\\
+\texttt{ConditionalFreqDist} & \texttt{cfreqdist}, \texttt{cfdist}\\
+\texttt{ConditionalProbDistI} & \texttt{cprobdist}, \texttt{cpdist}\\
+\texttt{TreeToken} & \texttt{treetok}\\
+\texttt{Tree} & \texttt{tree}\\
+\texttt{CFGProduction} & \texttt{prod}\\
+\texttt{aggedType} & \texttt{ttype}\\
+\end{tabular}
+
+These names are often prefixed by a descriptor, such as
+``\texttt{context\_token}''.
 
@@
 
-    WordInitialCaps
+Name classes with WordInitialCaps.  Name interfaces with a trailing
+``I'' (e.g. \texttt{ProbDistI} or \texttt{ParserI}).
 
@@
 
-    short, lower case names.
+Name modules with short lower-case names.  Modules that define data
+types are generally named after the primary data types they define
+(e.g., \texttt{nltk.cfg} and \texttt{nltk.token}).  Modules that
+define NLP tasks are generally named after the task that they define
+(e.g., \texttt{nltk.tagger} and  \texttt{nltk.parser}).
 
-\section{Packages and Modules}
-
-\section{Data Classes}
-
+%################################################################
+%#  Interfaces
+%################################################################
 \section{Interfaces}
 
-  use AssertionError for methods that must be implemented.
+An \textit{interface} is a special type of base class that specifies a
+set of methods that must be supported.  Each subclass of an interface
+provides an implementation for the methods defined by the interface.
 
-  use NotImplementedError for optional methods.
+Interfaces are implemented as classes, all of whose methods raise
+exceptions.  Required methods should raise \texttt{AssertionError}s;
+and optional methods should raise \texttt{NotImplementedError}s.  Each
+interface method defines a docstring that describes the behavior of
+that method.  See existing interfaces for examples.
 
+%################################################################
+%#  String Representations
+%################################################################
 \section{String Representations}
 
-  str is for multiline; repr is single-line, embeddable.  e.g., str
-  for production is something like ``A -> B C'' but for embeddable
-  (repr), it's more like ``[production: A -> B C]''..  etc.
+\begin{itemize}
+\item \texttt{repr} should return a single-line description of an
+  object, suitable for embedding in other descriptions.
 
-\section{Task Interfaces & Classes}
+\item \texttt{str} should produce a description that is suitable for
+  the output of \texttt{print}.  The \texttt{str} description may span
+  multiple lines, and does not need to have clear delimiters.
+\end{itemize}
+
+%################################################################
+%#  Task Interfaces
+%################################################################
+\section{Task Interfaces \& Classes}
 
@@
 
-    trace values: 0: no output; 1: less than a page for a typical
-    run; 2 or higher: more verbose.
+trace values: 0: no output; 1: less than a page for a typical run; 2
+or higher: more verbose.
 

### 2003-08-03 docbook/index_glossterms.py
a script to add <indexterm>s for each <glossterm> in a docbook file
svn/trunk@1051

@@ -1,0 +1,40 @@
+#
+# index_glossterms.py
+#
+# Add an <indexterm> for each <glossterm> in a docbook document.
+#
+
+import sys, re
+
+if len(sys.argv) != 3:
+    print 'Usage: %s <input> <output>' % sys.argv[0]
+    sys.exit(1)
+
+# Open the output file for writing.
+outfile = open(sys.argv[2], 'w')
+
+PARA_END_RE = re.compile(r'(</para>)\s*')
+
+# We will use re.sub to look for all the glossary terms on each
+# line, using the following regexp and sub function:
+GLOSSTERM_RE = re.compile('<glossterm>(.*?)</glossterm>')
+glossterms = []
+def glossterm_sub(match, glossterms=glossterms):
+    glossterms.append(match.group(1))
+
+for line in open(sys.argv[1]).readlines():
+    # Find all the glossary terms on this line.
+    GLOSSTERM_RE.sub(glossterm_sub, line)
+
+    # If a paragraph ended, then add any accumulated glossary
+    # terms right after it.
+    if glossterms and PARA_END_RE.search(line):
+        indexterms = '\n'
+        for t in glossterms:
+            indexterms += '<indexterm><primary>%s</primary></indexterm>\n' % t
+        line = PARA_END_RE.sub(r'\1'+indexterms, line)
+        glossterms[:] = []
+
+    # Write the line to the output file.
+    outfile.write(line)
+

### 2003-08-04 docbook/index_glossterms.py
- Replaced with a much simpler implementation that fixes a bug (it   missed <glossterm>s that spanned multiple lines before)
svn/trunk@1064

@@ -10,31 +10,13 @@ if len(sys.argv) != 3:
     print 'Usage: %s <input> <output>' % sys.argv[0]
     sys.exit(1)
 
-# Open the output file for writing.
-outfile = open(sys.argv[2], 'w')
+# Define a regexp that finds glossterms.
+GT_RE = re.compile('<glossterm>(.*?)</glossterm>', re.DOTALL)
 
-PARA_END_RE = re.compile(r'(</para>)\s*')
+# Add an indexterm after each glossterm.
+str = GT_RE.sub(r'<glossterm>\1</glossterm>'+
+                r'<indexterm><primary>\1</primary></indexterm>',
+                open(sys.argv[1]).read())
 
-# We will use re.sub to look for all the glossary terms on each
-# line, using the following regexp and sub function:
-GLOSSTERM_RE = re.compile('<glossterm>(.*?)</glossterm>')
-glossterms = []
-def glossterm_sub(match, glossterms=glossterms):
-    glossterms.append(match.group(1))
-
-for line in open(sys.argv[1]).readlines():
-    # Find all the glossary terms on this line.
-    GLOSSTERM_RE.sub(glossterm_sub, line)
-
-    # If a paragraph ended, then add any accumulated glossary
-    # terms right after it.
-    if glossterms and PARA_END_RE.search(line):
-        indexterms = '\n'
-        for t in glossterms:
-            indexterms += '<indexterm><primary>%s</primary></indexterm>\n' % t
-        line = PARA_END_RE.sub(r'\1'+indexterms, line)
-        glossterms[:] = []
-
-    # Write the line to the output file.
-    outfile.write(line)
-
+# Write to the output file.
+open(sys.argv[2], 'w').write(str)

### 2003-10-14 technical/interfaces/interfaces.tex
initial import
svn/trunk@1348

% Natural Language Toolkit Technical Report:
% Interfaces
%
% Copyright (C) 2003 University of Pennsylvania
% Author: Edward Loper <edloper@gradient.cis.upenn.edu>
% URL: <http://nltk.sf.net>
% For license information, see LICENSE.TXT
%
% $Id$

\documentclass[11pt]{article}
\usepackage{fullpage, alltt}

\begin{document}
\title{Interfaces\\
\Large Natural Language Processing Toolkit: Technical Report}
\author{Edward Loper}
\maketitle

%============= INTRODUCTION =============%
\section{Introduction}

An \textit{interface} is a specification of the behavior for a set of
related methods.  NLTK defines standard interfaces to ensure that
different components will work together; and to allow different
implementations of the same component to be easily substituted for
each other.  NLTK defines two types of interface:

\begin{itemize}
\item \textit{Processing interfaces} are used to define natural
  language processing tasks (such as parsing or classifying texts).
\item \textit{Data type interfaces} are used to define standard
  methods for accessing data types that have multiple implementations
  (such as probability distributions).
\end{itemize}

%============= BACKGROUND =============%
\section{Background}
\subsection{Protocols}

Python does not provide any language-level support for defining or
enforcing interfaces.  Instead, Python tends to rely on
\textit{protocols}, which are informal interfaces that are specified
by external documentation.  Protocols are typically defined relative
to existing data types or classes.  For example, the ``sequence''
protocol requires that an object implement the methods that tuples and
lists implement.\footnote{Or some subset of these methods; the exact
  requirements are not usually made explicit.}

This informal interface system allows Python functions to be used
polymorphically: you can call a function with a object type that it
doesn't expect, as long as you guarantee that the object behaves
correctly.  It also allows individual functions to only require
certain aspects of an protocol.  For example, a function that reads from
a file-like object might just require that its input implement the
\texttt{read} and \texttt{readline} methods.

\newpage\noindent
However, Python's informal protocol system has several disadvantages:
\begin{itemize}
\item In order to avoid duplicating documentation, interfaces must be
  documented externally (not in docstrings).  However, there is no
  standard place to doccument interfaces; and since the interfaces are
  separated from the code, it is easy for the interfaces to get
  out-of-sync with the code.
\item Since protocols are usually documented informally, different
  programmers may make different assumptions about what is required or
  guaranteed.  These differences in assumption can lead to subtle
  bugs, which can only be detected at run-time.
\item It is not possible to automatically check that classes correctly
  implement interfaces.
\item If a function requires only certain aspects of an protocol, then
  it can be more difficult to change that function's implementation.
  In the example above, it would be impossible to change the function
  that required a file-like object to an implementation that uses
  \texttt{seek} (without checking all calls to that function).
\end{itemize}

% Add other related work (e.g., zope ExtensionClasses?)

%============= INTERFACE CLASSES =============%
\section{Interface Classes}

For NLTK, we wanted to provide explicit formal definitions for all
interfaces.  We decided to encode these formal definitions using
special superclasses which we call ``interface classes.''  An
\textit{interface class} is a class that defines simple stubs for each
method that the interface specifies.  These stubs use docstrings to
describe the method's required behavior; and raise exceptions if
called.  This system has three primary advantages over external
documentation:

\begin{itemize}
\item For classes that implement an interface, the API documentation
  for methods can be inherited from the interface.\footnote{This
    assumes the use of an API documentation tool that supports
    documentation inheritance, such as Epydoc.}  This helps avoid
  duplicating documentation.
\item It is possible to automatically verify that classes define all
  of the required methods, using the required signatures.
\item It is easier to maintain consistancy between the interfaces'
  definitions and implementations in different versions of the
  toolkit.
\end{itemize}

\subsection{Naming Interface Classes}

Interface classes are always named with a trailing capital ``I,'' such
as \texttt{TokenizerI} or \texttt{EventI}.

\subsection{Interface Class Docstrings}

An interface class's docstring should describe the purpose of the
interface, and provide any necessary background information.  It
should also specify which methods (if any) are optional.

\subsection{Interface Class Constructors}

% Will this cause problems with the new super construct?
Interface classes should not define constructors.

\subsection{Interface Methods}

Every interface class defines a set of \textit{interface methods},
each of which specifies the behavior for a single method of the same
name.  An interface method consists of a declaration, followed by a
body containing a docstring and a \texttt{raise AssertionError(...)}
command:

\begin{alltt}
    def \textit{name}(\textit{parameters...}):
        """
        \textit{docstring}
        """
        raise AssertionError("\textit{name} not overridden")
\end{alltt}

\subsubsection{Optional Methods}

Interfaces may define ``optional methods,'' which should be
implemented when possible, but are not required.  The optionality of a
method is specified in two ways:

\begin{itemize}
\item Optional methods should raise a \texttt{NotImplementedError}
  exception (instead of an \texttt{AssertionError} exception).
\item The interface class docstring should provide a list of all
  optional methods.
\end{itemize}

\noindent
The following template should be used to write optional methods:

\begin{alltt}
    def \textit{name}(\textit{parameters...}):
        """
        \textit{docstring}
        """
        raise NotImplementedError("\textit{name} is not implemented")
\end{alltt}

\subsection{Implementing an Interface}

If a class implements an interface, then it must list that interface
as one of its base classes; and it must override each of the
interface's (required) methods.  Each overriding method must have a
compatible signature with the interface method\footnote{An overriding
  method is compatible with a base method if it will accept any set of
  parameters that the base method would accept.  For example, an
  overriding method may add a new optional argument; but it may not
  remove or rename an argument.}; and must have behavior that conforms
to the behavior described by the interface method's docstring.

Note that a single class can implement multiple interfaces, by
declaring them in its list of base classes, and overriding all of
their methods.

\subsection{Interface Inheritance Hierarchies}

Specialized interfaces can be derived from more general interfaces
using inheritance.  In particular, you can create a specialized
interface class by declaring it as a subclass of a more general
interface.  The documentation for any methods defined by the more
general interface will be inherited by the specialized interface.

For example, the \texttt{ParserI} interface class defines a general
interface for parsing; and its subclass \texttt{ProbabilisticParserI}
defines a more specialized interface for probabilistic parsing.

%============= TYPES OF INTERFACE =============%
\section{Types of Interface}

NLTK defines a large set of interfaces.  To help organize these
interfaces, we have categorized them according to a number of
attributes.  The most basic division is between \textit{processing
  interfaces}, which are used to define natural language processing
tasks (such as parsing or classifying texts); and \textit{data type
  interfaces}, which are used to define standard interfaces to data
types that have multiple implementations.

%============= PROCESSING INTERFACES =============%
\section{Processing Interfaces}

The primary use of interfaces in NLTK is to define standard
specifications for natural language processing tasks.  Each processing
interface is named after a single processing task, such as
\texttt{TokenizerI} (for tokenizing) and \texttt{ParserI} (for
parsing).\footnote{A processing interface's name should be based the
  deverbal noun formed from the task's verb.}

Each processing interface defines at least one method, which performs
the language processing task and returns the result.  If no result is
found, then this method should return \texttt{None}.  The name for
this method should be based on imperitive form of the interface name,
such as \texttt{tokenize} (for tokenizing) and \texttt{parse} (for
parsing).

When appropriate, each processing interface should also define a
method that returns a list of the possible results for the given task,
sorted in descending order of quality.\footnote{For some
  implementation-specific definition of quality; note that it is
  perfectly acceptable to define all possible results to have the same
  quality, in which case results will be returned in arbitrary order.}
This method is named by appending ``\texttt{\_n}'' to the imperitive
form of the interface name, such as \texttt{tokenize\_n} for
tokenizing; and \texttt{parse\_n} for parsing.  The signature for this
methods should be formed by adding an optional parameter \texttt{n} to
the signature of the basic language processing task method, which
controls the maximum number of results that should be returned.  If
\texttt{n} is not specified, then all available results should be
returned.

\subsection{Input and Output}

Almost all processing interfaces define tasks that take tokens or
collections of tokens as input; and return tokens or collections of
tokens as outputs.  Some advantages of working with tokens rather than
types include:

\begin{itemize}
\item It makes it easy to keep track of what information was derived
  from what source; and to connect together pieces of information that
  have been generated by different tasks.
\item It makes it easy to differentiate two occurances of the same
  type.
\item It often aids in the evaluation of system performance, by
  allowing direct token-by-token comparison of two algorithms' output
  without the need for any alignment algorithms.
\end{itemize}

A notable exception to the rule that most interfaces take tokens as
inputs is the \texttt{tokenizer} method, which is responsible for
creating the original list of tokens from a string.

\newpage
Processing interfaces can be divided into classes, based on their
input and output.  Currently, all NLTK processing classes fit into the
following categories:

\begin{itemize}
\item \textit{Token list generator}: creates a list of tokens.
  \texttt{TokenizerI} is an example of a token list generator.
\item \textit{Token map}: takes a single token as its input, and
  returns a token with the same location and a new type.
  \texttt{ClassifierI} is an example of a token map.
\item \textit{Token list map}: takes a list of tokens as its input,
  and returns a list of tokens with the same locations, and new types.
  \texttt{TaggerI} is an example of a token list map.
\item \textit{Token list collector}: takes a list of tokens as its
  input, and returns a single new token composed from the list.
  \texttt{ParserI} is an example of a token list collector.
\item \textit{Processing class factory}: creates new instances of
  processing classes.  C{ClassifierTrainerI} is an example of a
  processing class factory.
\end{itemize}

Additional classes are likely to be added in the future.  Some likely
candidates include the following classes:

\begin{itemize}
\item \textit{Token list filter}: takes a list of tokens as its
  input, and returns a list containing a subset of the tokens from the
  input list (preserving order).  A stop-word list might be defined
  with this class of processing interface.
\item \textit{Token expander}: takes a single token as its input, and
  returns a list of tokens derived from the single token.  A
  morphological analyzer might be defined with this class of
  processing interface.
\end{itemize}

\subsection{Function}

Processing interfaces can be also be divided into classes based on
their basic function.  This provides a partially orthoganal
classification to the the input/output categories.  Currently, all
NLTK processing classes fit into the following categories:

\begin{itemize}
\item \textit{Extender}: Returns data containing a superset of the
  information in the input.  An example of a token extender is
  \texttt{TaggerI}, which adds tags to each tokens' types; or
  \texttt{ParserI}, which adds constituancy information.
\item \textit{Selector}: Returns data containing a subset of the
  information in the input.  An example of a selector is a stop-word
  list, which filters out tokens with a given set of types.
\item \textit{Transformer}: Returns data that contains neither a
  superset nor a subset of the information in the input.  An example
  of a transformer is an automatic speech recognition processing
  interface.
\end{itemize}

Note that the distinction between these classes is often \textit{not}
determined by the processing task itself.  For example, we could
define a tagging interface to return either a list of tagged words (an
extender) or a list of tags (a transformer).

%============= DATA TYPE INTERFACES =============%
\section{Data type Interfaces}

Data type interfaces are used to provide specifications for data types
that are too general to be captured by a single class.  For example,
the \texttt{ProbDistI} interface class defines an interface for
probability distributions, which can be implemented using a wide
variety of techniques.

In NLTK, data type interfaces are not nearly as common as processing
interfaces.  Because they introduce extra complexity, they should only
be used to define a data type when it is not practical to use a single
class.  Currently, NLTK defines data type interfaces for two domains:
probability distributions (\texttt{ProbDistI} and
\texttt{ConditionalProbDistI}); and classification features
(\texttt{FeatureDetectorI}, \texttt{FeatureDetectorListI}, and
\texttt{FeatureValueListI}).  The classification feature data type
interfaces may be removed or revised when the classification system is
redesigned.

%============= DEFINING NEW INTERFACES =============%
\section{Defining New Interfaces}

When defining a new interface for NLTK, be sure to keep the following
questions and guidelines in mind:

\begin{itemize}
\item Is it a data interface or a processing interface?  Note that in
  some cases, this will be a design decision (e.g., a ``rule'' class
  could be reformulated as a ``decider'' class).
\item For processing interfaces, which input/output class does the
  interface belong to?
\item For processing interfaces, which function class does the
  interface belong to?
  \begin{itemize}
  \item If you're trying to decide whether a processing interface
    should be an extender or a transformer, think about typical uses
    of the interface's output.  Is the new information usually used in
    conjunction with the original information, or in place of it?
  \end{itemize}
\item What auxilliary data types are needed to define the interface?
\item Interfaces should be as general as is practical, and should not
  introduce any unnecessary restrictions on tasks.  For example,
  interfaces should try not to constrain what objects can be used for
  token types.
\item Interfaces should be named using standard terms, where possible.
  Interface names (including interface method names) should not
  conflict with other names that are already used by the toolkit.
  Neutral names are preferred over domain-specific names.
\item Think about what tasks are likely to be used to generate input
  for the interface, or to process its output; and make sure that the
  interface can be easily combined with the interfaces for those other
  tasks.
\item Is the new interface just a special case of an existing
  interface?  For example, there is probably no need for a ``word
  sense disambiguation'' interface, since word sense disambiguation is
  just a specific domain for classification.
\item Is the nomenclature used to define the interface internally
  consistant?  Is it consistant with the literature?\footnote{An
    example of inconsistant nomenclature that is currently in NLTK is
    the fact that classifiers assign ``labels'' instead of
    ``classifications.''}

\end{itemize}

%============= CURRENT INTERFACES =============%
\section{Current Interfaces}
\subsection{Processing Interfaces}
The following table lists the processing interfaces that are currently
defined by NLTK, and lists the I/O and functional classifications for
each interface.

\vspace{3mm}\noindent
\begin{tabular}{|p{.5\textwidth}|l|l|}
\hline
  \textbf{Interface} & \textbf{I/O Classification} &
  \textbf{Function} \\
\hline
\texttt{TokenizerI}

\textit{\qquad Divides a text into a list of tokens}
& token list generator & transformer \\
\hline
\texttt{TaggerI}

\textit{\qquad Adds a tag to each token in a list}
& token list map & extender \\
\hline
\texttt{StemmerI}

\textit{\qquad Strips morphological affixes}
& token list map & selector \\
\hline
\texttt{ParserI}

\textit{\qquad Identifies token constituants}
& token list collector & extender \\
\hline
\texttt{ChunkParserI}

\textit{\qquad Identifies non-overlapping groups of tokens}
& token list collector & extender \\
\hline
\texttt{ProbabilisticParserI}

\textit{\qquad Identifies token constituants}
& token list collector & extender \\
\hline
\texttt{ClassifierI}

\textit{\qquad Adds a label to a text}
& token list map & extender \\
\hline
\texttt{ClassifierTrainerI}

\textit{\qquad Creates a new classifier}
& processing class factory & transformer \\
% Leave this out for now: \texttt{FeatureSelectorI} & 
\hline
\texttt{CorpusReaderI}

\textit{\qquad Parses and reads standard corpora}
& generator & (special)\\
\hline
\end{tabular}

\subsection{Data Interfaces}

The following list desribes the data interfaces that are currently
defined by NLTK.

\begin{enumerate}
\item \textbf{Probability Distributions}
  \begin{itemize}
  \item \textbf{\texttt{ProbDistI}} is an interface for probability
    distributions.  Probability distributions are typically derived
    from frequency distributions (e.g., \texttt{MLEProbDist}), but can
    also be created analytically (e.g., \texttt{UniformProbDist}).
  \item \textbf{\texttt{ConditionalProbDistI}} is an interface for
    conditional probability distributions.  Probability distributions
    that are derived from frequency distributions are typically
    created using the \texttt{ConditionalProbDist} implementation; but
    implementations for analytic distributions (eg gaussians) may also
    be added.
  \end{itemize}
\newpage
\item \textbf{Chart Parsers}
  \begin{itemize}
  \item \textbf{\texttt{ChartRuleI}} is an interface for rules that
    controls what edges should be added to a chart in a
    \texttt{ChartParser}.
  \item \textbf{\texttt{IncrementalChartRuleI}} is an interface for
    rules that controls what edges should be added to a chart in an
    \texttt{IncrementalChartParser}.
  \item \textbf{\texttt{EdgeI}} is an interface for edges in a chart.
    This class is used as a superclass for \texttt{ProductionEdge} and
    \texttt{TokenEdge}.\footnote{Since it is very unlikely to ever get
      other implementations, it might make more sense to change this
      class from an interface to an abstract superclass.}
  \end{itemize}
\item \textbf{Classificaiton Features}\footnote{The classification
    system is being redesigned, and these interfaces are likely to be
    removed.}
  \begin{itemize}
  \item \textbf{\texttt{FeatureDetectorI}} is an interface for
    ``feature detectors,'' which map labeled texts to values.
  \item \textbf{\texttt{FeatureValueListI}} is an interface for sparse
    encodings of lists of feature values.
  \item \textbf{\texttt{FeatureDetectorListI}} is an interface for
    sparse encodings of lists of feature detectors.
  \end{itemize}
\end{enumerate}

\section{Open Questions}

\subsection{Extenders vs Transformers}

For a given task, how should we decide whether its interface should be
an extender or a transformer?  Some examples of the alternatives for
current interfaces include:

\vspace{3mm}
\begin{tabular}{|l|ll|}
  \hline
  Task & Extender Interface Output& Transformer Interface Output\\
  \hline
  Tagging &
    list of Token($\langle$word, tag$\rangle$, loc) &
    list of Token(tag, loc) \\
  Classifying & 
    Token($\langle$text, label$\rangle$, loc) &
    Token(label, loc) \\
  Stemming & 
    Token($\langle$word, stem$\rangle$, loc) &
    Token(stem, loc) \\
  \hline
\end{tabular}
\vspace{3mm}

Note that the output would always be a token; at issue is what
information would be included in the token's type.  Note also that
an extender can be converted to a transformer by discarding the
input word/text; and that a transformer can be converted to an
extender by aligning its output with its input (using locations),
and ``gluing'' the input word/text to the output.

Currently, we're using a mix of extenders and transformers, and
attempting to choose the class that will most likely be useful to
users.  However, the decision between extenders and transformers can
be fairly arbitrary in some cases, and might cause confusion.

\newpage
One alternative would be to eliminate extenders, and make everything
a transformer.  The user would ``glue'' information from different
stages together, where necessary.  E.g., a user who wanted to tag
some text and send it through a lexicalized parser could run:

\begin{alltt}
  >>> toks = my\_tokenizer.tokenize(text)
  >>> tags = my\_tagger.tag(toks)
  >>> tagged\_toks = [Token(TaggedType(tok.type(), tag.type()), tok.loc())
  ...                for (tok, tag) in zip(toks, tags)]
  >>> tree = my\_parser.parse(tagged\_toks)
\end{alltt}

Another option would be to eliminate transformers, and make
everything an extender.  The user would discard unnecessary
information, if they don't want it.  For example, a user who wanted
to tag and parse text with an unlexicalized parser could run:

\begin{alltt}
  >>> toks = my\_tokenizer.tokenize(text)
  >>> tagged\_toks = my\_tagger.tag(toks)
  >>> tags = [Token(tok.type().tag(), tok.loc()) for tok in tagged\_toks]
  >>> tree = my\_parser.parse(tags)
  >>> lexicalized\_tree = tree.lexicalize(toks)\footnote{This function\
is hypothetical.}
\end{alltt}

So, should we continue to use a mix of transformers and extenders?  Or
should we use one or the other exclusively?

\subsection{Accessing Context}

One issue with the token-based interfaces is that it can be difficult
to access a token's context.  For example, to disambiguate the sense
of a word, we might want to know what words are near it; but if a
processing class is only given the token, then it can't access this
information.

There are several ways we might address this issue, but each has its
own problems.

\begin{itemize}
\item Add pointers from tokens to the texts that they came from.
  Disadvantages: maintaining back pointers can be dangerous and error
  prone.  For example, if the user modifies the text, then the back
  pointers may become invalid.  This could potentially be solved by
  making all texts immutable (e.g., use tuples instead of lists for
  tokenized text); but that seems somewhat draconian.

\item Add an explicit feature-collection step, which acts on complete
  texts, and maps each token to a set of features that describes the
  token (e.g., words in its context).  Interfaces that need access to
  context would then act on these feature tokens, rather than on the
  original tokens.  Disadvantages: added complexity can be complex.
  Sometimes ``context'' is not best thought of as a set of features.

\item Add an extra ``text'' parameter to most interfaces, which can be
  used to look up the context of a token.  Disadvantages: finding the
  context for a token is likely to require linear search, and so would
  be very slow; interfaces become less simple and clean.
\end{itemize}

\end{document}

### 2004-04-04 technical/pycon2004/pycon2004.tex
- Initial import: Pycon 2004 presentation
svn/trunk@1689

\documentclass{article}
\usepackage{fullpage, alltt, epsfig, boxedminipage, url}
\twocolumn
\sloppy
\def\nogap{\setlength\itemsep{.0in}\setlength{\parskip}{0in}\setlength{\topsep}{0in}}

\begin{document}
\title{NLTK: Building a Pedagogical Toolkit in Python}
\author{\textbf{Edward Loper}\\
Department of Computer and Information Science\\
University of Pennsylvania, Philadelphia, PA 19104-6389, USA}
\date{}
\maketitle

\begin{abstract}

  Teachers of computational classes are faced with the challenge of
  setting up a practical programming component for student assignments
  and projects.  One solution is to construct a broad-coverage
  toolkit, eliminating overhead and leaving students free to think
  about the subject matter rather than low-level programming details.
  This article discusses the criteria and requirements that should
  guide the development of a programming toolkit for courseware, and
  explain how those criteria affected the design and implementation of
  the Natural Language Toolkit (NLTK), a broad-coverage toolkit
  designed for use as computational linguistics courseware.

\end{abstract}

%========================= Introduction =========================
\section{Introduction}

Computational linguistics is the study of the application of
computational methods to processing and analyzing natural language.
In an introductory course on computational linguistics, a practical
programming component provides an important tool for teaching students
about these methods.  However, different computational linguistics
domains require a variety of different data structures and functions,
and a diverse range of topics need to be included in the syllabus.
Furthermore, many of the students come from linguistics backgrounds,
and lack strong programming skills.  Therefore, setting up an
easy-to-use practical component can be quite a challenge.

A widespread practice is to employ multiple programming languages,
where each language provides native data structures and functions that
are a good fit for the task at hand.  For example, a typical course
might use Prolog for parsing; Perl for corpus processing; and a
finite-state toolkit for morphological analysis.  This approach allows
the teacher to draw on a wide variety of existing tools, and to avoid
developing a lot of software infrastructure.

But unfortunately, this approach also requires that a significant
portion of the course be devoted to teaching programming languages
instead of computational linguistics.  Furthermore, many interesting
projects span a variety of domains, and would require that multiple
languages be bridged.  For example, a student project that involved
syntactic parsing of corpus data from a morphologically rich language
might involve all three of the languages mentioned above: Perl for
string processing; a finite state toolkit for morphological analysis;
and Prolog for parsing.

These same challenges face the teacher of any computational subject
that requires a strong practical component (such as computational
biology).  In this article, we discuss the issues involved in building
a broad-coverage pedagogical toolkit; and show how we addressed those
issues in the Natural Language Toolkit (NLTK), a Python-based
programming toolkit developed in conjunction with a course we have
taught at the University of Pennsylvania.

%========================= Applications =========================
\section{Applications}

Before designing a pedagogical toolkit, it is important to consider
the uses that it will be put to.  We divided the toolkit's applications
into three basic categories: assignments, in-class demonstrations, and
advanced projects.

\subsection{Assignments}

We wanted the toolkit to support homework assignments of varying
difficulty and scope.  In the simples assignments, students would use
and experiment with existing components.  The toolkit should provide
broad coverage, to allow assignments to be created for a wide variety
of topics.  Once students became more familiar with the toolkit, they
could be asked to make minor changes or extensions to existing
components.  Finally, students can be asked to create complete systems
by combining existing components.

\subsection{In Class Demonstrations}

In class graphical demonstrations can be powerful tools for explaining
concepts and algorithms.  Interactive demonstration tools can be used
to display relevant data structures, and to show the step-by-step
execution of important algorithms.  It should be possible to modify
both data structures and control flow during the demonstration, in
response to questions from the class.  Finally, any graphical
demonstration tools should be accessible to the students.  This allows
students to experiment at home with the algorithms that they have seen
presented in class.

\subsection{Advanced Projects}

The toolkit should provide students with a flexible framework for
advanced projects.  Typical projects might involve implementing a new
algorithm, developing a new component, or adding support for a new
task.

%========================= Requirements =========================
\section{Requirements}

Based on these three applications, we defined the following set of
requirements for the toolkit's design, listed in decreasing of 
importance.

\paragraph{\textit{Ease of Use.}} The primary purpose of the toolkit is
to allow students to concentrate on building natural language
processing (NLP) systems.  The more time students must spend learning
to use the toolkit, the less useful it is.

\paragraph{\textit{Consistency.}} The toolkit should use consistent data
structures and interfaces.

\paragraph{\textit{Extensibility.}} The toolkit should easily
accommodate new components, whether those components replicate or
extend the toolkit's existing functionality.  The toolkit should be
structured in such a way that it is obvious where new extensions would
fit into the toolkit's infrastructure.

\paragraph{\textit{Documentation.}} The toolkit, its data structures,
and its implementation all need to be carefully and thoroughly
documented.  All nomenclature must be carefully chosen and
consistently used.

\paragraph{\textit{Simplicity.}} The toolkit should structure the
complexities of building NLP systems, not hide them.  Therefore, each
class defined by the toolkit should be simple enough that a student
could implement it by the time they finish an introductory course in
computational linguistics.

\paragraph{\textit{Modularity.}} The interaction between different
components of the toolkit should be kept to a minimum, using simple,
well-defined interfaces.  In particular, it should be possible to
complete individual projects using small parts of the toolkit, without
worrying about how they interact with the rest of the toolkit.  This
allows students to learn how to use the toolkit incrementally
throughout a course.  Modularity also makes it easier to change and
extend the toolkit.

\subsection{Non-Requirements}

It is equally important to specify properties that are \emph{not}
expected of the toolkit:

\paragraph{\textit{Comprehensiveness.}} Although the toolkit should
provide broad coverage, it is not intended to provide a comprehensive
set of tools.  Indeed, there should be a wide variety of ways in which
students can extend the toolkit.

\paragraph{\textit{Efficiency.}} The toolkit does not need to be highly
optimized for runtime performance.  However, it should be efficient
enough that students can use their NLP systems to perform real tasks.

\paragraph{\textit{Cleverness.}} Clear designs and implementations are
far preferable to ingenious yet indecipherable ones.

%========================= Why Python? =========================
\section{Why Python?}

The first step in designing a pedagogical toolkit is choosing a
suitable programming language.  Python is especially well suited to
the task, for a number of reasons:

\begin{itemize}
\item Python offers a shallow learning curve; it was designed to be
  easily learnt by children \cite{cp4e}.  This ensures that students
  without a strong computer science background can quickly get up to
  speed.

\item Python code is exceptionally readable, with transparent syntax
  and semantics; it has been praised as ``executable pseudocode.''
  Examples are therefore easy for students to follow.  Furthermore,
  students can learn by looking at the toolkit's source code, and not
  just by using it.

\item As an interpreted language, Python is suitable for interactive
  exploration.  Students can experiment with the toolkit, and get
  immediate feedback.

\item Python's light-weight object oriented system makes it easy to
  encapsulate data and methods in classes, without forcing students to
  use them where they're not necessary.

\item Python's recently added generator syntax makes it easy to create
  interactive implementations of algorithms.  These interactive
  implementations can be used to ``step through'' the algorithm,
  examining how its state changes as the algorithm progresses.

\item Python's extensive standard library provides a great deal of
  power, when needed.  For example, the \texttt{Numeric} library can
  be used to implement computationally intensive algorithms that would
  otherwise be too slow if implemented in Python.
\end{itemize}

%========================= Design =========================
\section{Design}

NLTK is implemented as a large collection of minimally interdependent
modules, organized into a shallow package hierarchy.  A set of core
modules defines basic data types that are used throughout the toolkit.  
The remaining modules are \emph{task modules}, each devoted to an
individual natural language processing task.  For example, the
\texttt{nltk.parser} package encompasses to the task of
\emph{parsing}, or deriving the syntactic structure of a sentence; and
the \texttt{nltk.tokenizer} module is devoted to the task of
\emph{tokenizing}, or dividing a text into its constituent parts.

\subsection{Core Data Types}

\subsubsection{Token}

To maximize interoperability between different modules, we use a
single class to encode information about natural language texts -- the
\texttt{Token} class.  Each \texttt{Token} instance represents a
single unit of text, such as a word, sentence, or document; and is
defined by (partial) mapping from property names to values.  For
example, the \texttt{TEXT} property is used to encode a token's text
content; and the \texttt{TAG} property is used to encode a word
token's part-of-speech tag.  The \texttt{LOC} property is used to
specify the location of a token in its containing text.  This location
can be used to decide whether two \texttt{Token} objects refer to the
same piece of text or not.

Natural language processing tasks are formulated as transformations on
\texttt{Tokens}.  In particular, each processing task takes a token,
and extends it to include new information.  Typically, these
modifications are \emph{monotonic}; in other words, new information is
added but existing information is not deleted or modified.  Thus,
tokens serve as a sort of \emph{blackboard}, where information about a
piece of text can be incrementally built up.  This architecture
contrasts with the more typical \emph{pipeline} architecture, where
each processing task's output discards its input information.  We
chose the ``blackboard'' approach over the ``pipeline'' approach
because it allows more flexibility when combining tasks into a single
system.  In particular:

\begin{itemize}
\item Since no information is discarded, and all information is
  accessed the same way, students don't need to worry about
  ``threading'' information through the system to the tasks that need
  it.

\item Because every processing tasks has the same basic interface, the
  order in which processing tasks are applied can be easily modified.

\item Monotonic modification allows tasks to transform to a piece of a
  structure without copying the remaining structure.

\item Monotonic modification ensures that the information seen by
  different tasks is consistent.
\end{itemize}

\noindent The primary disadvantage of the ``blackboard'' approach is
that it can be inefficient.  In particular, token properties are not
deleted even if they will not be used.  But for a pedagogical toolkit,
flexibility is more important than efficiency; and if necessary, the
unused properties can be explicitly deleted.

\subsubsection{Token Subclasses}
It is often useful to use a token as a key in a dictionary or an
element in a set.  NLTK therefore defines a hashable and immutable
token subclass, \texttt{FrozenToken}.

A number of other subclasses are used to support special token types.
For example, the \texttt{TreeToken} class adds methods for performing
tree manipulation operations; and the \texttt{ParentedTreeToken} class
automatically maintains consistent parent pointers for tree
structures.

\subsubsection{Other Core Data Types}

In addition to the \texttt{Token} class, NLTK defines a variety of
data types that are useful for language processing.  For example, the
\texttt{nltk.probability} module defines classes that encode frequency
distributions and probability distributions, including conditional
distributions and a variety of statistical smoothing techniques; the
\texttt{cfg} module defines classes for encoding context free grammars
and probabilistic context free grammars; and the \texttt{corpus}
module defines classes for reading and processing standard language
processing corpora.

\subsection{Task Modules}

Task modules define individual natural language processing tasks, such
as parsing and tokenizing.

\subsubsection{Task Classes}

Each processing task algorithm is encoded as a class.  For example,
the \texttt{ChartParser} and \texttt{Recursive\-Descent\-Parser} classes
each define a single algorithm for parsing a text.  Although it might
seem unintuitive at first, there are a number of advantages to using a
class (rather than a function) to encode an algorithm.  First, all
algorithm-specific options can be passed to the constructor, allowing
a consistent interface for actually applying the algorithms.  Second,
a number of algorithms need to have their state initialized before
they can be used.  For example, the \texttt{NthOrderTagger} class must
be initialized by training on a tagged corpus before it can be used.
Finally, subclassing can be used to create specialized versions of a
given algorithm.

\subsubsection{Task Interfaces}

In order to provide a formal specification of the interface for a
given task, each task module defines an \emph{interface} for its task.
These interfaces are encoded interfaces as skeletal base classes that
use docstrings to specify each method's behavior.  Using epydoc, these
docstrings can be inherited by derived classes \cite{epydoc}.  Each
required method raises an \texttt{AssertionError}, since any valid
implementation of the interface must override it; and any optional
methods raise \texttt{Not\-Implemented\-Error}.  Interfaces are
distinguished by naming them with a trailing capital ``\texttt{I},''
such as \texttt{ParserI} and \texttt{TokenizerI}.

Each interface defines a single \emph{action method}, which actually
performs the task defined by the interface.  For example, the
\texttt{ParserI} interface defines the \texttt{parse} method; and the
\texttt{Tokenizer} interface defines the \texttt{tokenize} method.
When appropriate, an interface will also define \emph{extended action
  methods}, which provide variations on the basic action method.  For
example, the \texttt{ParserI} interface defines the \texttt{parse\_n}
method, which finds all possible structures for a given sentence; and
the \texttt{TokenizerI} interface defines the \texttt{xtokenize}
method, which outputs an iterator over subtokens instead of a list of
subtokens.

\subsection{In-Class Demonstrations}

We used the Tkinter toolkit \cite{tkinter} to create several tools
that can be used to demonstrate the algorithms implemented by NLTK.
We chose Tkinter because it is almost universally distributed with
Python; it is therefore possible for students to play with the
demonstration tools at home.  These GUI demonstrations are built on
top of the basic task implementations; and use generators to step
through the algorithm's operations.  The demonstration tools also
modification and manual control over the basic algorithm.  This lets
teachers to use the tools to highlight specific points about an
algorithm; and makes it easy for students to experiment with the
algorithm at home.  Figure \ref{fig:chart} shows the chart parsing
demo, which is used to illustrate a variety of different chart parsing
algorithms.

\begin{figure}
\epsfig{file=chart.eps, width=8cm}
\caption{The Chart Parsing Demo}
\label{fig:chart}
\end{figure}

\section{Conclusions}

NLTK is a broad-coverage toolkit that provides a simple, extensible,
uniform framework for assignments, projects, and class demonstrations.
It is well documented, easy to learn, and simple to use.

NLTK is unique among computational linguistics tools in its
combination of three factors.  First, it was deliberately designed as
courseware and gives pedagogical goals primary status.  Second, its
target audience consists of both linguists and computer scientists,
and it is accessible and challenging at many levels of prior
computational skill.  Finally, it is based on an easy-to-learn and
easy-to-read programming language supporting rapid development and
literate programming.

We plan to continue extending the breadth of materials covered by the
toolkit.  We also plan to increase the number of algorithms
implemented by some of the existing modules, such as the text
classification module.

NLTK is an open source project, and we welcome any contributions.
Readers who are interested in contributing to NLTK, or who have
suggestions for improvements, are encouraged to contact the author.

\section{Acknowledgements}

We are indebted to our students for feedback on the toolkit, and to
all the independent contributors who helped add new content to the
toolkit.  We are grateful to Mitch Marcus and the Department of
Computer and Information Science at the University of Pennsylvania for
sponsoring the original development of the toolkit.

\nocite{nltk}
\bibliographystyle{apalike}
\bibliography{pycon}

\end{document}

### 2004-04-27 technical/acl2004/acl04.tex
first commit of ACL demo submission
svn/trunk@1777

\documentclass[11pt]{article}
\usepackage{acl04}
\usepackage{times}
\usepackage{latexsym}
\usepackage{url,alltt,epsfig,boxedminipage}

% hyphenation control
\pretolerance 250
\tolerance 500
\hyphenpenalty 200
\exhyphenpenalty 100
\doublehyphendemerits 7500
\finalhyphendemerits 7500
\brokenpenalty 10000
\lefthyphenmin 3
\righthyphenmin 3
\widowpenalty 10000
\clubpenalty 10000
\displaywidowpenalty 10000
\looseness 1

\title{NLTK: The Natural Language Toolkit}

\author{
  Steven Bird \\
  Department of Computer Science \\
  \indent and Software Engineering \\
  University of Melbourne \\
  Victoria 3010, Australia \\  
  {\tt sb@csse.unimelb.edu.au}
\And
  Edward Loper\\
  Department of Computer \\
  \indent and Information Science \\
  University of Pennsylvania\\
  Philadelphia PA 19104-6389, USA\\
  {\tt edloper@gradient.cis.upenn.edu}
}

\date{}

\newenvironment{sv}{\scriptsize\begin{alltt}}{\end{alltt}\normalsize}

\begin{document}

\maketitle

\begin{abstract}
  The Natural Language Toolkit is a suite of program modules, data
  sets, tutorials and exercises, covering symbolic and statistical
  natural language processing.  NLTK is written in Python and
  distributed under the GPL open source license.  Over the past three
  years, NLTK has become popular in teaching and research.  We
  describe the toolkit and report on its current state of development.
\end{abstract}

%========================= Introduction =========================
\section{Introduction}

The Natural Language Toolkit (NLTK) was developed in conjunction with
a computational linguistics course at the University of Pennsylvania
in 2001 \cite{LoperBird02}.  It was designed with three pedagogical
applications in mind: assignments, demonstrations, and projects.

\textbf{Assignments.}
NLTK supports assignments of varying difficulty
and scope.  In the simplest assignments, students experiment with
existing components for a wide variety of NLP tasks.  As students
become more familiar with the toolkit, they can be asked to modify
existing components, or to create complete systems out of existing
components.

\textbf{Demonstrations.}
NLTK's interactive graphical demonstrations have proven to be very
useful for students learning NLP concepts.
The demonstrations give a step-by-step execution of important
algorithms, displaying the current state of key data structures.
A screenshot of the chart parsing demonstration is shown in Figure~\ref{fig:chart}.

\textbf{Projects.}  NLTK provides students with a flexible framework
for advanced projects.  Typical projects might involve implementing a
new algorithm, developing a new component, or adding support for a new
task.

We chose Python because it has a shallow learning curve, its syntax
and semantics are transparent, and it has good string-handling
functionality.  As an interpreted language, Python facilitates
interactive exploration.  As an object-oriented language, data and
methods can be encapsulated easily.  Python's extensive standard
library is useful makes numerical and graphical programming as easy as
it could be, and the recently added generator syntax makes it easy to
create interactive implementations of algorithms
\cite{Loper04,Rossum03intro,Rossum03ref}.

\begin{figure}[bth]
\epsfig{file=../pycon2004/chart.eps, width=\linewidth}
\caption{Interactive Chart Parsing Demonstration}
\label{fig:chart}
\end{figure}

%========================= Design =========================

\section{Design}

NLTK is implemented as a large collection of minimally interdependent
modules, organized into a shallow package hierarchy.  A set of core
modules defines basic data types that are used throughout the toolkit.  
The remaining modules are \emph{task modules}, each devoted to an
individual natural language processing task.  For example, the
\texttt{nltk.parser} package encompasses to the task of
\emph{parsing}, or deriving the syntactic structure of a sentence; and
the \texttt{nltk.tokenizer} module is devoted to the task of
\emph{tokenizing}, or dividing a text into its constituent parts.

\subsection{Tokens and other core data types}

To maximize interoperability between different modules, we use a
single class to encode information about natural language texts -- the
\texttt{Token} class.  Each \texttt{Token} instance represents a
single unit of text, such as a word, sentence, or document; and is
defined by (partial) mapping from property names to values.  For
example, the \texttt{TEXT} property is used to encode a token's text
content:

\begin{alltt}
\textbf{>>> from nltk.token import *}
\textbf{>>> Token(TEXT="Hello World!")}
<Hello World!>
\end{alltt}
%
The \texttt{TAG} property is used to encode a token's part-of-speech
tag:

\begin{alltt}
\textbf{>>> Token(TEXT="python", TAG="NN")}
<python/NN>
\end{alltt}
%
The \texttt{SUBTOKENS} property is used to store a tokenized text:

\begin{alltt}
\textbf{>>> from nltk.tokenizer import *}
\textbf{>>> token = Token(TEXT="Hello World!")}
\textbf{>>> WSTokenizer().tokenize(token)} \textit{\# whitespace tokenizer}
\textbf{>>> print token['SUBTOKENS'])}
[<Hello>, <World!>]
\end{alltt}
%
In a similar fashion, other language processing tasks such as
word-sense disambiguation, chunking and parsing all add properties to
the \texttt{Token} data structure.

In general, language processing tasks are formulated as
annotations and transformations involving \texttt{Tokens}.  In
particular, each processing task takes a token, and extends it to
include new information.  Typically, these modifications are
\emph{monotonic}; in other words, new information is added but
existing information is not deleted or modified.  Thus, tokens serve
as a \emph{blackboard}, where information about a piece of text is
collated.  This architecture contrasts with the more typical
\emph{pipeline} architecture, where each processing task's output
discards its input information.  We chose the ``blackboard'' approach
over the ``pipeline'' approach because it allows more flexibility when
combining tasks into a single system.

In addition to the \texttt{Token} class and its derivatives, NLTK
defines a variety of other data types.  For instance, the
\texttt{nltk.probability} module defines classes for a variety of
probability distributions and statistical smoothing techniques; and
the \texttt{cfg} module defines classes for encoding context free
grammars and probabilistic context free grammars.

\subsection{The corpus module}

Many language processing tasks must be developed and tested using
annotated data sets, or corpora.  Several such corpora are distributed
with NLTK, as listed in Table~\ref{tab:data}.  The NLTK
\texttt{corpus} module defines classes for reading and processing
many of these corpora.  The following code fragment illustrates
how the Brown Corpus is accessed.

\begin{alltt}
\textbf{>>> from nltk.corpus import brown}
\textbf{>>> brown.items()}
['ca01', 'ca02', 'ca03', 'ca04', 'ca05', 'ca06',
'ca07', 'ca08', 'ca09', 'ca10', 'ca11', 'ca12', ... ]
\textbf{>>> brown.tokenize('ca01')}
<[<The/at>, <Fulton/np-tl>, <County/nn-tl>,
<Grand/jj-tl>, <Jury/nn-tl>, <said/vbd>, <Friday/nr>,
<an/at>, <investigation/nn>, <of/in>, <Atlanta's/np$>,
<recent/jj>, <primary/nn>, <election/nn>,
<produced/vbd>, <``/``>, <no/at>, <evidence/nn>,
<''/''>, <that/cs>, <any/dti>, <irregularities/nns>,
<took/vbd>, <place/nn>, <./.>, ...]>
\end{alltt}
%
A selection of 5\% of the Penn Treebank corpus is included with
NLTK, and it can be accessed as follows:

\begin{alltt}
\textbf{>>> from nltk.corpus import treebank}
\textbf{>>> treebank.items()}
['wsj_0001.prd', 'wsj_0002.prd', 'wsj_0003.prd', ...]
\textbf{>>> treebank.tokenize('parsed/wsj_0001.prd')}
[ (S: (NP-SBJ: (NP: <Pierre> <Vinken>) ...)
        (VP: will ...)),
  (S: (NP-SBJ: <Mr.> <Vinken>)
        (VP: <is> ...)) ]
\end{alltt}
%
[WHY CAN'T I PP() OR DRAW() THE RESULTING OBJECT?]

\begin{table*}
\small\noindent
\begin{boxedminipage}{\linewidth}
\begin{tabular}{llll}
\textbf{Corpus} &
\textbf{Contents and Wordcount} &
\textbf{Example Application} \\

20 Newsgroups (selection) &
3 newsgroups, 4000 posts, 780kw &
text classification \\

Brown Corpus &
15 genres, 1.15Mw, tagged &
training and testing taggers, text classification \\

CoNLL 2000 Chunking Data &
270kw, tagged and chunked &
training and testing chunk parsers \\

Project Gutenberg (selection) &
14 texts, 1.7Mw &
text classification, language modelling \\

NIST 1999 IEER (selection) &
63kw, named-entity markup &
training and testing named-entity recognizers \\

Levin Verb Index &
3k verbs with Levin classes &
parser development \\

Names Corpus &
8k male and female names &
text classification \\

PP Attachment Corpus &
28k prepositional phrases, tagged &
parser development \\

Roget's Thesaurus &
200kw, formatted text &
word-sense disambiguation \\

SEMCOR &
880kw, POS and sense tagged &
word-sense disambiguation \\

SENSEVAL 2 Corpus &
600kw, POS and sense tagged &
word-sense disambiguation \\

Stopwords Corpus &
2,400 stopwords for 11 lgs &
text retrieval \\

Penn Treebank (sample) &
40kw, tagged and parsed &
parser development \\

Wordnet 1.7 &
180kw in a semantic network &
word-sense disambiguation, NL understanding \\

Wordlist Corpus &
960kw and 20k affixes for 8 lgs &
spell checking
 \\
\end{tabular}
\caption{Corpora and Corpus Samples Distributed with NLTK}\label{tab:data}
\end{boxedminipage}
\end{table*}

\subsection{Processing modules}

Each language processing algorithm is implemented as a class.  For
example, the \texttt{ChartParser} and
\texttt{Recursive\-Descent\-Parser} classes each define a single
algorithm for parsing a text.  We implement language processing
algorithms using classes instead of functions for three reasons.
First, all algorithm-specific options can be passed to the
constructor, allowing a consistent interface for applying the
algorithms.  Second, a number of algorithms need to have their state
initialized before they can be used.  For example, the
\texttt{NthOrderTagger} class must be initialized by training on a
tagged corpus before it can be used.  Third, subclassing can be used
to create specialized versions of a given algorithm.

In order to provide a formal specification of the interface for a
given task, each processing module defines an \emph{interface} for its task.
Interface classes are distinguished by naming them with a trailing
capital ``\texttt{I},'' such as \texttt{ParserI} and
\texttt{TokenizerI}.
Each interface defines a single \emph{action method}, which actually
performs the task defined by the interface.  For example, the
\texttt{ParserI} interface defines the \texttt{parse} method; and the
\texttt{Tokenizer} interface defines the \texttt{tokenize} method.
When appropriate, an interface defines \emph{extended action
  methods}, which provide variations on the basic action method.  For
example, the \texttt{ParserI} interface defines the \texttt{parse\_n}
method, which finds at most $n$ parses for a given sentence; and
the \texttt{TokenizerI} interface defines the \texttt{xtokenize}
method, which outputs an iterator over subtokens instead of a list of
subtokens.

\subsection{Documentation}

Three different types of documentation are available.  Tutorials
explain how to use the toolkit, with detailed worked examples.  The
API documentation describes every module, interface, class, method,
function, and variable in the toolkit.  Technical reports explain and
justify the toolkit's design and implementation.  All are available
from \url{http://nltk.sourceforge.net/docs.html}.

\section{Installing NLTK}

NLTK is available from \url{nltk.sourceforge.net}, and is packaged for
easy installation under Unix, Mac OS X and Windows.  The full
distribution consists of four packages: the Python source code
(\texttt{nltk}); the corpora (\texttt{nltk-data}); the documentation
(\texttt{nltk-docs}); and third-party contributions
(\texttt{nltk-contrib}).  Before installing NLTK, it is necessary to
install Python version 2.3 or later, available from
\url{www.python.org}.  Full installation instructions and a quick
start guide are available from the NLTK homepage.

As soon as NLTK is installed, users can run the demonstrations.  On
Windows, the demonstrations can be run by double-clicking on their
Python source files.  Alternatively, from the Python interpreter, this
can be done as follows:

\begin{alltt}
\textbf{>>> import nltk.draw.rdparser}
\textbf{>>> nltk.draw.rdparser.demo()}
\textbf{>>> nltk.draw.srparser.demo()}
\textbf{>>> nltk.draw.chart.demo()}
\end{alltt}

\section{Using and contributing to NLTK}

\begin{table}[bth]
\small\noindent
\begin{boxedminipage}{\linewidth}
\begin{tabular}{l}
University of Pennsylvania, USA \\
\hspace{2ex}
\textit{Introduction to Computational Linguistics} \\[.5ex]

University of Melbourne, Australia \\
\hspace{2ex}
\textit{Human Language Technology} \\[.5ex]

University of Pittsburgh, USA \\
\hspace{2ex}
\textit{Artificial Intelligence Application Development} \\[.5ex]

Simon Fraser University, Canada \\
\hspace{2ex}
\textit{Computational Linguistics} \\[.5ex]

Macquarie University, Australia \\
\hspace{2ex}
\textit{Intelligent Text Processing} \\[.5ex]

University of Edinburgh, UK \\
\hspace{2ex}
\textit{Introduction to Computational Linguistics} \\[.5ex]

National Autonomous University of Mexico, Mexico \\
\hspace{2ex}
\textit{Introduction to Natural Language Processing in Python} \\[.5ex]

University of Magdeburg, Germany \\
\hspace{2ex}
\textit{Natural Language Systems} \\[.5ex]

Massachusetts Institute of Technology, USA \\
\hspace{2ex}
\textit{Natural Language Processing} \\[.5ex]

University of Colorado, USA \\
\hspace{2ex}
\textit{Natural Language Processing} \\[.5ex]

University of Amsterdam, Netherlands \\
\hspace{2ex}
\textit{Language Processing and Information Access} \\[.5ex]

Graz University of Technology, Austria \\
\hspace{2ex}
\textit{Information Search and Retrieval} \\[.5ex]

\end{tabular}
\caption{University Courses using NLTK}\label{tab:courses}
\end{boxedminipage}
\end{table}

NLTK has used at the University of Pennsylvania since 2001, and has
subsequently been adopted by several NLP courses at other
universities, including those listed in Table~\ref{tab:courses}.

Third party contributions to NLTK include: Brill tagger (Maloof),
hidden Markov model tagger (Cohn), GPSG-style feature-based grammar
and parser (Speer, Berwick), Kimmo finite-state morphological analyzer
(de Marcken, Yankama, Berwick) and decision list and decision tree
classifiers (Cohn).

NLTK is an open source project, and we welcome any contributions.
There are several ways to contribute: users can report bugs, suggest
features, or contribute patches on Sourceforge; users can participate
in discussions on the NLTK-Devel mailing list or in the NLTK public
forums; and users can submit their own NLTK-based projects for
inclusion in the nltk\_contrib directory.  New code modules that are
relevant, substantial, original and well-documented will be considered
for inclusion in NLTK proper.  Potential contributors should note that
all source code is distributed under the Gnu General Public License,
and all documentation is distributed under a Creative Commons
non-commercial license.  Thus they can be confident that their
work will remain freely available to all.  Further information about
contributing to NLTK is available at \url{http://nltk.sourceforge.net/contrib.html}.

\section{Conclusion}

NLTK is a broad-coverage toolkit that provides a simple, extensible,
uniform framework for assignments, demonstrations and projects.  It is
thoroughly documented, easy to learn, and simple to use.  Readers who
would like to receive occasional announcements about NLTK are
encouraged to sign up for the low-volume, moderated mailing list
\textit{NLTK-Announce}.

\section{Acknowledgements}

We are indebted to our students for feedback on the toolkit, and to
many independent contributors who helped add new content.

\bibliographystyle{acl}
\bibliography{nltk}

\end{document}

### 2004-04-30 technical/acl2004/acl04.tex
- Minor edits (noded by [el] comments) - Made code samples \small
svn/trunk@1780

@@
 and scope.  In the simplest assignments, students experiment with
-existing components for a wide variety of NLP tasks.  As students
+existing components to perform a wide variety of NLP tasks.  As students
 become more familiar with the toolkit, they can be asked to modify
@@
 interactive exploration.  As an object-oriented language, data and
-methods can be encapsulated easily.  Python's extensive standard
-library is useful makes numerical and graphical programming as easy as
-it could be, and the recently added generator syntax makes it easy to
-create interactive implementations of algorithms
-\cite{Loper04,Rossum03intro,Rossum03ref}.
+methods can be encapsulated easily.  Python comes with an extensive
+standard library, including tools for graphical programming and
+intensive numerical processing.  And the recently added generator
+syntax makes it easy to create interactive implementations of
+algorithms \cite{Loper04,Rossum03intro,Rossum03ref}.
 
@@
 \epsfig{file=../pycon2004/chart.eps, width=\linewidth}
+% [el] why the ``../pycon2004'' in the file name?
 \caption{Interactive Chart Parsing Demonstration}
@@
 
+% [el] I got rid of the package/module distinction here, because I 
+% think it'll just be confusing to this audience.  So let's just
+% call everything a module (and say we have a module hierarchy).
 NLTK is implemented as a large collection of minimally interdependent
-modules, organized into a shallow package hierarchy.  A set of core
+modules, organized into a shallow hierarchy.  A set of core
 modules defines basic data types that are used throughout the toolkit.
@@
 individual natural language processing task.  For example, the
-\texttt{nltk.parser} package encompasses to the task of
+\texttt{nltk.parser} module encompasses to the task of
 \emph{parsing}, or deriving the syntactic structure of a sentence; and
@@
 single unit of text, such as a word, sentence, or document; and is
-defined by (partial) mapping from property names to values.  For
+defined by a (partial) mapping from property names to values.  For
 example, the \texttt{TEXT} property is used to encode a token's text
@@
 
-\begin{alltt}
+\begin{alltt}\small
 \textbf{>>> from nltk.token import *}
@@
 
-\begin{alltt}
+\begin{alltt}\small
 \textbf{>>> Token(TEXT="python", TAG="NN")}
@@
 
-\begin{alltt}
+% [el] I'm assuming that 1.4 will be released by ACL, so use
+% WhitespaceTokenizer instead of WSTokenizer.
+\begin{alltt}\small
 \textbf{>>> from nltk.tokenizer import *}
 \textbf{>>> token = Token(TEXT="Hello World!")}
-\textbf{>>> WSTokenizer().tokenize(token)} \textit{\# whitespace tokenizer}
+\textbf{>>> WhitespaceTokenizer().tokenize(token)}
 \textbf{>>> print token['SUBTOKENS'])}
@@
 
-\begin{alltt}
+% [el] Changed to show off CorpusReader.groups().
+\begin{alltt}\small
 \textbf{>>> from nltk.corpus import brown}
-\textbf{>>> brown.items()}
-['ca01', 'ca02', 'ca03', 'ca04', 'ca05', 'ca06',
-'ca07', 'ca08', 'ca09', 'ca10', 'ca11', 'ca12', ... ]
-\textbf{>>> brown.tokenize('ca01')}
-<[<The/at>, <Fulton/np-tl>, <County/nn-tl>,
-<Grand/jj-tl>, <Jury/nn-tl>, <said/vbd>, <Friday/nr>,
-<an/at>, <investigation/nn>, <of/in>, <Atlanta's/np$>,
-<recent/jj>, <primary/nn>, <election/nn>,
-<produced/vbd>, <``/``>, <no/at>, <evidence/nn>,
-<''/''>, <that/cs>, <any/dti>, <irregularities/nns>,
-<took/vbd>, <place/nn>, <./.>, ...]>
+\textbf{>>> brown.groups()}
+['skill and hobbies', 'popular lore', 
+'humor', 'fiction: mystery', ...]
+\textbf{>>> brown.items('humor')}
+('cr01', 'cr02', 'cr03', 'cr04', 'cr05',
+'cr06', 'cr07', 'cr08', 'cr09')
+\textbf{>>> brown.tokenize('cr01')}
+<[<It/pps>, <was/bedz>, <among/in>,
+<these/dts>, <that/cs>, <Hinkle/np>,
+<identified/vbd>, <a/at>, ...]>
 \end{alltt}
@@
 
-\begin{alltt}
+% [el] Changed to show groups() and TreeToken.pp()
+% [el] This example wraps over a page at a weird place.  When
+% we've made all the other changes, maybe manually insert a
+% page break at a better location?
+\begin{alltt}\small
 \textbf{>>> from nltk.corpus import treebank}
-\textbf{>>> treebank.items()}
-['wsj_0001.prd', 'wsj_0002.prd', 'wsj_0003.prd', ...]
-\textbf{>>> treebank.tokenize('parsed/wsj_0001.prd')}
-[ (S: (NP-SBJ: (NP: <Pierre> <Vinken>) ...)
-        (VP: will ...)),
-  (S: (NP-SBJ: <Mr.> <Vinken>)
-        (VP: <is> ...)) ]
+\textbf{>>> treebank.groups()}
+('raw', 'tagged', 'parsed', 'merged')
+\textbf{>>> treebank.items('parsed')}
+['wsj_0001.prd', 'wsj_0002.prd', ...]
+\textbf{>>> item = 'parsed/wsj_0001.prd'}
+\textbf{>>> sentences = treebank.tokenize(item)}
+\textbf{>>> for sent in sentences['SUBTOKENS']:}
+\textbf{...     print sent.pp()} \emph{# pretty-print}
+(S:
+  (NP-SBJ:
+    (NP: <Pierre> <Vinken>)
+    ...
 \end{alltt}
 %
-[WHY CAN'T I PP() OR DRAW() THE RESULTING OBJECT?]
 
@@
 
+% [el] Changed the url here and elsewhere from nltk.sourceforge.net
+% to nltk.sf.net, because it fits better on small columns (and because 
+% it brought the length back down to 4 pages :) ).
 Three different types of documentation are available.  Tutorials
@@
 justify the toolkit's design and implementation.  All are available
-from \url{http://nltk.sourceforge.net/docs.html}.
+from \url{http://nltk.sf.net/docs.html}.
 
@@
 
-NLTK is available from \url{nltk.sourceforge.net}, and is packaged for
+NLTK is available from \url{nltk.sf.net}, and is packaged for
 easy installation under Unix, Mac OS X and Windows.  The full
@@
 
+% [el] We really should make a better interface for running demos.
+% But I don't think we can have that figured out by the deadline
+% for this paper.
 As soon as NLTK is installed, users can run the demonstrations.  On
@@
 
-\begin{alltt}
+\begin{alltt} \small
 \textbf{>>> import nltk.draw.rdparser}
@@
 features, or contribute patches on Sourceforge; users can participate
-in discussions on the NLTK-Devel mailing list or in the NLTK public
+% [el] Add a \cite with a url, or a url inline?  The url is:
+% http://lists.sourceforge.net/lists/listinfo/nltk-devel
+in discussions on the \textit{NLTK-Devel} mailing list or in the NLTK public
 forums; and users can submit their own NLTK-based projects for
@@
 for inclusion in NLTK proper.  Potential contributors should note that
+% [el] Should it be GNU instead of Gnu?
 all source code is distributed under the Gnu General Public License,
@@
 work will remain freely available to all.  Further information about
-contributing to NLTK is available at \url{http://nltk.sourceforge.net/contrib.html}.
+contributing to NLTK is available at \url{http://nltk.sf.net/contrib.html}.
 
@@
 encouraged to sign up for the low-volume, moderated mailing list
+% [el] Add a \cite with a url, or a url inline?  The url is:
+% http://lists.sourceforge.net/lists/listinfo/nltk-announce
 \textit{NLTK-Announce}.

### 2004-05-01 technical/acl2004/acl04.tex
further tweaking to respond to Edward's comments and fix some layout problems
svn/trunk@1781

@@
 \looseness 1
+
+\def\UrlFont{\tt\small}
+\def\object#1{\texttt{\small #1}}
 
@@
 
-\begin{abstract}
+\begin{abstract}\small
   The Natural Language Toolkit is a suite of program modules, data
@@
 \epsfig{file=../pycon2004/chart.eps, width=\linewidth}
-% [el] why the ``../pycon2004'' in the file name?
 \caption{Interactive Chart Parsing Demonstration}
@@
 
-%========================= Design =========================
-
 \section{Design}
 
-% [el] I got rid of the package/module distinction here, because I 
-% think it'll just be confusing to this audience.  So let's just
-% call everything a module (and say we have a module hierarchy).
 NLTK is implemented as a large collection of minimally interdependent
@@
 individual natural language processing task.  For example, the
-\texttt{nltk.parser} module encompasses to the task of
+\object{nltk.parser} module encompasses to the task of
 \emph{parsing}, or deriving the syntactic structure of a sentence; and
-the \texttt{nltk.tokenizer} module is devoted to the task of
+the \object{nltk.tokenizer} module is devoted to the task of
 \emph{tokenizing}, or dividing a text into its constituent parts.
@@
 
-To maximize interoperability between different modules, we use a
+To maximize interoperability between modules, we use a
 single class to encode information about natural language texts -- the
-\texttt{Token} class.  Each \texttt{Token} instance represents a
+\object{Token} class.  Each \object{Token} instance represents a
 single unit of text, such as a word, sentence, or document; and is
 defined by a (partial) mapping from property names to values.  For
-example, the \texttt{TEXT} property is used to encode a token's text
-content:
+example, the \object{TEXT} property is used to encode a token's text
+content:\footnote{Some code samples are specific to NLTK
+  version 1.4.}
 
@@
 %
-The \texttt{TAG} property is used to encode a token's part-of-speech
+The \object{TAG} property is used to encode a token's part-of-speech
 tag:
@@
 %
-The \texttt{SUBTOKENS} property is used to store a tokenized text:
+The \object{SUBTOKENS} property is used to store a tokenized text:
 
-% [el] I'm assuming that 1.4 will be released by ACL, so use
-% WhitespaceTokenizer instead of WSTokenizer.
 \begin{alltt}\small
@@
 word-sense disambiguation, chunking and parsing all add properties to
-the \texttt{Token} data structure.
+the \object{Token} data structure.
 
 In general, language processing tasks are formulated as
-annotations and transformations involving \texttt{Tokens}.  In
+annotations and transformations involving \object{Tokens}.  In
 particular, each processing task takes a token, and extends it to
@@
 
-In addition to the \texttt{Token} class and its derivatives, NLTK
+In addition to the \object{Token} class and its derivatives, NLTK
 defines a variety of other data types.  For instance, the
-\texttt{nltk.probability} module defines classes for a variety of
+\object{nltk.probability} module defines classes for
 probability distributions and statistical smoothing techniques; and
-the \texttt{cfg} module defines classes for encoding context free
+the \object{cfg} module defines classes for encoding context free
 grammars and probabilistic context free grammars.
@@
 \subsection{The corpus module}
-
-Many language processing tasks must be developed and tested using
-annotated data sets, or corpora.  Several such corpora are distributed
-with NLTK, as listed in Table~\ref{tab:data}.  The NLTK
-\texttt{corpus} module defines classes for reading and processing
-many of these corpora.  The following code fragment illustrates
-how the Brown Corpus is accessed.
-
-% [el] Changed to show off CorpusReader.groups().
-\begin{alltt}\small
-\textbf{>>> from nltk.corpus import brown}
-\textbf{>>> brown.groups()}
-['skill and hobbies', 'popular lore', 
-'humor', 'fiction: mystery', ...]
-\textbf{>>> brown.items('humor')}
-('cr01', 'cr02', 'cr03', 'cr04', 'cr05',
-'cr06', 'cr07', 'cr08', 'cr09')
-\textbf{>>> brown.tokenize('cr01')}
-<[<It/pps>, <was/bedz>, <among/in>,
-<these/dts>, <that/cs>, <Hinkle/np>,
-<identified/vbd>, <a/at>, ...]>
-\end{alltt}
-%
-A selection of 5\% of the Penn Treebank corpus is included with
-NLTK, and it can be accessed as follows:
-
-% [el] Changed to show groups() and TreeToken.pp()
-% [el] This example wraps over a page at a weird place.  When
-% we've made all the other changes, maybe manually insert a
-% page break at a better location?
-\begin{alltt}\small
-\textbf{>>> from nltk.corpus import treebank}
-\textbf{>>> treebank.groups()}
-('raw', 'tagged', 'parsed', 'merged')
-\textbf{>>> treebank.items('parsed')}
-['wsj_0001.prd', 'wsj_0002.prd', ...]
-\textbf{>>> item = 'parsed/wsj_0001.prd'}
-\textbf{>>> sentences = treebank.tokenize(item)}
-\textbf{>>> for sent in sentences['SUBTOKENS']:}
-\textbf{...     print sent.pp()} \emph{# pretty-print}
-(S:
-  (NP-SBJ:
-    (NP: <Pierre> <Vinken>)
-    ...
-\end{alltt}
-%
 
@@
 
+Many language processing tasks must be developed and tested using
+annotated data sets, or corpora.  Several such corpora are distributed
+with NLTK, as listed in Table~\ref{tab:data}.  The NLTK
+\object{corpus} module defines classes for reading and processing
+many of these corpora.  The following code fragment illustrates
+how the Brown Corpus is accessed.
+
+\begin{alltt}\small
+\textbf{>>> from nltk.corpus import brown}
+\textbf{>>> brown.groups()}
+['skill and hobbies', 'popular lore', 
+'humor', 'fiction: mystery', ...]
+\textbf{>>> brown.items('humor')}
+('cr01', 'cr02', 'cr03', 'cr04', 'cr05',
+'cr06', 'cr07', 'cr08', 'cr09')
+\textbf{>>> brown.tokenize('cr01')}
+<[<It/pps>, <was/bedz>, <among/in>,
+<these/dts>, <that/cs>, <Hinkle/np>,
+<identified/vbd>, <a/at>, ...]>
+\end{alltt}
+%
+A selection of 5\% of the Penn Treebank corpus is included with
+NLTK, and it is accessed as follows:
+
+\begin{alltt}\small
+\textbf{>>> from nltk.corpus import treebank}
+\textbf{>>> treebank.groups()}
+('raw', 'tagged', 'parsed', 'merged')
+\textbf{>>> treebank.items('parsed')}
+['wsj_0001.prd', 'wsj_0002.prd', ...]
+\textbf{>>> item = 'parsed/wsj_0001.prd'}
+\textbf{>>> sentences = treebank.tokenize(item)}
+\textbf{>>> for sent in sentences['SUBTOKENS']:}
+\textbf{...     print sent.pp()} \emph{# pretty-print}
+(S:
+  (NP-SBJ:
+    (NP: <Pierre> <Vinken>)
+    ...
+\end{alltt}
+%
+
 \subsection{Processing modules}
@@
 Each language processing algorithm is implemented as a class.  For
-example, the \texttt{ChartParser} and
-\texttt{Recursive\-Descent\-Parser} classes each define a single
+example, the \object{ChartParser} and
+\object{Recursive\-Descent\-Parser} classes each define a single
 algorithm for parsing a text.  We implement language processing
@@
 initialized before they can be used.  For example, the
-\texttt{NthOrderTagger} class must be initialized by training on a
+\object{NthOrderTagger} class must be initialized by training on a
 tagged corpus before it can be used.  Third, subclassing can be used
@@
 
-In order to provide a formal specification of the interface for a
-given task, each processing module defines an \emph{interface} for its task.
+Each processing module defines an \emph{interface} for its task.
 Interface classes are distinguished by naming them with a trailing
-capital ``\texttt{I},'' such as \texttt{ParserI} and
-\texttt{TokenizerI}.
+capital ``\object{I},'' such as \object{ParserI}.
 Each interface defines a single \emph{action method}, which actually
 performs the task defined by the interface.  For example, the
-\texttt{ParserI} interface defines the \texttt{parse} method; and the
-\texttt{Tokenizer} interface defines the \texttt{tokenize} method.
+\object{ParserI} interface defines the \object{parse} method; and the
+\object{Tokenizer} interface defines the \object{tokenize} method.
 When appropriate, an interface defines \emph{extended action
   methods}, which provide variations on the basic action method.  For
-example, the \texttt{ParserI} interface defines the \texttt{parse\_n}
+example, the \object{ParserI} interface defines the \object{parse\_n}
 method, which finds at most $n$ parses for a given sentence; and
-the \texttt{TokenizerI} interface defines the \texttt{xtokenize}
+the \object{TokenizerI} interface defines the \object{xtokenize}
 method, which outputs an iterator over subtokens instead of a list of
@@
 
-% [el] Changed the url here and elsewhere from nltk.sourceforge.net
-% to nltk.sf.net, because it fits better on small columns (and because 
-% it brought the length back down to 4 pages :) ).
 Three different types of documentation are available.  Tutorials
@@
 distribution consists of four packages: the Python source code
-(\texttt{nltk}); the corpora (\texttt{nltk-data}); the documentation
-(\texttt{nltk-docs}); and third-party contributions
-(\texttt{nltk-contrib}).  Before installing NLTK, it is necessary to
+(\object{nltk}); the corpora (\object{nltk-data}); the documentation
+(\object{nltk-docs}); and third-party contributions
+(\object{nltk-contrib}).  Before installing NLTK, it is necessary to
 install Python version 2.3 or later, available from
@@
 
-% [el] We really should make a better interface for running demos.
-% But I don't think we can have that figured out by the deadline
-% for this paper.
 As soon as NLTK is installed, users can run the demonstrations.  On
@@
 
-\begin{table}[bth]
+NLTK has used at the University of Pennsylvania since 2001, and has
+subsequently been adopted by several NLP courses at other
+universities, including those listed in Table~\ref{tab:courses}.
+
+Third party contributions to NLTK include: Brill tagger (Maloof),
+hidden Markov model tagger (Cohn), GPSG-style feature-based grammar
+and parser (Speer, Berwick), Kimmo finite-state morphological analyzer
+(de Marcken, Yankama, Berwick) and decision list and decision tree
+classifiers (Cohn).
+
+NLTK is an open source project, and we welcome any contributions.
+There are several ways to contribute: users can report bugs, suggest
+features, or contribute patches on Sourceforge; users can participate
+in discussions on the \textit{NLTK-Devel} mailing
+list\footnote{\url{http://lists.sourceforge.net/lists/listinfo/nltk-devel}}
+or in the NLTK public
+forums; and users can submit their own NLTK-based projects for
+inclusion in the nltk\_contrib directory.  New code modules that are
+relevant, substantial, original and well-documented will be considered
+for inclusion in NLTK proper.  Potential contributors should note that
+all source code is distributed under the GNU General Public License,
+and all documentation is distributed under a Creative Commons
+non-commercial license.  Thus they can be confident that their
+work will remain freely available to all.  Further information about
+contributing to NLTK is available at \url{http://nltk.sf.net/contrib.html}.
+
+\begin{table}[bt]
 \small\noindent
@@
 
-NLTK has used at the University of Pennsylvania since 2001, and has
-subsequently been adopted by several NLP courses at other
-universities, including those listed in Table~\ref{tab:courses}.
-
-Third party contributions to NLTK include: Brill tagger (Maloof),
-hidden Markov model tagger (Cohn), GPSG-style feature-based grammar
-and parser (Speer, Berwick), Kimmo finite-state morphological analyzer
-(de Marcken, Yankama, Berwick) and decision list and decision tree
-classifiers (Cohn).
-
-NLTK is an open source project, and we welcome any contributions.
-There are several ways to contribute: users can report bugs, suggest
-features, or contribute patches on Sourceforge; users can participate
-% [el] Add a \cite with a url, or a url inline?  The url is:
-% http://lists.sourceforge.net/lists/listinfo/nltk-devel
-in discussions on the \textit{NLTK-Devel} mailing list or in the NLTK public
-forums; and users can submit their own NLTK-based projects for
-inclusion in the nltk\_contrib directory.  New code modules that are
-relevant, substantial, original and well-documented will be considered
-for inclusion in NLTK proper.  Potential contributors should note that
-% [el] Should it be GNU instead of Gnu?
-all source code is distributed under the Gnu General Public License,
-and all documentation is distributed under a Creative Commons
-non-commercial license.  Thus they can be confident that their
-work will remain freely available to all.  Further information about
-contributing to NLTK is available at \url{http://nltk.sf.net/contrib.html}.
-
 \section{Conclusion}
@@
 encouraged to sign up for the low-volume, moderated mailing list
-% [el] Add a \cite with a url, or a url inline?  The url is:
-% http://lists.sourceforge.net/lists/listinfo/nltk-announce
-\textit{NLTK-Announce}.
+\textit{NLTK-Announce}\footnote{\url{http://lists.sourceforge.net/lists/listinfo/nltk-announce}}.
 
@@
 We are indebted to our students for feedback on the toolkit, and to
-many independent contributors who helped add new content.
+many contributors listed on the NLTK website.
 

### 2004-05-01 technical/acl2004/acl04.tex
final polishing
svn/trunk@1783

@@
   Victoria 3010, Australia \\  
-  {\tt sb@csse.unimelb.edu.au}
+  {\tt\small sb@csse.unimelb.edu.au}
 \And
@@
   Philadelphia PA 19104-6389, USA\\
-  {\tt edloper@gradient.cis.upenn.edu}
+  {\tt\small edloper@gradient.cis.upenn.edu}
 }
@@
 for advanced projects.  Typical projects might involve implementing a
-new algorithm, developing a new component, or adding support for a new
+new algorithm, developing a new component, or implementing a new
 task.
@@
 interactive exploration.  As an object-oriented language, data and
-methods can be encapsulated easily.  Python comes with an extensive
+methods can be encapsulated and re-used easily.  Python comes with an extensive
 standard library, including tools for graphical programming and
-intensive numerical processing.  And the recently added generator
+numerical processing.  And the recently added generator
 syntax makes it easy to create interactive implementations of
@@
 annotations and transformations involving \object{Tokens}.  In
-particular, each processing task takes a token, and extends it to
-include new information.  Typically, these modifications are
+particular, each processing task takes a token and extends it to
+include new information.  These modifications are typically
 \emph{monotonic}; in other words, new information is added but
@@
 defines a variety of other data types.  For instance, the
-\object{nltk.probability} module defines classes for
+\object{probability} module defines classes for
 probability distributions and statistical smoothing techniques; and
@@
 annotated data sets, or corpora.  Several such corpora are distributed
-with NLTK, as listed in Table~\ref{tab:data}.  The NLTK
+with NLTK, as listed in Table~\ref{tab:data}.  The
 \object{corpus} module defines classes for reading and processing
@@
 capital ``\object{I},'' such as \object{ParserI}.
-Each interface defines a single \emph{action method}, which actually
+Each interface defines a single \emph{action method} which
 performs the task defined by the interface.  For example, the
-\object{ParserI} interface defines the \object{parse} method; and the
+\object{ParserI} interface defines the \object{parse} method and the
 \object{Tokenizer} interface defines the \object{tokenize} method.
@@
 example, the \object{ParserI} interface defines the \object{parse\_n}
-method, which finds at most $n$ parses for a given sentence; and
+method which finds at most $n$ parses for a given sentence; and
 the \object{TokenizerI} interface defines the \object{xtokenize}
@@
 encouraged to sign up for the low-volume, moderated mailing list
-\textit{NLTK-Announce}\footnote{\url{http://lists.sourceforge.net/lists/listinfo/nltk-announce}}.
+\textit{NLTK-Announce}.\footnote{\url{http://lists.sourceforge.net/lists/listinfo/nltk-announce}}
 
@@
 
-We are indebted to our students for feedback on the toolkit, and to
-many contributors listed on the NLTK website.
+We are indebted to our students and colleagues for feedback on the
+toolkit, and to many contributors listed on the NLTK website.
 

### 2004-05-25 technical/acl2004/acl04.tex
minor fixes in response to reviewers' comments and final proofread, plus formatting tweaks
svn/trunk@1822

@@
 \documentclass[11pt]{article}
-\usepackage{acl04}
+\usepackage{colacl}
 \usepackage{times}
@@
 }
-
-\date{}
 
@@
 functionality.  As an interpreted language, Python facilitates
-interactive exploration.  As an object-oriented language, data and
-methods can be encapsulated and re-used easily.  Python comes with an extensive
+interactive exploration.  As an object-oriented language, Python
+permits data and methods to be encapsulated and re-used easily.  Python comes with an extensive
 standard library, including tools for graphical programming and
-numerical processing.  And the recently added generator
+numerical processing.  The recently added generator
 syntax makes it easy to create interactive implementations of
@@
 \object{Token} class.  Each \object{Token} instance represents a
-single unit of text, such as a word, sentence, or document; and is
+unit of text such as a word, sentence, or document, and is
 defined by a (partial) mapping from property names to values.  For
@@
 \textbf{>>> from nltk.tokenizer import *}
-\textbf{>>> token = Token(TEXT="Hello World!")}
-\textbf{>>> WhitespaceTokenizer().tokenize(token)}
-\textbf{>>> print token['SUBTOKENS'])}
+\textbf{>>> tok = Token(TEXT="Hello World!")}
+\textbf{>>> WhitespaceTokenizer().tokenize(tok)}
+\textbf{>>> print tok['SUBTOKENS'])}
 [<Hello>, <World!>]
@@
 include new information.  These modifications are typically
-\emph{monotonic}; in other words, new information is added but
+\emph{monotonic}; new information is added but
 existing information is not deleted or modified.  Thus, tokens serve
@@
 collated.  This architecture contrasts with the more typical
-\emph{pipeline} architecture, where each processing task's output
-discards its input information.  We chose the ``blackboard'' approach
-over the ``pipeline'' approach because it allows more flexibility when
+\emph{pipeline} architecture where each processing task's output
+discards its input information.  We chose the blackboard approach
+over the pipeline approach because it allows more flexibility when
 combining tasks into a single system.
@@
 \begin{tabular}{llll}
+\setlength{\tabcolsep}{0.5\tabcolsep}
 \textbf{Corpus} &
@@
 15 genres, 1.15Mw, tagged &
-training and testing taggers, text classification \\
+training \& testing taggers, text classification \\
 
@@
 270kw, tagged and chunked &
-training and testing chunk parsers \\
+training \& testing chunk parsers \\
 
@@
 63kw, named-entity markup &
-training and testing named-entity recognizers \\
+training \& testing named-entity recognizers \\
 
@@
 Names Corpus &
-8k male and female names &
+8k male \& female names &
 text classification \\
@@
 SEMCOR &
-880kw, POS and sense tagged &
+880kw, POS \& sense tagged &
 word-sense disambiguation \\
@@
 SENSEVAL 2 Corpus &
-600kw, POS and sense tagged &
+600kw, POS \& sense tagged &
 word-sense disambiguation \\
@@
 Penn Treebank (sample) &
-40kw, tagged and parsed &
+40kw, tagged \& parsed &
 parser development \\
@@
 180kw in a semantic network &
-word-sense disambiguation, NL understanding \\
+WSD, NL understanding \\
 
@@
 Many language processing tasks must be developed and tested using
-annotated data sets, or corpora.  Several such corpora are distributed
+annotated data sets or corpora.  Several such corpora are distributed
 with NLTK, as listed in Table~\ref{tab:data}.  The
@@
     (NP: <Pierre> <Vinken>)
+    (ADJP:
+      (NP: <61> <years>)
+      <old>
+    )
     ...
@@
 subtokens.
+
+NLTK includes the following modules:
+\object{cfg},
+\object{corpus},
+\object{draw}
+(\object{cfg},
+\object{chart},
+\object{corpus},
+\object{featurestruct},
+\object{fsa},
+\object{graph},
+\object{plot},
+\object{rdparser},
+\object{srparser},
+\object{tree}),
+\object{eval},
+\object{featurestruct},
+\object{parser}
+(\object{chart},
+\object{chunk},
+\object{probabilistic}),
+\object{probability},
+\object{sense},
+\object{set},
+\object{stemmer}
+(\object{porter}),
+\object{tagger},
+\object{test},
+\object{token},
+\object{tokenizer},
+\object{tree}, and
+\object{util}.
+Please see the online documentation for details.
 
@@
 
-NLTK has used at the University of Pennsylvania since 2001, and has
+NLTK has been used at the University of Pennsylvania since 2001, and has
 subsequently been adopted by several NLP courses at other
@@
 
-Third party contributions to NLTK include: Brill tagger (Maloof),
-hidden Markov model tagger (Cohn), GPSG-style feature-based grammar
-and parser (Speer, Berwick), Kimmo finite-state morphological analyzer
-(de Marcken, Yankama, Berwick) and decision list and decision tree
-classifiers (Cohn).
+Third party contributions to NLTK include: Brill tagger (Chris Maloof),
+hidden Markov model tagger (Trevor Cohn, Phil Blunsom), GPSG-style feature-based grammar
+and parser (Rob Speer, Bob Berwick), finite-state morphological analyzer
+(Carl de Marcken, Beracah Yankama, Bob Berwick), decision list and decision tree
+classifiers (Trevor Cohn), and Discourse Representation Theory
+implementation (Edward Ivanovic).
 
@@
 relevant, substantial, original and well-documented will be considered
-for inclusion in NLTK proper.  Potential contributors should note that
-all source code is distributed under the GNU General Public License,
+for inclusion in NLTK proper.
+All source code is distributed under the GNU General Public License,
 and all documentation is distributed under a Creative Commons
-non-commercial license.  Thus they can be confident that their
+non-commercial license.  Thus, potential contributors can be confident that their
 work will remain freely available to all.  Further information about
@@
 \begin{tabular}{l}
-University of Pennsylvania, USA \\
+
+Graz University of Technology, Austria \\
+\hspace{2ex}
+\textit{Information Search and Retrieval} \\[.5ex]
+
+Macquarie University, Australia \\
+\hspace{2ex}
+\textit{Intelligent Text Processing} \\[.5ex]
+
+Massachusetts Institute of Technology, USA \\
+\hspace{2ex}
+\textit{Natural Language Processing} \\[.5ex]
+
+National Autonomous University of Mexico, Mexico \\
+\hspace{2ex}
+\textit{Introduction to Natural Language Processing}\\
+\hspace{2ex}
+\textit{in Python} \\[.5ex]
+
+Ohio State University, USA \\
+\hspace{2ex}
+\textit{Statistical Natural Language Processing}\\[.5ex]
+
+University of Amsterdam, Netherlands \\
+\hspace{2ex}
+\textit{Language Processing and Information Access} \\[.5ex]
+
+University of Colorado, USA \\
+\hspace{2ex}
+\textit{Natural Language Processing} \\[.5ex]
+
+University of Edinburgh, UK \\
 \hspace{2ex}
 \textit{Introduction to Computational Linguistics} \\[.5ex]
+
+University of Magdeburg, Germany \\
+\hspace{2ex}
+\textit{Natural Language Systems} \\[.5ex]
+
+University of Malta, Malta \\
+\hspace{2ex}
+\textit{Natural Language Algorithms}\\[.5ex]
 
@@
 \textit{Human Language Technology} \\[.5ex]
+
+University of Pennsylvania, USA \\
+\hspace{2ex}
+\textit{Introduction to Computational Linguistics} \\[.5ex]
 
@@
 
-Macquarie University, Australia \\
-\hspace{2ex}
-\textit{Intelligent Text Processing} \\[.5ex]
-
-University of Edinburgh, UK \\
-\hspace{2ex}
-\textit{Introduction to Computational Linguistics} \\[.5ex]
-
-National Autonomous University of Mexico, Mexico \\
-\hspace{2ex}
-\textit{Introduction to Natural Language Processing in Python} \\[.5ex]
-
-University of Magdeburg, Germany \\
-\hspace{2ex}
-\textit{Natural Language Systems} \\[.5ex]
-
-Massachusetts Institute of Technology, USA \\
-\hspace{2ex}
-\textit{Natural Language Processing} \\[.5ex]
-
-University of Colorado, USA \\
-\hspace{2ex}
-\textit{Natural Language Processing} \\[.5ex]
-
-University of Amsterdam, Netherlands \\
-\hspace{2ex}
-\textit{Language Processing and Information Access} \\[.5ex]
-
-Graz University of Technology, Austria \\
-\hspace{2ex}
-\textit{Information Search and Retrieval} \\[.5ex]
-
 \end{tabular}
@@
 
-NLTK is a broad-coverage toolkit that provides a simple, extensible,
+NLTK is a broad-coverage natural language toolkit that provides a simple, extensible,
 uniform framework for assignments, demonstrations and projects.  It is
-thoroughly documented, easy to learn, and simple to use.  Readers who
+thoroughly documented, easy to learn, and simple to use.  NLTK is now widely
+used in research and teaching.  Readers who
 would like to receive occasional announcements about NLTK are
@@
 
+\pagebreak
+
 \bibliographystyle{acl}

### 2006-10-28 slides/data.tex
updated for toolbox version
svn/trunk@3648

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Accessing and Analyzing Linguistic Field Data}

\author{Steven Bird}
\institute{
  University of Melbourne, AUSTRALIA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
  \item variety of data types, e.g.
  \item \textbf{lexicon:} database of words, minimally containing part
    of speech information and glosses
  \item \textbf{paradigm:} any kind of rational
    tabulation of words or phrases to illustrate contrasts and systematic
    variation
  \item \textbf{text:} any larger unit such as a narrative or a
    conversation
  \item descriptions: field notes, grammars and analytical papers
  \item complex inter-relations (new word; lexicon update, paradigm;
    add to field notes; extend grammar, ...)
  \item tasks: sorting, tabulating, garnering statistics, searching,
  verifying transcriptions
  \item principle challenge is computational
  \end{itemize}

\end{frame}

\begin{frame}
\frametitle{Tools and Technologies}
\begin{itemize}
\item General-purpose tools
  \begin{itemize}
  \item word processors
  \item spreadsheets
  \item databases
  \end{itemize}
\item Special prupose tools
  \begin{itemize}
  \item Toolbox
  \end{itemize}
\item broader question: data management in support of best practices
\item Bird, Steven and Gary Simons (2003).  Seven Dimensions of Portability
    for Language Documentation and Description, \textit{Language} 79: 557-582.
\end{itemize}
\end{frame}

\section{General Purpose Tools}

\subsection{Word Processors}

\begin{frame}
\frametitle{General Purpose Tools: Word Processors}

\begin{itemize}
\item office software widely used in computer-based language
  documentation work
\item familiar, ready availability (though may be too expensive?)
\item Word processors for managing dictionaries:
  \begin{itemize}
  \item consistency of content and format --- time consuming
  \item how would we check the following constraint: \\
    \textit{each entry must have a part-of-speech field, drawn from a set of 20
     possibilities, displayed after the pronunciation field, and rendered
     in 11-point bold}
  \item generally a bad idea!
  \end{itemize}
\item save to a non-proprietary format (e.g. RTF, HTML, or XML)
\item write programs to do this checking automatically
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: Microsoft Word}
\begin{itemize}
\item sleep \verb|[sli:p]| \textbf{vi} \textit{condition of body and mind...}
\item We can enter this in MSWord, then "Save as Web Page"
\item resulting HTML:

\begin{verbatim}
<p class=MsoNormal>sleep <span style='mso-spacerun:yes'> </span>[<span
class=SpellE>sli:p</span>]<span style='mso-spacerun:yes'>  </span><b><span
style='font-size:11.0pt'>vi</span></b><span style='mso-spacerun:yes'>  </span><i>a
condition of body and mind ...o:p></o:p></i></p>
\end{verbatim}

\item Observations:
  \begin{itemize}
  \item entry: HTML paragraph using the \verb|<p>| element
  \item pos: inside a \verb|<span style='font-size:11.0pt'>| element
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: Validating Dictionaries stored in MSWord}
\scriptsize

\begin{verbatim}
>>> import re
>>> legal_pos = set(['n', 'v.t.', 'v.i.', 'adj', 'det'])
>>> pattern = re.compile(r"'font-size:11.0pt'>([a-z.]+)<")
>>> document = open("dict.htm").read()
>>> used_pos = set(re.findall(pattern, document))
>>> illegal_pos = used_pos.difference(legal_pos)
>>> print list(illegal_pos)
['v.intr', 'v.i', 'intrans']
\end{verbatim}
\end{frame}

\subsection{Spreadsheets}

\begin{frame}[fragile]
\frametitle{Example: Converting MSWord dictionary to Excel}
\small

\begin{verbatim}
>>> import re
>>> document = open("dict.htm").read()
>>> document = re.sub("[\r\n]", "", document)
>>> word_pattern = re.compile(r">([\w]+)")
>>> pron_pattern = re.compile(r"\[.*>([a-z:]+)<.*\]")
>>> for entry in document.split("<p"):
...     word_match = word_pattern.search(entry)
...     pron_match = pron_pattern.search(entry)
...     if word_match and pron_match:
...         lex = word_match.group(1)
...         pos = pron_match.group(1)
...         print '"%s","%s"' % (lex, pos)
"sleep","sli:p"
"walk","wo:k"
"wake","weik"
\end{verbatim}
\end{frame}

\begin{frame}
\frametitle{Spreadsheets}
\begin{itemize}
\item wordlists, paradigms
\item e.g. comparative wordlist (\url{http://www.rosettaproject.org/}):
  \begin{itemize}
  \item row for each cognate set
  \item column for each language
  \end{itemize}
\item export spreadsheets in \textit{CSV format}
\item read them using Python's \texttt{csv} module
\item e.g. find cognates having an edit-distance of at least three from each other
\end{itemize}
\end{frame}

\subsection{Databases}

\begin{frame}
\frametitle{Databases}
\begin{itemize}
  \item lexicons are stored in a full-fledged relational databases
  \item databases can implement many well-formedness constraints\\
    e.g. ensure that all parts-of-speech come from an
    \textit{enumerated type}
  \item However, the relational model is often too restrictive
  \item optional, repeatable fields (e.g.
    dictionary sense definitions and example sentences).
  \item SQL cannot express many linguistically-motivated queries\\
    \textit{Find all words that appear in example sentences for which no dictionary entry is provided}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Database Example}  
\scriptsize

\begin{verbatim}
>>> import csv
>>> lexemes = []
>>> defn_words = []
>>> for row in csv.reader(open("dict.csv")):
...     lexeme, pron, pos, defn = row
...     lexemes.append(lexeme)
...     defn_words += defn.split()
>>> undefined = list(set(defn_words).difference(set(lexemes)))
>>> undefined.sort()
>>> print undefined
['...', 'a', 'and', 'body', 'by', 'cease', 'condition', 'down',
'each', 'foot', 'lifting', 'mind', 'of', 'progress', 'setting', 'to']
\end{verbatim}
\end{frame}

\section{Processing Toolbox Data}

\begin{frame}
\frametitle{Processing Toolbox Data}

\begin{itemize}
\item single most popular tool for managing linguistic field data
\item many kinds of validation and formatting not supported by Toolbox software
\item simple file format
\item each file is a collection of \textit{entries} (aka
  \textit{records})
\item each entry is made up of one or more \textit{fields}
\item challenges:
  \begin{enumerate}
  \item different entry types, implications for which fields are present
  \item field order only sometimes significant
  \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Toolbox Example}
\small

\begin{verbatim}
\lx kaa
\ps N.M
\cl isi
\ge cooking banana
\gp banana bilong kukim
\sf FLORA
\dt 12/Feb/2005
\ex Taeavi iria kaa isi kovopaueva kaparapasia.
\xp Taeavi i bin planim gaden banana bilong kukim tasol long paia.
\xe Taeavi planted banana in order to cook it.
\end{verbatim}
\end{frame}

\subsection{Accessing Toolbox Data}

\begin{frame}[fragile]
\frametitle{Accessing Toolbox Data}

\begin{itemize}
\item scan the file, convert into tree object
\item preserves order of fields, gives array and XPath-style access
\item example: calculate average number of fields per entry
\end{itemize}

\begin{verbatim}
>>> from nltk_lite.corpora import toolbox
>>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
>>> sum(len(entry) for entry in lexicon) / len(lexicon)
10
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Accessing with Indexes}

\begin{verbatim}
>>> lexicon[3][0] 
<Element lx at 77bd28>
>>> lexicon[3][0].tag
'lx'
>>> lexicon[3][0].text
'kaa'
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Accessing with Indexes (cont)}

\scriptsize
\begin{verbatim}
>>> for field in lexicon[3]:
...     print "\\%s %s" % (field.tag, field.text)
\lx kaa
\ps N.M
\cl isi
\ge cooking banana
\gp banana bilong kukim
\sf FLORA
\dt 12/Feb/2005
\ex Taeavi iria kaa isi kovopaueva kaparapasia.
\xp Taeavi i bin planim gaden banana bilong kukim tasol long paia.
\xe Taeavi planted banana in order to cook it.
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Accessing with Paths}

\begin{itemize}
\item lexicon is a series of \texttt{record} objects
\item each contains field objects, such as \texttt{lx} and \texttt{ps}
\item address all the lexemes: \texttt{record/lx}
\end{itemize}

\small
\begin{verbatim}
>>> [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
['kaa', 'kaa', 'kaa', 'kaakaaro', 'kaakaaviko', 'kaakaavo', 'kaakaoko',
'kaakasi', 'kaakau', 'kaakauko', 'kaakito', 'kaakuupato', ..., 'kuvuto']
\end{verbatim}
\end{frame}

\subsection{Adding and Removing Fields}

\begin{frame}[fragile]
\frametitle{Adding New Fields}
\small

\begin{itemize}
\item Example: add CV field
\item Aside: utility function to do CV template
\end{itemize}

\begin{verbatim}
>>> import re
>>> def cv(s):
...     s = s.lower()
...     s = re.sub(r'[^a-z]',     r'_', s)
...     s = re.sub(r'[aeiou]',    r'V', s)
...     s = re.sub(r'[^V_]',      r'C', s)
...     return (s)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Adding New Fields (cont)}
\small

\begin{verbatim}
>>> from nltk_lite.etree.ElementTree import Element
>>> for entry in lexicon:
...     for i in range(len(entry)):
...         if entry[i].tag == 'lx':
...             cv_field = Element("cv")
...             cv_field.text = cv(entry[i].text)
...             entry.insert(i+1, cv_field)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Adding New Fields (cont)}
\small
\begin{verbatim}
>>> for field in lexicon[50]:
...     print "\\%s %s" % (field.tag, field.text)
\lx kaeviro
\cv CVVCVCV
\ps V.A
\ge lift off
\ge take off
\gp go antap
\nt used to describe action of plane
\dt 12/Feb/2005
\ex Pita kaeviroroe kepa kekesia oa vuripierevo kiuvu.
\xp Pita i go antap na lukim haus win i bagarapim.
\xe Peter went to look at the house that the wind destroyed.
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Removing Fields}
\scriptsize

\begin{itemize}
\item sanitise our data before giving it to others
\end{itemize}

\small
\begin{verbatim}
>>> retain = ('lx', 'ps', 'ge')
>>> for entry in lexicon:
...     entry[:] = [f for f in entry if f.tag in retain]
>>> for field in lexicon[50]:
...     print "\\%s %s" % (field.tag, field.text)
\lx kaeviro
\ps V.A
\ge lift off
\ge take off
\end{verbatim}
\end{frame}

\subsection{Formatting}

\begin{frame}[fragile]
\frametitle{Formatting Entries}

\begin{itemize}
\item request specific fields without concern for ordering
\end{itemize}
\scriptsize

\begin{verbatim}
>>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
>>> for entry in lexicon[70:80]:
...     lx = entry.findtext('lx')
...     ps = entry.findtext('ps')
...     ge = entry.findtext('ge')
...     print "%s (%s) '%s'" % (lx, ps, ge)
kakapikoto (N.N2) 'newborn baby'
kakapu (V.B) 'place in sling for purpose of carrying'
kakapua (N.N) 'sling for lifting'
kakara (N.N) 'arm band'
Kakarapaia (N.PN) 'village name'
kakarau (N.F) 'frog'
Kakarera (N.PN) 'name'
Kakareraia (N.???) 'name'
kakata (N.F) 'cockatoo'
kakate (N.F) 'bamboo water tube'
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Generating HTML Tables}

\begin{itemize}
\item Publish shoebox lexicon on the web
\end{itemize}
\scriptsize

\begin{verbatim}
>>> html = "<table>\n"
>>> for entry in lexicon[70:80]:
...     lx = entry.findtext('lx')
...     ps = entry.findtext('ps')
...     ge = entry.findtext('ge')
...     html += "  <tr><td>%s</td><td>%s</td><td>%s</td></tr>\n" % (lx, ps, ge)
>>> html += "</table>"
>>> print html
\end{verbatim}

\tiny

\begin{verbatim}
<table>
  <tr><td>kakapikoto</td><td>N.N2</td><td>newborn baby</td></tr>
  <tr><td>kakapu</td><td>V.B</td><td>place in sling for purpose of carrying</td></tr>
  <tr><td>kakapua</td><td>N.N</td><td>sling for lifting</td></tr>
  <tr><td>kakara</td><td>N.N</td><td>bracelet</td></tr>
  <tr><td>Kakarapaia</td><td>N.PN</td><td>village name</td></tr>
  <tr><td>kakarau</td><td>N.F</td><td>stingray</td></tr>
  <tr><td>Kakarera</td><td>N.PN</td><td>name</td></tr>
  <tr><td>Kakareraia</td><td>N.???</td><td>name</td></tr>
  <tr><td>kakata</td><td>N.F</td><td>cockatoo</td></tr>
  <tr><td>kakate</td><td>N.F</td><td>bamboo tube for water</td></tr>
</table>
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Generating XML}

\scriptsize
\begin{verbatim}
>>> import sys
>>> from nltk_lite.etree.ElementTree import ElementTree
>>> tree = ElementTree(lexicon[3])
>>> tree.write(sys.stdout)
<record>
  <lx>kaakaaro</lx>
  <ps>N.N</ps>
  <ge>mixtures</ge>
  <gp>???</gp>
  <eng>mixtures</eng>
  <eng>charm used to keep married men and women youthful and attractive</eng>
  <cmt>Check vowel length. Is it kaakaaro or kaakaro?</cmt>
  <dt>14/May/2005</dt>
  <ex>Kaakaroto ira purapaiveira aue iava opita, voeao-pa airepa oraouirara, ra va aiopaive.</ex>
  <xp>Kokonas ol i save wokim long ol kain samting bilong ol nupela marit, bai ol i ken kaikai.</xp>
  <xe>Mixtures are made from coconut, ???.</xe>
</record>
\end{verbatim}
\end{frame}

\section{Linguistic Exploration}

\subsection{Reduplication}

\begin{frame}[fragile]
\frametitle{Reduplication}
\scriptsize
\begin{itemize}
\item create a table of lexemes and their glosses

\begin{verbatim}
>>> lexgloss = {}
>>> for entry in lexicon:
...     lx = entry.findtext('lx')
...     if lx and entry.findtext('ps')[0] == 'V':
...         lexgloss[lx] = entry.findtext('ge')
\end{verbatim}

\item For each lexeme, check if the lexicon contains the reduplicated form:

\begin{verbatim}
>>> for lex in lexgloss:
...     if lex+lex in lexgloss:
...         print "%s (%s); %s (%s)" % (lex, lexgloss[lex], lex+lex, lexgloss[lex+lex])
\end{verbatim}

\tiny
\begin{verbatim}
kuvu (fill.up); kuvukuvu (stamp the ground)
kitu (save); kitukitu (scrub clothes)
kopa (ingest); kopakopa (gulp.down)
kasi (burn); kasikasi (angry)
koi (high pitched sound); koikoi (groan with pain)
kee (chip); keekee (shattered)
kauo (jump); kauokauo (jump up and down)
kea (deceived); keakea (lie)
kove (drop); kovekove (drip repeatedly)
kape (unable to meet); kapekape (grip with arms not meeting)
kapo (fasten.cover.strip); kapokapo (fasten.cover.strips)
koa (skin); koakoa (remove the skin)
kipu (paint); kipukipu (rub.on)
koe (spoon out a solid); koekoe (spoon out)
kovo (work); kovokovo (surround)
kiru (have sore near mouth); kirukiru (crisp)
kotu (bite); kotukotu (grind teeth together)
kavo (collect); kavokavo (work black magic)
kuri (scrape); kurikuri (scratch repeatedly)
karu (unhook); karukaru (open)
kare (return); karekare (return)
kari (break); karikari (shred)
kiro (write); kirokiro (write)
kae (carry); kaekae (tempt)
koru (make return); korukoru (obstruct)
ku (finished with); kuku (spoonfeed)
kosi (exit); kosikosi (exit)
\end{verbatim}
\end{itemize}
\end{frame}

\subsection{Complex Search Criteria}

\begin{frame}[fragile]
\frametitle{Complex Search Criteria}
\begin{itemize}
\item for phonological description, identify segments, alternations,
  syllable canon...
\item what syllable types occur in lexemes (MSC, conspiracies)?
\scriptsize

\begin{verbatim}
>>> from nltk_lite.tokenize import regexp
>>> from nltk_lite.probability import FreqDist
>>> fd = FreqDist()
>>> lexemes = [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
>>> for lex in lexemes:
...     for syl in regexp(lex, pattern=r'[^aeiou][aeiou]'):
...         fd.inc(syl)
\end{verbatim}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Complex Search Criteria (cont)}

\begin{itemize}
\item Tabulate the results:

{\scriptsize
\begin{verbatim}
>>> for vowel in 'aeiou':
...     for cons in 'ptkvsr':
...          print '%s%s:%4d ' % (cons, vowel, fd.count(cons+vowel)),
...     print
pa:  84  ta:  43  ka: 414  va:  87  sa:   0  ra: 185 
pe:  32  te:   8  ke: 139  ve:  25  se:   1  re:  62 
pi:  97  ti:   0  ki:  88  vi:  96  si:  95  ri:  83 
po:  31  to: 140  ko: 403  vo:  42  so:   3  ro:  86 
pu:  49  tu:  35  ku: 169  vu:  44  su:   1  ru:  72 
\end{verbatim}}
\item NB \texttt{t} and \texttt{s} columns
\item \texttt{ti} not attested, while \texttt{si} is frequent: palatalization?
\item which lexeme contains \texttt{su}?  \textit{kasuari}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Prosodically-motivated search}
\small
\begin{itemize}
\item segmental and prosodic constraints
\item well-formed morphemes and lexemes
\item highly-specific query (e.g. find things predicted not to exist
  by our theoretical model?)
\item \textit{Find all trisyllabic lexemes ending in a long vowel}
\item bottom-up design: need to count syllables, and test for vowels
\end{itemize}

\begin{verbatim}
>>> def num_syls(word):
...     template = cv(word)
...     num_cons = template.count('C')
...     return num_cons
>>> def is_vowel(char):
...     return (char in 'aeiou')
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Prosodically-motivated search (cont)}
\tiny

\begin{verbatim}
>>> for entry in lexicon:
...     lx = entry.findtext('lx')
...     if lx:
...         ps = entry.findtext('ps')
...         if num_syls(lx) == 3 and ps[0] == 'V'\
...           and is_vowel(lx[-1]) and is_vowel(lx[-2]):
...             ge = entry.findtext('ge')
...             print "%s (%s) '%s'" % (lx, ps, ge)
kaetupie (V.B) 'tighten'
kakupie (V.B) 'shout'
kapatau (V.B) 'add to'
kapuapie (V.B) 'wound'
kapupie (V.B) 'close tight'
kapuupie (V.B) 'close'
karepie (V.B) 'return'
karivai (V.A) 'have an appetite'
kasipie (V.B) 'care for'
kaukaupie (V.B) 'shine intensely'
kavorou (V.A) 'covet'
kavupie (V.B) 'leave.behind'
...
kuverea (V.A) 'incomplete'
\end{verbatim}
\end{frame}

\subsection{Minimal Sets}

\begin{frame}
\frametitle{Finding Minimal Sets}
\begin{itemize}
\item E.g. mace vs maze, face vs faze
\item minimal set parameters: context, target, display
\vfil
\small

\begin{tabular}{l|lll}
\textbf{Minimal Set} &     \textbf{Context} &  \textbf{Target} & \textbf{Display} \\ \hline
\textit{bib, bid, big} &   first two letters&  third letter    & word\\
\textit{deal (N), deal (V)} & whole word    &  pos             & word (pos)
\end{tabular}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Finding Minimal Sets: Example 1}
\scriptsize

\begin{verbatim}
>>> from nltk_lite.utilities import MinimalSet
>>> pos = 1
>>> ms = MinimalSet((lex[:pos] + '_' + lex[pos+1:], lex[pos], lex)
...                 for lex in lexemes if len(lex) == 4)
>>> for context in ms.contexts(3):
...     print context + ':',
...     for target in ms.targets():
...         print "%-4s" % ms.display(context, target, "-"),
...     print
k_si: kasi -    kesi -    kosi
k_ru: karu kiru keru kuru koru
k_pu: kapu kipu -    -    kopu
k_ro: karo kiro -    -    koro
k_ri: kari kiri keri kuri kori
k_pa: kapa -    kepa -    kopa
k_ra: kara kira kera -    kora
k_ku: kaku -    -    kuku koku
k_ki: kaki kiki -    -    koki
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Finding Minimal Sets: Example 2}
\scriptsize

\begin{verbatim}
>>> entries = [(e.findtext('lx'), e.findtext('ps'), e.findtext('ge'))
...               for e in lexicon
...               if e.findtext('lx') and e.findtext('ps') and e.findtext('ge')]
>>> ms = MinimalSet((lx, ps[0], "%s (%s)" % (ps[0], ge))
...          for (lx, ps, ge) in entries)
>>> for context in ms.contexts()[:10]:
...     print "%10s:" % context, "; ".join(ms.display_all(context))
  kokovara: N (unripe coconut); V (unripe)
     kapua: N (sore); V (have sores)
      koie: N (pig); V (get pig to eat)
      kovo: C (garden); N (garden); V (work)
    kavori: N (crayfish); V (collect crayfish or lobster)
    korita: N (cutlet?); V (dissect meat)
      keru: N (bone); V (harden like bone)
  kirokiro: N (bush used for sorcery); V (write)
    kaapie: N (hook); V (snag)
       kou: C (heap); V (lay egg)
\end{verbatim}
\end{frame}

\subsection{Example Application}

\begin{frame}
\frametitle{Example Application: Improving Access to Lexical Resources}
\begin{itemize}
\item lexicon as a \textit{language resource}
\item existence of standardized writing system?  literacy?
\item lookup by approximate spelling
\item soundex
\item edit distance
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Computing Soundex}
\scriptsize

\begin{verbatim}
>>> group = {
...     ' ':0,                     # blank (for short words)
...     'p':1,  'b':1,  'v':1,     # labials
...     't':2,  'd':2,  's':2,     # alveolars
...     'l':3,  'r':3,             # sonorant consonants
...     'i':4,  'e':4,             # high front vowels
...     'u':5,  'o':5,             # high back vowels
...     'a':6                      # low vowels
... }
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Computing Soundex (cont)}
\scriptsize

\begin{verbatim}
>>> def soundex(word):
...     if len(word) == 0: return word  # sanity check
...     word += '    '                  # ensure word long enough
...     c0 = word[0].upper()
...     c1 = group[word[1]]
...     cons = filter(lambda x: x in 'pbvtdslr ', word[2:])
...     c2 = group[cons[0]]
...     c3 = group[cons[1]]
...     return "%s%d%d%d" % (c0, c1, c2, c3)
>>> print soundex('kalosavi')
K632
>>> print soundex('ti')
T400
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Soundex Index}
\begin{verbatim}
>>> soundex_idx = {}
>>> for lex in lexemes:
...     code = soundex(lex)
...     if code not in soundex_idx:
...         soundex_idx[code] = set()
...     soundex_idx[code].add(lex)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Sorting and Lookup}
\scriptsize
\begin{verbatim}
>>> from nltk_lite.utilities import edit_dist
>>> def fuzzy_spell(target):
...     scored_candidates = []
...     code = soundex(target)
...     for word in soundex_idx[code]:
...         dist = edit_dist(word, target)
...         scored_candidates.append((dist, word))
...     scored_candidates.sort()
...     return [w for (d,w) in scored_candidates[:10]]
\end{verbatim}

\begin{verbatim}
>>> fuzzy_spell('kokopouto')
['kokopeoto', 'kokopuoto', 'kokepato', 'koovoto', 'koepato',
 'kooupato', 'kopato', 'kopiito', 'kovuto', 'koavaato']
>>> fuzzy_spell('kogou')
['kogo', 'koou', 'kokeu', 'koko', 'kokoa', 'kokoi', 'kokoo',
 'koku', 'kooe', 'kooku']
\end{verbatim}
\end{frame}

\section{Generating Reports}

\subsection{Summaries}

\begin{frame}[fragile]
\frametitle{Generating Summary Reports}
\begin{itemize}
\item overall picture of the quality and organisation of the data
\item E.g. entry patterns
\end{itemize}

\small
\begin{verbatim}
>>> fd = FreqDist()
>>> for entry in lexicon:
...     for field in entry:
...	        fd.inc(field.tag)
>>> fd.sorted_samples()[:10]
['ge', 'ex', 'xe', 'xp', 'gp', 'lx', 'ps', 'dt', 'rt', 'eng']
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Generating Summary Reports (cont)}
\small

\begin{verbatim}
>>> fd = FreqDist()
>>> for entry in lexicon:
...	    fd.inc(':'.join(field.tag for field in entry))
>>> top_ten = fd.sorted_samples()[:10]
>>> print '\n'.join(top_ten)
lx:rt:ps:ge:gp:dt:ex:xp:xe
lx:ps:ge:gp:dt:ex:xp:xe
lx:ps:ge:gp:dt:ex:xp:xe:ex:xp:xe
lx:rt:ps:ge:gp:dt:ex:xp:xe:ex:xp:xe
lx:ps:ge:gp:nt:dt:ex:xp:xe
lx:ps:ge:gp:dt
lx:ps:ge:ge:gp:dt:ex:xp:xe:ex:xp:xe
lx:rt:ps:ge:ge:gp:dt:ex:xp:xe:ex:xp:xe
lx:ps:ge:ge:gp:dt:ex:xp:xe
lx:rt:ps:ge:ge:gp:dt:ex:xp:xe
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Generating Summary Reports (cont)}
\small

\begin{verbatim}
>>> fd = FreqDist()
>>> for entry in lexicon:
...     previous = "0"
...     for field in entry:
...         current = field.tag
...         fd.inc("%s->%s" % (previous, current))
...         previous = current
>>> fd.sorted_samples()[:10]
['ex->xp', 'xp->xe', '0->lx', 'ge->gp', 'ps->ge', 'dt->ex', 'lx->ps',
'gp->dt', 'xe->ex', 'lx->rt']
\end{verbatim}
\end{frame}

\subsection{Grammar}

\begin{frame}[fragile]
\frametitle{Grammar}
\small

\begin{verbatim}
>>> from nltk_lite import parse
>>> grammar = parse.cfg.parse_grammar('''
...   S -> Head "ps" Glosses Comment "dt" Examples
...   Head -> "lx" | "lx" "rt"
...   Glosses -> Gloss Glosses
...   Glosses ->
...   Gloss -> "ge" | "gp"
...   Examples -> Example Examples
...   Examples ->
...   Example -> "ex" "xp" "xe"
...   Comment -> "cmt"
...   Comment ->
...   ''')
>>> rd_parser = parse.RecursiveDescent(grammar)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Grammar}
\small

\begin{verbatim}
>>> for entry in lexicon[10:20]:
...     marker_list = [field.tag for field in entry]
...     if rd_parser.get_parse_list(marker_list):
...         print "+", ':'.join(marker_list)
...     else:
...         print "-", ':'.join(marker_list)
- lx:ps:ge:gp:sf:nt:dt:ex:xp:xe:ex:xp:xe
- lx:rt:ps:ge:gp:nt:dt:ex:xp:xe:ex:xp:xe
- lx:ps:ge:gp:nt:dt:ex:xp:xe:ex:xp:xe
- lx:ps:ge:gp:nt:sf:dt
- lx:ps:ge:gp:dt:cmt:ex:xp:xe:ex:xp:xe
+ lx:ps:ge:ge:ge:gp:cmt:dt:ex:xp:xe
+ lx:rt:ps:ge:gp:cmt:dt:ex:xp:xe:ex:xp:xe
+ lx:rt:ps:ge:ge:gp:dt
- lx:rt:ps:ge:ge:ge:gp:dt:cmt:ex:xp:xe:ex:xp:xe:ex:xp:xe
+ lx:rt:ps:ge:gp:dt:ex:xp:xe
\end{verbatim}
\end{frame}

\subsection{Timestamps}

\begin{frame}[fragile]
\frametitle{Looking at Timestamps}
\scriptsize
\begin{verbatim}
>>> fd = FreqDist()
>>> from string import split
>>> for entry in shoebox.dictionary('rotokas'):
...     if 'dt' in entry:
...         (day, month, year) = split(entry['dt'], '/')
...         fd.inc((month, year))
>>> for time in fd.sorted_samples():
...     print time[0], '/', time[1], ':', fd.count(time)
Feb / 2005 : 307
Dec / 2004 : 151
Jan / 2005 : 123
Feb / 2004 : 64
Sep / 2004 : 49
May / 2005 : 46
Mar / 2005 : 37
Apr / 2005 : 29
Jul / 2004 : 14
Nov / 2004 : 5
Oct / 2004 : 5
Aug / 2004 : 4
May / 2003 : 2
Jan / 2004 : 1
May / 2004 : 1
\end{verbatim}
\end{frame}

\end{document}

### 2006-10-31 slides/data.tex
updated with simplified examples from chapter
svn/trunk@3664

@@
   \end{itemize}
-\item Special prupose tools
+\item Special purpose tools
   \begin{itemize}
@@
 \begin{verbatim}
-<p class=MsoNormal>sleep <span style='mso-spacerun:yes'> </span>[<span
-class=SpellE>sli:p</span>]<span style='mso-spacerun:yes'>  </span><b><span
-style='font-size:11.0pt'>vi</span></b><span style='mso-spacerun:yes'>  </span><i>a
-condition of body and mind ...o:p></o:p></i></p>
+<p class=MsoNormal>
+  sleep
+  <span style='mso-spacerun:yes'> </span>
+  [<span class=SpellE>sli:p</span>]
+  <span style='mso-spacerun:yes'> </span>
+  <b><span style='font-size:11.0pt'>vi</span></b>
+  <span style='mso-spacerun:yes'> </span>
+  <i>a condition of body and mind ...</i>
+</p>
 \end{verbatim}
@@
 \item preserves order of fields, gives array and XPath-style access
-\item example: calculate average number of fields per entry
 \end{itemize}
@@
 >>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
->>> sum(len(entry) for entry in lexicon) / len(lexicon)
-10
 \end{verbatim}
@@
 \begin{verbatim}
->>> for field in lexicon[3]:
-...     print "\\%s %s" % (field.tag, field.text)
+>>> toolbox.to_sfm_string(lexicon[3])
 \lx kaa
@@
 \begin{verbatim}
->>> from nltk_lite.etree.ElementTree import Element
+>>> from nltk_lite.etree.ElementTree import SubElement
 >>> for entry in lexicon:
-...     for i in range(len(entry)):
-...         if entry[i].tag == 'lx':
-...             cv_field = Element("cv")
-...             cv_field.text = cv(entry[i].text)
-...             entry.insert(i+1, cv_field)
+...     for field in entry:
+...	    if field.tag == 'lx':
+...         cv_field = SubElement(entry, 'cv')
+...         cv_field.text = cv(field.text)
 \end{verbatim}
@@
 \begin{verbatim}
->>> for field in lexicon[50]:
-...     print "\\%s %s" % (field.tag, field.text)
+>>> toolbox.to_sfm_string(lexicon[50])
 \lx kaeviro
@@
 ...     entry[:] = [f for f in entry if f.tag in retain]
->>> for field in lexicon[50]:
-...     print "\\%s %s" % (field.tag, field.text)
+>>> toolbox.to_sfm_string(lexicon[50])
 \lx kaeviro
@@
 \begin{verbatim}
->>> def num_syls(word):
+>>> def num_cons(word):
 ...     template = cv(word)
-...     num_cons = template.count('C')
-...     return num_cons
+...     return template.count('C')
 >>> def is_vowel(char):
@@
 ...         ps = entry.findtext('ps')
-...         if num_syls(lx) == 3 and ps[0] == 'V'\
+...         if num_cons(lx) == 3 and ps[0] == 'V'\
 ...           and is_vowel(lx[-1]) and is_vowel(lx[-2]):
@@
 \item overall picture of the quality and organisation of the data
+\item e.g. average number of fields per record
+\end{itemize}
+
+\begin{verbatim}
+>>> sum(len(entry) for entry in lexicon) / len(lexicon)
+10
+\end{verbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Generating Summary Reports (cont)}
+\begin{itemize}
 \item E.g. entry patterns

### 2006-10-31 slides/data.tex
further polishing
svn/trunk@3665

@@
 \frametitle{Example: Microsoft Word}
+\small
 \begin{itemize}
@@
 
+\small
 \begin{verbatim}
@@
 
+\small
 \begin{itemize}
@@
 \frametitle{Adding New Fields (cont)}
-\small
+\scriptsize
 \begin{verbatim}
@@
 ...     ge = entry.findtext('ge')
-...     html += "  <tr><td>%s</td><td>%s</td><td>%s</td></tr>\n" % (lx, ps, ge)
+...     html += "  <tr><td>%s</td><td>%s</td><td>%s</td></tr>\n" %
+...       (lx, ps, ge)
 >>> html += "</table>"
@@
 
-\scriptsize
+\tiny
 \begin{verbatim}
@@
 \end{verbatim}
+\end{itemize}
+\end{frame}
 
-\tiny
+\begin{frame}[fragile]
+\frametitle{Reduplication (cont)}
+\scriptsize
 \begin{verbatim}
@@
 kavo (collect); kavokavo (work black magic)
-kuri (scrape); kurikuri (scratch repeatedly)
-karu (unhook); karukaru (open)
-kare (return); karekare (return)
-kari (break); karikari (shred)
-kiro (write); kirokiro (write)
-kae (carry); kaekae (tempt)
-koru (make return); korukoru (obstruct)
-ku (finished with); kuku (spoonfeed)
-kosi (exit); kosikosi (exit)
+...
 \end{verbatim}
-\end{itemize}
 \end{frame}
@@
 \item what syllable types occur in lexemes (MSC, conspiracies)?
+\end{itemize}
+
 \scriptsize
-
 \begin{verbatim}
@@
 \end{verbatim}
-\end{itemize}
 \end{frame}
@@
 
-{\scriptsize
+\scriptsize
 \begin{verbatim}
@@
 ...     for cons in 'ptkvsr':
-...          print '%s%s:%4d ' % (cons, vowel, fd.count(cons+vowel)),
+...          print '%s%s:%4d ' %
+...            (cons, vowel, fd.count(cons+vowel)),
 ...     print
@@
 pu:  49  tu:  35  ku: 169  vu:  44  su:   1  ru:  72 
-\end{verbatim}}
+\end{verbatim}
+
+\small
 \item NB \texttt{t} and \texttt{s} columns
@@
 >>> fd.sorted_samples()[:10]
-['ge', 'ex', 'xe', 'xp', 'gp', 'lx', 'ps', 'dt', 'rt', 'eng']
+['ge', 'ex', 'xe', 'xp', 'gp', 'lx', 'ps',
+'dt', 'rt', 'eng']
 \end{verbatim}
@@
 >>> fd.sorted_samples()[:10]
-['ex->xp', 'xp->xe', '0->lx', 'ge->gp', 'ps->ge', 'dt->ex', 'lx->ps',
-'gp->dt', 'xe->ex', 'lx->rt']
+['ex->xp', 'xp->xe', '0->lx', 'ge->gp', 'ps->ge',
+'dt->ex', 'lx->ps', 'gp->dt', 'xe->ex', 'lx->rt']
 \end{verbatim}
@@
 \frametitle{Grammar}
-\small
+\scriptsize
 
@@
 \frametitle{Looking at Timestamps}
+
 \scriptsize
@@
 Oct / 2004 : 5
-Aug / 2004 : 4
-May / 2003 : 2
-Jan / 2004 : 1
-May / 2004 : 1
+...
 \end{verbatim}

### 2007-05-03 slides/programming.py
demonstration examples from book chapter
svn/trunk@4469

@@ -1,0 +1,371 @@
+# 2.1 Python the Calculator
+
+3 + 2 * 5 - 1
+3/3
+1/3
+1 +
+
+### 2.2 Understanding the Basics: Strings and Variables
+
+### 2.2.1 Representing Text
+
+Hello World
+'Hello World'
+"Hello World"
+'Hello' + 'World'
+'Hi' + 'Hi' + 'Hi'
+'Hi' * 3
+
+### 2.2.2 Storing and reusing values
+
+msg = 'Hello World'
+msg
+
+msg = 'Hi'
+num = 3
+msg * num
+
+msg = msg * num
+msg
+
+### 2.2.3 Printing and inspecting strings
+
+msg = 'Hello World'
+msg
+
+Now do this using IDLEs text editor
+
+================================ RESTART ================================
+
+print msg
+
+msg2 = 'Goodbye'
+print msg, msg2
+
+msg = "I like NLP!"
+msg = 'Hello World'
+
+### 2.3 Slicing and Dicing
+
+### 2.3.1 Accessing individual characters
+
+msg[0]
+msg[3]
+msg[5]
+msg[11]
+len(msg)
+msg[-1]
+msg[-3]
+msg[-6]
+
+### 2.3.2 Accessing substrings
+
+msg[1:4]
+len(msg)
+msg[0:11]
+msg[0:-6]
+msg[:3]
+msg[6:]
+msg[:-1]
+msg[:]
+msg[6:11:2]
+msg[10:5:-2]
+
+### 2.4 Strings, Sequences, and Sentences
+
+sent = 'colorless green ideas sleep furiously'
+sent[16:21]
+len(sent)
+
+### 2.4.1 Lists
+
+phrase1 = ['colorless', 'green', 'ideas']
+phrase1
+len(phrase1)
+
+phrase1[0]
+phrase1[-1]
+phrase1[-5]
+
+phrase1[1:3]
+phrase1[-2:]
+
+phrase2 = phrase1 + ['sleep', 'furiously']
+phrase2
+phrase1
+
+phrase1[0] = 'colorful'
+phrase1
+
+msg[0] = 'J'
+
+phrase2.sort()
+phrase2
+
+phrase2.reverse()
+phrase2
+
+phrase2[::-1]
+phrase2
+
+phrase2.append('said')
+phrase2.append('Chomsky')
+phrase2
+
+phrase2.index('green')
+
+bat = ['bat', [[1, 'n', 'flying mammal'], [2, 'n', 'striking instrument']]]
+
[…]

### 2007-05-03 slides/advanced.tex
updates to slides based on revised tutorials; mention availability of slides on teaching page; other misc fixes
svn/trunk@4472

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Advanced Programming in Python}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
  University of Pennsylvania, USA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\section{Regular Expressions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Program Development}

\begin{frame}
\frametitle{Program Development}
\begin{itemize}
\item High-level Design
\item Algorithm Design
\item Structured Programming
\item Debugging
\item Performance Tuning
\end{itemize}
\end{frame}

\subsection{Defining Functions and Modules}

\begin{frame}[fragile]
\frametitle{Defining Functions}
\scriptsize
\begin{itemize}
\item part of a program needs to be used several times over
\item e.g. form plural of singular noun, to be done in several places
  in a program
\item localize this work inside a \textit{function}
\item also helps readability, reusability

\begin{verbatim}
  >>> def plural(word):
  ...     if word[-1] == 'y':
  ...         return word[:-1] + 'ies'
  ...     elif word[-1] in 'sx':
  ...         return word + 'es'
  ...     elif word[-2:] in ['sh', 'ch']:
  ...         return word + 'es'
  ...     elif word[-2:] == 'an':
  ...         return word[:-2] + 'en'
  ...     return word + 's'
  >>> plural('fairy')
  'fairies'
  >>> plural('woman')
  'women'
\end{verbatim}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Defining Modules}

\begin{verbatim}
def parsed(files):
    for file in files:
        path = os.path.join(get_basedir(), "treebank", file)
        s = open(path).read()
        for t in tokenize.blankline(s):
            yield tree.bracket_parse(t)
\end{verbatim}
\end{frame}

\subsection{Algorithm Design}

\begin{frame}
  \frametitle{Algorithm Design}
  \small

  \begin{itemize}
  \item \textit{algorithm}: a "recipe" for solving a problem
  \item e.g. to multiply 16 by 12 we might use any of the following methods:

    \begin{enumerate}
    \item Add 16 to itself 12 times over
    \item Perform "long multiplication", starting with the least-significant
      digits of both numbers
    \item Look up a multiplication table
    \item Repeatedly halve the first number and double the second,
      16*12 = 8*24 = 4*48 = 2*96 = 192
    \item Do 10*12 to get 120, then add 6*12
    \end{enumerate}

  \item computation time, intermediate storage
  \item brute-force, divide-and-conquer, dynamic programming, greedy search
  \item Textbook: Levitin, Anany (2003) \textit{Introduction to the Design and
      Analysis of Algorithms}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Sorting Algorithms}

\begin{itemize}
\item Many algorithms for sorting
\item Illustrates the difference in algorithm complexity

\begin{verbatim}
  >>> from random import shuffle
  >>> from nltk_lite.misc import sort
  >>> a = range(1000)
  >>> shuffle(a); sort.bubble(a)
  250918
  >>> shuffle(a); sort.merge(a)
  6175
  >>> shuffle(a); sort.quick(a)
  2378
\end{verbatim}
\end{itemize}
\end{frame}

\subsection{Top-Down Design}

\begin{frame}
\frametitle{Top-Down Design}
\begin{itemize}
\item Break down high-level task into manageable components
\item Build and test each component
\item Assemble them into a complex system
\item Example: identify adjective-noun collocations in text
  \begin{enumerate}
  \item read in the corpus
  \item count events (each adjective, noun, adj-noun combination)
  \item compute $\chi^2$ statistics
  \item sort adj-noun pairs in decreasing order
  \item output $n$-most significant collocations
  \end{enumerate}
\item what are the inputs and outputs of each step (i.e. \textit{interfaces})
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Top-Down Design (cont)}

\begin{enumerate}
\item Define top-level function:

\begin{verbatim}
def colloc(corpus_name, num):
    corpus = load_corpus(corpus_name)
    a_counts, n_counts, a_n_counts = count(corpus, 'JJ', 'NN')
    a_n_scores = chisq(a_counts, n_counts, a_n_counts)
    a_n_ranked = sort_by_score(a_n_scores)
    return a_n_ranked[:num]
\end{verbatim}

\item Iteratively define lower-level functions, e.g. \texttt{chisq()}

\item Bottom-up testing
\begin{itemize}
\item standard test cases for components, known output
\item only assemble once each component is known to work correctly
  on the test cases
\end{itemize}
\end{enumerate}
\end{frame}

\subsection{Debugging}

\begin{frame}
  \frametitle{Debugging}

  \begin{itemize}
  \item What's in the name:
    \begin{itemize}
    \item problems are small relative to their impact
    \item hard-to-find
    \item seem to take on a life of their own as the programmer tries to
      hunt them down
    \end{itemize}
  \item steps:
    \begin{itemize}
    \item isolate the problem
    \item syntax error: see error messages, linked to line numbers
    \item run-time error: \textit{stack-trace}
    \item add diagnostic print statements
    \item display values of variables just before problem
    \item Python debugger: \texttt{pdb}
    \end{itemize}
  \end{itemize}
\end{frame}

\subsection{Next Steps...}

\begin{frame}
\frametitle{Next Steps: Learning to Program}
\begin{itemize}
\item experimental
\item Python's \texttt{help} and \texttt{dir} commands
\item tutorial exercises
\end{itemize}
\end{frame}

### 2007-05-03 slides/chunk.tex
updates to slides based on revised tutorials; mention availability of slides on teaching page; other misc fixes
svn/trunk@4472

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Chunk Parsing}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
  University of Pennsylvania, USA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\frame{\titlepage}

\section{Introduction}

\subsection{What is it?}

\begin{frame}
\begin{itemize}
\item chunk parsing:
  \begin{itemize}
    \item efficient and robust approach to parsing natural language
    \item a popular alternative to the full parsing
  \end{itemize}
\item chunks:
  \begin{itemize}
  \item non-overlapping regions of text
  \item contain a head word (e.g. noun)
  \item adjacent modifiers and function words
  \end{itemize}
\item motivations:
  \begin{itemize}
  \item extract information
  \item ignore information
  \end{itemize}
\end{itemize}
\end{frame}
\subsection{Motivation}

\pgfdeclareimage[width=12cm]{chunk-coref}{../images/chunk-coref}
\begin{frame}
  \frametitle{Extracting Information: Coreference Annotation}
  \centerline{\pgfuseimage{chunk-coref}}
\end{frame}

\pgfdeclareimage[width=12cm]{chunk-muc}{../images/chunk-muc}
\begin{frame}
  \frametitle{Extracting Information: Message Understanding}
  \centerline{\pgfuseimage{chunk-muc}}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Ignoring Information: Lexical Acquisition}

  \begin{itemize}
  \item studying syntactic patterns, e.g. finding verbs in a corpus, displaying possible arguments
  \item e.g. \texttt{gave}, in 100 files of the Penn Treebank corpus
  \item replaced internal details of each noun phrase with \texttt{NP}

\begin{verbatim}
  gave NP
  gave up NP in NP
  gave NP up
  gave NP help
  gave NP to NP
\end{verbatim}

  \item use in lexical acquisition, grammar development
  \end{itemize}
\end{frame}

\subsection{Analogy with Tokenization and Tagging}

\pgfdeclareimage[width=8cm]{chunk-segmentation}{../images/chunk-segmentation}
\begin{frame}[fragile]
  \frametitle{Analogy with Tokenization and Tagging}

  \begin{itemize}
  \item fundamental in NLP: segmentation and labelling
  \item tokenization and tagging \\[2ex]
  \centerline{\pgfuseimage{chunk-segmentation}}
  \item other similarities: skipping material; finite-state; application specific
  \end{itemize}
\end{frame}

\subsection{Chunking vs Parsing}

\begin{frame}[fragile]
  \frametitle{Chunking vs Parsing}
  \scriptsize

  \begin{enumerate}
  \item Parsing
\begin{verbatim}
  [
    [ G.K. Chesterton ],
    [
      [ author ] of
      [
        [ The Man ] who was
        [ Thursday ]
      ]
    ]
  ]
\end{verbatim}

  \item Chunking:
\begin{verbatim}
  [ G.K. Chesterton ],
  [ author ] of
  [ The Man ] who was
  [ Thursday ]
\end{verbatim}
\end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{Chunking vs Parsing}
  \begin{enumerate}
  \item flat vs nested
  \item context
  \item robustness
  \item efficiency
  \item methodology
  \end{enumerate}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Perfection is unattainable}

\begin{verbatim}
  1. Prepositional phrase:
  [
    [ I ]
    [ turned ]
    [ off the spectroroute ]
  ]

  2. Verb-particle construction:
  [
    [ I ]
    [ turned off ]
    [ the spectroroute ]
  ]
\end{verbatim}
\end{frame}

\section{Accessing Chunked Corpora}

\subsection{Representing Chunks: Tags vs Trees}

\pgfdeclareimage[width=8cm]{chunk-tagrep}{../images/chunk-tagrep}
\begin{frame}
  \frametitle{Tag Representation}
  \centerline{\pgfuseimage{chunk-tagrep}}
\end{frame}

\pgfdeclareimage[width=8cm]{chunk-treerep}{../images/chunk-treerep}
\begin{frame}
  \frametitle{Tree Representation}
  \centerline{\pgfuseimage{chunk-treerep}}
\end{frame}

\subsection{Chunk Structures}

\begin{frame}[fragile]
  \frametitle{Chunk Structures}
\begin{verbatim}
  (S: (NP: 'I')
      'saw'
      (NP: 'the' 'big' 'dog')
      'on'
      (NP: 'the' 'hill'))
\end{verbatim}
  \begin{itemize}
  \item Demonstration: reading chunk structures from Treebank and CoNLL-2000 corpora
  \end{itemize}
\end{frame}

\section{Chunk Parsing}

\subsection{Chunking with Regular Expressions}

\begin{frame}[fragile]
  \frametitle{Chunk Parsing}
  \small

  \begin{itemize}
  \item regular expressions over part-of-speech tags: \texttt{parse.RegexpChunk}
  \item \textit{Tag string:} a string consisting of tags
    delimited with angle-brackets,
    e.g. \verb/<DT><JJ><NN><VBD><DT><NN>/
  \item \textit{Tag pattern:} regular expression over tag strings
    \begin{itemize}
    \item \verb/<DT><JJ>?<NN>/
    \item \verb/<NN|JJ>+/
    \item \verb/<NN.*>/
    \end{itemize}
  \item chunk a sequence of words matching a tag pattern: \texttt{parse.ChunkRule}

\begin{verbatim}
    >>> grammar = "NP: {<DT|NN>+}  # Chunk sequences of DT and NN'"
\end{verbatim}

  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{A Simple NP Chunker}
  \scriptsize

\begin{verbatim}
    grammar = r"""
    NP:
      {<DT>?<JJ>*<NN>}    # chunk determiners, adjectives and nouns
      {<NNP>+}            # chunk sequences of proper nouns
    """
    cp = chunk.Regexp(grammar)
    tagged_tokens = [("the", "DT"), ("little", "JJ"), ("cat", "NN"),
         ("sat", "VBD"), ("on", "IN"), ("the", "DT"), ("mat", "NN")]

    >>> cp.parse(tagged_tokens)
    (S:
      (NP: ('the', 'DT') ('little', 'JJ') ('cat', 'NN'))
      ('sat', 'VBD')
      ('on', 'IN')
      (NP: ('the', 'DT') ('mat', 'NN')))
\end{verbatim}
\end{frame}

\subsection{Developing Chunkers}

\begin{frame}[fragile]
  \frametitle{Developing Chunkers}
  \tiny

\begin{verbatim}
    cp1 = chunk.Regexp(r"""
    NP: {<DT><JJ><NN>}      # Chunk det+adj+noun
        {<DT|NN>+}          # Chunk sequences of NN and DT
    """)
    cp2 = chunk.Regexp(r"""
    NP: {<DT|NN>+}          # Chunk sequences of NN and DT
        {<DT><JJ><NN>}      # Chunk det+adj+noun
    """)

    >>> print cp1.parse(tagged_tokens, trace=1)
    # Input:
     <DT>  <JJ>  <NN>  <VBD>  <IN>  <DT>  <NN> 
    # Chunk det+adj+noun:
    {<DT>  <JJ>  <NN>} <VBD>  <IN>  <DT>  <NN> 
    # Chunk sequences of NN and DT:
    {<DT>  <JJ>  <NN>} <VBD>  <IN> {<DT>  <NN>}
\end{verbatim}

\begin{itemize}
  \item tracing; rule ordering; overlapping contexts
\end{itemize}
\end{frame}

\subsection{More Chunking Rules}

\begin{frame}[fragile]
  \frametitle{More Chunking Rules: Chinking}
  \small

  \begin{itemize}
  \item \textit{chink:} sequence of stopwords
  \item \textit{chinking:} process of removing tokens from a chunk
  \end{itemize}

  \tiny

\begin{tabular}{llll}
& \textbf{Entire chunk}
& \textbf{Middle of a chunk}
& \textbf{End of a chunk} \\
\textbf{Input:}
& \verb|[a/DT  big/JJ  cat/NN]|
& \verb|[a/DT  big/JJ  cat/NN]|
& \verb|[a/DT  big/JJ  cat/NN]| \\
\textbf{Operation:}
& Chink \verb|a/DT big/JJ cat/NN|
& Chink \verb|big/JJ|
& Chink \verb|cat/NN| \\
\textbf{Output:}
& \verb|a/DT  big/JJ  cat/NN|
& \verb|[a/DT] big/JJ [cat/NN]|
& \verb|[a/DT  big/JJ] cat/NN|
\end{tabular}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Chinking Example}
  \scriptsize

\begin{verbatim}
    >>> grammar = r"""
    ... NP:
    ...   {<.*>+}          # Chunk everything
    ...   }<VBD|IN>+{      # Chink sequences of VBD and IN
    ... """
    >>> cp = chunk.Regexp(grammar)
    >>> print cp.parse(tagged_tokens)
    (S:
      (NP: ('the', 'DT') ('little', 'JJ') ('cat', 'NN'))
      ('sat', 'VBD')
      ('on', 'IN')
      (NP: ('the', 'DT') ('mat', 'NN')))
    >>> print chunk.accuracy(cp, conll2000.chunked(files='test', chunk_types=('NP',)))
    0.581041433607
\end{verbatim}
\end{frame}

\section{Evaluating Chunk Parsers}

\subsection{Precision and Recall}

\begin{frame}
  \frametitle{Evaluating Chunk Parsers}

  \begin{itemize}
  \item Process:
    \begin{enumerate}
    \item take some already chunked text
    \item strip off the chunks
    \item rechunk it
    \item compare the result with the original chunked text
    \end{enumerate}
  \item \texttt{ChunkScore.score()}
    \begin{itemize}
    \item \textit{precision}: what fraction of the returned chunks were correct?
    \item \textit{recall}: what fraction of correct chunks were returned?
    \end{itemize}
  \end{itemize}
\end{frame}

\pgfdeclareimage[width=8cm]{precision-recall}{../images/precision-recall}
\begin{frame}
  \frametitle{Precision and Recall}
  \centerline{\pgfuseimage{precision-recall}}
\end{frame}

\subsection{Evaluation in NLTK}

\begin{frame}[fragile]
  \frametitle{Chunker evaluation in NLTK}
  \scriptsize

\begin{verbatim}
>>> rule = parse.ChunkRule('<DT|JJ|NN>+', "Chunk sequences of DT, JJ, and NN")
>>> chunkparser = parse.RegexpChunk([rule], chunk_node='NP', top_node='S')
>>> chunkscore = parse.ChunkScore()
>>> for chunk_struct in islice(treebank.chunked(), 10):
...     test_sent = chunkparser.parse(chunk_struct.leaves())
...     chunkscore.score(chunk_struct, test_sent)
>>> print chunkscore
ChunkParse score:
    Precision:  48.6%
    Recall:     34.0%
    F-Measure:  40.0%
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Error Analysis: Missed Chunks}
  \scriptsize

\begin{verbatim}
>>> from random import randint
>>> missed = chunkscore.missed()
>>> for i in range(15):
...     print missed[randint(0,len(missed)-1)]
(('A', 'DT'), ('Lorillard', 'NNP'), ('spokewoman', 'NN'))
(('it', 'PRP'),)
(('symptoms', 'NNS'),)
(('even', 'RB'), ('brief', 'JJ'), ('exposures', 'NNS'))
(('its', 'PRP$'), ('Micronite', 'NN'), ('cigarette', 'NN'), ('filters', 'NNS'))
(('30', 'CD'), ('years', 'NNS'))
(('workers', 'NNS'),)
(('preliminary', 'JJ'), ('findings', 'NNS'))
(('Medicine', 'NNP'),)
(('cancer', 'NN'), ('deaths', 'NNS'))
(('Consolidated', 'NNP'), ('Gold', 'NNP'), ('Fields', 'NNP'), ('PLC', 'NNP'))
(('Medicine', 'NNP'),)
(('its', 'PRP$'), ('Micronite', 'NN'), ('cigarette', 'NN'), ('filters', 'NNS'))
(('a', 'DT'), ('forum', 'NN'))
(('researchers', 'NNS'),)
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Error Analysis: Incorrect Chunks}
  \scriptsize

\begin{verbatim}
>>> incorrect = chunkscore.incorrect()
>>> for i in range(15):
...     print incorrect[randint(0,len(incorrect)-1)]
(('New', 'JJ'), ('York-based', 'JJ'))
(('Micronite', 'NN'), ('cigarette', 'NN'))
(('a', 'DT'), ('forum', 'NN'), ('likely', 'JJ'))
(('later', 'JJ'),)
(('later', 'JJ'),)
(('brief', 'JJ'),)
(('preliminary', 'JJ'),)
(('New', 'JJ'), ('York-based', 'JJ'))
(('resilient', 'JJ'),)
(('group', 'NN'),)
(('cancer', 'NN'),)
(('the', 'DT'),)
(('cancer', 'NN'),)
(('Micronite', 'NN'), ('cigarette', 'NN'))
(('A', 'DT'),)
\end{verbatim}
\end{frame}

\subsection{Evaluation Methodology}

\begin{frame}
  \frametitle{Evaluation Methodology}
  \begin{itemize}
  \item Baseline:
    \begin{itemize}
    \item How hard is chunking?
    \item What is a good baseline for evaluation?
    \end{itemize}
  \item Bake-off
  \end{itemize}
\end{frame}

\subsection{Development Methodology}

\begin{frame}
  \frametitle{Development Methodology}
  \begin{itemize}
  \item approaches
    \begin{itemize}
    \item different rules and combinations
    \item hand-crafted vs automatic
    \end{itemize}
  \item focus on diagnosis:
    \begin{itemize}
    \item manual
    \item utility functions
    \item error analysis
    \item evaluation
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
  \item light-weight methods: as seen in tagging
  \item applications: extraction, lexical acquisition\\
    (aside: chunking as a utility method in parsing)
  \item next: parsing
  \item but first: switch to application focus
  \end{itemize}
\end{frame}

\end{document}

### 2007-05-03 slides/introduction.tex
updates to slides based on revised tutorials; mention availability of slides on teaching page; other misc fixes
svn/trunk@4472

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Introduction to Natural Language Processing}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
   University of Pennsylvania, USA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Introduction}

\begin{frame}
  \titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Questions}
  \begin{itemize}
    \item How do we write programs to manipulate natural language?
    \item What questions about language could we answer?
    \item How would the programs work?
    \item What data would they need?
    \item First: what do they look like?
  \end{itemize}
\end{frame}

\subsection{Searching Pronunciation Dictionary}

\begin{frame}[fragile]
  \frametitle{Searching Pronunciation Dictionary}
  \tiny

\begin{verbatim}
    >>> from nltk_lite.corpora import cmudict
    >>> from string import join
    >>> for word, num, pron in cmudict.raw():
    ...     stress_pattern = join(c for c in join(pron) if c in "012")
    ...     if stress_pattern.endswith("1 0 0 0 0"):
    ...         print word, "/", join(pron)
    ACCUMULATIVELY / AH0 K Y UW1 M Y AH0 L AH0 T IH0 V L IY0
    AGONIZINGLY / AE1 G AH0 N AY0 Z IH0 NG L IY0
    CARICATURIST / K EH1 R AH0 K AH0 CH ER0 AH0 S T
    CIARAMITARO / CH ER1 AA0 M IY0 T AA0 R OW0
    CUMULATIVELY / K Y UW1 M Y AH0 L AH0 T IH0 V L IY0
    DEBENEDICTIS / D EH1 B EH0 N AH0 D IH0 K T AH0 S
    DELEONARDIS / D EH1 L IY0 AH0 N AA0 R D AH0 S
    FORMALIZATION / F AO1 R M AH0 L AH0 Z EY0 SH AH0 N
    GIANNATTASIO / JH AA1 N AA0 T AA0 S IY0 OW0
    HYPERSENSITIVITY / HH AY2 P ER0 S EH1 N S AH0 T IH0 V AH0 T IY0
    IMAGINATIVELY / IH2 M AE1 JH AH0 N AH0 T IH0 V L IY0
    INSTITUTIONALIZES / IH2 N S T AH0 T UW1 SH AH0 N AH0 L AY0 Z AH0 Z
    INSTITUTIONALIZING / IH2 N S T AH0 T UW1 SH AH0 N AH0 L AY0 Z IH0 NG
    MANGIARACINA / M AA1 N JH ER0 AA0 CH IY0 N AH0
    SPIRITUALIST / S P IH1 R IH0 CH AH0 W AH0 L AH0 S T
    SPIRITUALISTS / S P IH1 R IH0 CH AH0 W AH0 L AH0 S T S
    SPIRITUALISTS / S P IH1 R IH0 CH AH0 W AH0 L AH0 S S
    SPIRITUALISTS / S P IH1 R IH0 CH AH0 W AH0 L AH0 S
    SPIRITUALLY / S P IH1 R IH0 CH AH0 W AH0 L IY0
    UNALIENABLE / AH0 N EY1 L IY0 EH0 N AH0 B AH0 L
    UNDERKOFFLER / AH1 N D ER0 K AH0 F AH0 L ER0
\end{verbatim}
\end{frame}

\subsection{Minimal Sets from Lexicon}
\begin{frame}[fragile]
  \frametitle{Minimal Sets from Lexicon}
  \tiny
\begin{verbatim}
    >>> from nltk_lite.corpora import shoebox
    >>> from nltk_lite.utilities import MinimalSet
    >>> length, position, min = 4, 1, 3
    >>> lexemes = [field[1].lower() for entry in shoebox.raw('rotokas')
    ...		   for field in entry if field[0] == 'lx']
    >>> ms = MinimalSet()
    >>> for lex in lexemes:
    ...     if len(lex) == length:
    ...         context = lex[:position] + '_' + lex[position+1:]
    ...         target = lex[position]
    ...         ms.add(context, target, lex)
    >>> for context in ms.contexts(3):
    ...     for target in ms.targets():
    ...         print "%-4s" % ms.display(context, target, "-"),
    ...     print
    kasi -    kesi kusi kosi
    kava -    -    kuva kova
    karu kiru keru kuru koru
    kapu kipu -    -    kopu
    karo kiro -    -    koro
    kari kiri keri kuri kori
    kapa -    kepa -    kopa
    kara kira kera -    kora
    kaku -    -    kuku koku
    kaki kiki -    -    koki
\end{verbatim}
\end{frame}

\subsection{Modelling Text Genres}
\begin{frame}[fragile]
  \frametitle{Modelling Text Genres}
  \tiny
\begin{verbatim}
    >>> from nltk_lite.corpora import genesis
    >>> from nltk_lite.probability import ConditionalFreqDist
    >>> from nltk_lite.utilities import print_string
    >>> cfdist = ConditionalFreqDist()
    >>> prev = None
    >>> for word in genesis.raw():
    ...     word = word.lower()
    ...     cfdist[prev].inc(word)
    ...     prev = word
    >>> words = []
    >>> prev = 'lo,'
    >>> for i in range(99):
    ...     words.append(prev)
    ...     for word in cfdist[prev].sorted_samples():
    ...         if word not in words:
    ...             break
    ...     prev = word
    >>> print_string(join(words))
    lo, it came to the land of his father and he said, i will not be a
    wife unto him, saying, if thou shalt take our money in their kind,
    cattle, in thy seed after these are my son from off any more than all
    that is this day with him into egypt, he, hath taken away unawares to
    pass, when she bare jacob said one night, because they were born two
    hundred years old, as for an altar there, he had made me out at her
    pitcher upon every living creature after thee shall come near her:
    yea,
\end{verbatim}
\end{frame}

\subsection{Exploring Syntax}
\begin{frame}[fragile]
  \frametitle{Exploring Syntax}
  \tiny
\begin{verbatim}
    >>> from nltk_lite.corpora import treebank
    >>> from string import join
    >>> def vp_conj(tree):
    ...     if tree.node == 'VP' and len(tree) == 3 and tree[1].leaves() == ['but']:
    ...         return True
    ...     else:
    ...         return False
    >>> for tree in treebank.parsed():
    ...     for vp1,conj,vp2 in tree.subtrees(vp_conj):
    ...         print join(child.node for child in vp1), "*BUT*", join(child.node for child in vp2)
    VBP ADVP-TMP PP-PRD PP *BUT* VBP VP
    VBZ VP *BUT* VBZ NP PP-CLR
    PP-TMP VBZ VP *BUT* VBD ADVP-TMP S
    VBZ SBAR *BUT* VBZ SBAR
    VBD SBAR *BUT* VBD RB VP
    VBD SBAR *BUT* VBD S
    VBP NP-PRD *BUT* VBP RB ADVP-TMP VP
    VBN PP PP-TMP *BUT* ADVP-TMP VBN NP
    MD VP *BUT* VBZ NP SBAR-ADV
    VBD ADVP-CLR *BUT* VBD NP
    VBN NP PP *BUT* VBN NP PP SBAR-PRP
    VBD NP *BUT* MD RB VP
    VBD NP PP-CLR *BUT* VBD PRT NP
    VBZ S *BUT* MD VP
\end{verbatim}
\end{frame}

\section{The Language Challenge}

\subsection{The Richness of Language}

\begin{frame}
\frametitle{The Richness of Language}
\begin{itemize}
\item basic needs and lofty aspirations; technical know-how and
  flights of fantasy
\item ideas are shared over great separations of distance and time
\end{itemize}

\scriptsize

\begin{enumerate}
\item Overhead the day drives level and grey, hiding the sun by a flight
   of grey spears.  (William Faulkner, \textit{As I Lay Dying}, 1935)
\item When using the toaster please ensure that the exhaust fan is turned
   on. (sign in dormitory kitchen)
\item Amiodarone weakly inhibited CYP2C9, CYP2D6, and CYP3A4-mediated
   activities with Ki values of 45.1-271.6 $\mu$M (Medline)
\item Iraqi Head Seeks Arms (spoof headline, \url{http://www.snopes.com/humor/nonsense/head97.htm}
\item The earnest prayer of a righteous man has great power and wonderful
   results. (James 5:16b)
\item Twas brillig, and the slithy toves did gyre and gimble in the wabe
   (Lewis Carroll, \textit{Jabberwocky}, 1872)
\item There are two ways to do this, AFAIK :smile:  (internet discussion archive)
\end{enumerate}
\end{frame}

\begin{frame}[fragile]
\frametitle{Disciplines Studying Language}
\begin{enumerate}
\item linguistics
\item translation
\item literary criticism
\item philosophy
\item anthropology
\item psychology
\item law
\item hermeneutics
\item forensics
\item telephony
\item pedagogy
\item archaeology
\item cryptanalysis
\item speech pathology
\end{enumerate}
\end{frame}

\subsection{Language and the Internet}
\begin{frame}
\frametitle{Language and the Internet}
\small
\begin{itemize}
\item unprecedented volume of information:\\
  \emph{mostly unstructured text}
\item 8 Tb books in 2003
\item 24 hours of scientific literature would take 5 years to read
\item fraction of work/leisure time spent navigating this information
\item a great challenge for natural language processing
\item despite success of web search engines, we need skill, knowledge, and luck to answer the following questions:
  \begin{enumerate}
  \item \textit{What tourist sites can I visit between Philadelphia and Pittsburgh on a
    limited budget?}
  \item \textit{What do expert critics say about Canon digital cameras?}
  \item \textit{What predictions about the steel market were made by
      credible commentators in the past week?}
  \end{enumerate}
\item requires a combination of language processing tasks, e.g.
  information extraction, inference, and summarisation
\end{itemize}
\end{frame}

\subsection{The Promise of NLP}
\begin{frame}
\frametitle{The Promise of NLP}
\begin{itemize}
\item importance in scientific, economic, social and cultural arenas
\item growing rapidly as its theories and methods are deployed in new technologies
\item therefore a wide range of people should have a working knowledge of NLP
  \begin{itemize}
  \item academia: humanities computing, corpus linguistics, computer science, artificial intelligence
  \item industry: HCI, business information analysis, web software development
  \end{itemize}
\item the goal of the book is to open the field of NLP to a broad audience.
\end{itemize}
\end{frame}

\section{Language and Computation}

\subsection{NLP and Intelligence}

\begin{frame}
\frametitle{NLP and Intelligence}
\begin{itemize}
\item long-standing challenge to build intelligent machines
\item chief measure of machine intelligence has been linguistic: Turing test
\item research on spoken dialogue systems, also MT\\
  \emph{--- integrated NLP systems which future users would regard as highly intelligent}
\item Example human-machine dialogue illustrates a typical application:
\small
\begin{tabular}{ll}
S: & How may I help you?\\
U: & When is Saving Private Ryan playing?\\
S: & For what theater?\\
U: & The Paramount theater.\\
S: & Saving Private Ryan is not playing at the Paramount theater,\\
   & but it's playing at the Madison theater at 3:00, 5:30, 8:00, and 10:30. 
\end{tabular}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{NLP and Intelligence (cont)}
\begin{itemize}
\item today's systems limited to narrowly defined domains
\item couldn't ask above system for other information, e.g.:
  \begin{itemize}
  \item driving instructions
  \item details of nearby restaurants
  \end{itemize}
\item to add such support we would have to:
  \begin{itemize}
  \item store the required information
  \item incorporate suitable questions and answers into the system
  \end{itemize}
\item common-sense reasoning vs business logic
\item need to make progress on natural linguistic interaction without
  recourse to this unrestricted knowledge and reasoning capability
\end{itemize}
\end{frame}

\subsection{Language and Symbol Processing}

\begin{frame}
\frametitle{Language and Symbol Processing}
\begin{itemize}
\item origin of the idea that natural language could be treated computationally: philosophy of language work in early 1900s, to reconstruct mathematical reasoning using logic
\item language as a formal system
\item three further developments:
  \begin{enumerate}
  \item formal language theory
  \item symbolic logic
  \item principle of compositionality
  \end{enumerate}
\item more recent developments:
  \begin{enumerate}
  \item data-intensive NLP
  \item machine learning in NLP
  \item evaluation-led methodologies
  \end{enumerate}
\item many interesting philosophical issues (see book)
\item key: balancing act between symbolic and statistical approaches
\end{itemize}
\end{frame}

\subsection{The Balancing Act}

\begin{frame}
\frametitle{Web as Corpus: Absolutely vs Definitely}
{\small
\begin{tabular}{l|llll}
\textbf{Google hits}
& \textbf{adore}
& \textbf{love}
& \textbf{like}
& \textbf{prefer} \\ \hline
absolutely  & 289,000 & 905,000 & 16,200  & 644 \\
definitely  & 1,460   & 51,000  & 158,000 & 62,600 \\
ratio       & 198/1   & 18/1    & 1/10    & 1/97
\end{tabular}}
\begin{itemize}
\item useful information for statistical language models
\item statistical evidence for binary-valued features in lexical items
\end{itemize}
\end{frame}

\section{The Python Programming Language}

\subsection{Key Features}

\begin{frame}
\frametitle{Python: Key Features}
\begin{itemize}
\item simple yet powerful, shallow learning curve
\item object-oriented: encapsulation, re-use
\item scripting language, facilitates interactive exploration
\item excellent functionality for processing linguistic data
\item extensive standard library, incl graphics, web, numerical processing
\item downloaded for free from \url{http://www.python.org/}
\end{itemize}
\end{frame}

\subsection{Python Example}
\begin{frame}[fragile]
\frametitle{Python Example}

\begin{verbatim}
 import sys
 for line in sys.stdin.readlines():
     for word in line.split():
         if word.endswith('ing'):
             print word
\end{verbatim}

\begin{enumerate}
\item whitespace: nesting lines of code; scope
\item object-oriented: attributes, methods (e.g. \texttt{line})
\item readable
\end{enumerate}
\end{frame}

\subsection{Comparison with Perl}
\begin{frame}[fragile]
\frametitle{Comparison with Perl}

\begin{verbatim}
  while (<>) {
      foreach my $word (split) {
          if ($word =~ /ing$/) {
              print "$word\n";
          }
      }
  }
\end{verbatim}

\begin{enumerate}
\item syntax is obscure: \textit{what are:} \verb|<> $ my split| ?
\item ``it is quite easy in Perl to write programs that simply
  look like raving gibberish, even to experienced Perl programmers''
  (Hammond \textit{Perl Programming for Linguists} 2003:47)
\item large programs difficult to maintain, reuse
\end{enumerate}
\end{frame}

\subsection{What NLTK adds to Python}
\begin{frame}
\frametitle{What NLTK adds to Python}

NLTK defines a basic infrastructure that can be used to build NLP
programs in Python.  It provides:

\begin{itemize}
\item Basic classes for representing data relevant to natural language
  processing
\item Standard interfaces for performing tasks, such
  as tokenization, tagging, and parsing
\item Standard implementations for each task, which
  can be combined to solve complex problems
\item Extensive documentation, including tutorials
  and reference documentation
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Installing Python and NLTK}
\begin{enumerate}
\item Install Python, Numeric
\item Install NLTK-Lite, NLTK-Lite-Corpora
\item Optional: Matplotlib, WordNet
\item Set environment variable \verb|NLTK_LITE_CORPORA|
\end{enumerate}

\textit{For detailed instructions, see:}

\begin{itemize}
\item  http://nltk.sourceforge.net/install.html
\item  CDROM: /install.html
\end{itemize}

\end{frame}

\end{document}

### 2007-05-03 slides/programming.tex
updates to slides based on revised tutorials; mention availability of slides on teaching page; other misc fixes
svn/trunk@4472

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Programming Fundamentals and Python}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
  University of Pennsylvania, USA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Introduction: Programming Basics}

\begin{frame}
  \titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
    \item non-technical overview
    \item many working program fragments
    \item try them for yourself as we go along
    \item many online tutorials (see \url{www.python.org})
    \item Textbook: Zelle, John (2004) \textit{Python Programming: An Introduction
        to Computer Science}
  \end{itemize}
\end{frame}

\subsection{Lists}

\begin{frame}[fragile]
  \frametitle{Defining Lists}
  \small

\begin{itemize}
\item list: ordered sequence of items
\item item: string, number, complex object (e.g. a list)
\item list representation: comma separated items: \verb|['John', 14, 'Sep', 1984]|
\item list initialization:

\begin{verbatim}
    >>> a = ['colourless', 'green', 'ideas']
\end{verbatim}

\item sets the value of variable \texttt{a}
\item to see the its value, do: \texttt{print a}
\item in interactive mode, just type the variable name:

\begin{verbatim}
  >>> a
  ['colourless', 'green', 'ideas']
\end{verbatim}
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Simple List Operations}
  \small

\begin{enumerate}
\item length: \texttt{len()}
\item indexing: \texttt{a[0]}, \texttt{a[1]}
\item indexing from right: \texttt{a[-1]}
\item slices: \texttt{a[1:3]}, \texttt{a[-2:]}
\item concatenation: \texttt{b = a + ['sleep', 'furiously']}
\item sorting: \texttt{b.sort()}
\item reversing: \texttt{b.reverse()}
\item iteration: \texttt{for item in a:}
\item all the above applies to strings as well
\item double indexing: \texttt{b[2][1]}
\item finding index: \texttt{b.index('green')}
\end{enumerate}
\end{frame}

\subsection{Strings}

\begin{frame}
  \frametitle{Simple String Operations}
  \small

\begin{enumerate}
\item joining: \texttt{c = ' '.join(b)}
\item splitting: \texttt{c.split('r')}
\item lambda expressions: \texttt{lambda x: len(x)}
\item maps: \texttt{map(lambda x: len(x), b)}
\item list comprehensions: \texttt{[(x, len(x)) for x in b]}
\item getting help: \texttt{help(list)}, \texttt{help(str)}
\end{enumerate}
\end{frame}

\subsection{Dictionaries}

\begin{frame}[fragile]
  \frametitle{Dictionaries}
  \small

\begin{itemize}
\item accessing items by their names, e.g. dictionary
\item defining entries:

\begin{verbatim}
  >>> d = {}
  >>> d['colourless'] = 'adj'
  >>> d['furiously'] = 'adv'
  >>> d['ideas'] = 'n'
\end{verbatim}

\item accessing:

\begin{verbatim}
  >>> d.keys()
  ['furiously', 'colourless', 'ideas']
  >>> d['ideas']
  'n'
  >>> d
  {'furiously': 'adv', 'colourless': 'adj', 'ideas': 'n'}
\end{verbatim}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Dictionaries: Iteration}

\begin{verbatim}
  >>> for w in d:
  ...    print "%s [%s]," % (w, d[w]),
  furiously [adv], colourless [adj], ideas [n],
\end{verbatim}

\begin{itemize}
\item rule of thumb: dictionary entries are like variable names
\item \textit{create} them by assigning to them\\
  \verb|x = 2| (variable), \verb|d['x'] = 2| (dictionary entry)
\item \textit{access} them by reference\\
  \verb|print x| (variable), \verb|print d['x']| (dictionary entry)
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Dictionaries: Example: Counting Word Occurrences}
\small

\begin{verbatim}
  >>> from nltk_lite.corpora import gutenberg
  >>> count = {}
  >>> for word in gutenberg.raw('shakespeare-macbeth'):
  ...     word = word.lower()
  ...     if word not in count:
  ...         count[word] = 0
  ...     count[word] += 1
\end{verbatim}

Now inspect the dictionary:

\begin{verbatim}
  >>> print count['scotland']
  12
  >>> frequencies = [(freq, word) for (word, freq) in count.items()]
  >>> frequencies.sort()
  >>> frequencies.reverse()
  >>> print frequencies[:20]
  [(1986, ','), (1245, '.'), (692, 'the'), (654, "'"), (567, 'and'), (482, ':'), (399, 'to'), (365, 'of'), (360, 'i'), (255, 'a'), (246, 'that'), (242, '?'), (224, 'd'), (218, 'you'), (213, 'in'), (207, 'my'), (198, 'is'), (170, 'not'), (165, 'it'), (156, 'with')]
\end{verbatim}
\end{frame}

\section{Regular Expressions}

\subsection{Introduction}

\begin{frame}
\frametitle{Regular Expressions}
\begin{itemize}
\item string matching
\item substitution
\item patterns, classes
\item Python's regular expression module: \texttt{re}
\item NLTK's utility function: \texttt{re\_show}
\end{itemize}
\end{frame}

\subsection{Matching and Substituting}

\begin{frame}[fragile]
\frametitle{Loading module, Matching}

\begin{itemize}
\item Set up:

\begin{verbatim}
  >>> import re
  >>> from nltk_lite.utilities import re_show
  >>> sent = "colourless green ideas sleep furiously"
\end{verbatim}

\item Matching:

\begin{verbatim}
  >>> re_show('l', sent)
  co{l}our{l}ess green ideas s{l}eep furious{l}y
  >>> re_show('green', sent)
  colourless {green} ideas sleep furiously
\end{verbatim}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Substitutions}

\begin{itemize}
\item E.g. replace all instances of \texttt{l} with \texttt{s}.
\item Creates an output string (doesn't modify input)

\begin{verbatim}
  >>> re.sub('l', 's', sent)
  'cosoursess green ideas sseep furioussy'
\end{verbatim}

\item Work on substrings (NB not words)

\begin{verbatim}
  >>> re.sub('green', 'red', sent)
  'colourless red ideas sleep furiously'
\end{verbatim}
\end{itemize}
\end{frame}

\subsection{More Complex Patterns}

\begin{frame}[fragile]
\frametitle{More Complex Patterns}

\begin{itemize}
\item Disjunction:

\begin{verbatim}
  >>> re_show('(green|sleep)', sent)
  colourless {green} ideas {sleep} furiously
  >>> re.findall('(green|sleep)', sent)
  ['green', 'sleep']
\end{verbatim}

\item Character classes, e.g. non-vowels followed by vowels:

\begin{verbatim}
  >>> re_show('[^aeiou][aeiou]', sent)
  {co}{lo}ur{le}ss g{re}en{ i}{de}as s{le}ep {fu}{ri}ously
  >>> re.findall('[^aeiou][aeiou]', sent)
  ['co', 'lo', 'le', 're', ' i', 'de', 'le', 'fu', 'ri']
\end{verbatim}
\end{itemize}
\end{frame}

\subsection{Structured Results}

\begin{frame}[fragile]
\frametitle{Structured Results}
\begin{itemize}
\item Select a sub-part to be returned
\item e.g. non-vowel characters which appear before a vowel:

\begin{verbatim}
  >>> re.findall('([^aeiou])[aeiou]', sent)
  ['c', 'l', 'l', 'r', ' ', 'd', 'l', 'f', 'r']
\end{verbatim}

\item generate \textit{tuples}, for later tabulation

\begin{verbatim}
  >>> re.findall('([^aeiou])([aeiou])', sent)
  [('c', 'o'), ('l', 'o'), ('l', 'e'), ('r', 'e'), (' ', 'i'), ('d', 'e'), ('l', 'e'), ('f', 'u'), ('r', 'i')]
\end{verbatim}
\end{itemize}
\end{frame}

\section{Accessing Data and Programs}

\subsection{Accessing Files}

\begin{frame}[fragile]
\frametitle{Accessing Files and the Web}
\begin{itemize}
\item accessing local files (create \texttt{corpus.txt} first)

\begin{verbatim}
  >>> print open('corpus.txt').read() 
  Hello world.  This is a test file.
\end{verbatim}

\item Accessing URLs on the Web:

\begin{verbatim}
  >>> from urllib import urlopen
  >>> page = urlopen("http://news.bbc.co.uk/").read()
  >>> page = re.sub('<[^>]*>', '', page)
  >>> page = re.sub('\s+', ' ', page)
  >>> print page[:60]
   BBC NEWS | News Front Page News Sport Weather World Service
\end{verbatim}
\end{itemize}
\end{frame}

\subsection{Accessing NLTK}

\begin{frame}[fragile]
\frametitle{Accessing NLTK}

\begin{itemize}
\item modules: classes, functions
\item data structures, algorithms
\item importing, e.g. \texttt{from nltk\_lite.utilities import *}
\item conflicts
\item explicit naming: \texttt{from nltk\_lite.utilities import re\_show}
\item higher level module import:

\begin{verbatim}
  >>> from nltk_lite import utilities
  >>> utilities.re_show('green', s)
  colourless {green} ideas sleep furiously
\end{verbatim}
\end{itemize}
\end{frame}

\subsection{Accessing Corpora}

\begin{frame}[fragile]
\frametitle{Texts from Project Gutenberg}

\begin{verbatim}
  >>> from nltk_lite.corpora import gutenberg
  >>> gutenberg.items
  ['austen-emma', 'austen-persuasion', 'austen-sense', 'bible-kjv', 'blake-poems', 'blake-songs', 'chesterton-ball', 'chesterton-brown', 'chesterton-thursday', 'milton-paradise', 'shakespeare-caesar', 'shakespeare-hamlet', 'shakespeare-macbeth', 'whitman-leaves']
  >>> count = 0
  >>> for word in gutenberg.raw('whitman-leaves'):
  ...     count += 1
  >>> print count
  154873
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Brown Corpus}
\small

\begin{verbatim}
  >>> from nltk_lite.corpora import brown
  >>> brown.items
  ['a', 'b', 'c', 'd', 'e', 'f', 'g', 'h', 'j', 'k', 'l', 'm', 'n', 'p', 'r']
  >>> from nltk_lite.corpora import extract
  >>> print extract(0, brown.raw())
  ['The', 'Fulton', 'County', 'Grand', 'Jury', 'said', 'Friday', 'an', 'investigation', 'of', "Atlanta's", 'recent', 'primary', 'election', 'produced', '``', 'no', 'evidence', "''", 'that', 'any', 'irregularities', 'took', 'place', '.']
  >>> print extract(0, brown.tagged())
  [('The', 'at'), ('Fulton', 'np-tl'), ('County', 'nn-tl'), ('Grand', 'jj-tl'), ('Jury', 'nn-tl'), ('said', 'vbd'), ('Friday', 'nr'), ('an', 'at'), ('investigation', 'nn'), ('of', 'in'), ("Atlanta's", 'np$'), ('recent', 'jj'), ('primary', 'nn'), ('election', 'nn'), ('produced', 'vbd'), ('``', '``'), ('no', 'at'), ('evidence', 'nn'), ("''", "''"), ('that', 'cs'), ('any', 'dti'), ('irregularities', 'nns'), ('took', 'vbd'), ('place', 'nn'), ('.', '.')]
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Penn Treebank}
\small

\begin{verbatim}
  >>> from nltk_lite.corpora import treebank
  >>> print extract(0, treebank.parsed())
  (S:
    (NP-SBJ:
      (NP: (NNP: 'Pierre') (NNP: 'Vinken'))
      (,: ',')
      (ADJP: (NP: (CD: '61') (NNS: 'years')) (JJ: 'old'))
      (,: ','))
    (VP:
      (MD: 'will')
      (VP:
        (VB: 'join')
        (NP: (DT: 'the') (NN: 'board'))
        (PP-CLR:
          (IN: 'as')
          (NP: (DT: 'a') (JJ: 'nonexecutive') (NN: 'director')))
        (NP-TMP: (NNP: 'Nov.') (CD: '29'))))
    (.: '.'))
\end{verbatim}
\end{frame}

\end{document}

### 2007-05-03 slides/tag.tex
updates to slides based on revised tutorials; mention availability of slides on teaching page; other misc fixes
svn/trunk@4472

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Tagging}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
  University of Pennsylvania, USA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\frame{\titlepage}

\section{Introduction}

\subsection{Parts of Speech}

\begin{frame}
  \frametitle{Parts of speech}

  \begin{itemize}
  \item How can we predict the bahaviour of a previously unseen word?
  \item Words can be divided into classes that behave similarly.
  \item Traditionally eight parts of speech: noun, verb,
    pronoun, preposition, adverb, conjunction, adjective and article.
  \item More recently larger sets have been used: eg Penn Treebank (45
    tags), Susanne (353 tags).
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Parts of Speech}

  \begin{alertblock}{What use are parts of speech?}
    They tell us a lot about a word (and the words near it).
  \end{alertblock}

 g \pause
  \begin{itemize}
  \item Tell us what words are likely to occur in the neighbourhood
    (eg adjectives often followed by nouns, personal pronouns often
    followed by verbs, possessive pronouns by nouns)
  \item Pronunciations can be dependent on part of speech, eg
    \textcolor{green}{object, content, discount} (useful for speech
    synthesis and speech recognition)
  \item Can help information retrieval and extraction (stemming,
    partial parsing) 
  \item Useful component in many NLP systems
  \end{itemize}

\end{frame}

\subsection{Open and closed classes}

\begin{frame}
  \frametitle{Closed and open classes}
  \begin{itemize}
  \item Parts of speech may be categorised as \emph{open} or
    \emph{closed} classes
  \item Closed classes have a fixed membership of words (more or
    less), eg determiners, pronouns, prepositions
  \item Closed class words are usually \emph{function words} ---
    frequently occurring, grammatically important, often short (eg
    \textcolor{green}{of,it,the,in})
  \item The major open classes are \emph{nouns}, \emph{verbs},
    \emph{adjectives} and \emph{adverbs}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Closed classes in English}
  \begin{tabular}{ll}
  \textbf{prepositions} & on, under, over, to, with, by\\
  \textbf{determiners} & the, a, an, some\\
  \textbf{pronouns} & she, you, I, who\\
  \textbf{conjunction}s & and, but, or, as, when, if\\
  \textbf{auxiliary} verbs & can, may, are\\
  \textbf{particles} & up, down, at, by\\
  \textbf{numeral}s & one, two, first, second\\
  \end{tabular}
\end{frame}

\begin{frame}
  \frametitle{Open classes}
  \begin{description}
  \item[nouns] Proper nouns (\textcolor{green}{Scotland}, \textcolor{green}{BBC}),\\
    common nouns:
    \begin{itemize}
    \item count nouns (\textcolor{green}{goat},
      \textcolor{green}{glass})
    \item mass nouns (\textcolor{green}{snow},
      \textcolor{green}{pacifism})
    \end{itemize}
  \item[verbs] actions and processes (\textcolor{green}{run},
    \textcolor{green}{hope}), also auxiliary verbs
  \item[adjectives] properties and qualities (age, colour, value)
  \item[adverbs] modify verbs, or verb phrases, or other adverbs:\\
    \textcolor{green}{\emph{Unfortunately} John walked \emph{home
        extremely slowly yesterday}}
  \end{description}
\end{frame}

\subsection{Tagsets}

\begin{frame}
  \frametitle{The Penn Treebank tagset (1)}
  {\footnotesize
    \begin{tabular}{|lll|lll|}
      \hline
      CC & Coord Conjuncn & \emph{and,but,or} &  
      NN & Noun, sing. or mass & \emph{dog} \\
      CD & Cardinal number & \emph{one,two} & 
      NNS & Noun, plural & \emph{dogs} \\
      DT & Determiner & \emph{the,some} &  
      NNP & Proper noun, sing. & \emph{Edinburgh} \\
      EX & Existential there & \emph{there} &  
      NNPS & Proper noun, plural & \emph{Orkneys} \\
      FW & Foreign Word & \emph{mon dieu} &  
      PDT & Predeterminer & \emph{all, both}\\
      IN & Preposition & \emph{of,in,by} & 
      POS & Possessive ending & \emph{'s} \\
      JJ & Adjective & \emph{big} & 
      PP & Personal pronoun & \emph{I,you,she} \\
      JJR & Adj., comparative & \emph{bigger} & 
      PP\$ & Possessive pronoun & \emph{my,one's} \\
      JJS & Adj., superlative & \emph{biggest} &  
      RB & Adverb & \emph{quickly} \\
      LS & List item marker & \emph{1,One} &  
      RBR & Adverb, comparative & \emph{faster} \\
      MD & Modal & \emph{can,should} &
      RBS & Adverb, superlative & \emph{fastest} \\
      \hline
    \end{tabular}
  }
\end{frame}

\begin{frame}
  \frametitle{The Penn Treebank tagset (2)}
  {\footnotesize
    \begin{tabular}{|lll|lll|}
      \hline
      RP & Particle & \emph{up,off} &
      WP\$ & Possessive-Wh & \emph{whose} \\ 
      SYM & Symbol & \emph{+,\%,\&}  &
      WRB & Wh-adverb & \emph{how,where} \\
      TO & ``to'' & \emph{to} &
      \$ & Dollar sign & \emph{\$}\\
      UH & Interjection & \emph{oh, oops} &
      \# & Pound sign & \emph{\#} \\
      VB & verb, base form &     \emph{eat} &
      `` & Left quote & ` , `` \\
      VBD & verb, past tense & \emph{ate} &
      '' & Right quote & ', '' \\
      VBG & verb, gerund & \emph{eating} &
      ( & Left paren & ( \\
      VBN & verb, past part & \emph{eaten} &
      ) & Right paren & ) \\
      VBP & Verb, non-3sg, pres & \emph{eat} &
      , & Comma & , \\
      VBZ & Verb, 3sg, pres & \emph{eats} &
      . & Sent-final punct & . ! ? \\
      WDT & Wh-determiner &  \emph{which,that} &
      : & Mid-sent punct. & : ; --- ... \\
      WP & Wh-pronoun & \emph{what,who} & & & \\
      \hline
    \end{tabular}
  }
\end{frame}

\subsection{Tagging}

\begin{frame}
  \frametitle{Tagging}
  \begin{itemize}
  \item Definition: POS Tagging is the assignment of a single
    part-of-speech tag to each word (and punctuation marker) in a
    corpus.  For example:\\
    {\scriptsize
    \textcolor{green}{``/``  The/DT  guys/NNS  that/WDT  make/VBP
      traditional/JJ  hardware/NN  are/VBP  really/RB  being/VBG
      obsoleted/VBN  by/IN  microprocessor-based/JJ  machines/NNS  ,/,
      ''/''  said/VBD  Mr./NNP  Benton/NNP  ./.}}
  \item Non-trivial: POS tagging must resolve ambiguities since the
    same word can have different tags in different contexts
  \item In the Brown corpus 11.5\% of word types and 40\% of word
    tokens are ambiguous
  \item In many cases one tag is much more likely for a given word
    than any other
  \item Limited scope: only supplying a tag for each word, no larger
    structures created (eg prepositional phrase attachment)
  \end{itemize}  
\end{frame}

%There                  EX
%are                    VBP
%11                     CD
%players                        NNS
%in                     IN
%a                      DT
%football               NN
%team                   NN
%.                      .

\begin{frame}
  \frametitle{Information sources for tagging}

  What information can help decide the correct PoS tag for a word?
  \begin{description}
  \item[Other PoS tags] Even though the PoS tags of other words may be
    uncertain too, we can use information that some tag sequences are
    more likely than others (eg \emph{the/AT red/JJ drink/NN} vs \emph{the/AT
      red/JJ drink/VBP}).\newline
    Using \emph{only} information about the most likely PoS tag
    sequence does not result in an accurate tagger (about 77\%
    correct)
  \item[The word identity] Many words can gave multiple possible tags,
    but some are more likely than others (eg \emph{fall/VBP} vs
    \emph{fall/NN}) \newline
    Tagging each word with its most common tag results in a tagger
    with about 90\% accuracy
  \end{description}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Simple Taggers and Evaluation}

\subsection{Simple taggers}

\begin{frame}[fragile]
  \frametitle{Tagging in NLTK}
The simplest possible tagger tags everything as a noun:
{\small
\begin{verbatim}
from nltk_lite import tokenize
text = 'There are 11 players in a football team'
text_tokens = list(tokenize.whitespace(text))
# ['There', 'are', '11', 'players', 'in', 'a', 'football', 'team']
\end{verbatim}
  \pause
\begin{verbatim}
from nltk_lite import tag
mytagger = tag.Default('nn')
for t in mytagger.tag(text_tokens):
    print t
# ('There', 'NN')
# ('are', 'NN')
# ...
\end{verbatim}}
\end{frame}

\begin{frame}[fragile]
  \frametitle{A regular expression tagger}
We can use regular expressions to tag tokens based on regularities in
the text, eg numerals:

{\small
\begin{verbatim}
default_pattern = (r'.*', 'NN')
cd_pattern = (r' ^[0-9]+(.[0-9]+)?$', 'CD')
patterns = [cd_pattern, default_pattern]
NN_CD_tagger = tag.Regexp(patterns)
re_tagged = list(NN_CD_tagger.tag(text_tokens))
# [('There', 'NN'), ('are', 'NN'), ('11', 'NN'), ('players', 'NN'), 
('in', 'NN'), ('a', 'NN'), ('football', 'NN'), ('team', 'NN')]
\end{verbatim}
}%$
\end{frame}

\subsection{Unigram taggers}

\begin{frame}[fragile]
  \frametitle{A unigram tagger}
The NLTK UnigramTagger class implements a tagging algorithm based on a
table of unigram probabilities:
\[ \mbox{tag}(w) = \arg\max_{t_i} P(t_i|w) \]

\pause
Training a UnigramTagger on the Penn Treebank:
{\small
\begin{verbatim}
from nltk_lite.corpora import treebank
from itertools import islice

# sentences 0-2999
train_sents = list(islice(treebank.tagged(), 3000))
# from sentence 3000 to the end
test_sents = list(islice(treebank.tagged(), 3000, None))

unigram_tagger = tag.Unigram()
unigram_tagger.train(train_sents)
\end{verbatim}}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Unigram tagging}
{\small
\begin{verbatim}
>>> list(unigram_tagger.tag(tokenize.whitespace("Mr. Jones saw 
the book on the shelf")))
[('Mr.', 'NNP'), ('Jones', 'NNP'), ('saw', 'VBD'), ('the', 'DT'), 
('book', 'NN'), ('on', 'IN'), ('the', 'DT'), ('shelf', None)]
\end{verbatim}}

  The UnigramTagger assigns the default tag \texttt{None} to words
  that are not in the training data (eg \emph{shelf})

  \pause
  We can combine taggers to ensure every word is tagged:
{\small
\begin{verbatim}
>>> unigram_tagger = tag.Unigram(cutoff=0,backoff=NN_CD_tagger)
>>> unigram_tagger.train(train_sents)
>>> list(unigram_tagger.tag(tokenize.whitespace("Mr. Jones saw 
the book on the shelf")))
[('Mr.', 'NNP'), ('Jones', 'NNP'), ('saw', 'VBD'), ('the', 'DT'), 
('book', 'VB'), ('on', 'IN'), ('the', 'DT'), ('shelf', 'NN')]
\end{verbatim}}
\end{frame}

\subsection{Evaluating taggers}

\begin{frame}
  \frametitle{Evaluating taggers}
  \begin{itemize}
  \item Basic idea: compare the output of a tagger with a
    human-labelled \emph{gold standard}
  \item Need to compare how well an automatic method does with the
    agreement between people
  \item The best automatic methods have an accuracy of about 96-97\%
    when using the (small) Penn treebank tagset (but this is still an
    average of one error every couple of sentences...)
  \item Inter-annotator agreement is also only about 97\% 
  \item A good unigram baseline (with smoothing) can obtain 90-91\%!
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Evaluating taggers in NLTK}
  NLTK provides a function \texttt{tag.accuracy} to automate
  evaluation.  It needs to be provided with a tagger, together with
  some text to be tagged and the gold standard tags.

  \pause
  We can make
  print more prettily:
\begin{verbatim}
def print_accuracy(tagger, data):
    print '%3.1f%%' % (100 * tag.accuracy(tagger, data))
\end{verbatim}

  \pause
\begin{verbatim}
>>> print_accuracy(NN_CD_tagger, test_sents)
15.0%
>>> print_accuracy(unigram_tagger, train_sents)
93.8%
>>> print_accuracy(unigram_tagger, test_sents)
82.8%
\end{verbatim}
\end{frame}

\subsection{Error analysis}

\begin{frame}
  \frametitle{Error analysis}
  \begin{itemize}
  \item The \% correct score doesn't tell you everything --- it is
    useful know what is misclassified as what
  \item \emph{Confusion matrix}: A matrix (ntags x ntags) where the rows
    correspond to the correct tags and the columns correspond to the
    tagger output.  Cell $(i,j)$ gives the count of the number of
    times tag $i$ was classified as tag $j$
  \item The leading diagonal elements correspond to correct
    classifications
  \item Off diagonal elements correspond to misclassifications 
  \item Thus a confusion matrix gives information on the major
    problems facing a tagger (eg NNP vs. NN vs. JJ)
  \item See section 3 of the NLTK tutorial on Tagging
  \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{More Powerful Taggers}

\subsection{N-gram Taggers}

\begin{frame}
  \frametitle{N-gram taggers}
  \begin{itemize}
  \item Basic idea:  Choose the tag  that maximises:
    \[ P(\mbox{word} | \mbox{tag}) \cdot 
       P(\mbox{tag} | \mbox{previous n tags}) \]

  \item For a bigram model the best tag at position $i$ is:
    \[ t_i = \arg\max_{t_j} P(w_i|t_j) P(t_j | t_{i-1}) \]
    Assuming that you know the previous tag, $t_{i-1}$.
  \item Interpretation: choose the tag $t_i$ that is most likely to
    \alert{generate} word $w_i$ given that the previous tag was $t_{i-1}$ 
  \end{itemize}
\end{frame}

\pgfdeclareimage[width=8cm]{tag-context}{../images/tag-context}
\begin{frame}
  \frametitle{N-gram taggers}
  \pgfuseimage{tag-context}
\end{frame}

\begin{frame}
  \frametitle{Example (J+M, p304)}
  {\small
  Secretariat/NNP is/VBZ expected/VBZ to/TO \alert{race/VB} tomorrow/NN

  \medskip
  People/NNS continue/VBP to/TO inquire/VB the/DT reason/NN for/IN the/DT
  \alert{race/NN} for/IN outer/JJ space/NN
  }

  \begin{itemize}
  \item<2-> ``race'' is a verb in the first, a noun in the second.
  \item<3-> Assume that race is the only untagged word, so we can assume
    the tags of the others.
  \item<4-> Probabilities of ``race'' being a verb, or race being a noun
    in the first example:
    \begin{align*}
      P(\mbox{race is }VB) &= P(VB|TO) P(\mbox{race}|VB) \\
      P(\mbox{race is }NN) &= P(NN|TO) P(\mbox{race}|NN) 
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Example (continued)}
  \begin{align*}
    P(NN|TO) &= 0.021 \\
    P(VB|TO) &= 0.34 \\
    \\
    P(\mbox{race}|NN) &= 0.00041\\
    P(\mbox{race}|VB) &= 0.00003\\
    \\
    P(\mbox{race is }VB) &= P(VB|TO) P(\mbox{race}|VB) \\
    &= 0.34 \times 0.00003 = 0.00001 \\
    P(\mbox{race is }NN) &= P(NN|TO) P(\mbox{race}|NN) \\
    &= 0.021 \times 0.00041 = 0.000007
   \end{align*}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Simple bigram tagging in NLTK}
{\small
\begin{verbatim}
>>> default_pattern = (r'.*', 'NN')
>>> cd_pattern = (r' ^[0-9]+(.[0-9]+)?$', 'CD')
>>> patterns = [cd_pattern, default_pattern]
>>> NN_CD_tagger = tag.Regexp(patterns)
>>> unigram_tagger = tag.Unigram(cutoff=0, backoff=NN_CD_tagger)
>>> unigram_tagger.train(train_sents)
>>> bigram_tagger = tag.Bigram(backoff=unigram_tagger)
>>> bigram_tagger.train(train_sents)
\end{verbatim}%$

\begin{verbatim}
>>> print_accuracy(bigram_tagger, train_sents)
95.6%
>>> print_accuracy(bigram_tagger, test_sents)
84.2%
\end{verbatim}
}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Limitation of NLTK n-gram taggers}

  \begin{itemize}
  \item Does not find the most likely sequence of tags, simply works
    left to right always assigning the most probable single tag (given
    the previous tag assignments)
  \item Does not cope with zero probability problem well (no smoothing
    or discounting)
  \item see module \verb|nltk_lite.tag.hmm|
  \end{itemize}
\end{frame}

\subsection{The Brill Tagger}

\begin{frame}
  \frametitle{Brill Tagger}
  \begin{itemize}
  \item Problem with n-gram taggers: size
  \item A rule-based system...
  \item ...but the rules are learned from a corpus
  \item Basic approach: start by applying general rules, then
    successively refine with additional rules that correct the
    mistakes (painting analogy)
  \item Learn the rules from a corpus, using a set of rule templates,
    eg:\\
    Change tag \textbf{a} to \textbf{b} when the following word is
    tagged \textbf{z}
  \item Choose the best rule each iteration    
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Brill Tagger: Example}

{\scriptsize
\begin{tabular}{lllp{4cm}p{4cm}}
\textbf{Sentence}
& \textbf{Gold}
& \textbf{Unigram}
& \textbf{Replace \texttt{nn} with \texttt{vb}}
& \textbf{Replace \texttt{to} with \texttt{in}} \\
&&& \textbf{when the previous word is \texttt{to}}
& \textbf{when the next tag is \texttt{nns}} \\ \hline
The            & at      & at \\
President      & nn-tl   & nn-tl \\
said           & vbd     & vbd \\
he             & pps     & pps \\
will           & md      & md \\
ask            & vb      & vb \\
Congress       & np      & np \\
to             & to      & to \\
increase       & vb      & \textit{nn}     &    \textit{vb} \\
grants         & nns     & nns \\
to             & in      & \textit{to}     &    \textit{to} &   \textit{in} \\
states         & nns     & nns \\
for            & in      & in \\
vocational     & jj      & jj \\
rehabilitation & nn      & nn
\end{tabular}
}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Summary}

\begin{frame}
  \frametitle{Summary}

  \begin{itemize}
  \item \textbf{Reading:} 
    Jurafsky and Martin, chapter 8 (esp. sec 8.5); 
    Manning and Sch\"utze, chapter 10; 
  \item Rule-based and statistical tagging
  \item HMMs and n-grams for statistical tagging
  \item Operation of a simple bigram tagger
  \item TnT --- an accurate trigram-based tagger
  \end{itemize}
\end{frame}

\end{document}

### 2007-05-08 slides/tag.tex
tweaks to confusion matrix code (replaced dict with set); initial notes on confusion matrix added to tag.txt
svn/trunk@4481

@@
 
- g \pause
+  \pause
   \begin{itemize}
@@
   \frametitle{Closed classes in English}
+  \textcolor{black}{
   \begin{tabular}{ll}
@@
   \textbf{conjunction}s & and, but, or, as, when, if\\
-  \textbf{auxiliary} verbs & can, may, are\\
+  \textbf{auxiliary verbs} & can, may, are\\
   \textbf{particles} & up, down, at, by\\
   \textbf{numeral}s & one, two, first, second\\
-  \end{tabular}
+  \end{tabular}}
 \end{frame}
@@
   \frametitle{Open classes}
+  \textcolor{black}{
   \begin{description}
@@
         extremely slowly yesterday}}
-  \end{description}
+  \end{description}}
 \end{frame}
@@
   \frametitle{The Penn Treebank tagset (1)}
-  {\footnotesize
+  \textcolor{black}{\footnotesize
     \begin{tabular}{|lll|lll|}
@@
   \frametitle{The Penn Treebank tagset (2)}
-  {\footnotesize
+  \textcolor{black}{\footnotesize
     \begin{tabular}{|lll|lll|}
@@
   \frametitle{Information sources for tagging}
-
+  \textcolor{black}{
   What information can help decide the correct PoS tag for a word?
@@
     with about 90\% accuracy
-  \end{description}
+  \end{description}}
 \end{frame}
@@
   \frametitle{Tagging in NLTK}
-The simplest possible tagger tags everything as a noun:
-{\small
+\textcolor{black}{The simplest possible tagger tags everything as a noun:}
+\small
 \begin{verbatim}
@@
 # ...
-\end{verbatim}}
+\end{verbatim}
 \end{frame}
@@
   \frametitle{A regular expression tagger}
+\textcolor{black}{
 We can use regular expressions to tag tokens based on regularities in
-the text, eg numerals:
-
-{\small
+the text, eg numerals:}
+\small
 \begin{verbatim}
@@
 \end{verbatim}
-}%$
 \end{frame}

### 2007-05-09 rst.py
doc/en/introduction.txt * polishing * added index terms * systematic use of numbered examples * adjusted table dimensions * consistent use of double quotes * citations to SPE, field-wide surveys
doc/en/preface-nltk.txt:
doc/en/chunk.txt:
doc/en/nlp-python.txt:
nltk_lite/*/*.py
* removed redundant .readlines()

doc/images/dialogue.graffle
doc/images/dialogue.png
* new diagram, rotated 90% for better aspect ratio
* incorporated linguistic module names

doc/book.bib
* citations to SPE, field-wide surveys

doc/definitions.txt
nltk_lite/__init__.py
* changed to version 0.8 beta 1

doc/rst.py
* adde

@@ -75,7 +75,7 @@ EXTERN_REFERENCE_FILES = []
 """A list of .ref files, for crossrefering to external documents (used
    when building one chapter at a time)."""
 
-BIBTEX_FILE = 'book.bib'
+BIBTEX_FILE = '../book.bib'
 """The name of the bibtex file used to generate bibliographic entries.
    """
 
@@ -544,6 +544,10 @@ class Citations(Transform):
             m = re.match(r'(?i)author\s*=\s*(.*)$', line)
             if m and key:
                 bibliography[key][0] = self.bib_authors(m.group(1))
+            else:
+                m = re.match(r'(?i)editor\s*=\s*(.*)$', line)
+                if m and key:
+                    bibliography[key][0] = self.bib_authors(m.group(1))
                 
             #   year = <year>,
             m = re.match(r'(?i)year\s*=\s*(.*)$', line)

### 2007-06-11 slides/data.tex
*****************************   Changed nltk_lite to nltk *****************************
This has been done in preparation for 1.0,
in which the '-Lite' part of the name will
go away.  In the meantime the distributions
will still be called NLTK-Lite, but the package
itself will be just 'nltk'.  Code, demos, tests,
tutorials, contrib are all updated.


svn/trunk@4647

@@
 \begin{verbatim}
->>> from nltk_lite.corpora import toolbox
+>>> from nltk.corpora import toolbox
 >>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
@@
 \begin{verbatim}
->>> from nltk_lite.etree.ElementTree import SubElement
+>>> from nltk.etree.ElementTree import SubElement
 >>> for entry in lexicon:
@@
 >>> import sys
->>> from nltk_lite.etree.ElementTree import ElementTree
+>>> from nltk.etree.ElementTree import ElementTree
 >>> tree = ElementTree(lexicon[3])
@@
 \begin{verbatim}
->>> from nltk_lite.tokenize import regexp
->>> from nltk_lite.probability import FreqDist
+>>> from nltk.tokenize import regexp
+>>> from nltk.probability import FreqDist
 >>> fd = FreqDist()
@@
 \begin{verbatim}
->>> from nltk_lite.utilities import MinimalSet
+>>> from nltk.utilities import MinimalSet
 >>> pos = 1
@@
 \begin{verbatim}
->>> from nltk_lite.utilities import edit_dist
+>>> from nltk.utilities import edit_dist
 >>> def fuzzy_spell(target):
@@
 \begin{verbatim}
->>> from nltk_lite import parse
+>>> from nltk import parse
 >>> grammar = parse.cfg.parse_grammar('''

### 2007-06-11 slides/introduction.tex
*****************************   Changed nltk_lite to nltk *****************************
This has been done in preparation for 1.0,
in which the '-Lite' part of the name will
go away.  In the meantime the distributions
will still be called NLTK-Lite, but the package
itself will be just 'nltk'.  Code, demos, tests,
tutorials, contrib are all updated.


svn/trunk@4647

@@
 \begin{verbatim}
-    >>> from nltk_lite.corpora import cmudict
+    >>> from nltk.corpora import cmudict
     >>> from string import join
@@
 \begin{verbatim}
-    >>> from nltk_lite.corpora import shoebox
-    >>> from nltk_lite.utilities import MinimalSet
+    >>> from nltk.corpora import shoebox
+    >>> from nltk.utilities import MinimalSet
     >>> length, position, min = 4, 1, 3
@@
 \begin{verbatim}
-    >>> from nltk_lite.corpora import genesis
-    >>> from nltk_lite.probability import ConditionalFreqDist
-    >>> from nltk_lite.utilities import print_string
+    >>> from nltk.corpora import genesis
+    >>> from nltk.probability import ConditionalFreqDist
+    >>> from nltk.utilities import print_string
     >>> cfdist = ConditionalFreqDist()
@@
 \begin{verbatim}
-    >>> from nltk_lite.corpora import treebank
+    >>> from nltk.corpora import treebank
     >>> from string import join
@@
 \begin{enumerate}
-\item Install Python, Numeric
-\item Install NLTK-Lite, NLTK-Lite-Corpora
-\item Optional: Matplotlib, WordNet
-\item Set environment variable \verb|NLTK_LITE_CORPORA|
+\item Install Python, Install NLTK, NLTK-Corpora
+\item Optional: Numeric, Matplotlib
+\item Set environment variable \verb|NLTK_CORPORA|
 \end{enumerate}

### 2007-06-11 slides/programming.tex
*****************************   Changed nltk_lite to nltk *****************************
This has been done in preparation for 1.0,
in which the '-Lite' part of the name will
go away.  In the meantime the distributions
will still be called NLTK-Lite, but the package
itself will be just 'nltk'.  Code, demos, tests,
tutorials, contrib are all updated.


svn/trunk@4647

@@
 \begin{verbatim}
-  >>> from nltk_lite.corpora import gutenberg
+  >>> from nltk.corpora import gutenberg
   >>> count = {}
@@
   >>> import re
-  >>> from nltk_lite.utilities import re_show
+  >>> from nltk.utilities import re_show
   >>> sent = "colourless green ideas sleep furiously"
@@
 \begin{verbatim}
-  >>> from nltk_lite import utilities
+  >>> from nltk import utilities
   >>> utilities.re_show('green', s)
@@
 \begin{verbatim}
-  >>> from nltk_lite.corpora import gutenberg
+  >>> from nltk.corpora import gutenberg
   >>> gutenberg.items
@@
 \begin{verbatim}
-  >>> from nltk_lite.corpora import brown
+  >>> from nltk.corpora import brown
   >>> brown.items
   ['a', 'b', 'c', 'd', 'e', 'f', 'g', 'h', 'j', 'k', 'l', 'm', 'n', 'p', 'r']
-  >>> from nltk_lite.corpora import extract
+  >>> from nltk.corpora import extract
   >>> print extract(0, brown.raw())
@@
 \begin{verbatim}
-  >>> from nltk_lite.corpora import treebank
+  >>> from nltk.corpora import treebank
   >>> print extract(0, treebank.parsed())

### 2007-06-11 slides/tag.tex
*****************************   Changed nltk_lite to nltk *****************************
This has been done in preparation for 1.0,
in which the '-Lite' part of the name will
go away.  In the meantime the distributions
will still be called NLTK-Lite, but the package
itself will be just 'nltk'.  Code, demos, tests,
tutorials, contrib are all updated.


svn/trunk@4647

@@
 \begin{verbatim}
-from nltk_lite import tokenize
+from nltk import tokenize
 text = 'There are 11 players in a football team'
@@
 \begin{verbatim}
-from nltk_lite import tag
+from nltk import tag
 mytagger = tag.Default('nn')
@@
 \begin{verbatim}
-from nltk_lite.corpora import treebank
+from nltk.corpora import treebank
 from itertools import islice
@@
     or discounting)
-  \item see module \verb|nltk_lite.tag.hmm|
+  \item see module \verb|nltk.tag.hmm|
   \end{itemize}

### 2007-06-11 images/ambig01-a.py

@@ -1,4 +1,4 @@
-from nltk_lite.parse import bracket_parse
+from nltk.parse import bracket_parse
 sent = '(S (S Kim arrived) (conj or) (S (S Dana left) (conj and) (S everyone cheered)))'
 tree =  bracket_parse(sent)
 tree.draw()

### 2007-06-11 images/ambig01-b.py

@@ -1,4 +1,4 @@
-from nltk_lite.parse import bracket_parse
+from nltk.parse import bracket_parse
 sent = '(S (S (S Kim arrived) (conj or) (S Dana left)) (conj and) (S everyone cheered))'
 tree =  bracket_parse(sent)
 tree.draw()

### 2007-06-11 images/ambig02.py

@@ -1,4 +1,4 @@
-from nltk_lite.parse import bracket_parse
+from nltk.parse import bracket_parse
 sent = '(S (NP the policeman)(VP (V saw)(NP (NP the burglar)(PP with a gun))))'
 tree =  bracket_parse(sent)
 tree.draw()

### 2007-06-11 images/ambig03.py

@@ -1,4 +1,4 @@
-from nltk_lite.parse import bracket_parse
+from nltk.parse import bracket_parse
 sent = '(S (NP the policeman)(VP (V saw)(NP the burglar)(PP with a telescope)))'
 tree =  bracket_parse(sent)
 tree.draw()

### 2007-06-11 slides/programming.py

@@ -235,7 +235,7 @@ pos['colorless'] = 'adj'
 pos['furiously'] = 'adv'
 pos['ideas'] = 'n'
 
-from nltk_lite.utilities import SortedDict
+from nltk.utilities import SortedDict
 pos = SortedDict(pos)
 
 pos['ideas']
@@ -292,7 +292,7 @@ sorted(set(sentence))
 
 ### 2.6.4 Scaling it up
 
-from nltk_lite.corpora import gutenberg
+from nltk.corpora import gutenberg
 count = {}                                        # initialize a dictionary
 for word in gutenberg.raw('shakespeare-macbeth'): # tokenize Macbeth
     word = word.lower()                           # normalize to lowercase
@@ -322,7 +322,7 @@ for n in range(len(text)):
 		print n, n+1
 
 import re
-from nltk_lite.utilities import re_show
+from nltk.utilities import re_show
 re_show('very very', str)
 re_show('o+', str)
 re_show('oo+', str)

### 2007-06-11 tree2image.py

@@ -2,7 +2,7 @@
 #
 # Natural Language Toolkit: tree->image rendering script
 #
-# Copyright (C) 2001-2006 University of Pennsylvania
+# Copyright (C) 2001-2007 University of Pennsylvania
 # Author: Edward Loper <edloper@gradient.cis.upenn.edu>
 # URL: <http://nltk.sf.net>
 # For license information, see LICENSE.TXT
@@ -38,8 +38,8 @@ Requires Imagemagick C{convert}.
 """
 
 import re, sys, os
-from nltk_lite.draw import *
-from nltk_lite.draw.tree import *
+from nltk.draw import *
+from nltk.draw.tree import *
 import tkFont
 import tempfile
 import pickle
@@ -183,7 +183,7 @@ def tree_to_ps(s, outfile):
     return bbox[2:]
 
 # # hmm.. a bit of a hack..  e.g., no backslashing allowed:
-# from nltk_lite.parse.tree import bracket_parse, Tree
+# from nltk.parse.tree import bracket_parse, Tree
 # def tree_to_qtree(s):
 #     return '\Tree %s' % _tree_to_qtree(bracket_parse(s))
 # def _tree_to_qtree(tree):

### 2007-07-04 slides/programming.tex
added new diagram on linguistic data types
svn/trunk@4816

@@
 \begin{verbatim}
-  >>> from nltk.corpora import gutenberg
+  >>> from nltk.corpus import gutenberg
   >>> count = {}
-  >>> for word in gutenberg.raw('shakespeare-macbeth'):
+  >>> for word in gutenberg.read('shakespeare-macbeth'):
   ...     word = word.lower()
@@
 \begin{verbatim}
-  >>> from nltk.corpora import gutenberg
+  >>> from nltk.corpus import gutenberg
   >>> gutenberg.items
@@
 \begin{verbatim}
-  >>> from nltk.corpora import brown
+  >>> from nltk.corpus import brown
   >>> brown.items
   ['a', 'b', 'c', 'd', 'e', 'f', 'g', 'h', 'j', 'k', 'l', 'm', 'n', 'p', 'r']
-  >>> from nltk.corpora import extract
-  >>> print extract(0, brown.raw())
+  >>> print brown.read('a')[0]
   ['The', 'Fulton', 'County', 'Grand', 'Jury', 'said', 'Friday', 'an', 'investigation', 'of', "Atlanta's", 'recent', 'primary', 'election', 'produced', '``', 'no', 'evidence', "''", 'that', 'any', 'irregularities', 'took', 'place', '.']
-  >>> print extract(0, brown.tagged())
+  >>> print brown.tagged('a')[0]
   [('The', 'at'), ('Fulton', 'np-tl'), ('County', 'nn-tl'), ('Grand', 'jj-tl'), ('Jury', 'nn-tl'), ('said', 'vbd'), ('Friday', 'nr'), ('an', 'at'), ('investigation', 'nn'), ('of', 'in'), ("Atlanta's", 'np$'), ('recent', 'jj'), ('primary', 'nn'), ('election', 'nn'), ('produced', 'vbd'), ('``', '``'), ('no', 'at'), ('evidence', 'nn'), ("''", "''"), ('that', 'cs'), ('any', 'dti'), ('irregularities', 'nns'), ('took', 'vbd'), ('place', 'nn'), ('.', '.')]
@@
 \begin{verbatim}
-  >>> from nltk.corpora import treebank
-  >>> print extract(0, treebank.parsed())
+  >>> from nltk.corpus import treebank
+  >>> print treebank.parsed('wsj_0001')[0]
   (S:
@@
 
+\section{Miscellaneous Python Features}
+
+\subsection{Character Encoding}
+
+\begin{frame}
+\begin{itemize}
+
+\end{itemize}
+\end{frame}
+
 \end{document}

### 2007-07-12 slides/baypiggies.tex
template for baypiggies talk
svn/trunk@4870

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Introduction to Natural Language Processing}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
   University of Pennsylvania, USA
}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{What is NLTK?}

\begin{frame}
  \titlepage
\end{frame}

\subsection{Introduction}

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
    \item What is NLP?
    \item What is NLTK?
    \item ...
    \item ...
  \end{itemize}
\end{frame}

\subsection{Data Formats}

\begin{frame}
  \frametitle{Data Formats}
  \begin{itemize}
    \item ...
    \item ...
    \item ...
    \item ...
    \item ...
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Overview of Talk}
  \begin{enumerate}
    \item ...
    \item ...
    \item ...
    \item ...
    \item ...
  \end{enumerate}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{What is NLP?}

\subsection{Notes}

\begin{frame}
  \frametitle{Notes}
  \begin{itemize}
    \item QA, Summarization, Dialog (NL understanding; holy grail of NLP)
    \item RTE
    \item Low-level components
    \item ambiguity, why NLP is hard, PP attachment, combinatorial explosion
    \item limitations of NLP, avoiding AI-completeness
    \item statistical vs rule-based
    \item limitations of just throwing more data at a problem
  \end{itemize}
\end{frame}

\pgfdeclareimage[width=8cm]{tag-context}{../images/tag-context}
\begin{frame}
  \frametitle{Example of an Image}
  \pgfuseimage{tag-context}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Case Studies}

\subsection{??}

\begin{frame}
  \frametitle{Notes}
  \begin{itemize}
    \item Text classification
    \item Distributional similarity
    \item Parsing
  \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Getting Involved}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2007-07-12 slides/baypiggies.tex
incomplete slides on RTE
svn/trunk@4871

@@
 
+\begin{frame}
+  \frametitle{NLP}
+
+\begin{block}{Major sub-domains of NLP}
+  \begin{itemize}
+    \item Understanding
+    \item Translation
+    \item Generation
+  \end{itemize}
+\end{block}
+
+\begin{block}{Modalities}
+  \begin{itemize}
+    \item Text
+    \item Speech
+    \item Images
+  \end{itemize}
+\end{block}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Textual Inference}
+
+\begin{exampleblock}{Example 1}
+  \begin{description}
+    \item [Text] Never before had ski racing, a sport dominated by
+      monosyllabic mountain men, seen the likes of Alberto Tomba, the
+      flamboyant Bolognese flatlander who at 21 captured two gold
+      medals at the Calgary Olympics.
+    \item [Hypothesis] Alberto Tomba won a race.
+  \end{description}
+\end{exampleblock}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Textual Inference}
+
+\begin{exampleblock}{Example 2}
+  \begin{description}
+    \item [Text] Researchers at the Harvard School of Public Health
+      say that people who drink coffee may be doing a lot more than
+      keeping themselves awake---this kind of consumption apparently
+      also can help reduce the risk of diseases.
+    \item [Hypothesis] Coffee drinking has health benefits.
+  \end{description}
+\end{exampleblock}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Why is it useful?}
+
+  \begin{description}
+    \item[Information Extraction] 
+    \item[Spoken Dialogue] 
+    \item[Question Answering] Who won races at the Calgary Olympics?
+    \item[Text Summarization] 
+  \end{description}
+
+\end{frame}
+
 \subsection{Notes}

### 2007-07-12 slides/baypiggies.tex
fixed latex bug
svn/trunk@4872
added intro slides
svn/trunk@4873

@@
 \begin{frame}
-  \frametitle{Introduction}
+  \frametitle{Natural Language Processing}
   \begin{itemize}
-    \item What is NLP?
-    \item What is NLTK?
-    \item ...
-    \item ...
+    \item text and information
+      \note[item]{world is awash with text; information locked up in text}
+      \note[item]{sources: news, blogs, transcribed speech, other languages}
+    \item Natural Language Processing (NLP)
+      \note[item]{automatically process text, extracting structure and content, to give people access to the information}
+    \item Engineering goals
+      \note[build a system to provide a service (translation, summarization)]
+    \item Scientific goals
+      \note[analyze the structure of language, computational linguistics]
+    \item Data
   \end{itemize}
@@
 
-\begin{frame}
-  \frametitle{Data Formats}
-  \begin{itemize}
-    \item ...
-    \item ...
-    \item ...
-    \item ...
-    \item ...
-  \end{itemize}
+\begin{frame}[fragile]
+  \frametitle{Chunked Tagged WSJ Text}
+{\scriptsize\begin{verbatim}
+[ Pierre/NNP Vinken/NNP ]
+,/,
+[ 61/CD years/NNS ]
+old/JJ ,/, will/MD join/VB
+[ the/DT board/NN ]
+as/IN
+[ a/DT nonexecutive/JJ director/NN Nov./NNP 29/CD ]
+./.
+
+[ Mr./NNP Vinken/NNP ]
+is/VBZ
+[ chairman/NN ]
+of/IN
+[ Elsevier/NNP N.V./NNP ]
+,/,
+[ the/DT Dutch/NNP publishing/VBG group/NN ]
+./.
+\end{verbatim}}
 \end{frame}
 
+\begin{frame}[fragile]
+  \frametitle{Parsed WSJ Text}
+{\scriptsize\begin{verbatim}
+( (S
+    (NP-SBJ
+      (NP (NNP Pierre) (NNP Vinken) )
+      (, ,)
+      (ADJP
+        (NP (CD 61) (NNS years) )
+        (JJ old) )
+      (, ,) )
+    (VP (MD will)
+      (VP (VB join)
+        (NP (DT the) (NN board) )
+        (PP-CLR (IN as)
+          (NP (DT a) (JJ nonexecutive) (NN director) ))
+        (NP-TMP (NNP Nov.) (CD 29) )))
+    (. .) ))
+\end{verbatim}}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{IOB-Tagged Text (I=inside, O=outside, B=begin)}
+{\scriptsize\begin{verbatim}
+Characteristically RB B-ADVP
+, , O
+he PRP B-NP
+did VBD B-VP
+n't RB I-VP
+waste VB I-VP
+much JJ B-NP
+time NN I-NP
+getting VBG B-VP
+started VBN I-VP
+. . O
+\end{verbatim}}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Named-Entity Classification}
+{\tiny\begin{verbatim}
+<b_enamex type="LOCATION">NAIROBI<e_enamex>, <b_enamex type="LOCATION">Kenya<e_enamex>
+(<b_enamex type="ORGANIZATION">AP<e_enamex>) _ <b_numex type="CARDINAL">Thousands<e_numex>
+of laborers, students and opposition politicians on <b_timex type="DATE">Saturday<e_timex>
+protested tax hikes imposed by their cash-strapped government, which they accused of
+failing to provide basic services.  Beneath a scorching sun, they sang anti-government
+songs and chanted "<b_enamex type="PERSON">Moi<e_enamex> must go," showing their derision
+for President <b_enamex type="PERSON">Daniel arap Moi<e_enamex>,
+<b_enamex type="LOCATION">Kenya<e_enamex>'s ruler for <b_timex type="DURATION">20 years<e_timex>.
+By voice vote, the <b_numex type="CARDINAL">5,000<e_numex> protesters approved a resolution
+calling for the government to scrap new taxes, convene a convention to write a new
+Constitution, stop harassing students and street vendors, and halt ethnic violence.
+\end{verbatim}}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Tagged Biomedical Text in XML}
+{\tiny\begin{verbatim}
+<sentence id="@@0001"> <interactor pos="NN">Coexpression</interactor>
+<token pos="IN">of</token> <token pos="NN">mouse</token> <gene>PKR ( 1-515 ) WT</gene>
+<token pos="IN">as</token> <token pos="DT">a</token> <gene>Gal4</gene>
+<token pos="VBG">DNA-binding</token> <token pos="NN">domain</token>
+<token pos="NN">fusion</token> <token pos="IN">with</token>
+<token pos="RB">either</token> <token pos="DT">the</token>
+<token pos="JJ">catalytic-deficient</token> <token pos="JJ">human</token>
+<gene>PKR ( 1-551 )</gene> <token pos="IN">K296R</token>
+<token pos="JJ">mutant</token> <token pos=",">,</token> <token pos="DT">the</token>
+<token pos="NNP">RNA-binding-deficient</token> <token pos="JJ">human</token>
+<gene>PKR ( 1-551 ) K64E / K296R</gene> <token pos="RB">double</token>
+<token pos="JJ">mutant</token> <token pos=",">,</token> <token pos="CC">or</token>
+<token pos="JJ">wild-type</token> <token pos="NN">mouse</token>
+<gene>PKR ( 1-515 ) WT</gene> <token pos="IN">as</token>
+<token pos="JJ">full-length</token> <gene>PKR-Gal4</gene>
+<interactor pos="NN">activation</interactor> <token pos="NN">domain</token>
+<token pos="NNS">fusions</token> <token pos="VBD">resulted</token>
+<token pos="IN">in</token> <interactor pos="NN">activation</interactor>
+<token pos="IN">of</token> <token pos="DT">the</token> <gene>HIS3</gene>
+<token pos="CC">and</token> <gene>lacZ</gene> <token pos="NNS">reporters</token>
+<token pos=".">.</token></sentence>
+\end{verbatim}}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{CMU Pronunciation Lexicon}
+{\scriptsize\begin{verbatim}
+APPROVE 1 AH0 P R UW1 V
+APPROVED 1 AH0 P R UW1 V D
+APPROVES 1 AH0 P R UW1 V Z
+APPROVING 1 AH0 P R UW1 V IH0 NG
+APPROVINGLY 1 AH0 P R UW1 V IH0 NG L IY0
+APPROXIMATE 1 AH0 P R AA1 K S AH0 M AH0 T
+APPROXIMATE 2 AH0 P R AA1 K S AH0 M EY2 T
+APPROXIMATED 1 AH0 P R AA1 K S AH0 M EY2 T AH0 D
+APPROXIMATELY 1 AH0 P R AA1 K S AH0 M AH0 T L IY0
+APPROXIMATES 1 AH0 P R AA1 K S AH0 M AH0 T S
+APPROXIMATES 2 AH0 P R AA1 K S AH0 M EY2 T S
+APPROXIMATING 1 AH0 P R AA1 K S AH0 M EY2 T IH0 NG
+APPROXIMATION 1 AH0 P R AA2 K S AH0 M EY1 SH AH0 N
+APPROXIMATIONS 1 AH0 P R AA2 K S AH0 M EY1 SH AH0 N Z
+\end{verbatim}}
+\end{frame}
+
+\subsection{NLTK}
+
 \begin{frame}
-  \frametitle{Overview of Talk}
+  \frametitle{Natural Language Toolkit}
   \begin{enumerate}
-    \item ...
-    \item ...
-    \item ...
-    \item ...
-    \item ...
+    \item facilitate access to data
+      \note[item]{only write a format reader once, then share}
+      \note[item]{reading a format correctly is harder than you think, cf csv}
+    \item standard algorithms
+      \note[item]{getting into NLP requires knowledge of linguistics, CS, AI which is hard to get}
+      \note[item]{clarity of Python: people can read the reference implementation, try it, reimplement it...}
+    \item tutorials
+      \note[item]{major textbooks are formidable, encyclopedic}
+      \note[item]{we've selected useful core, and provided practical learning materials (book, demos, source)}
+      \note[item]{practical: learning-by-doing; theoretically sound: stepping stone to research literature}
+    \item contents
+      \note[item]{50k lines of code, 300Mb datasets, 380pp book}
+    \item evolution: beginnings, redesign
+      \note[item]{teaching NLP, got too heavy using nested dictionaries, people ended up "programming in NLTK", rather than programming in Python using NLTK where needed}
+      \note[item]{NLTK-Lite: completely reimplemented, simpler structure, uses new language features like set, generator expressions, defaultdict, elementtree}
+      \note[item]{Design philosophy: interface definitions, lightweight reference implementations, utilities for common tasks}
   \end{enumerate}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Find 20 Most Frequent Words in a Text}
+
+{\scriptsize\begin{verbatim}
+>>> counts = defaultdict(int)
+>>> for word in text.split():
+...     counts[word] += 1
+>>> ranked_words = sorted(counts.items, key=itemgetter(1))
+>>> print [word for (word,count) in reversed(ranked_words)][:20]
+\end{verbatim}}
+
+\note{defaultdict; sorting tuples, list comprehensions}
+
+{\scriptsize\begin{verbatim}
+>>> print FreqDist(tokenize.wordpunct(text)).sorted()[:20]
+\end{verbatim}}
+
 \end{frame}
@@
   \begin{description}
-    \item[Information Extraction] 
-    \item[Spoken Dialogue] 
-    \item[Question Answering] Who won races at the Calgary Olympics?
-    \item[Text Summarization] 
+    \item [Information Extraction] blah
+    \item [Spoken Dialogue] blah
+    \item [Question Answering] Who won races at the Calgary Olympics?
+    \item [Text Summarization] bkah
   \end{description}

### 2007-07-12 slides/baypiggies.tex
- Added some content for part 3 (case studies)
svn/trunk@4875

@@
 \documentclass{beamer}             % for slides
+\usepackage{alltt}
 % \documentclass[handout]{beamer}    % for handout
@@
 
-\subsection{??}
-
 \begin{frame}
@@
     \item Parsing
+  \end{itemize}
+\end{frame}
+
+\subsection{Text Classificaiton}
+
+\begin{frame}
+  \frametitle{Text Classification}
+  \begin{itemize}
+    \item Choose the most appropriate label for each piece of text.
+    \begin{itemize}
+      \item Topic classification
+      \item Spam Filtering
+      \item Word Sense Disambiguation
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Features}
+  \begin{itemize}
+    \item To decide which label is appropriate for a text, 
+      classifiers look at \emph{features} of that text.
+    \item Features indicate what aspects of the text are
+      relevant for the classification task.
+    \item Features are typically hand-crafted for a given task.
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Classifying in NLTK}
+  \begin{itemize}
+    \item Classifiers are applied to \emph{featuresets}:
+\begin{alltt}\small
+>>> animal = dict(fur=True, legs=4, 
+                  size='large', spots=True)
+>>> my_classifier.classify(animal)
+'leopard'
+\end{alltt}
+    \item Features are found with a \emph{feature extractor}:
+\begin{alltt}\small
+>>> def document_features(document):
+...     return dict([('contains-word(%s)'%w,True) 
+...                  for w in document])
+\end{alltt}
+    \item Training:
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Demo -- Predict Genders of Names}
+  \begin{itemize}
+  \item Given a new name, that we've never seen before, predict
+    whether it is male of female.
+  \item Use the \texttt{names} corpus.
+  \end{itemize}
+\end{frame}
+
+\subsection{Distributional Similarity}
+
+\begin{frame}
+  \frametitle{Word Similarity?}
+  \begin{itemize}
+  \item It's useful to know whether two words are similar
+    \begin{itemize}
+    \item ...
+    \end{itemize}
+  \item Different notions of \emph{similarity}:
+    \item What words tend to co-occur?
+    \item What words occur in similar contexts?
   \end{itemize}

### 2007-07-12 slides/baypiggies.tex
slide updates
svn/trunk@4876

@@
 \documentclass{beamer}             % for slides
+%\documentclass[notes,handout]{beamer} % presenter notes
+% \documentclass[handout]{beamer}    % for handout
 \usepackage{alltt}
-% \documentclass[handout]{beamer}    % for handout
 \input{beamer}
@@
     \item Engineering goals
-      \note[build a system to provide a service (translation, summarization)]
+      \note[item]{build a system to provide a service (translation, summarization)}
     \item Scientific goals
-      \note[analyze the structure of language, computational linguistics]
+      \note[item]{analyze the structure of language, computational linguistics}
     \item Data
@@
 \begin{frame}[fragile]
-  \frametitle{IOB-Tagged Text (I=inside, O=outside, B=begin)}
+  \frametitle{IOB-Tagged Text}
 {\scriptsize\begin{verbatim}

### 2007-07-12 slides/demos/names.py
added demos
svn/trunk@4877

@@ -1,0 +1,13 @@
+from nltk.classify.naivebayes import NaiveBayesClassifier
+from nltk.classify.util import names_demo
+
+# Feature Extraction:
+def name_features(name):
+    features = {}
+    return features
+
+# Test the classifier:
+classifier = names_demo(NaiveBayesClassifier.train, name_features)
+
+# Feature Analysis:
+#classifier.show_most_informative_features()

### 2007-07-12 slides/demos/parse.py

@@ -1,0 +1,2 @@
+import nltk.draw.chart
+nltk.draw.chart.demo()

### 2007-07-12 slides/demos/similar_words.py

@@ -1,0 +1,39 @@
+from nltk.book import *
+from collections import defaultdict
+
+######################################################################
+def build_context_map():
+    """Build a dictionary mapping words in the brown corpus to lists
+    of local lexical contexts, where a context is encoded as a tuple
+    (prevword, nextword)."""
+    context_map = defaultdict(list)
+    for item in corpus.brown.items:
+        words = corpus.brown.tokenized(item, group_by_sent=False)
+        words = [word.lower() for word in words]
+        for i in range(1, len(words)-1):
+            prevword, word, nextword = words[i-1:i+2]
+            context_map[word].append( (prevword, nextword) )
+    return context_map
+
+if 'context_map' not in globals():
+    context_map = build_context_map()
+    
+######################################################################
+def dist_sim(context_map, word, num=6):
+    """Display words that appear in similar contexts to the given
+    word, based on the given context map."""
+    contexts = set(context_map.get(word, ()))
+    fd = nltk.FreqDist(w for w in context_map
+                           for c in context_map[w]
+                             if c in contexts and w!=word)
+    
+    print 'Words similar to %r:' % word
+    print ' '.join('%10s' % wd for wd in fd.sorted()[:num])
+    print ' '.join('%10s' % fd[wd] for wd in fd.sorted()[:num])
+
+######################################################################
+    
+dist_sim(context_map, 'man')
+dist_sim(context_map, 'woman')
+dist_sim(context_map, 'walk')
+dist_sim(context_map, 'in')

### 2007-07-12 slides/demos/similar_words_2.py

@@ -1,0 +1,38 @@
+from nltk.probability import ConditionalFreqDist
+from nltk.corpus import brown
+
+######################################################################
+def build_association_distribution():
+    assoc = ConditionalFreqDist()
+    
+    # For each document in the "Brown Corpus"...
+    for document in brown.documents:
+        words = brown.tagged(document, group_by_sent=False)
+        
+        # For each word that's a noun...
+        for index, (word, tag) in enumerate(words):
+            if tag.startswith('N'):
+
+                # Look at any nouns in the next 5 words...
+                window = words[index+1:index+5]
+                for (window_word, window_tag) in window:
+                    if window_tag.startswith('N'):
+
+                        # And add them to our freq. distribution
+                        assoc[word].inc(window_word.lower())
+
+    return assoc
+
+if 'associations' not in globals():
+    associations = build_association_distribution()
+    
+######################################################################
+def assoc(word):
+    print '%20s -> %s' % (word, associations[word].max())
+
+######################################################################
+assoc('man')
+assoc('woman')
+assoc('level')
+
+

### 2007-07-12 slides/demos/names.py
added comments
svn/trunk@4878

@@ -1,3 +1,8 @@
+######################################################################
+##
+##  Guess an unseen name's gender!
+##
+
 from nltk.classify.naivebayes import NaiveBayesClassifier
 from nltk.classify.util import names_demo
 

### 2007-07-12 slides/demos/parse.py

@@ -1,2 +1,7 @@
+######################################################################
+##
+##  Chart Parsing Demo
+##
+
 import nltk.draw.chart
 nltk.draw.chart.demo()

### 2007-07-12 slides/demos/similar_words.py

@@ -1,3 +1,8 @@
+######################################################################
+##
+##  What words occur in similar contexts?
+##
+
 from nltk.book import *
 from collections import defaultdict
 

### 2007-07-12 slides/demos/similar_words_2.py

@@ -1,3 +1,8 @@
+######################################################################
+##
+##  What words tend to co-occur?
+##
+
 from nltk.probability import ConditionalFreqDist
 from nltk.corpus import brown
 

### 2007-07-12 slides/baypiggies.tex
more on getting involved; more on case studies
svn/trunk@4879

@@
     \item Text classification
-    \item Distributional similarity
+    \item Word similarity
     \item Parsing
@@
     \item Training:
+\begin{alltt}\small
+>>> labeled_featuresets = [(document_features(doc), label)
+..       .                 for (doc, label) in labeled_docs]
+>>> classifier = NaiveBayesClassifier.train(labeled_docs)
+\end{alltt}
+
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Word Similarity?}
+  \frametitle{Word Similarity}
   \begin{itemize}
@@
     \begin{itemize}
-    \item ...
+    \item Search: don't just find the exact terms you specify.
+    \item Question Answering: handle synonyms.
+    \item Parsing: what words act similarly to one another?
     \end{itemize}
@@
 
+\subsection{Parsing}
+
+\begin{frame}
+  \frametitle{Parsing}
+  \begin{itemize}
+  \item Chart Parsing Demo
+  \end{itemize}
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{frame}
+  \frametitle{Getting Involved}
+  \begin{itemize}
+  \item http://nltk.org/
+    \begin{itemize}
+    \item http://nltk.org/index.php/Development
+    \item http://nltk.org/index.php/Chatroom
+    \end{itemize}
+  \item Mailing lists:
+    \begin{itemize}
+    \item nltk-devel@lists.sourceforge.net
+    \item nltk-announce@lists.sourceforge.net
+    \item nltk-users@lists.sourceforge.net
+    \end{itemize}
+  \end{itemize}
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2007-07-12 slides/baypiggies.tex
fixed size of python snippets
svn/trunk@4880

@@
     \item Classifiers are applied to \emph{featuresets}:
-\begin{alltt}\small
+{\small\begin{alltt}
 >>> animal = dict(fur=True, legs=4, 
@@
 'leopard'
-\end{alltt}
+\end{alltt}}
     \item Features are found with a \emph{feature extractor}:
-\begin{alltt}\small
+{\small\begin{alltt}
 >>> def document_features(document):
@@
 ...                  for w in document])
-\end{alltt}
+\end{alltt}}
     \item Training:
-\begin{alltt}\small
->>> labeled_featuresets = [(document_features(doc), label)
-..       .                 for (doc, label) in labeled_docs]
->>> classifier = NaiveBayesClassifier.train(labeled_docs)
-\end{alltt}
+{\small\begin{alltt}
+>>> train = [(document_features(doc), label)
+..           for (doc, label) in labeled_docs]
+>>> classifier = NaiveBayesClassifier.train(train)
+\end{alltt}}
 

### 2007-07-16 slides/data.tex
reorganised slides, updated somewhat for new data chapter
svn/trunk@4909

@@
 
-\title{Accessing and Analyzing Linguistic Field Data}
+\title{Linguistic Data Management}
 
@@
   \begin{itemize}
-  \item variety of data types, e.g.
-  \item \textbf{lexicon:} database of words, minimally containing part
-    of speech information and glosses
-  \item \textbf{paradigm:} any kind of rational
-    tabulation of words or phrases to illustrate contrasts and systematic
-    variation
-  \item \textbf{text:} any larger unit such as a narrative or a
-    conversation
-  \item descriptions: field notes, grammars and analytical papers
-  \item complex inter-relations (new word; lexicon update, paradigm;
-    add to field notes; extend grammar, ...)
-  \item tasks: sorting, tabulating, garnering statistics, searching,
-  verifying transcriptions
-  \item principle challenge is computational
+  \item language resources, types, proliferation
+  \item role in NLP, CL
+  \item enablers: storage/XML/Unicode; digital publication; resource catalogues
+  \item obstacles: discovery, access, format, tool
+  \item data types: texts and lexicons
+  \item useful ways to access data using Python: csv, html, xml
   \end{itemize}
@@
 \begin{frame}
-\frametitle{Tools and Technologies}
+\frametitle{Linguistic Databases}
 \begin{itemize}
-\item General-purpose tools
-  \begin{itemize}
-  \item word processors
-  \item spreadsheets
-  \item databases
-  \end{itemize}
-\item Special purpose tools
-  \begin{itemize}
-  \item Toolbox
-  \end{itemize}
-\item broader question: data management in support of best practices
-\item Bird, Steven and Gary Simons (2003).  Seven Dimensions of Portability
-    for Language Documentation and Description, \textit{Language} 79: 557-582.
+\item Field linguistics
+\item Corpora
+\item Reference Corpus
 \end{itemize}
@@
 
-\section{General Purpose Tools}
-
-\subsection{Word Processors}
+\pgfdeclareimage[width=8cm]{datatypes}{../images/datatypes}
+\begin{frame}
+\frametitle{Fundamental Data Types}
+  \pgfuseimage{datatypes}
+\end{frame}
 
 \begin{frame}
-\frametitle{General Purpose Tools: Word Processors}
+\frametitle{Example: TIMIT}
+\begin{itemize}
+\item TI (Texas Instruments) + MIT
+\item balance
+\item sentence selection
+\item layers of annotation
+\item speaker demographics, lexicon
+\item combination of time-series and record-structured data
+\item programs for speech corpus
+\end{itemize}
+\end{frame}
+
+\pgfdeclareimage[width=0.7\textwidth]{timit}{../images/timit}
+\begin{frame}
+\frametitle{Example: TIMIT}
+  \pgfuseimage{timit}
+\end{frame}
+
+\pgfdeclareimage[width=\textwidth]{timit-structure}{../images/timit-structure}
+\begin{frame}
+\frametitle{Example: TIMIT}
+  \pgfuseimage{timit-structure}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: TIMIT}
+\scriptsize
+\begin{verbatim}
+    >>> phonetic = list(corpus.timit.phonetic())
+    >>> phonetic[0]
+    ['h#', 'sh', 'iy', 'hv', 'ae', 'dcl', 'y', 'ix', 'dcl', 'd', 'aa', 'kcl',
+    's', 'ux', 'tcl', 'en', 'gcl', 'g', 'r', 'iy', 's', 'iy', 'w', 'aa',
+    'sh', 'epi', 'w', 'aa', 'dx', 'ax', 'q', 'ao', 'l', 'y', 'ih', 'ax', 'h#']
+    >>> tokenized = list(corpus.timit.tokenized(offset=True))
+    >>> tokenized[0]
+    [('she', 7812, 10610), ('had', 10610, 14496), ('your', 14496, 15791),
+    ('dark', 15791, 20720), ('suit', 20720, 25647), ('in', 25647, 26906),
+    ('greasy', 26906, 32668), ('wash', 32668, 37890), ('water', 38531, 42417),
+    ('all', 43091, 46052), ('year', 46052, 50522)]
+\end{verbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Example: TIMIT}
+\scriptsize
+\begin{verbatim}
+    >>> timitdict = corpus.timit.dictionary()
+    >>> timitdict['greasy'] + timitdict['wash'] + timitdict['water']
+    ['g', 'r', 'iy1', 's', 'iy', 'w', 'ao1', 'sh', 'w', 'ao1', 't', 'axr']
+    >>> phonetic[0][17:30]
+    ['g', 'r', 'iy', 's', 'iy', 'w', 'aa', 'sh', 'epi', 'w', 'aa', 'dx', 'ax']
+
+    >>> spkrinfo = corpus.timit.spkrinfo()
+    >>> spkrinfo['dr1-fvmh0']
+    {'use': 'TRN', 'birthdate': '01/08/60', 'sex': 'F', 'race': 'WHT',
+    'ht': '5\'05"', 'edu': 'BS', 'recdate': '03/11/86', 'dr': '1',
+    'id': 'VMH0', 'comments': 'BEST NEW ENGLAND ACCENT SO FAR'}
+\end{verbatim}
+\end{frame}
+
+\section{Creating Data}
+
+\begin{frame}
+\frametitle{Lifecycle}
 
 \begin{itemize}
-\item office software widely used in computer-based language
-  documentation work
-\item familiar, ready availability (though may be too expensive?)
-\item Word processors for managing dictionaries:
+\item create
+\item annotate texts
+\item refine lexicon
+\item organize structure
+\item publish
+\end{itemize}
+\end{frame}
+
+\pgfdeclareimage[width=\textwidth]{evolution}{../images/evolution}
+\begin{frame}
+\frametitle{Evolution}
+  \pgfuseimage{evolution}
+\end{frame}
+
+\subsection{Primary Data}
+
+\begin{frame}
+\frametitle{Creating Data: Primary Data}
+
 \begin{itemize}
-  \item consistency of content and format --- time consuming
-  \item how would we check the following constraint: \\
-    \textit{each entry must have a part-of-speech field, drawn from a set of 20
-     possibilities, displayed after the pronunciation field, and rendered
-     in 11-point bold}
-  \item generally a bad idea!
-  \end{itemize}
-\item save to a non-proprietary format (e.g. RTF, HTML, or XML)
-\item write programs to do this checking automatically
+\item spiders
+\item recording
+\item texts
 \end{itemize}
@@
 \begin{frame}[fragile]
-\frametitle{Example: Microsoft Word}
+\frametitle{Data Cleansing: Accessing Spreadsheets}
+\scriptsize
+\begin{verbatim}
+dict.csv:
+"sleep","sli:p","v.i","a condition of body and mind ..."
+"walk","wo:k","v.intr","progress by lifting and setting down each foot ..."
+"wake","weik","intrans","cease to sleep"
+
+    >>> import csv
+    >>> file = open("dict.csv", "rb")
+    >>> for row in csv.reader(file):
+    ...     print row
+    ['sleep', 'sli:p', 'v.i', 'a condition of body and mind ...']
+    ['walk', 'wo:k', 'v.intr', 'progress by lifting and setting down each foot ...']
+    ['wake', 'weik', 'intrans', 'cease to sleep']
+\end{verbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Data Cleansing: Validation}
 \small
+\begin{verbatim}
+    def undefined_words(csv_file):
+        import csv
+        lexemes = set()
+        defn_words = set()
+        for row in csv.reader(open(csv_file)):
+            lexeme, pron, pos, defn = row
+            lexemes.add(lexeme)
+            defn_words.union(defn.split())
+        return sorted(defn_words.difference(lexemes))
+
+    >>> print undefined_words("dict.csv")
+    ['...', 'a', 'and', 'body', 'by', 'cease',
+     'condition', 'down', 'each', 'foot',
+     'lifting', 'mind', 'of', 'progress',
+     'setting', 'to']
+\end{verbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Data Cleansing: Accessing Web Text}
+
+\scriptsize
+\begin{verbatim}
+>>> import urllib, nltk
+>>> html = urllib.urlopen('http://en.wikipedia.org/').read()
+>>> text = nltk.clean_html(html)
+>>> text.split()
+['Wikimedia', 'Error', 'WIKIMEDIA', 'FOUNDATION', 'Fout', 'Fel',
+'Fallo', '\xe9\x94\x99\xe8\xaf\xaf', '\xe9\x8c\xaf\xe8\xaa\xa4',
+'Erreur', 'Error', 'Fehler', '\xe3\x82\xa8\xe3\x83\xa9\xe3\x83\xbc',
+'B\xc5\x82\xc4\x85d', 'Errore', 'Erro', 'Chyba', 'EnglishThe',
+'Wikimedia', 'Foundation', 'servers', 'are', 'currently',
+'experiencing', 'technical', 'difficulties.The', 'problem', 'is',
+'most', 'likely', 'temporary', 'and', 'will', 'hopefully', 'be',
+'fixed', 'soon.', 'Please', 'check', 'back', 'in', 'a', 'few',
+'minutes.For', 'further', 'information,', 'you', 'can', 'visit',
+'the', 'wikipedia', 'channel', 'on', 'the', 'Freenode', 'IRC', ...
+\end{verbatim}
+\end{frame}
+
+\subsection{Annotation}
+
+\begin{frame}[fragile]
+\frametitle{Creating Data: Annotation}
+
 \begin{itemize}
-\item sleep \verb|[sli:p]| \textbf{vi} \textit{condition of body and mind...}
-\item We can enter this in MSWord, then "Save as Web Page"
-\item resulting HTML:
-
-\begin{verbatim}
-<p class=MsoNormal>
-  sleep
-  <span style='mso-spacerun:yes'> </span>
-  [<span class=SpellE>sli:p</span>]
-  <span style='mso-spacerun:yes'> </span>
-  <b><span style='font-size:11.0pt'>vi</span></b>
-  <span style='mso-spacerun:yes'> </span>
-  <i>a condition of body and mind ...</i>
-</p>
-\end{verbatim}
-
-\item Observations:
-  \begin{itemize}
-  \item entry: HTML paragraph using the \verb|<p>| element
-  \item pos: inside a \verb|<span style='font-size:11.0pt'>| element
-  \end{itemize}
+\item linguistic annotation
+\item Tools: \verb|http://www.exmaralda.org/annotation/|
 \end{itemize}
@@
 
-\begin{frame}[fragile]
-\frametitle{Example: Validating Dictionaries stored in MSWord}
-\scriptsize
-
-\begin{verbatim}
->>> import re
->>> legal_pos = set(['n', 'v.t.', 'v.i.', 'adj', 'det'])
->>> pattern = re.compile(r"'font-size:11.0pt'>([a-z.]+)<")
->>> document = open("dict.htm").read()
->>> used_pos = set(re.findall(pattern, document))
->>> illegal_pos = used_pos.difference(legal_pos)
->>> print list(illegal_pos)
-['v.intr', 'v.i', 'intrans']
-\end{verbatim}
-\end{frame}
-
-\subsection{Spreadsheets}
-
-\begin{frame}[fragile]
-\frametitle{Example: Converting MSWord dictionary to Excel}
-\small
-
-\begin{verbatim}
->>> import re
->>> document = open("dict.htm").read()
->>> document = re.sub("[\r\n]", "", document)
->>> word_pattern = re.compile(r">([\w]+)")
->>> pron_pattern = re.compile(r"\[.*>([a-z:]+)<.*\]")
->>> for entry in document.split("<p"):
-...     word_match = word_pattern.search(entry)
-...     pron_match = pron_pattern.search(entry)
-...     if word_match and pron_match:
-...         lex = word_match.group(1)
-...         pos = pron_match.group(1)
-...         print '"%s","%s"' % (lex, pos)
-"sleep","sli:p"
-"walk","wo:k"
-"wake","weik"
-\end{verbatim}
-\end{frame}
-
-\begin{frame}
-\frametitle{Spreadsheets}
-\begin{itemize}
-\item wordlists, paradigms
-\item e.g. comparative wordlist (\url{http://www.rosettaproject.org/}):
-  \begin{itemize}
-  \item row for each cognate set
-  \item column for each language
-  \end{itemize}
-\item export spreadsheets in \textit{CSV format}
-\item read them using Python's \texttt{csv} module
-\item e.g. find cognates having an edit-distance of at least three from each other
-\end{itemize}
-\end{frame}
-
-\subsection{Databases}
-
-\begin{frame}
-\frametitle{Databases}
-\begin{itemize}
-  \item lexicons are stored in a full-fledged relational databases
-  \item databases can implement many well-formedness constraints\\
-    e.g. ensure that all parts-of-speech come from an
-    \textit{enumerated type}
-  \item However, the relational model is often too restrictive
-  \item optional, repeatable fields (e.g.
-    dictionary sense definitions and example sentences).
-  \item SQL cannot express many linguistically-motivated queries\\
-    \textit{Find all words that appear in example sentences for which no dictionary entry is provided}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Database Example}  
-\scriptsize
-
-\begin{verbatim}
->>> import csv
->>> lexemes = []
->>> defn_words = []
->>> for row in csv.reader(open("dict.csv")):
-...     lexeme, pron, pos, defn = row
-...     lexemes.append(lexeme)
-...     defn_words += defn.split()
->>> undefined = list(set(defn_words).difference(set(lexemes)))
->>> undefined.sort()
->>> print undefined
-['...', 'a', 'and', 'body', 'by', 'cease', 'condition', 'down',
-'each', 'foot', 'lifting', 'mind', 'of', 'progress', 'setting', 'to']
-\end{verbatim}
-\end{frame}
-
-\section{Processing Toolbox Data}
+\subsection{Lexicons}
 
@@
 \item many kinds of validation and formatting not supported by Toolbox software
-\item simple file format
 \item each file is a collection of \textit{entries} (aka
@@
 \item each entry is made up of one or more \textit{fields}
-\item challenges:
-  \begin{enumerate}
-  \item different entry types, implications for which fields are present
-  \item field order only sometimes significant
-  \end{enumerate}
+\item we can apply our programming methods, including chunking and parsing
 \end{itemize}
@@
 
-\subsection{Accessing Toolbox Data}
-
 \begin{frame}[fragile]
@@
 
-\small
+\scriptsize
 \begin{verbatim}
->>> from nltk.corpora import toolbox
+>>> from nltk.corpus import toolbox
 >>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
@@
 
+\scriptsize
+\begin{verbatim}
+>>> [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
+['kaa', 'kaa', 'kaa', 'kaakaaro', 'kaakaaviko', 'kaakaavo', 'kaakaoko',
+'kaakasi', 'kaakau', 'kaakauko', 'kaakito', 'kaakuupato', ..., 'kuvuto']
+\end{verbatim}
+
 \small
@@
 
-\small
-\begin{verbatim}
->>> [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
-['kaa', 'kaa', 'kaa', 'kaakaaro', 'kaakaaviko', 'kaakaavo', 'kaakaoko',
-'kaakasi', 'kaakau', 'kaakauko', 'kaakito', 'kaakuupato', ..., 'kuvuto']
-\end{verbatim}
 \end{frame}
 
-\subsection{Adding and Removing Fields}
+\begin{frame}
+\frametitle{Data Cleansing: Toolbox}
+
+\begin{itemize}
+\item parsing (Listing 4)
+\item chunking (Listing 5)
+\item adding missing fields (next)
+\end{itemize}
+\end{frame}
 
@@
 \begin{frame}[fragile]
-\frametitle{Removing Fields}
+\frametitle{Generating HTML Tables from Toolbox Data}
+
 \scriptsize
-
-\begin{itemize}
-\item sanitise our data before giving it to others
-\end{itemize}
-
-\small
-\begin{verbatim}
->>> retain = ('lx', 'ps', 'ge')
->>> for entry in lexicon:
-...     entry[:] = [f for f in entry if f.tag in retain]
->>> toolbox.to_sfm_string(lexicon[50])
-\lx kaeviro
-\ps V.A
-\ge lift off
-\ge take off
-\end{verbatim}
-\end{frame}
-
-\subsection{Formatting}
-
-\begin{frame}[fragile]
-\frametitle{Formatting Entries}
-
-\begin{itemize}
-\item request specific fields without concern for ordering
-\end{itemize}
-\scriptsize
-
-\begin{verbatim}
->>> lexicon = toolbox.parse_corpus('rotokas.dic', key='lx')
->>> for entry in lexicon[70:80]:
-...     lx = entry.findtext('lx')
-...     ps = entry.findtext('ps')
-...     ge = entry.findtext('ge')
-...     print "%s (%s) '%s'" % (lx, ps, ge)
-kakapikoto (N.N2) 'newborn baby'
-kakapu (V.B) 'place in sling for purpose of carrying'
-kakapua (N.N) 'sling for lifting'
-kakara (N.N) 'arm band'
-Kakarapaia (N.PN) 'village name'
-kakarau (N.F) 'frog'
-Kakarera (N.PN) 'name'
-Kakareraia (N.???) 'name'
-kakata (N.F) 'cockatoo'
-kakate (N.F) 'bamboo water tube'
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Generating HTML Tables}
-
-\begin{itemize}
-\item Publish shoebox lexicon on the web
-\end{itemize}
-\scriptsize
-
 \begin{verbatim}
@@
 
-\tiny
+\small
 \begin{verbatim}
@@
 >>> tree.write(sys.stdout)
+\end{verbatim}
+
+\scriptsize
+\begin{verbatim}
 <record>
@@
 
-\section{Linguistic Exploration}
-
-\subsection{Reduplication}
-
 \begin{frame}[fragile]
-\frametitle{Reduplication}
+\frametitle{Analysis: Reduplication}
 \scriptsize
@@
 
-\subsection{Complex Search Criteria}
+\subsection{Analysis: Complex Search Criteria}
 
@@
 \frametitle{Complex Search Criteria}
+\scriptsize
+\begin{verbatim}
+>>> from nltk import tokenize, FreqDist
+>>> fd = FreqDist()
+>>> lexemes = [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
+>>> for lex in lexemes:
+...     for syl in tokenize.regexp(lex, pattern=r'[^aeiou][aeiou]'):
+...         fd.inc(syl)
+\end{verbatim}
+
+\normalsize
 \begin{itemize}
@@
 
-\scriptsize
-\begin{verbatim}
->>> from nltk.tokenize import regexp
->>> from nltk.probability import FreqDist
->>> fd = FreqDist()
->>> lexemes = [lexeme.text.lower() for lexeme in lexicon.findall('record/lx')]
->>> for lex in lexemes:
-...     for syl in regexp(lex, pattern=r'[^aeiou][aeiou]'):
-...         fd.inc(syl)
-\end{verbatim}
 \end{frame}
@@
 \begin{frame}[fragile]
-\frametitle{Complex Search Criteria (cont)}
+\frametitle{Analysis: Complex Search Criteria (cont)}
 
@@
 
-\begin{frame}[fragile]
-\frametitle{Prosodically-motivated search}
-\small
-\begin{itemize}
-\item segmental and prosodic constraints
-\item well-formed morphemes and lexemes
-\item highly-specific query (e.g. find things predicted not to exist
-  by our theoretical model?)
-\item \textit{Find all trisyllabic lexemes ending in a long vowel}
-\item bottom-up design: need to count syllables, and test for vowels
-\end{itemize}
-
-\begin{verbatim}
->>> def num_cons(word):
-...     template = cv(word)
-...     return template.count('C')
->>> def is_vowel(char):
-...     return (char in 'aeiou')
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Prosodically-motivated search (cont)}
-\tiny
-
-\begin{verbatim}
->>> for entry in lexicon:
-...     lx = entry.findtext('lx')
-...     if lx:
-...         ps = entry.findtext('ps')
-...         if num_cons(lx) == 3 and ps[0] == 'V'\
-...           and is_vowel(lx[-1]) and is_vowel(lx[-2]):
-...             ge = entry.findtext('ge')
-...             print "%s (%s) '%s'" % (lx, ps, ge)
-kaetupie (V.B) 'tighten'
-kakupie (V.B) 'shout'
-kapatau (V.B) 'add to'
-kapuapie (V.B) 'wound'
-kapupie (V.B) 'close tight'
-kapuupie (V.B) 'close'
-karepie (V.B) 'return'
-karivai (V.A) 'have an appetite'
-kasipie (V.B) 'care for'
-kaukaupie (V.B) 'shine intensely'
-kavorou (V.A) 'covet'
-kavupie (V.B) 'leave.behind'
-...
-kuverea (V.A) 'incomplete'
-\end{verbatim}
-\end{frame}
-
-\subsection{Minimal Sets}
-
 \begin{frame}
-\frametitle{Finding Minimal Sets}
+\frametitle{Analysis: Finding Minimal Sets}
 \begin{itemize}
@@
 \begin{verbatim}
->>> from nltk.utilities import MinimalSet
+>>> from nltk import MinimalSet
 >>> pos = 1
@@
 
-\subsection{Example Application}
+\section{Publishing}
 
 \begin{frame}
-\frametitle{Example Application: Improving Access to Lexical Resources}
+\frametitle{Publishing}
+
 \begin{itemize}
-\item lexicon as a \textit{language resource}
-\item existence of standardized writing system?  literacy?
-\item lookup by approximate spelling
-\item soundex
-\item edit distance
-\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Computing Soundex}
-\scriptsize
-
-\begin{verbatim}
->>> group = {
-...     ' ':0,                     # blank (for short words)
-...     'p':1,  'b':1,  'v':1,     # labials
-...     't':2,  'd':2,  's':2,     # alveolars
-...     'l':3,  'r':3,             # sonorant consonants
-...     'i':4,  'e':4,             # high front vowels
-...     'u':5,  'o':5,             # high back vowels
-...     'a':6                      # low vowels
-... }
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Computing Soundex (cont)}
-\scriptsize
-
-\begin{verbatim}
->>> def soundex(word):
-...     if len(word) == 0: return word  # sanity check
-...     word += '    '                  # ensure word long enough
-...     c0 = word[0].upper()
-...     c1 = group[word[1]]
-...     cons = filter(lambda x: x in 'pbvtdslr ', word[2:])
-...     c2 = group[cons[0]]
-...     c3 = group[cons[1]]
-...     return "%s%d%d%d" % (c0, c1, c2, c3)
->>> print soundex('kalosavi')
-K632
->>> print soundex('ti')
-T400
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Soundex Index}
-\begin{verbatim}
->>> soundex_idx = {}
->>> for lex in lexemes:
-...     code = soundex(lex)
-...     if code not in soundex_idx:
-...         soundex_idx[code] = set()
-...     soundex_idx[code].add(lex)
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Sorting and Lookup}
-\scriptsize
-\begin{verbatim}
->>> from nltk.utilities import edit_dist
->>> def fuzzy_spell(target):
-...     scored_candidates = []
-...     code = soundex(target)
-...     for word in soundex_idx[code]:
-...         dist = edit_dist(word, target)
-...         scored_candidates.append((dist, word))
-...     scored_candidates.sort()
-...     return [w for (d,w) in scored_candidates[:10]]
-\end{verbatim}
-
-\begin{verbatim}
->>> fuzzy_spell('kokopouto')
-['kokopeoto', 'kokopuoto', 'kokepato', 'koovoto', 'koepato',
- 'kooupato', 'kopato', 'kopiito', 'kovuto', 'koavaato']
->>> fuzzy_spell('kogou')
-['kogo', 'koou', 'kokeu', 'koko', 'kokoa', 'kokoi', 'kokoo',
- 'koku', 'kooe', 'kooku']
-\end{verbatim}
-\end{frame}
-
-\section{Generating Reports}
-
-\subsection{Summaries}
-
-\begin{frame}[fragile]
-\frametitle{Generating Summary Reports}
-\begin{itemize}
-\item overall picture of the quality and organisation of the data
-\item e.g. average number of fields per record
+\item metadata: DC, OLAC
+\item repositories
+\item search
+\item demo
 \end{itemize}
 
-\small
-\begin{verbatim}
->>> sum(len(entry) for entry in lexicon) / len(lexicon)
-10
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Generating Summary Reports (cont)}
-\begin{itemize}
-\item E.g. entry patterns
-\end{itemize}
-
-\small
-\begin{verbatim}
->>> fd = FreqDist()
->>> for entry in lexicon:
-...     for field in entry:
-...	        fd.inc(field.tag)
->>> fd.sorted_samples()[:10]
-['ge', 'ex', 'xe', 'xp', 'gp', 'lx', 'ps',
-'dt', 'rt', 'eng']
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Generating Summary Reports (cont)}
-\small
-
-\begin{verbatim}
->>> fd = FreqDist()
->>> for entry in lexicon:
-...	    fd.inc(':'.join(field.tag for field in entry))
->>> top_ten = fd.sorted_samples()[:10]
->>> print '\n'.join(top_ten)
-lx:rt:ps:ge:gp:dt:ex:xp:xe
-lx:ps:ge:gp:dt:ex:xp:xe
-lx:ps:ge:gp:dt:ex:xp:xe:ex:xp:xe
-lx:rt:ps:ge:gp:dt:ex:xp:xe:ex:xp:xe
-lx:ps:ge:gp:nt:dt:ex:xp:xe
-lx:ps:ge:gp:dt
-lx:ps:ge:ge:gp:dt:ex:xp:xe:ex:xp:xe
-lx:rt:ps:ge:ge:gp:dt:ex:xp:xe:ex:xp:xe
-lx:ps:ge:ge:gp:dt:ex:xp:xe
-lx:rt:ps:ge:ge:gp:dt:ex:xp:xe
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Generating Summary Reports (cont)}
-\small
-
-\begin{verbatim}
->>> fd = FreqDist()
->>> for entry in lexicon:
-...     previous = "0"
-...     for field in entry:
-...         current = field.tag
-...         fd.inc("%s->%s" % (previous, current))
-...         previous = current
->>> fd.sorted_samples()[:10]
-['ex->xp', 'xp->xe', '0->lx', 'ge->gp', 'ps->ge',
-'dt->ex', 'lx->ps', 'gp->dt', 'xe->ex', 'lx->rt']
-\end{verbatim}
-\end{frame}
-
-\subsection{Grammar}
-
-\begin{frame}[fragile]
-\frametitle{Grammar}
-\small
-
-\begin{verbatim}
->>> from nltk import parse
->>> grammar = parse.cfg.parse_grammar('''
-...   S -> Head "ps" Glosses Comment "dt" Examples
-...   Head -> "lx" | "lx" "rt"
-...   Glosses -> Gloss Glosses
-...   Glosses ->
-...   Gloss -> "ge" | "gp"
-...   Examples -> Example Examples
-...   Examples ->
-...   Example -> "ex" "xp" "xe"
-...   Comment -> "cmt"
-...   Comment ->
-...   ''')
->>> rd_parser = parse.RecursiveDescent(grammar)
-\end{verbatim}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Grammar}
-\scriptsize
-
-\begin{verbatim}
->>> for entry in lexicon[10:20]:
-...     marker_list = [field.tag for field in entry]
-...     if rd_parser.get_parse_list(marker_list):
-...         print "+", ':'.join(marker_list)
-...     else:
-...         print "-", ':'.join(marker_list)
-- lx:ps:ge:gp:sf:nt:dt:ex:xp:xe:ex:xp:xe
-- lx:rt:ps:ge:gp:nt:dt:ex:xp:xe:ex:xp:xe
-- lx:ps:ge:gp:nt:dt:ex:xp:xe:ex:xp:xe
-- lx:ps:ge:gp:nt:sf:dt
-- lx:ps:ge:gp:dt:cmt:ex:xp:xe:ex:xp:xe
-+ lx:ps:ge:ge:ge:gp:cmt:dt:ex:xp:xe
-+ lx:rt:ps:ge:gp:cmt:dt:ex:xp:xe:ex:xp:xe
-+ lx:rt:ps:ge:ge:gp:dt
-- lx:rt:ps:ge:ge:ge:gp:dt:cmt:ex:xp:xe:ex:xp:xe:ex:xp:xe
-+ lx:rt:ps:ge:gp:dt:ex:xp:xe
-\end{verbatim}
-\end{frame}
-
-\subsection{Timestamps}
-
-\begin{frame}[fragile]
-\frametitle{Looking at Timestamps}
-
-\scriptsize
-\begin{verbatim}
->>> fd = FreqDist()
->>> from string import split
->>> for entry in shoebox.dictionary('rotokas'):
-...     if 'dt' in entry:
-...         (day, month, year) = split(entry['dt'], '/')
-...         fd.inc((month, year))
->>> for time in fd.sorted_samples():
-...     print time[0], '/', time[1], ':', fd.count(time)
-Feb / 2005 : 307
-Dec / 2004 : 151
-Jan / 2005 : 123
-Feb / 2004 : 64
-Sep / 2004 : 49
-May / 2005 : 46
-Mar / 2005 : 37
-Apr / 2005 : 29
-Jul / 2004 : 14
-Nov / 2004 : 5
-Oct / 2004 : 5
-...
-\end{verbatim}
 \end{frame}

### 2007-07-24 rst.py
updated to correctly build doctest html files
svn/trunk@4990

@@ -104,6 +104,8 @@ CALLOUT_IMG = '<img src="callouts/callout%s.gif" alt="[%s]" class="callout" />'
 REF_EXTENSION = '.ref'
 """File extension for reference files."""
 
+CSS_STYLESHEET = '../nltkdoc.css'
+
 ######################################################################
 #{ Reference files
 ######################################################################
@@ -511,6 +513,10 @@ directives.register_directive('gloss', gloss_directive)
 class Citations(Transform):
     default_priority = 500 # before footnotes.
     def apply(self):
+        if not os.path.exists(BIBTEX_FILE):
+            warning('Warning bibtex file %r not found.  '
+                    'Not linking citations.' % BIBTEX_FILE)
+            return
         bibliography = self.read_bibinfo(BIBTEX_FILE)
         for k, citation_refs in self.document.citation_refs.items():
             for citation_ref in citation_refs[:]:
@@ -1356,7 +1362,7 @@ epydoc.docwriter.html_colorize .PYSRC_EXPANDTO_JAVASCRIPT = ''
 class CustomizedHTMLWriter(HTMLWriter):
     settings_defaults = HTMLWriter.settings_defaults.copy()
     settings_defaults.update({
-        'stylesheet': '../nltkdoc.css',
+        'stylesheet': CSS_STYLESHEET,
         'stylesheet_path': None,
         'output_encoding': 'ascii',
         'output_encoding_error_handler': 'xmlcharrefreplace',
@@ -2306,12 +2312,16 @@ def parse_args():
     optparser.add_option("--letter",
         action="store_const", dest="papersize", const="letterpaper",
         help="Use letter paper size.")
+    optparser.add_option("--css",
+        action="store", dest="css", help="CSS stylesheet")
     optparser.add_option("--bibliography",
         action="store_const", dest="bibliography", const=True,
         help="Include a bibliography (LaTeX only).")
 
     optparser.set_defaults(action='html', documentclass='report',
-                           papersize='letterpaper', bibliography=False)
+                           papersize='letterpaper',
+                           bibliography=False,
+                           css=CSS_STYLESHEET)
 
     options, filenames = optparser.parse_args()
     return options, filenames
@@ -2327,6 +2337,10 @@ def main():
     if docutils.writers.html4css1.Image is None:
         warning('Cannot scale images in HTML unless Python '
              'Imaging\n         Library (PIL) is installed!')
+
+    if options.css:
[…]

### 2007-07-24 doctests/update_list.py
moved makefile for converting doctests->html to /doc/doctests
svn/trunk@4994

@@ -1,0 +1,68 @@
+import os, os.path, re, sys
+
+DOCTEST_SRC = '../../nltk/test'
+
+HEAD = (".. ==========================================================\n"
+        ".. AUTO-GENERATED LISTING -- DO NOT EDIT!:\n")
+FOOT = (".. END AUTO-GENERATED LISTING\n"
+        ".. ==========================================================\n")
+LISTING_RE = re.compile(re.escape(HEAD)+'.*'+re.escape(FOOT), re.DOTALL)
+
+TITLE_REGEXPS = (
+    '\A\s*----+[ ]*\n(.*)\n----+[ ]*\n',
+    '\A\s*====+[ ]*\n(.*)\n====+[ ]*\n',
+    '\A\s*(.*)\n====+[ ]*\n',
+    '\A\s*(.*)\n----+[ ]*\n')
+
+def find_title(basename):
+    filename = os.path.join(DOCTEST_SRC, basename + '.doctest')
+    head = open(filename).read(400)
+    for regexp in TITLE_REGEXPS:
+        m = re.match(regexp, head)
+        if m: return m.group(1)
+    print 'Warning: no title found for %s' % basename
+    return basename
+
+def doctest_listing():
+    listing = ''
+    
+    files = [f for f in os.listdir(DOCTEST_SRC) if f.endswith('.doctest')]
+    for filename in files:
+        basename = filename.replace('.doctest', '')
+        if basename == 'temp': continue
+        
+        if (os.path.exists(basename+'.errs') and 
+            not re.search(r'OK\s*\Z', open(basename+'.errs').read())):
+            errs = '[`Test(s) failed! <%s.errs>`__]' % basename
+            print 'test %s failed' % basename
+        else:
+            errs = ''
+
+        title = find_title(basename)
+        listing += ('* `%s <%s.html>`__ -- %s %s\n' %
+                    (basename, basename, title, errs))
+    return listing
+
+def splice_listing(filename, listing):
+    # Read index:
+    index = open(filename).read()
+    
+    # Sanity check:
+    matches = len(LISTING_RE.findall(index))
+    if matches == 0:
+        print >>sys.stderr, "Splice location not found!"
+        sys.exit(-1)
+    if matches > 1:
+        print >>sys.stderr, "Multiple splice locations found!"
+        sys.exit(-1)
+
+    # Write new index:
+    out = open(filename, 'w')
[…]

### 2007-07-24 rst.py
added -o option; fixed bug in removing blocktest blocks for doctests; strip doctest directives from doctests
svn/trunk@4997

@@ -411,6 +411,10 @@ callout_directive.arguments = (0,1,0) # 1 optional arg, no whitespace
 callout_directive.content = True
 directives.register_directive('callouts', callout_directive)
 
+_OPTION_DIRECTIVE_RE = re.compile(
+    r'(\n[ ]*\.\.\.[ ]*)?#\s*doctest:\s*([^\n\'"]*)$', re.MULTILINE)
+def strip_doctest_directives(text):
+    return _OPTION_DIRECTIVE_RE.sub('', text)
 
 ######################################################################
 #{ RST In/Out table
@@ -1347,10 +1351,11 @@ class UnindentDoctestVisitor(docutils.nodes.NodeVisitor):
         docutils.nodes.NodeVisitor.__init__(self, document)
     def unknown_visit(self, node): pass
     def unknown_departure(self, node): pass
-    def visit_doctest_block(self, node):
-        if (isinstance(node.parent, docutils.nodes.block_quote) and
-            len(node.parent.children) == 1):
-            node.parent.replace_self(node)
+    def visit_block_quote(self, node):
+        if (len(node) == sum([1 for c in node if
+                              isinstance(c, docutils.nodes.doctest_block)])):
+            node.replace_self(list(node))
+        raise docutils.nodes.SkipNode()
         
 ######################################################################
 #{ HTML Output
@@ -1391,6 +1396,7 @@ class CustomizedHTMLTranslator(HTMLTranslator):
     def visit_doctest_block(self, node):
         # Collect the text content of the doctest block.
         text = ''.join(str(c) for c in node)
+        text = strip_doctest_directives(text)
 
         # Colorize the contents of the doctest block.
         colorizer = HTMLDoctestColorizer(self.encode, node['callouts'])
@@ -1754,6 +1760,7 @@ class CustomizedLaTeXTranslator(LaTeXTranslator):
 
     def visit_doctest_block(self, node):
         text = ''.join(str(c) for c in node)
+        text = strip_doctest_directives(text)
         colorizer = LaTeXDoctestColorizer(self.encode, wrap=False,
                                           callouts=node['callouts'])
         self.literal = True
@@ -1892,38 +1899,40 @@ class CustomizedLaTeXTranslator(LaTeXTranslator):
     def depart_pylisting(self, node):
         self.body.append( self.context.pop() )
 
-    def visit_pysrc_block(self, node):
-        self.literal = True
-        colorizer = LaTeXDoctestColorizer(self.encode, wrap=False, 
[…]

### 2007-08-30 slides/introduction.tex
/doc/slides/*.tex * updated for 0.9
/nltk/wordnet/browse.py
* accepted Jussi Salmela's fixes


svn/trunk@5325

@@
   \frametitle{Searching Pronunciation Dictionary}
-  \tiny
+  \scriptsize
 
 \begin{verbatim}
-    >>> from nltk.corpora import cmudict
-    >>> from string import join
-    >>> for word, num, pron in cmudict.raw():
-    ...     stress_pattern = join(c for c in join(pron) if c in "012")
-    ...     if stress_pattern.endswith("1 0 0 0 0"):
-    ...         print word, "/", join(pron)
 ACCUMULATIVELY / AH0 K Y UW1 M Y AH0 L AH0 T IH0 V L IY0
@@
   \frametitle{Minimal Sets from Lexicon}
-  \tiny
+  \small
 \begin{verbatim}
-    >>> from nltk.corpora import shoebox
-    >>> from nltk.utilities import MinimalSet
-    >>> length, position, min = 4, 1, 3
-    >>> lexemes = [field[1].lower() for entry in shoebox.raw('rotokas')
-    ...		   for field in entry if field[0] == 'lx']
-    >>> ms = MinimalSet()
-    >>> for lex in lexemes:
-    ...     if len(lex) == length:
-    ...         context = lex[:position] + '_' + lex[position+1:]
-    ...         target = lex[position]
-    ...         ms.add(context, target, lex)
-    >>> for context in ms.contexts(3):
-    ...     for target in ms.targets():
-    ...         print "%-4s" % ms.display(context, target, "-"),
-    ...     print
 kasi -    kesi kusi kosi
@@
   \frametitle{Modelling Text Genres}
-  \tiny
+  \small
 \begin{verbatim}
-    >>> from nltk.corpora import genesis
-    >>> from nltk.probability import ConditionalFreqDist
-    >>> from nltk.utilities import print_string
-    >>> cfdist = ConditionalFreqDist()
-    >>> prev = None
-    >>> for word in genesis.raw():
-    ...     word = word.lower()
-    ...     cfdist[prev].inc(word)
-    ...     prev = word
-    >>> words = []
-    >>> prev = 'lo,'
-    >>> for i in range(99):
-    ...     words.append(prev)
-    ...     for word in cfdist[prev].sorted_samples():
-    ...         if word not in words:
-    ...             break
-    ...     prev = word
-    >>> print_string(join(words))
 lo, it came to the land of his father and he said, i will not be a
@@
   \frametitle{Exploring Syntax}
-  \tiny
+  \scriptsize
 \begin{verbatim}
-    >>> from nltk.corpora import treebank
-    >>> from string import join
-    >>> def vp_conj(tree):
-    ...     if tree.node == 'VP' and len(tree) == 3 and tree[1].leaves() == ['but']:
-    ...         return True
-    ...     else:
-    ...         return False
-    >>> for tree in treebank.parsed():
-    ...     for vp1,conj,vp2 in tree.subtrees(vp_conj):
-    ...         print join(child.node for child in vp1), "*BUT*", join(child.node for child in vp2)
 VBP ADVP-TMP PP-PRD PP *BUT* VBP VP
@@
 \begin{enumerate}
-\item Install Python, Install NLTK, NLTK-Corpora
-\item Optional: Numeric, Matplotlib
-\item Set environment variable \verb|NLTK_CORPORA|
+\item Install Python, Install NLTK, NLTK-Data
+\item Optional: Numpy, Matplotlib
+\item Set environment variable \verb|NLTK_DATA|
 \end{enumerate}
@@
 \begin{itemize}
-\item  http://nltk.sourceforge.net/install.html
+\item  http://nltk.org/index.php/Installation
 \item  CDROM: /install.html

### 2007-08-30 slides/programming.tex
fixes to get slides to build
svn/trunk@5031
/doc/slides/*.tex * updated for 0.9
/nltk/wordnet/browse.py
* accepted Jussi Salmela's fixes


svn/trunk@5325

--- 2007-06-11 slides/programming.tex
@@
 \begin{verbatim}
-  >>> from nltk.corpora import gutenberg
+  >>> import nltk
   >>> count = {}
-  >>> for word in gutenberg.raw('shakespeare-macbeth'):
+  >>> for word in nltk.corpus.gutenberg.words('shakespeare-macbeth'):
   ...     word = word.lower()
@@
 \begin{verbatim}
-  >>> import re
-  >>> from nltk.utilities import re_show
+  >>> import nltk, re
   >>> sent = "colourless green ideas sleep furiously"
@@
 \begin{verbatim}
-  >>> re_show('l', sent)
+  >>> nltk.re_show('l', sent)
   co{l}our{l}ess green ideas s{l}eep furious{l}y
-  >>> re_show('green', sent)
+  >>> nltk.re_show('green', sent)
   colourless {green} ideas sleep furiously
@@
 \begin{verbatim}
-  >>> re_show('(green|sleep)', sent)
+  >>> nltk.re_show('(green|sleep)', sent)
   colourless {green} ideas {sleep} furiously
@@
 \begin{verbatim}
-  >>> re_show('[^aeiou][aeiou]', sent)
+  >>> nltk.re_show('[^aeiou][aeiou]', sent)
   {co}{lo}ur{le}ss g{re}en{ i}{de}as s{le}ep {fu}{ri}ously
@@
   >>> page = urlopen("http://news.bbc.co.uk/").read()
-  >>> page = re.sub('<[^>]*>', '', page)
-  >>> page = re.sub('\s+', ' ', page)
-  >>> print page[:60]
+  >>> text = nltk.clean_html(page)
+  >>> print text[:60]
    BBC NEWS | News Front Page News Sport Weather World Service
@@
 \item data structures, algorithms
-\item importing, e.g. \texttt{from nltk\_lite.utilities import *}
-\item conflicts
-\item explicit naming: \texttt{from nltk\_lite.utilities import re\_show}
-\item higher level module import:
+\item importing, e.g. \texttt{import nltk}
 
@@
 \begin{verbatim}
-  >>> from nltk.corpora import gutenberg
-  >>> gutenberg.items
+  >>> nltk.corpus.gutenberg.items
   ['austen-emma', 'austen-persuasion', 'austen-sense', 'bible-kjv', 'blake-poems', 'blake-songs', 'chesterton-ball', 'chesterton-brown', 'chesterton-thursday', 'milton-paradise', 'shakespeare-caesar', 'shakespeare-hamlet', 'shakespeare-macbeth', 'whitman-leaves']
   >>> count = 0
-  >>> for word in gutenberg.raw('whitman-leaves'):
+  >>> for word in nltk.corpus.gutenberg.words('whitman-leaves'):
   ...     count += 1
@@
 \begin{verbatim}
-  >>> from nltk.corpora import brown
-  >>> brown.items
+  >>> nltk.corpus.brown.items
   ['a', 'b', 'c', 'd', 'e', 'f', 'g', 'h', 'j', 'k', 'l', 'm', 'n', 'p', 'r']
-  >>> from nltk.corpora import extract
-  >>> print extract(0, brown.raw())
+  >>> print nltk.corpus.brown.words('a')
   ['The', 'Fulton', 'County', 'Grand', 'Jury', 'said', 'Friday', 'an', 'investigation', 'of', "Atlanta's", 'recent', 'primary', 'election', 'produced', '``', 'no', 'evidence', "''", 'that', 'any', 'irregularities', 'took', 'place', '.']
-  >>> print extract(0, brown.tagged())
+  >>> print nltk.corpus.brown.tagged_sents('a')
   [('The', 'at'), ('Fulton', 'np-tl'), ('County', 'nn-tl'), ('Grand', 'jj-tl'), ('Jury', 'nn-tl'), ('said', 'vbd'), ('Friday', 'nr'), ('an', 'at'), ('investigation', 'nn'), ('of', 'in'), ("Atlanta's", 'np$'), ('recent', 'jj'), ('primary', 'nn'), ('election', 'nn'), ('produced', 'vbd'), ('``', '``'), ('no', 'at'), ('evidence', 'nn'), ("''", "''"), ('that', 'cs'), ('any', 'dti'), ('irregularities', 'nns'), ('took', 'vbd'), ('place', 'nn'), ('.', '.')]
@@
 \begin{verbatim}
-  >>> from nltk.corpora import treebank
-  >>> print extract(0, treebank.parsed())
+  >>> print nltk.corpus.treebank.parsed_sents('wsj_0001')[0]
   (S:
@@
 
+% \section{Miscellaneous Python Features}
+
+% \subsection{Character Encoding}
+
+% \begin{frame}
+% \begin{itemize}
+
+% \end{itemize}
+% \end{frame}
+
 \end{document}

### 2007-09-04 slides/preface.tex
/doc/slides/preface.tex * new introductory slides * moved material from introduction.tex
/nltk/test.portuguese2.doctest
* Portuguese translation by Tiago Tresoldi
  (unfortunately rst.py breaks on non-ascii characters)

/data/corpora/mac_morpho/*
* POS-tagged corpus for Brazilian Portuguese
  contributed by the NILC group from the
  University of Sao Paulo in Sao Carlos

svn/trunk@5335

\documentclass{beamer}             % for slides
% \documentclass[handout]{beamer}    % for handout
\input{beamer}

\title{Introduction to Natural Language Processing}

\author{Steven Bird \and Ewan Klein \and Edward Loper}
\institute{
  University of Melbourne, AUSTRALIA
  \and
  University of Edinburgh, UK
  \and
   University of Pennsylvania, USA
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Preface}

\begin{frame}
  \titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Overview}

\begin{frame}
  \frametitle{Knowledge and Communication in Language}
  \begin{itemize}
    \item human knowledge, human communication, expressed in language
    \item language technologies: process human language automatically
    \item handheld devices: predictive text, handwriting recognition
    \item web search engines: access to information locked up in text
    \item two facets of the multilingual information society:
    \begin{itemize}
      \item natural human-machine interfaces
      \item access to stored information
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Problem}
  \begin{itemize}
    \item awash with language data
    \item inadequate tools (will this ever change?)
    \item overheads: Perl, Prolog, Java
    \item Natural Language Toolkit (NLTK) as a solution
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{NLTK: What you get...}
  \begin{itemize}
    \item Book
    \item Documentation
    \item FAQ
    \item Installation instructions for Python, NLTK, data
    \item Distributions:
      \textit{Windows, Mac OSX, Unix, data, documentation}
    \item CD-ROM:
      \textit{Python, NLTK, documentation, third-party libraries for numerical
        processing and visualization, instructions}
    \item Mailing lists:\\
      \texttt{nltk-announce}, 
      \texttt{nltk-devel}, 
      \texttt{nltk-users}, 
      \texttt{nltk-portuguese}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{NLTK: Who it is for...}
  \begin{itemize}
    \item people who want to learn how to:
    \begin{itemize}
      \item write programs
      \item to analyze written language
    \end{itemize}
    \item does not presume programming abilities:
    \begin{itemize}
      \item working examples
      \item graded exercises
    \end{itemize}
    \item experienced programmers:
    \begin{itemize}
      \item quickly learn Python (if necessary)
      \item Python features for NLP
      \item NLP algorithms and data structures
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{NLTK: What you will learn...}
  \begin{enumerate}
    \item how to analyze language data
    \item key concepts from linguistic description and analysis
    \item how linguistic knowledge is used in NLP components
    \item data structures and algorithms used in NLP and linguistic data management
    \item standard corpora and their use in formal evaluation
    \item organization of the field of NLP
    \item skills in Python programming for NLP
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{NLTK: Your likely goals...}
\small
\begin{tabular}{p{0.18\textwidth}|p{0.38\textwidth}|p{0.38\textwidth}}
\textbf{Goals} & \multicolumn{2}{c}{\textbf{Background}} \\
& \textit{Arts and Humanities} & \textit{Science and Engineering} \\ \hline
\textbf{Language Analysis}
& Programming to manage language data, explore linguistic models, and test empirical claims
& Language as a source of interesting problems in data modeling, data mining, and knowledge discovery\\ \hline
\textbf{Language Technology}
& Learning to program, with applications to familiar problems, to work in language technology or other technical field
& Knowledge of linguistic algorithms and data structures for high quality, maintainable language processing software\\
\end{tabular}

\end{frame}

\section{NLTK Book}

\begin{frame}
  \frametitle{Philosophy}
  \begin{itemize}
    \item practical
    \item programming
    \item principled
    \item pragmatic
    \item pleasurable
    \item portal    
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Structure}
  \begin{itemize}
    \item Three parts:
    \begin{enumerate}
      \item \textbf{Basics:} text processing, tokenization, tagging,
        lexicons, simple engineering
      \item \textbf{Parsing:} phrase structure, trees, grammars, parsing
      \item \textbf{Advanced Topics:} selected topics in greater depth:
        feature-based grammar, semantics, linguistic data management
    \end{enumerate}
    \item each part: chapter on programming; three chapters on NLP
    \item each chapter: motivation, sections, graded exercises, summary, further reading
  \end{itemize}
\end{frame}

\section{The Python Programming Language}

\subsection{Key Features}

\begin{frame}
  \frametitle{Python: Key Features}
  \begin{itemize}
    \item simple yet powerful, shallow learning curve
    \item object-oriented: encapsulation, re-use
    \item scripting language, facilitates interactive exploration
    \item excellent functionality for processing linguistic data
    \item extensive standard library, incl graphics, web, numerical processing
    \item downloaded for free from \url{http://www.python.org/}
  \end{itemize}
\end{frame}

\subsection{Python Example}

\begin{frame}[fragile]
  \frametitle{Python Example}

\begin{verbatim}
 import sys
 for line in sys.stdin.readlines():
     for word in line.split():
         if word.endswith('ing'):
             print word
\end{verbatim}

\begin{enumerate}
\item whitespace: nesting lines of code; scope
\item object-oriented: attributes, methods (e.g. \texttt{line})
\item readable
\end{enumerate}
\end{frame}

\subsection{Comparison with Perl}
\begin{frame}[fragile]
\frametitle{Comparison with Perl}

\begin{verbatim}
  while (<>) {
      foreach my $word (split) {
          if ($word =~ /ing$/) {
              print "$word\n";
          }
      }
  }
\end{verbatim}

\begin{enumerate}
\item syntax is obscure: \textit{what are:} \verb|<> $ my split| ?
\item ``it is quite easy in Perl to write programs that simply
  look like raving gibberish, even to experienced Perl programmers''
  (Hammond \textit{Perl Programming for Linguists} 2003:47)
\item large programs difficult to maintain, reuse
\end{enumerate}
\end{frame}

\subsection{What NLTK adds to Python}
\begin{frame}
\frametitle{What NLTK adds to Python}

NLTK defines a basic infrastructure that can be used to build NLP
programs in Python.  It provides:

\begin{itemize}
\item Basic classes for representing data relevant to natural language
  processing
\item Standard interfaces for performing tasks, such
  as tokenization, tagging, and parsing
\item Standard implementations for each task, which
  can be combined to solve complex problems
\item Extensive documentation, including tutorials
  and reference documentation
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{NLTK Design: Requirements}
  \begin{enumerate}
    \item \textbf{simplicity:} intuitive framework with substantial building blocks
    \item \textbf{consistency:} uniform data structures, interfaces --- predictability
    \item \textbf{extensibility:} accommodates new components
      (replicate vs extend exiting functionality)
    \item \textbf{modularity:} interaction between components
    \item \textbf{well-documented:} substantial documentation
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{NLTK Design: Non-requirements}
  \begin{enumerate}
    \item \textbf{encyclopedic:} has many gaps; opportunity for students to extend it
    \item \textbf{efficiency:} not highly optimised for runtime performance
    \item \textbf{programming tricks:} avoid in preference for clear implementations
      (replicate vs extend exiting functionality)
    \item \textbf{modularity:} interaction between components
    \item \textbf{well-documented:} substantial documentation
  \end{enumerate}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Corpora Distributed with NLTK}
  \tiny
  \begin{itemize}
    \item Australian ABC News, 2 genres, 660k words, sentence-segmented
    \item Brown Corpus, 15 genres, 1.15M words, tagged
    \item CMU Pronouncing Dictionary, 127k entries
    \item CoNLL 2000 Chunking Data, 270k words, tagged and chunked
    \item CoNLL 2002 Named Entity, 700k words, pos- and named-entity-tagged (Dutch, Spanish)
    \item Floresta Treebank, 9k sentences (Portuguese)
    \item Genesis Corpus, 6 texts, 200k words, 6 languages
    \item Gutenberg (sel), 14 texts, 1.7M words
    \item Indian POS-Tagged Corpus, 60k words pos-tagged (Bangla, Hindi, Marathi, Telugu)
    \item NIST 1999 Info Extr (sel), 63k words, newswire and named-entity SGML markup
    \item Names Corpus, 8k male and female names
    \item PP Attachment Corpus, 28k prepositional phrases, tagged as noun or verb modifiers
    \item Presidential Addresses, 485k words, formatted text
    \item Roget's Thesaurus, 200k words, formatted text
    \item SEMCOR, 880k words, part-of-speech and sense tagged
    \item SENSEVAL 2, 600k words, part-of-speech and sense tagged
    \item Shakespeare XML Corpus (sel), 8 books
    \item Stopwords Corpus, 2,400 stopwords for 11 languages
    \item Switchboard Corpus (sel), 36 phonecalls, transcribed, parsed
    \item Univ Decl Human Rights, 480k words, 300+ languages
    \item US Pres Addr Corpus, 480k words
    \item Penn Treebank (sel), 40k words, tagged and parsed
    \item TIMIT Corpus (sel), audio files and transcripts for 16 speakers
    \item Wordlist Corpus, 960k words and 20k affixes for 8 languages
    \item WordNet, 145k synonym sets
  \end{itemize}
\end{frame}

\end{document}

### 2008-11-22 docbook.py
doc/ch05.rst * commented out notes and empty sections * empty sections not allowed by docbook * changed code to use nltk.trigrams() * removed instances of *emphasis* inside a   bulletted list, which docbook complained about   (as emphasis inside a definition) * indented code blocks
doc/ch12.rst
* fixed xref to ch01

doc/ch*.rst
* changed sect-* to sec-* for consistent xrefs

doc/docbook.py
* added fix to get <informalfigure> markup just in
  those cases where a media object occurs outside
  a figure or table.



svn/trunk@6942

@@ -806,6 +806,11 @@ class DocBookTranslator(nodes.NodeVisitor):
         # else. - boney.
 
         # self.body.append('<para>')
+
+        # ensure image is contained in a figure so that it is allocated page space
+        if element == 'mediaobject' and not self.figure_tag_stack and not self.table_tag_stack:
+            self.body.append('<informalfigure>')
+
         self.body.append('<%s>' % element)
         self.body.append('<imageobject>')
         self.body.append(self.emptytag(node, 'imagedata', **atts))
@@ -814,6 +819,9 @@ class DocBookTranslator(nodes.NodeVisitor):
             self.body.append('<textobject><phrase>%s</phrase></textobject>\n' % node.attributes['alt'])
         self.body.append('</%s>' % element)
         # self.body.append('</para>')
+
+        if element == 'mediaobject' and not self.figure_tag_stack and not self.table_tag_stack:
+            self.body.append('</informalfigure>')
 
     def depart_image(self, node):
         pass

### 2008-11-22 docbook.py
Solved problem with duplicate identifiers appearing in XML output by prefixing figure labels consistently with "fig-".
svn/trunk@6944

@@ -668,7 +668,6 @@ class DocBookTranslator(nodes.NodeVisitor):
         atts = {}
         if 'ids' in node.attributes and node.attributes['ids']:
             atts['id'] = node.attributes['ids'][0]
-        
         try:
             last_child = node.children[-1]
             if isinstance(last_child, nodes.caption) and \

### 2008-11-22 docbook.py
* fixed typo in code * trying to fix problem with images inside tables
svn/trunk@6947

@@ -678,7 +678,7 @@ class DocBookTranslator(nodes.NodeVisitor):
             else:
                 figure_tag = "informalfigure"
         except IndexError:
-            rigure_tag = "informalfigure"
+            figure_tag = "informalfigure"
                 
         self.body.append(self.starttag(node, figure_tag, **atts))
         self.stack_push(self.figure_tag_stack, figure_tag)
@@ -807,7 +807,7 @@ class DocBookTranslator(nodes.NodeVisitor):
         # self.body.append('<para>')
 
         # ensure image is contained in a figure so that it is allocated page space
-        if element == 'mediaobject' and not self.figure_tag_stack and not self.table_tag_stack:
+        if element == 'mediaobject' and not self.figure_tag_stack: # and not self.table_tag_stack:
             self.body.append('<informalfigure>')
 
         self.body.append('<%s>' % element)
@@ -819,7 +819,7 @@ class DocBookTranslator(nodes.NodeVisitor):
         self.body.append('</%s>' % element)
         # self.body.append('</para>')
 
-        if element == 'mediaobject' and not self.figure_tag_stack and not self.table_tag_stack:
+        if element == 'mediaobject' and not self.figure_tag_stack: # and not self.table_tag_stack:
             self.body.append('</informalfigure>')
 
     def depart_image(self, node):

### 2008-11-25 docbook.py
Makefile * Fix for 0.9.6b1 release; remove data from dmg distribution,   complete command lines for syncing to googlecode
doc/docbook.py
* reinstated condition for mediaobjects inside tables

nltk/__init__.py
* changed version to 0.9.6b1

doc/en/ch06.rst
* removing redundant content
* updated example to use enumerate instead
  of loop counter

nltk/test/tag.doctest
* fixed formatting problem that caused doctest to fail



svn/trunk@6972

@@ -807,7 +807,7 @@ class DocBookTranslator(nodes.NodeVisitor):
         # self.body.append('<para>')
 
         # ensure image is contained in a figure so that it is allocated page space
-        if element == 'mediaobject' and not self.figure_tag_stack: # and not self.table_tag_stack:
+        if element == 'mediaobject' and not self.figure_tag_stack and not self.table_tag_stack:
             self.body.append('<informalfigure>')
 
         self.body.append('<%s>' % element)
@@ -819,7 +819,7 @@ class DocBookTranslator(nodes.NodeVisitor):
         self.body.append('</%s>' % element)
         # self.body.append('</para>')
 
-        if element == 'mediaobject' and not self.figure_tag_stack: # and not self.table_tag_stack:
+        if element == 'mediaobject' and not self.figure_tag_stack and not self.table_tag_stack:
             self.body.append('</informalfigure>')
 
     def depart_image(self, node):
